---
title: "Introduction to Statistical Methods"
subtitle: "Topic 3: Numerical Measures and Boxplots"
author: "Longhai Li"
date: "`r Sys.Date()`"
output: 
   html_document:
       toc: true
       number_sections: true
       highlight: tango
       fig_width: 10
       fig_height: 8
---
 Researcher was interested in the side effects of radiation therapy on
 patents suffering cancerous lesions. In particular he was interested in the
 effect of radiation therapy on Mental dexterity. Each patient was scored
 (Pre.test) on a mental dexterity test prior to treatment. They were again
 scored on a similar test (Post.Test) one month after treatment began.  
 
 N1 = 16 were not given radiation and served as controls,
 
 N2 = 19 were given radiation dosage in the range 25-50, 
 
 N3 = 18 were given radiation dosage in the range 75-125, and 
 
 N4 = 21 were given radiation dosage in the range 125-250. 
 
 In the data set,
 
 Pre.test: mental dexterity score before test
 
 Post.test: mental dexterity score after test
 Treat: indicator of dosage level 

# Import a dataset called "mental.csv"
```{r }
mental <- read.csv ("mental.csv", stringsAsFactors = TRUE)
```

# Preview the dataset

```{r }
library(rmarkdown)
paged_table(mental)
plot(mental$Pre.test, col=as.numeric(mental$Treat))
plot(mental$Post.test, col=as.numeric(mental$Treat))

```

# Numerical measures 

```{r }
mean (mental$Pre.test)

median (mental$Pre.test)

range (mental$Pre.test)

var (mental$Pre.test)

sd (mental$Pre.test)

IQR (mental$Pre.test)

x <- c(1,2,3,4,5)

plot (ecdf(x))

plot (ecdf (mental$Pre.test))
abline (h = seq (0,1, by = 0.05), lty = 1, col = "grey")


quantile (mental$Pre.test)

quantile (mental$Pre.test, probs = 0.34)

quantile (mental$Pre.test, probs = seq (0, 1, by = 0.1))

quantile (mental$Pre.test, probs = seq (0, 1, by = 0.01))
```

# Boxplot

```{r }
boxplot (mental$Post.test)

boxplot (Post.test ~ Treat, data = mental)
```

# Change the default order of treatment

```{r }
levels(mental$Treat)
mental$Treat2 <- factor(mental$Treat, levels =  levels(mental$Treat)[c(4,1:3)])
levels(mental$Treat2)

paged_table(mental[, c("Treat", "Treat2")])
```

# Comparison Boxplot for Looking at the Treatment Effect

```{r }

boxplot (Pre.test ~ Treat2, data = mental)
boxplot (Post.test ~ Treat2, data = mental)
plot ( mental$Pre.test,mental$Post.test)
mental$dex.reduction <- mental$Pre.test - mental$Post.test
paged_table(mental)
boxplot (dex.reduction ~ Treat2, data = mental)
```

