---
title: "Introduction to Statistical Methods"
subtitle: "Topic 2: Frequency Graph and Scatterplot"
author: "Longhai Li"
date: "`r Sys.Date()`"
output: 
   html_document: 
     toc: yes
     number_sections: yes
     highlight: tango
     fig_width: 10
     fig_height: 8
editor_options: 
  chunk_output_type: inline
---

```{r include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE)
```


# Load the dataset

```{r }
survey <-  read.csv ("survey.csv", stringsAsFactors = TRUE)
library("rmarkdown")
paged_table(survey)
```
# Frequency 
get frequency table of variable 'Smoke' of data set 'survey' 

```{r }
smoke_freq <- table (survey$Smoke)
smoke_freq
```

# Bar graph
Draw bar graph of variable 'Smoke' of data set 'survey'

```{r }
barplot (smoke_freq)
```

we can also use a compact expression

```{r }
barplot (table (survey$Smoke))
```

Sort in the order of frequency

```{r }
barplot(sort(smoke_freq))
```

barplot can produce barplot of any named vector as we created

```{r }
afreq <- c(10,20,30,40,60); names(afreq) <- paste0("A", 1:5);afreq
barplot(afreq)
```

# Pie chart
Draw pie chart

```{r }
pie (smoke_freq)
```

or you can use a compact expression

```{r }
pie (table (survey$Smoke))
```

# Relative frequency and percentage

```{r }
smoke_relfreq <- smoke_freq / nrow (survey)
smoke_freq
```


```{r }
smoke_perc <- smoke_freq / nrow (survey) * 100
smoke_perc
```

find angles used in pie chart

```{r }
smoke_pie <- smoke_freq / nrow (survey) * 360
smoke_pie
```
# Histogram

draw histogram of variable 'Pulse' of data set 'survey'

```{r }
hist (survey$Pulse)
hist (survey$Pulse, nclass = 20)
hist (survey$Pulse, breaks = seq (35, 110, by = 5)) 
```


work with a vector of numbers 

```{r }

x <- scan (text = "2 3 4 2 1 2 3 4 4 2 2 2 2 2 1 1 1 2 4 5 5")
hist (x)

x <- scan ("numbers.txt")
hist (x)
```


# Cumulative frequency for a histogram

```{r }
pulse_hist <- hist (survey$Pulse, breaks = seq (35, 110, by = 5))
```

look at the results of hist function

```{r }
pulse_hist
```

draw culumative frequency

```{r }
plot (pulse_hist$mids, cumsum(pulse_hist$counts), type = "b")
```

draw culumative relative frequency

```{r }
n <- sum (pulse_hist$counts)
plot (pulse_hist$mids, cumsum(pulse_hist$counts)/n, type = "b")
```

# Empirical distribution function

Empirical distribution function computed with observations $x_1,\ldots,x_n$ is defined as
$$
\hat{F}(x) = \frac{\sum_{i=1}^n I(x_i\leq x)}{n} = \frac{\mbox{number of } x_i\leq x }{n}
$$
where, $I()$ is the indicator function, equal to 1 if the condition in () is true and 0 otherwise. 


```{r }
x <- c(1,2,3,4,5)

plot (ecdf(x))

plot(ecdf (survey$Pulse))
abline (h = seq (0,1, by = 0.05), v = seq(0,100,by=5), 
        lty = 2, col = "grey")

```

# Stem-leaf plot

```{r }
stem (survey$Pulse)
```

# Scatterplot of two or three variables 

```{r }
plot(survey$Height)
plot(survey$Height, col = as.factor(survey$Sex))
plot(survey$Pulse, col = survey$Sex)
plot(survey$Height~survey$Pulse)
plot(survey$Height~survey$Pulse, col = survey$Sex)
plot(survey$Wr.Hnd, col = survey$Sex)
plot(survey$Wr.Hnd~survey$Height, col = survey$Sex)
plot(survey$Wr.Hnd~survey$NW.Hnd, col = survey$Sex)
```

