---
title: "Unit0: Introduction to R"
date: "January 2020"
author: "Longhai Li"
output:
  html_document:
       theme: united
       toc: true
       toc_float: no
       number_sections: true
       highlight: tango
       fig_width: 10
       fig_height: 8
  pdf_document:
    toc: yes
    fig_height: 6
    fig_width: 10
    number_sections: yes
editor_options: 
  chunk_output_type: console
---

### Some Settings {-}

To find the working directory, type in R console: 

```{r}
getwd()
```


```{r setup, include=TRUE, echo=TRUE}
options(digits = 4)
library("knitr")
#Set working directory for "knitr":
#opts_knit$set(root.dir = "~/OneDrive - University of Saskatchewan/teaching/stat345-2001/rdemo")
#You can also choose the "Project Direcotry" beside "Knitr" button
```


# Basic R Objects and Operations

## Vector
```{r }
# create a vector
x <- 1:10
x <- seq (3,30, by = 2)
a <- c(66.32, 69.87, 70.12, 90.37, 50.08, 61.20, 65.00, 57.65)
d <- a [1]
a [1] <- 85.34
```

Find the average of numbers in a:
```{r }
mean (a)
```

Save the mean of a as another object:

```{r }
ma <- mean (a)
```

One can also read numbers using scan like this:

```{r }
y <- scan(text = "7  8  9 10 11 12 13 13 14 17 17 45")
y
```

Read a vector of numbers from a file:

```{r }
x <- scan("data/numbers.txt")
```

## Matrix

```{r }
# create a matrix
A <- matrix (0, 4, 2)
A

A <- matrix (1:8, 4,2)
A

D <- matrix (1:8, 4, 2, byrow=T)
D

D <- matrix(1:8, 2, 4)
D

# create another matrix with all entry 0
B <- matrix (0, 100, 50)

# assign a number to B
B[1,4] <- 4.5
B[1,] <- 1:50
head(B)
```

## List (Output of Most Data Analysis Functions)

Create a list

```{r }
alist <- list (newa = a, newA = A)
```

List the names of components:
```{r }
names (alist)
str (alist)
```

Look at the component of alist:
```{r }
alist$newA
```

Modify a component:

```{r }
alist$newa <- 10:17
```

## Data Frame (Input of Most Data Analysis Functions)

```{r }
# create a dataframe
scores <- c (30, 45, 50)
names <- c("Peter", "John", "Alice")
stat245_scores <- data.frame (names, scores)
stat245_scores$names
stat245_scores$scores [2] <- 17
stat245_scores$perc <- stat245_scores$score/50 * 100 + 10

stat245_scores
names(stat245_scores)
```

# Import a dataset into R and Simple Operation

```{r }

# import myagpop.csv into an R data frame called 'myagpop'
agpop <- read.csv ("data/agpop.csv")

# Now, we can use the data:

# preview agpop
head (agpop)

# look at the variable name
colnames (agpop) 

# find number of cols
ncol (agpop) 

# find number of rows
nrow (agpop) 

# access a certain row 
agpop [2, ]
# access a certain column
agpop [1:20, "acres92"] ## equivalent to 
agpop$acres92[1:20]
agpop$largef92[1:20]
# find mean of acres92
mean (agpop $acres92)
# find sd of acres92
sd (agpop $acres92)

agpop_AK  <- agpop [agpop$state == "AK", ]

agpop_AK <- subset (agpop, state == "AK")

agpop_W <- subset (agpop, region == "W")

agpop_largefarm <- subset (agpop, largef92 > 10)

## simple analysis
summary (agpop)



```

Produce Plots

Histogram of farming acreages in 1992:
```{r }
hist (agpop$acres92)
```

Scatterplot of farming  acreages in 1987 versus 1992
```{r }
plot (agpop$acres87, agpop$acres92)
abline (a = 0, b = 1)

#dev.off()## this is used to close the jpeg file
```

# Create your own function

```{r }
## data is a matrix or data.frame
means_col <- function (data)
{
    n <- ncol (data)
    cmeans <- rep (NA, n)
    for (j in 1:n)
    {
        cmeans[j] <- mean (data[,j])
        
    }
    names(cmeans) <- colnames(data)
    cmeans
}

## apply function
means_col (agpop[, 3:13])
## R built-in function
colMeans (agpop[, 3:13])
```


