---
title: '[STAT 345/834: Design and Analysis of Experiments](.)'
author: "Longhai Li"
date: "March 2018"
output:
  html_document:
    fig_height: 6
    fig_width: 10
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_float: no
  pdf_document:
    toc: yes
    number_sections: yes
    fig_height: 6
    fig_width: 10
subtitle: 'Unit 8: Introduction Response Surface Design'
---
# Draw Some Response Surfaces
## An Additive Model
```{r }
add.model = function(x1,x2)
{ 
  x1+x2
}

# response surface plot
x2<- x1 <- seq(0,1,by=0.01)
z <- outer(x1,x2, add.model)

persp (x1, x2, z, theta = 30)
# contour plot
contour(x1,x2,z)
```

## An Interaction Model
```{r }
inter.model = function(x1,x2)
{ 
  x1+x2+5*x1*x2
}

# response surface plot
x2<- x1 <- seq(0,1,by=0.01)
z <- outer(x1,x2, inter.model)

persp(x1, x2, z, theta=45)
# contour plot
contour(x1,x2,z)
```

## A Circular Quadratic Model
```{r fig.height=8, fig.width=8}
quad.model = function(x1,x2)
{ 
  x1+x2-10*(x1-0.5)^2-10*(x2-0.5)^2
}

# response surface plot
x2<- x1 <- seq (0,1,by=0.01)
z <- outer (x1,x2, quad.model)

persp(x1, x2, z, theta=45)
# contour plot
contour(x1,x2,z)
```

## An Elliptical Quadratic Model
```{r fig.height=8, fig.width=8}
quad.model = function(x1,x2)
{ 
  -x1-x2-10*(x1-0.5)^2-40*(x2-0.5)^2
}

# response surface plot
x2<- x1 <- seq(0,1,by=0.01)
z <- outer(x1,x2, quad.model)

persp(x1, x2, z, theta=45)
# contour plot
contour(x1,x2,z)
```


## A Complex Model
```{r fig.height=8, fig.width=8}
quad.model = function(x1,x2)
{ 
  x1+x2+5*x1*x2-10*(x1-.5)^2-20*(x2-.5)^2
}

# response surface plot
x2<- x1 <- seq (0,1,by=0.01)
z <- outer (x1,x2, quad.model)

persp (x1, x2, z, theta = 45)
# contour plot
contour (x1,x2,z)
```

# Design and Analysis of Second-order response surface around optimum

## Read in the data
```{r}
yield<- read.table("data/yield.txt", header=T)
yield
attach(yield)
```

## Without Central Composite Design
### Fit a 2nd-order model
```{r}


# define x1 and x2
x1<-c(-1,-1,1,1,rep(0,5))
x2<-c(-1,1,-1,1,rep(0,5))
data.frame (x1,x2, y=y[1:9]) # using only the first 9 observations

g<-lm(y[1:9]~x1+x2+I(x1^2)+x1*x2,data=yield)
summary(g)

```
### ANOVA
```{r}
anova (g)

```



## Analysis of Central Composite Design

In CCD,we take 4 more observations so that we can estimate pure quadratic effect of $x_2$. 

### Fit a 2nd order model with pure quadratic terms
```{r}

x1<-c(-1,-1,1,1,rep(0,5),1.414,-1.414,0,0)
x2<-c(-1,1,-1,1,rep(0,5),0,0,1.414,-1.414)

data.frame (x1,x2, y)

h<-lm(y~x1+x2+I(x1^2)+I(x2^2)+x1*x2,data=yield)

summary(h)

```
### ANOVA

```{r}
anova (h)

```

### Analysis after removing interaction term (non-significant)
```{r}
h2<-lm(y~x1+x2+I(x1^2)+I(x2^2),data=yield)
summary(h2)
anova(h2)
```

```{r}
detach ("yield")
```

