---
title: "Unit 7: Two-Level Fractional Factorial Design"
author: "Longhai Li"
date: "March 2020"
output: 
  html_document:
       theme: united
       toc: true
       toc_float: no
       number_sections: true
       highlight: tango
       fig_width: 10
       fig_height: 6
       df_print: default
  pdf_document:
    toc: yes
    keep_tex: no
    number_sections: yes
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

# Functions and Package{-}

```{r include=TRUE}
library(BsMD)## LenthPlot function from this package
library (FrF2)## FrF2 function is provided by this package

qqnorm.lenth <- function (effects)
{
    abseffects<- abs(effects)
    m <- length (abseffects)
    # Psuedo sd based all abs(effect)
    s0<-1.5*median(abseffects) 
    non.sig <- abseffects<2.5*s0
    # Refined estimate of sd 
    pse<-1.5*median(abseffects[non.sig]) 

    sme<-qt(1- (1-0.95^(1/m))/2,m/3)*pse;
    sig <- abseffects>sme
    
    hqq<-qqnorm(effects, type = "n")
    text(hqq$x, hqq$y, labels = names (effects), col = sig + 1)
    qqline(effects)
    cat ("Significant Factors Selected by Lenth's Method:\n",
         names(abseffects)[sig], "\n")

    names(effects)[sig]
}

```
# One Half Factorial Design and Analysis

## Full $2^5$ Factorial  Design
```{r}


des1 <- FrF2 (nfactors = 5, nruns = 32); des1

## look at the full design matrix
model.des1 <- data.frame( model.matrix(~A*B*C*D*E, des1))

model.des1 [with(model.des1, order (-A1.B1.C1.D1.E1, D1, C1, B1, A1)), 
            c("A1","B1", "C1", "D1", "E1", "A1.B1.C1.D1", "A1.C1.D1.E1",
            "A1.B1.C1", "D1.E1", "A1.B1.C1.D1.E1" ) ]

```
## $2^{5-1}$ design with E = ABCD (ABCDE=I)
```{r}
des2 <- FrF2 (nfactors = 5, nruns=16, generators = c("ABCD"), alias.info = 3)
des2 [with (des2, order (D,C,B,A)), ]
design.info(des2)$aliased
```


Aliase Structure: 
E = ABCD, A = BCDE, B=ACDE, C = ABDE, D = ABCE,   
AE = BCD, BE = ACD, CE = ABD, AB = CDE, AC = BDE, AD = BCE, BC = ADE, BD = ACE, CD = ABE, that is, all third order interactions are confounded with 2nd order interactions. Fourth-order interactions are also confounded with 1st-order main effect. 

## $2^{5-1}$ design with E = -ABCD (ABCDE=-I)
```{r}
des3 <- FrF2 (nfactors = 5, nruns=16, generators = c("-ABCD"), alias.info = 3)
des3[with (des3, order (D,C,B,A)), ]
design.info(des3)$aliased
```

Aliase Structure: 
E = -ABCD, A = -BCDE, B = -ACDE, C = -ABDE, D = -ABCE,   
AE = -BCD, BE = -ACD, CE = -ABD, AB = -CDE, AC = -BDE, AD = -BCE, BC = - ADE, BD = - ACE, CD = - ABE, that is, all third order interactions are confounded with 2nd order interactions. Fourth-order interactions are also confounded with 1st-order main effect. 

## Analysis of a Dataset Collected with $2^{5-1}$ Design


A $2^{5-1}$ design used for process improvement
Five factors in a manufacturing process for an integrated circuit were investigated in a $2^{5−1}$ design with the objective of improving the process yield. The five factors.



```{r }
# read in the data
process<- read.table("data/process.txt", header=T)

process

attach(process)
```
```{r}
# define the design matrix
A<-rep(c(-1,1),8)
B<-rep(c(-1,-1,1,1),4)
C<-rep(c(rep(-1,4),rep(1,4)),2)
D<-c(rep(-1,8),rep(1,8))
E<-A*B*C*D
```

 

```{r}

# alternatively, we can generate these design with FrF2:

des2 <- FrF2 (nfactors = 5, nruns=16, generators = c("ABCD"), alias.info = 3)
des2 [with (des2, order (D,C,B,A)), ]
design.info(des2)$aliased

```

```{r}

g <- lm(y~(A+B+C+D+E)^2, data = process)
summary(g)
anova(g)
effects<-g$coef[-1]

# using lenth method to screen effects 
qqnorm.lenth(effects)
LenthPlot(effects)

# Fit the model with significant effects
h<-lm(y~A+B+C+A*B,data=process)
anova(h)
summary(h)
par (mfrow = c(2,2))
boxplot(y~A)
boxplot(y~B)
boxplot(y~C)


interaction.plot (factor (A), factor(B), y, ylim = range (y)) #
points (factor(A), y, pch = B+2)

```

```{r}
detach ("process")
```

# One Quarter Factorial Design and Analysis

## $2^{6-2}$ design with E=ABC and F=BCD

```{r}

des5<- FrF2 (nfactors = 6, nruns=16, generators = c("ABC", "BCD"), 
             alias.info = 3); 
des5[with (des5,order(D,C,B,A)),]
design.info(des5)$aliased


```

```{r}
#Let's look at the full factorial
des6<- FrF2 (nfactors = 6, nruns=64)
model_des6 <- data.frame(model.matrix (~A*B*C*D*E*F, des6))
model_des6[ with(model_des6, order (-B1.C1.D1.F1, -A1.B1.C1.E1,  D1, C1, B1, A1)),
            c("A1","B1", "C1","D1", "A1.B1.C1.E1", "B1.C1.D1.F1")]


```

```{r include=FALSE}
## you may try this package too
library(dplyr)
model_des6 %>% arrange(desc(B1.C1.D1.F1), desc(A1.B1.C1.E1),  D1, C1, B1, A1) %>% select(c("A1","B1", "C1","D1", "A1.B1.C1.E1", "B1.C1.D1.F1")) 
```

## Analysis of a Dataset Collected with $2^{6-2}$ Design

```{r }
# read in the data
shrinkage<- read.table("data/shrinkage.txt", header=T)
knitr::kable(shrinkage, align="c")

attach(shrinkage)

# define the design matrix
A<-rep(c(-1,1),8)
B<-rep(c(-1,-1,1,1),4)
C<-rep(c(rep(-1,4),rep(1,4)),2)
D<-c(rep(-1,8),rep(1,8))
E<-A*B*C
F<-B*C*D

g <- lm(y~(A+B+C+D+E+F)^2+A*B*D+A*B*F,data=shrinkage)
anova(g)
summary(g)
effects<-g$coef[-1]
```



```{r}
# using lenth's method to screen effect
qqnorm.lenth(effects[!is.na(effects)])
```
```{r}
# Fit the model with significant effects
h<-lm(y~(A+B+D)^2,data=shrinkage)
anova(h)
```
```{r fig.height=8}
## look at the fitting results

par (mfrow = c(2,2))

plot (y~factor(A))
plot (y~factor(B))

interaction.plot(factor (A), factor (B), y, ylim = range (y))
points (factor(A), y, pch = B + 2)

interaction.plot(factor (A), factor (D), y, ylim = range (y))
points (factor(A), y, pch = D + 2)

```

```{r}
## residual analysis
qqnorm(h$res)
qqline(h$res)
plot(h$fit,h$res,xlab="fitted value",ylab="residual")
```

```{r fig.height=12}

par(mfrow=c(3,2))
plot(A,h$res,xlab="A",ylab="residual")
plot(B,h$res,xlab="B",ylab="residual")
plot(C,h$res,xlab="C",ylab="residual")
plot(D,h$res,xlab="D",ylab="residual")
plot(E,h$res,xlab="E",ylab="residual")
plot(F,h$res,xlab="F",ylab="residual")



```

```{r}
detach ("shrinkage")
```

