---
title: "Unit 1: Analysis of Simple Comparative Experiments"
date: "Jan 2020"
author: "Longhai Li"
output:
  html_document:
       theme: united
       df_print: kable
       toc: true
       toc_float: false
       number_sections: true
       highlight: tango
       fig_width: 10
       fig_height: 8
editor_options: 
  chunk_output_type: console
---
# Unpaired Design
Example of Comparing Modified and Unmodified Cement 

## Input Data for Analysis

```{r }
mod <- c(16.85, 16.40, 17.21, 16.35, 16.52, 17.04, 16.96, 17.15, 16.59, 16.57)
unmod <- c(17.50, 17.63, 18.25, 18.00, 17.86, 17.75, 18.22, 17.90, 17.96, 18.15)
```

## Draw boxplot 

```{r }
boxplot(mod,unmod)
```

## Perform two sample t-test (with equal variance assumption)

```{r }
t.test(mod,unmod,var.equal=TRUE)
t.test(mod,unmod) #(Welch approximation)
```


## Checking normality assumptions

```{r }
# normality check
qqnorm(mod)
qqline(mod,col=2)
qqnorm(unmod)
qqline(unmod,col=4)
```

## Testing equality of variance

```{r }
var.test(mod,unmod)

# alternatively we can using these two steps
f <- var (mod)/var (unmod);f
pv <- 2* pf (f, df1 = 9, df2=9, lower.tail = F); pv
```

#  Paired Design
Consider an experiment to compare two methods, MSI and SIB, to determine chlorine content in sewage effluents. Eight samples were collected at different doses and contact times. Two methods were applied to each of the eight samples. It is a paired comparison design because the pair of treatments are applied to the same samples (or units).

## Read Data

```{r }
chlorine <- read.table ("data/chlorine.txt", header = T)

chlorine

MSI <- chlorine$MSI
SIB <- chlorine$SIB
Sample <- chlorine$Sample
```

## Drawing boxplot

```{r }
boxplot (MSI, SIB)
```

## Drawing matplot

```{r }
matplot(chlorine[,c("MSI","SIB")],
        type = "b", xlab = "Sample", ylab = "Chlorine")
```

## Drawing difference vs sample

```{r }
plot (SIB - MSI ~ Sample, type = "h")
abline (h = 0)
```

## Testing for the difference in means

```{r }
t.test (MSI, SIB)

t.test (MSI, SIB, paired = T)

t.test (MSI-SIB)

```


---
title: "unit1.R"
author: "longhaili"
date: "Mon Mar  5 22:02:27 2018"
---
