---
title: "[STAT 812: Computational Statistics](../stat812.html)"
subtitle: "Rejection Sampling"
date: "September 2018"
author: "Longhai Li"
output:
  pdf_document:
    toc: yes
    fig_height: 6
    fig_width: 10
    number_sections: yes
  html_document:
       theme: united
       toc: true
       toc_float: true
       number_sections: true
       highlight: tango
       fig_width: 10
       fig_height: 8
---
# Rejection Sampling for Gamma Distribution
## Envolop Function
```{r}
# Note: this is only a toy example demonstrating how to program rejection sampling
# this is not a good sampling scheme for gamma distribution. 
# Do not use it for serious applications that demand high efficiency. 


#log of a function which is always above the Gamma density function  
# alpha must be > 2
log_g_gamma <- function(x, alpha)
{
   (alpha-1) * (log(alpha-1) - 1) - log( 1 + (x-(alpha-1))^2 / (2*alpha-1) )  
}

## look at the approximating functionn
xvec <- seq (0, 3, by = 0.0001)
alpha <- 1.5
log_g_gamma_val <- log_g_gamma(xvec, alpha = alpha)
log_gamma_val <- dgamma (xvec, shape = alpha, log =T)
ylim <- range (log_g_gamma_val, log_gamma_val, finite = T)
plot (xvec, log_g_gamma_val, col = "black", type = "l", ylim = ylim )
points (xvec,log_gamma_val, col = "red",type = "l")

## look at the approximating functionn
xvec <- seq (0, 3, by = 0.0001)
alpha <- 2.1
log_g_gamma_val <- log_g_gamma(xvec, alpha = alpha)
log_gamma_val <- dgamma (xvec, shape = alpha, log =T)
ylim <- range (log_g_gamma_val, log_gamma_val, finite = T)
plot (xvec, log_g_gamma_val, col = "black", type = "l", ylim = ylim )
points (xvec,log_gamma_val, col = "red",type = "l")

## look at the approximating functionn
xvec <- seq (0, 3, by = 0.0001)
alpha <- 4.5
log_g_gamma_val <- log_g_gamma(xvec, alpha = alpha)
log_gamma_val <- dgamma (xvec, shape = alpha, log =T)
ylim <- range (log_g_gamma_val, log_gamma_val, finite = T)
plot (xvec, log_g_gamma_val, col = "black", type = "l", ylim = ylim )
points (xvec,log_gamma_val, col = "red",type = "l")
```

## Rejection Sampling Function
```{r}
#sampling from Gamma distribution with rejection sampling
sample_gamma_rej <- function(n,alpha)
{  sample_gamma <- rep(0,n)
   no.draw <- 0  
   for(i in 1:n)
   {   rejected <- TRUE
       
       while(rejected)
       {  sample_gamma[i] <- rcauchy(1) * sqrt(2*alpha-1) + (alpha -1)
          no.draw <- no.draw + 1
          U <- runif(1)
          rejected <- (log(U) > dgamma(sample_gamma[i],shape=alpha,log=TRUE) - 
	                            log_g_gamma(sample_gamma[i],alpha) )
       }
   }
   attr(sample_gamma, "accept.rate") <- n/no.draw
   sample_gamma
}

```
## Test 1

```{r}
alpha <- 2.1
gammarn <- sample_gamma_rej (2000,alpha); attr (gammarn, "accept.rate")
hist (gammarn)
qqplot(gammarn, rgamma (2000, alpha))
abline (a = 0, b=1)

```

## Test 2

```{r}
alpha <- 2.5
gammarn <- sample_gamma_rej (2000,alpha); attr (gammarn, "accept.rate")
hist (gammarn)
qqplot(gammarn, rgamma (2000, alpha))
abline (a = 0, b=1)

```

## Test 3

```{r}
alpha <- 4.5
gammarn <- sample_gamma_rej (2000,alpha); attr (gammarn, "accept.rate")
hist (gammarn)
qqplot(gammarn, rgamma (2000, alpha))
abline (a = 0, b=1)

```

We see that when alpha is larger than 2, the overall acceptance rate is very low


If you requires highly efficient gamma random numbers generators, read a paper in Computational Statistics and Data Analysis (2007) (http://www.sciencedirect.com/science/article/pii/S0167947306003616)

# Adaptive Rejection Sampling for Truncated Normal

```{r cache=TRUE}
library (ars) 

## a direct rejection sampling for truncated normal
sample_tnorm_drs <- function (n, lb = -Inf, ub = Inf)
{
    x <- rep (0, n)
    for (i in 1:n)
    {
        rej <- TRUE
        while (rej)
        {
            x[i] <- rnorm (1)
            if (x[i] >= lb & x[i] <= ub) rej <- FALSE
        }
    }
    x
}

## sample from truncated normal using ars package
sample_tnorm_ars <- function (n, lb, ub)
{
    logf <- function (x) dnorm (x, log = TRUE) ## define log density
    fprima <- function (x) -x ## define derivative of log density
    
    ars (n, f = logf, fprima = fprima, 
         x = c(lb, (lb + ub )/2, ub), # starting points
         lb = TRUE, ub = TRUE, xlb = lb, xub = ub) # boundary of log density
}
n <- 1000
system.time (
    rn_tnorm_ars <- sample_tnorm_ars (n, -5, -4)
)

system.time(
 rn_tnorm_rej <- sample_tnorm_drs (n, -5, -4)
)

par (mfrow = c(1,3))
hist (rn_tnorm_ars, main = "Adaptive Rejection Sampling")
hist (rn_tnorm_rej, main = "Naive Rejection")
qqplot (rn_tnorm_ars, rn_tnorm_rej)
abline (a = 0, b = 1)

# Draw truncated normal sample on the far tail
n <- 1000
system.time (
    rn_tnorm_ars2 <- sample_tnorm_ars (n, -50, -40)
)
system.time (
    rn_tnorm_ars3 <- sample_tnorm_ars (n, 100, 110)
)

par (mfrow = c(1,2))
hist (rn_tnorm_ars2, main = "Adaptive Rejection Sampling")
hist (rn_tnorm_ars3, main = "Adaptive Rejection Sampling")

```

