---
title: "CURRICULUM VITAE of LONGHAI LI (`r format(Sys.Date(), '%b %d, %Y')`)"
engine: knitr
metadata-files:
  - _website.yaml 
format: 
  html: default
---
```{r}
#| label: setup-ritem
#| include: false
#| cache: false

.current_item_count <- 0

ritem <- function(n = NULL) {
  if (!is.null(n)) {
    .current_item_count <<- n
  }
  
  num <- .current_item_count
  .current_item_count <<- .current_item_count - 1
  
  # Fetch the pandoc target format safely
  out_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
  
  # The \n\n forces Pandoc to start a brand new paragraph for EVERY item
  if (knitr::is_latex_output()) {
    out <- sprintf("\n\n\\hangindent=2.5em \\hangafter=1 \\makebox[2.5em][l]{%d.}", num)
    return(knitr::asis_output(out))
  } else if (knitr::is_html_output()) {
    out <- sprintf("\n\n<span class='ritem-num'>[%d]</span>", num)
    return(knitr::asis_output(out))
  } else if (!is.null(out_format) && out_format == "typst") {
    # Typst output block: 
    # Use a raw inline block so Pandoc doesn't intercept it as a Markdown list. 
    # Typst natively recognizes lines starting with "N. " as explicitly 
    # numbered ordered list items with built-in hanging indents.
    out <- sprintf("\n\n`[%d]`{=typst} ", num)
    return(knitr::asis_output(out))
  } else {
    return(paste0("\n\n[", num, "]"))
  }
}
```

## 1. PERSONAL

* Born in Nov. 1979. 
* Citizenship: Canadian
* Official webpage: [https://artsandscience.usask.ca/profile/LLi](https://artsandscience.usask.ca/profile/LLi) 
* Professional web site: [https://longhaisk.github.io](https://longhaisk.github.io) 
* Phone: +1 (306) 966-6095 
* Email: [longhai.li@usask.ca](mailto:longhai.li@usask.ca)
* Address:
  Department of Mathematics & Statistics
  University of Saskatchewan
  106 Wiggins RD
  Saskatoon, SK, S7W0G8 
  CANADA

## 2. DEGREES

* Ph.D.,  University of Toronto, 2007, Statistics 
  Supervisor: [Radford M. Neal](https://glizen.com/radfordneal/)
  Thesis: Bayesian Classification and Regression with High Dimensional Features
* M.Sc., University of Toronto, 2003, Statistics
* B.Sc.,  University of Science and Technology of China, 2002, Statistics.

## 4. Employment History

* Full Professor, July 1st, 2018, Dept. of Math & Stat., Univ. of Saskatchewan, SK, Canada
* Associate Professor, July 1st, 2012, Dept. of Math & Stat., Univ. of Saskatchewan, SK, Canada
* Assistant Professor, July 1st, 2007, Dept. of Math & Stat., Univ. of Saskatchewan, SK, Canada
* Research Intern, Nov. 2006 to Feb. 2007, Microsoft Research, Redmond, WA, USA
* Sessional Instructor, 2006-2007, University of Toronto, Toronto, ON, Canada

## 7. LEAVES

* Sabbatical Leave, Jan. 1st, 2025 to June 30th, 2025.
* Sabbatical Leave, July 1st, 2020 to June 30th, 2021.
* Parental Leave, Jan. 1st, 2019 to June 30th, 2019.
* Sabbatical Leave, July 1st, 2013 to June 30th, 2014.

## 9. TEACHING ACTIVITIES

### 9.1 Scheduled Instructional Activity

**2025-2026**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 851 | T2 | Linear Statistical Models | LEC | 4 | 39 | 156 |
| STAT 443 | T2 | Linear Models | LEC | 3 | 39 | 117 |
| STAT 442 | T2 | Statistical Inference | LEC | 3 | 39 | 117 |
| STAT 850 | T2 | Mathematical Statistics and Inference | LEC | 1 | 39 | 39 |
| STAT 845 | T1 | Statistical Methods for Research | LEC | 10 | 39 | 390 |
| MATH 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |

**2024-2025**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 348 | T1 | Sampling Techniques | LEC | 15 | 39 | 585 |
| STAT 812 | T1 | Computational Statistics | LEC | 5 | 39 | 195 |
| STAT 420 | T1 | Topics in Computational Statistics | LEC | 1 | 39 | 39 |
| BIOS 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| BIOS 996 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |

**2023-2024**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 245 | T1 | Introduction to Statistical Methods | LEC | 91 | 39 | 3549 |
| STAT 812 | T1 | Computational Statistics | LEC | 5 | 39 | 195 |
| STAT 420 | T1 | Topics in Computational Statistics | LEC | 1 | 39 | 39 |
| STAT 443 | T2 | Linear Statistical Models | LEC | 4 | 39 | 156 |
| STAT 851 | T2 | Linear Models | LEC | 4 | 39 | 156 |
| BIOS 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| BIOS 996 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |

**2022-2023**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 348 | T1 | Sampling Techniques | LEC | 35 | 39 | 1365 |
| STAT 812 | T1 | Computational Statistics | LEC | 3 | 39 | 117 |
| STAT 420 | T1 | Topics in Computational Statistics | LEC | 1 | 39 | 39 |
| STAT 443 | T2 | Linear Statistical Models | LEC | 1 | 39 | 39 |
| STAT 851 | T2 | Linear Models | LEC | 1 | 39 | 39 |
| BIOS 996 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |
| BIOS 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |

**2021-2022**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 244 | T2 | Elementary Statistical Concepts | LEC | 72 | 39 | 2808 |
| STAT 342 | T1 | Mathematical Statistics | LEC | 8 | 39 | 312 |
| STAT 443 | T2 | Linear Statistical Models | LEC | 3 | 39 | 117 |
| STAT 851 | T2 | Linear Models | LEC | 5 | 39 | 195 |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 2 | | |
| MATH 994 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| BIOS 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| ENGR 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |

**2020-2021**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 3 | | |
| BIOS 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| ENGR 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |

**2019-2020**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 245 | T1 | Introduction to Statistical Methods | LEC | 128 | 39 | 4992 |
| STAT 345 | T2 | Design and Analysis of Experiments | LEC | 27 | 39 | 1053 |
| STAT 834 | T2 | Advanced Experimental Design | LEC | 1 | 39 | 39 |
| STAT 812 | T1 | Computational Statistics | LEC | 5 | 39 | 195 |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |
| BIOS 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| ENGR 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |

**2018-2019**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 812 | T1 | Computational Statistics | LEC | 11 | 39 | 429 |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 2 | | |
| BIOS 994 | T1T2 | Research Supervision (M.Sc.) | RES | 2 | | |
| BIOS 994 | T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| MATH 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |

**2017-2018**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 241 | T1 | Probability Theory | LEC | 70 | 39 | 2730 |
| STAT 345 | T2 | Design and Analysis of Experiments | LEC | 25 | 39 | 975 |
| STAT 834 | T2 | Advanced Experimental Design | LEC | 9 | 39 | 351 |
| STAT 841 | T2 | Probability Theory | LEC | 5 | 39 | 195 |
| MATH 994 | T1T2 | M.Sc. Research Supervision | RES | 5 | | |

**2016-2017**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 812 | T1 | Computational Statistics | LEC | 12 | 39 | 468 |
| STAT 348 | T2 | Sampling Techniques | LEC | 33 | 39 | 1287 |
| STAT 442 | T2 | Statistical Inference | LEC | 1 | 39 | 39 |
| STAT 846 | T2 | Sp. Topics (Statistical Inference) | LEC | 8 | 39 | 312 |
| MATH 994 | T1T2 | M.Sc. Research Supervision | RES | 6 | | |

**2015-2016**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 245 | T1 | Intro. to Stat. Methods | LEC | 157 | 39 | 6123 |
| STAT 342 | T1 | Mathematical Statistics | LEC | 7 | 39 | 273 |
| STAT 242 | T2 | Stat. Theory & Methodology | LEC | 11 | 39 | 429 |
| STAT 245 | T2 (2014SS) | Intro to Statistical Methods | LEC | 30 | 39 | 1170 | 
| MATH 994 | T1T2 | M.Sc. Research Supervision | RES | 5 | | |
| MATH 996 | T1T2 | Ph.D. Research Supervision | RES | 1 | | |

**2014-2015**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 342 | T1 | Mathematical Statistics | LEC | 9 | 39 | 351 |
| STAT 812 | T1 | Computational Statistics | LEC | 10 | 39 | 390 |
| STAT 348 | T2 | Sampling Techniques | LEC | 21 | 39 | 819 |
| STAT 442 | T2 | Statistical Inference | LEC | 5 | 39 | 195 |
| STAT 846 | T2 | Sp. Topics (Statistical Inference) | LEC | 8 | 39 | 312 |
| STAT 244 | T2 | Elementary Statistical Concepts | LEC | 14 | 39 | 546 |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 4 | | |
| MATH 996 | T1 | Research Supervision (Ph.D.) | RES | 2 | | |

**2012-2013**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 241 | T1 | Probability Theory | LEC | 51 | 39 | 1989 |
| STAT 348 | T2 | Sampling Techniques | LEC | 15 | 39 | 585 |
| STAT 245 | T2 | Intro to Statistical Methods | LEC | 135 | 39 | 5265 |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 2 | | |
| MATH 996 | T1T2 | Research Supervision (Ph.D.) | RES | 2 | | |

**2011-2012**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 241 | T1 | Probability Theory | LEC | 43 | 39 | 1677 |
| STAT 342 | T1 | Mathematical Statistics | LEC | 3 | 39 | 117 |
| STAT 841 | T1 | Probability Theory | LEC | 10 | 39 | 390 |
| STAT 245 | T2 | Intro to Statistical Methods | LEC | 138 | 39 | 5382 |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 2 | | |

**2010-2011**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 241 | T1 | Probability Theory | LEC | 42 | 39 | 1638 |
| STAT 846 | T1 | Computational Statistics | LEC | 8 | 39 | 195 |
| STAT 242 | T2 | Stat. Theory & Methodology | LEC | 13 | 39 | 507 |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 2 | | |
| MATH 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |

**2009-2010**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 342 | T1 | Mathematical Statistics | LEC | 4 | 39 | 156 |
| STAT 841 | T1 | Probability Theory | LEC | 6 | 39 | 234 |
| STAT 241 | T2 | Probability Theory | LEC | 28 | 39 | 1092 |
| STAT 244 | T2 | Elem. Stat. Concepts | LEC | 83 | 39 | 3237 |
| STAT 848 | T2 | Multivariate Data Analysis | READ | 2 | 39 | 78 |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 2 | | |
| MATH 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |

**2008-2009**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 342 | T1 | Mathematical Statistics | LEC | 19 | 39 | 741 |
| STAT 848 | T2 | Multivariate Data Analysis | LEC | 6 | 39 | 234 |

**2007-2008**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|---|--|----------|--|--|--|--|
| STAT 342 | T1 | Mathematical Statistics | LEC | 7 | 39 | 273 |
| STAT 846 | T2 | Computational Statistics | LEC | 5 | 39 | 195 |

### 9.2 Unscheduled Instructional Activity

* Creating and assessing PhD qualifying exam on Mathematical Statistics, July 2021.
* Creating and assessing PhD qualifying exam on Mathematical Statistics, May 2021.

### 9.3 Course and Program Development

**2022-2023**

`r ritem(3)` Renewal of the University of Saskatchewan Courses for SSC Accreditation.
`r ritem()` Participated in creating new M.Sc. and Ph.D. programs in Statistics, approved on May 18, 2023.

**2020-2021**

`r ritem()` Participation in creating Certificate in Statistical Methods, 2021.

### 9.4 Teaching Materials

**2025-2026**

`r ritem(4)` STAT 442/850: a textbook entitled "Theory of Statistical Inference" in PDF ($\ge$ 200 pages) and HTML.
`r ritem()`  STAT 845: a textbook entitled "Statistical Inference and Learning Methods for Research" in PDF ($\ge$ 200 pages) and HTML
`r ritem()`  STAT 443/851: a textbook entitled "Theory of Linear Model" in PDF ($\ge$ 200 pages) and HTML.

**2024-2025**

`r ritem()` STAT 443/851: 273 pages of handwritten lecture notes were developed and posted to students.

### 9.5 Other Teaching-Related Activities

**2025-2026**

`r ritem(10)` Peer evaluation for Matthew Schmirler, Mar. 2026.

**2023-2024**

`r ritem()` Data Science Bootcamp, a case study, June 19, 2023.

**2022-2023**

`r ritem()` Peer evaluation for Raj Srinivasan, April 2023.
`r ritem()` Peer evaluation for Saima Khosa, Dec. 2022.

**2021-2022**

`r ritem()` Peer Teaching Evaluation for Shahedul Khan, Nov. 2021.
`r ritem()` Peer Teaching Evaluation for Li Xing, Nov. 2021.

**2020-2021**

`r ritem()` Peer Teaching Evaluation for Shahedul Khan, STAT 812, Nov. 12, 2020.

**2019-2020**

`r ritem()` Peer Teaching Evaluation for Annalizer McGillvray, Nov. 29, 2019.

**2017-2018**

`r ritem()` Peer Teaching Evaluation for Shahedul Khan, 2017-2018.
`r ritem()` Peer Teaching Evaluation for Lawrence Chang, 2017-2018.

## 10. SUPERVISION AND ADVISORY ACTIVITIES

### 10.1 Undergraduate Student Supervision

**2025-2026**

`r ritem(16)` George Chen, B.Sc., Computer Science, Simon Fraser University, June 1, 2025 -- Aug. 31, 2025, Supervised.
`r ritem()` Shruti Kaur, B.Sc., Computer Science and Statistics, May -- Aug., 2025, Supervised.

**2023-2024**

`r ritem()` George Chen, B.Sc., Computer Science, Simon Fraser University, May 7, 2023 -- Aug. 25, 2023, Supervised.

**2022-2023**

`r ritem()` Noah Little, B.Sc., Anatomy and Cell Biology, April 2022 -- Aug. 31, 2022, Supervised.

**2020-2021**

`r ritem()` Noah Little, B.Sc., Anatomy and Cell Biology, Aug. 2020 -- Feb. 2021, Supervised.
`r ritem()` Lifang Lei, B.Sc., Mechanical Engineering, Oct. 2020 -- March 2021, Supervised.
`r ritem()` Yanping Li, B.Sc., Business, Edward Business School, Feb. 2021 -- June 2021, Supervised.
`r ritem()` Lina Li, B.Sc., Statistics, May 2020 -- June 2020, Summer Research Assistant, Supervised.

**2019-2020**

`r ritem()` Hao Hu, B.Sc., Statistics, 2019--2020, Undergraduate Thesis Supervision, Supervised.
`r ritem()` Steven Liu, B.Sc., Computer Science, July 2019 -- Oct. 2020, Summer Research Assistant, Supervised.
`r ritem()` Steven Liu, B.Sc., Computer Science, June 17, 2019 -- Aug. 31, 2019, Supervised.

**2018-2019**

`r ritem()` Jian Su, B.Sc., Computer Science, May 1, 2018 -- Aug. 31, 2018, Supervised.

**2016-2017**

`r ritem()` Jiaqi Xiao, B.Sc., Economics, Research Assistant, May 2016 -- Aug. 2016, Supervised.

**2015-2016**

`r ritem()` Jiaqi Xiao, B.Sc., Economics, May 2015 -- Aug. 2015, Supervised.

**2014-2015**

`r ritem()` Zhouji Zhang, B.Sc., Mathematics, June 1, 2014 -- June 30, 2014, Supervised.

**2013-2014**

`r ritem()` Bei Zhang, B.Sc., Statistics, May 1, 2013 -- Aug. 31, 2013, Supervised.

### 10.2 Graduate Student Supervision

`r ritem(26)` Jing Wang, Ph.D., Biostatistics, School of Public Health, Co-supervised with Prof. Li Xing, 2023--2026 (Transferred to other supervisor)
`r ritem()` Dananji Egodage, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Cindy Feng, 2023--2025 (Defended: Aug. 30, 2025)
`r ritem()` Wuqian Effie Gao, M.Sc., Biostatistics, School of Public Health, Co-supervised with Prof. Cindy Feng, 2022--2024 (Defended: Aug. 30, 2024)
`r ritem()` Lina Li, M.Sc., Statistics, Math & Stat, MITACS Project supervisor, Supervised, 2022--2022 (Completed: August 2022)
`r ritem()` Hao Hu, Ph.D., Statistics, Math & Stat, Co-supervised with Prof. Li Xing, 2021--2022 (Transferred to M.Sc. July 2022)
`r ritem()` Hao Hu, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Li Xing, 2021--2022 (Defended: Sept. 15, 2022)
`r ritem()` Man Chen, M.Sc., Statistics, Math & Stat, Supervised, 2019--2021 (Defended: April 30, 2021)
`r ritem()` Wutao Yin, Ph.D., Biomedical Engineering, Co-supervised with Prof. FangXiang Wu, 2019--2021 (Defended: Dec. 17, 2021)
`r ritem()` Wutao Yin, Ph.D., Statistics, Math & Stats, Supervised, 2018--2019 (Transferred to Engineering May 2019)
`r ritem()` Tingxuan Wu, Ph.D., Biostatistics, School of Public Health, Co-supervised with Prof. Cindy Feng, 2018--2023 (Defended: May 24, 2023)
`r ritem()` Tingxuan Wu, M.Sc., Biostatistics, School of Public Health, Co-supervised with Prof. Cindy Feng, 2017--2018 (Defended: Dec. 4, 2018)
`r ritem()` Mei Dong, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Lloyd Balbuena, 2017--2019 (Defended: May 23, 2019)
`r ritem()` Alireza Sadeghpour, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Cindy Feng, 2016--2017 (Defended: Sept. 19, 2017)
`r ritem()` Xiaoying Wang, M.Sc., Statistics, Math & Stat, Supervised, 2016--2019 (Defended: March 12, 2019)
`r ritem()` Wei Bai, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Cindy Feng, 2016--2018 (Defended: July 12, 2018)
`r ritem()` Yunyang Wang, M.Sc., Statistics, Math & Stat, Supervised, 2014--2017 (Defended: Nov. 18, 2017)
`r ritem()` Naorin Islam, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Shahedul Khan, 2014--2017 (Defended: Nov. 28, 2017)
`r ritem()` Arash Shamloo, M.MATH., Statistics, Math & Stat, Supervised, 2016--2017 (Project Completed: August 31, 2017)
`r ritem()` Arash Shamloo, M.Sc., Biostatistics, School of Public Health, Supervised, 2014--2016 (Defended: 2016)
`r ritem()` Setu Chandra Kar, M.Sc., Statistics, Math & Stat, Supervised, 2014--2016 (Defended: 2016)
`r ritem()` Shi Qiu, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Cindy Feng, 2012--2015 (Defended: March 26, 2015)
`r ritem()` Masud Rana, M.Sc., Statistics, Math & Stats, Co-supervised with Prof. Shahedul Khan, 2010--2012 (Defended: Sept. 2012)
`r ritem()` Saima Khan Khosa, Ph.D., Statistics, Math & Stats, Supervised, 2010--2012 (Transferred to other supervisor)
`r ritem()` Lai Jiang, Ph.D., Statistics, Math & Stat, Supervised, 2009--2015 (Defended: Sept. 14, 2015)
`r ritem()` Lai Jiang, M.Sc., Statistics, Math & Stats, Supervised, 2008--2009 (Transferred to Ph.D.; M.Sc. supervision ended)
`r ritem()` Zhengrong Li, M.Sc., Statistics, Math & Stats, Supervised, 2007--2012 (Defended: May 2012)

### 10.3 Graduate Theses Supervised

**2025-2026**

`r ritem(20)` Dananji Egodage, 2025, Component-wise Z-residuals for Checking Bayesian Hurdle Models, M.Sc., defended on Aug. 30, 2025.

**2024-2025**

`r ritem()` Effie Wuqian Gao, 2024, Z-residuals for Checking Bayesian Hurdle Models, M.Sc., defended on Aug. 30, 2024.

**2023-2024**

`r ritem()` Tingxuan Wu, 2023, Residual Diagnostics and Statistical Inference for Shared Frailty Models, Ph.D., defended on May 24, 2023.

**2022-2023**

`r ritem()` Hao Hu, 2022, Identifying Risk Factors for Cognitive Decline using Statistical Learning Techniques and Functional Data Analysis, M.Sc., defended on Sept. 15, 2022.

**2021-2022**

`r ritem()` Wutao Yin, 2021, Artificial Intelligence Based Methods for Autism Spectrum Disorder Diagnosis from fMRI Data, Ph.D., defended on Dec. 17, 2021.
`r ritem()` Man Chen, 2021, Association Between Gut Microbiome and Parkinson's Disease Revealed by Sparse Learning, M.Sc., defended on April 30, 2021.

**2019-2020**

`r ritem()` Dong Mei, 2019, Feature Selection Bias in Assessing the Predictivity of SNPs for Alzheimer's Disease, M.Sc., defended on May 23, 2019.
`r ritem()` Xiaoying Wang, 2019, Comparison of Statistical Testing and Predictive Analysis Methods for Feature Selection in Zero-inflated Microbiome Data, M.Sc., defended on March 12, 2019.

**2018-2019**

`r ritem()` Tingxuan Wu, 2018, Randomized Survival Probability Residuals for Assessing Parametric Survival Models, M.Sc., defended on Dec. 4, 2018.
`r ritem()` Wei Bai, 2018, Randomized Quantile Residual for Assessing Generalized Linear Mixed Models with Application to Zero-Inflated Microbiome Data, M.Sc., defended on July 12, 2018.

**2016-2017**

`r ritem()` Yunyang Wang, 2016, Comparison of Stochastic Volatility Models Using Integrated Information Criteria, M.Sc., defended on Nov. 18, 2016.
`r ritem()` Alireza Sadeghpour, 2016, Empirical Investigation of Randomized Quantile Residuals for Diagnosis of Non-Normal Regression Models, M.Sc., defended on Sept. 19, 2016.
`r ritem()` Naorin Islam, 2016, Substance Abuse and Health: A Structural Equation Modeling Approach to Assess Latent Health Effects, M.Sc., defended on Nov. 28, 2016.

**2015-2016**

`r ritem()` Lai Jiang, 2015, Fully Bayesian T-probit Regression with Heavy-tailed Priors for Selection in High-Dimensional Features with Grouping Structure, Ph.D., defended on Sept. 14, 2015.
   
**2014-2015**

`r ritem()` Shi Qiu, 2015, Cross-validatory Model Comparison and Divergent Regions Detection using iIS and iWAIC for Disease Mapping, M.Sc., defended on March 26, 2015.

**2012-2013**

`r ritem()` Masud Rana, 2012, Spatial-Longitudinal Bent-Cable Model with An Application to Atmospheric CFC Data, M.Sc., defended in Sept. 2012.
`r ritem()` Zhengrong Li, 2012, A Non-MCMC Procedure for Fitting Dirichlet Process Mixture Models, M.Sc., defended in May 2012.

### 10.4 Supervision of Post-Doctoral Fellows and Research Associates

* Tingxuan Wu, University of Saskatchewan, Postdoc Research Associate,  Jan. 2025 – Dec. 2026.
* Tingxuan Wu, University of Saskatchewan, Postdoc Fellow, June 2023 – April 2024.
* Ming Ming Zhang, MITACS Postdoc, 2022-2025.
* Jinhong Shi, team-supervised for CFREF projects, Sept. 2016 - Aug. 2019.

### 10.5 Staff Supervision

* Saima Khosa, faculty mentoring, Sept. 2022- Dec. 2022.

### 10.6 Thesis Committee Memberships
```{r}
#| label: tbl-thesis-committee
#| tbl-cap: "Thesis Committee Memberships"
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(gt)

# Create the dataframe using tribble for easy row-by-row data entry
committee_data <- tribble(
  ~`#`, ~NAME, ~DEG, ~PROGRAM, ~`TIME FRAME`, ~ROLE,
  31, "Tiansui Wu", "M.Sc.", "Biostatistics", "2025--Present", "Chair",
  30, "Prabhawi Kahatapitiye", "Ph.D.", "Statistics", "2025--Present", "Member",
  29, "Rasel Kabir", "Ph.D.", "Biostatistics", "2025--Present", "Chair",
  28, "Mohammad Toranjsimin", "M.Sc.", "Biostatistics", "2023--2025 (Def. 09/2025)", "Member",
  27, "Lina Li", "Ph.D.", "Biostatistics", "2023--Present", "Member",
  26, "Kyle Gardiner", "M.Sc.", "Statistics", "2023--2024 (Def. 09/2024)", "Member",
  25, "Mangladeep Bhullar", "Ph.D.", "Physics", "2023--2025 (Def. 11/2025)", "Cognate",
  24, "Hammed Jimoh", "M.Sc.", "Statistics", "2022--2023", "Member",
  23, "Qi Zhang", "M.Sc.", "Sociology", "2021--2022", "External",
  22, "Han Wang", "Ph.D.", "Sociology", "2020--Present", "Cognate",
  21, "Yanzhao Cheng", "Ph.D.", "Biostatistics", "2020--2022 (Def. 10/2021)", "Chair",
  20, "Naeima Ashleik", "Ph.D.", "Statistics", "2017--2018 (Def. 03/2018)", "Member",
  19, "Mehdi Rostami", "M.Sc.", "Biostatistics", "2014--2016 (Def. 06/2016)", "Member",
  18, "Sanjeev Rijal", "M.Sc.", "Statistics", "2014--2018 (Def. 07/2017)", "Member",
  17, "Farhad Maleki", "Ph.D.", "Bioinformatics", "2014--2019", "Cognate",
  16, "Saima Khan Khosa", "Ph.D.", "Statistics", "2014--2017", "Member",
  15, "Yue Dong", "M.Sc.", "Statistics", "2014--2016 (Def. 06/2016)", "Member",
  14, "Temitope Adesina", "M.Sc.", "Biostatistics", "2014--2015", "Chair",
  13, "Sudhakar Achath", "M.Sc.", "Statistics", "2014--2017 (Def. 05/2017)", "Member",
  12, "Xiaolei Yu", "Ph.D.", "Geography", "2012--2022 (Def. 06/2022)", "Cognate",
  11, "Matthew Schmirler", "Ph.D.", "Statistics", "2012--2022 (Def. 07/2022)", "Member",
  10, "Masha Naseri", "Ph.D.", "Computer Sci.", "2012--2014 (Def. 02/2014)", "Cognate",
  9, "Chel Hee Lee", "Ph.D.", "Statistics", "2012--2013", "Member",
  8, "Weiwei Fan", "M.Sc.", "Bioinformatics", "2012--2014 (Def. 01/2014)", "Member",
  7, "Zhaoqin Li", "Ph.D.", "Geography", "2011--2017 (Def. 04/2017)", "Cognate",
  6, "Courtney Kendall", "M.Sc.", "Statistics", "2011--2014 (Def. 08/2014)", "Member",
  5, "Michael Janzen", "Ph.D.", "Computer Sci.", "2011--2012 (Def. 03/2012)", "Cognate",
  4, "Matthew Schmirler", "M.Sc.", "Statistics", "2010--2013 (Def. 09/2012)", "Member",
  3, "Mohammed Obeidat", "Ph.D.", "Statistics", "2010--2014 (Def. 07/2014)", "Member",
  2, "Lingling Jin", "Ph.D.", "Bioinformatics", "2010--2018 (Def. 08/2017)", "Cognate",
  1, "Tolulope Sajobi", "Ph.D.", "Biostatistics", "2008--2012 (Def. 03/2012)", "Member"
)

# Render the gt table
committee_data |> 
  gt() |> 
  tab_options(
    table.width = pct(100),
    
    table.layout = "fixed",
    table.font.size = px(11), 
    container.overflow.x = TRUE,
    # --- remove all lines ---
    table.border.top.style = "hidden",
    table.border.bottom.style = "hidden",
    column_labels.border.top.style = "hidden",
    column_labels.border.bottom.style = "hidden",
    table_body.hlines.style = "none",
    table_body.vlines.style = "none",
    table_body.border.top.style = "hidden",
    table_body.border.bottom.style = "hidden",
    stub.border.style = "hidden"
  ) |> 
  cols_width(
    `#` ~ pct(5),
    NAME ~ pct(25),
    DEG ~ pct(10),
    PROGRAM ~ pct(20),
    `TIME FRAME` ~ pct(30),
    ROLE ~ pct(10)
  ) |> 
  opt_row_striping() |> 
  # --- gray header background + bold ---
  tab_style(
    style = list(
      cell_fill(color = "#d9d9d9"),
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  ) |> 
  cols_align(
    align = "center",
    columns = c(`#`, DEG, ROLE)
  )
```

## 11. BOOKS AND CHAPTERS IN BOOKS

### 11.1 Authored Books

`r ritem(2)` Soltanifar, M., Li, L., and Rosenthal, J., 2010. A Collection of Exercises in Advanced Probability Theory - the solutions manual of all even-numbered exercises from “A First Look at Rigorous Probability Theory”, World Scientific Publishing, (Second Edition, 2006), Singapore.
`r ritem()` Li, L., 2007. Bayesian Classification and Regression with High Dimensional Features. (Ph.D. Thesis), Toronto: University of Toronto.

### 11.3 Chapters in Books

* Feng, C. X. and  Li, L ., 2016. Modeling Zero Inflation and Overdispersion in the Length of Hospital Stay for Patients with Ischaemic Heart Disease, in the book Advanced Statistical Methods in Big-Data Sciences, edited by D. Chen, J. Chen, X. Lu, G. Yi and H. Yu, Springer, Chapter 3, pp. 35-53.

## 12. PAPERS IN REFERRED JOURNALS

**2025-2026**

`r ritem(33)` Wu, T., Gao, WE, Feng, C., and Li, L., Z-residuals for Diagnosing Bayesian Models, *Journal of American Statistical Association*, under revision. 
`r ritem()` Wu, T., Li, L., and Feng, C., $Z$-residuals Diagnostics for Cox Proportional Hazards Models: Distinguishing Functional Form Misspecification from Nonproportional Hazards, with an Application to Biliary Cirrhosis Survival Times, *Canadian Journal of Statistics*, Accepted on June 5, 2026.
`r ritem()` Nolan, J., Su, C., Li, L., 2025. Evaluating Railroad Duopoly Behavior: A Market Level Analysis. *Review of Network Economics* 24, 87–111. [https://doi.org/10.1515/rne-2025-0034](https://doi.org/10.1515/rne-2025-0034)
                   
**2024-2025**

`r ritem()` Wu, T., Feng, C., Li, L., 2025. Cross-validatory Z-Residual for Diagnosing Shared Frailty Models. *The American Statistician*, 79(2), 198–211. [https://doi.org/10.1080/00031305.2024.2421370](https://doi.org/10.1080/00031305.2024.2421370)
`r ritem()` Wu, T., Li, L., Feng, C., 2025. Z-residual diagnostic tool for assessing covariate functional form in shared frailty models. *Journal of Applied Statistics*, 52(1), 28–58. [https://doi.org/10.1080/02664763.2024.2355551](https://doi.org/10.1080/02664763.2024.2355551)
`r ritem()` Wu, T., Feng, C., Li, L., 2024+. A Comparison of Estimation Methods for Shared Gamma Frailty Models. *Statistics in Biosciences*, accepted 24 June 2024. [https://doi.org/10.1007/s12561-024-09444-7](https://doi.org/10.1007/s12561-024-09444-7)
  
**2023-2024**

`r ritem()` Feng, C., Li, L., Xu, C., 2023. Advancements in predicting and modeling rare event outcomes for enhanced decision-making. *BMC Medical Research Methodology* 23, Article 243 (pp. 1-3). [https://doi.org/10.1186/s12874-023-02060-x](https://doi.org/10.1186/s12874-023-02060-x)

**2021-2022**

`r ritem()` Yin, W., Li, L., Wu, F.-X., 2022. A semi-supervised autoencoder for autism disease diagnosis. *Neurocomputing*, 483, 140–147. [https://doi.org/10.1016/j.neucom.2022.02.017](https://doi.org/10.1016/j.neucom.2022.02.017)
`r ritem()` Cheng, H., Wang, W., Li, L., 2022. Determinants of Citizen Acceptance of White-Collar Crime in China. *Journal of Asian and African Studies*, 59(3), 826-843. [https://doi.org/10.1177/00219096221123742](https://doi.org/10.1177/00219096221123742) (Published OnlineFirst; final pagination may vary.)
`r ritem()` Yin, W., Li, L., Wu, F.-X., 2022. Corrigendum to “Deep learning for brain disorder diagnosis based on fMRI images, Neurocomputing 469 (2022) 332–345”. *Neurocomputing* 509, 271. [https://doi.org/10.1016/j.neucom.2022.08.074](https://doi.org/10.1016/j.neucom.2022.08.074)
`r ritem()` Yin, W., Li, L., Wu, F.-X., 2022. Deep learning for brain disorder diagnosis based on fMRI images. *Neurocomputing* 469, 332–345. [https://doi.org/10.1016/j.neucom.2020.05.113](https://doi.org/10.1016/j.neucom.2020.05.113)

**2020-2021**

`r ritem()` Bai, W., Dong, M., Li, L., Feng, C., Xu, W., 2021. Randomized quantile residuals for diagnosing zero-inflated generalized linear mixed models with applications to microbiome count data. *BMC Bioinformatics* 22, Article 564 (pp. 1-17). [https://doi.org/10.1186/s12859-021-04371-6](https://doi.org/10.1186/s12859-021-04371-6)
`r ritem()` Li, L., Wu, T., Feng, C., 2021. Model Diagnostics for Censored Regression via Randomized Survival Probabilities. *Statistics in Medicine* 40(6), 1482–1497. [https://doi.org/10.1002/sim.8852](https://doi.org/10.1002/sim.8852)
`r ritem()` Dagasso, G., Yan, Y., Wang, L., Li, L., Kutcher, R., Zhang, W., Jin, L., 2021. Leveraging Machine Learning to Advance Genome-Wide Association Studies. *International Journal of Data Mining and Bioinformatics*, 25(1/2), 17–36. [https://doi.org/10.1504/ijdmb.2021.116881](https://doi.org/10.1504/ijdmb.2021.116881)

**2019-2020**

`r ritem()` Dong, M., Li, L., Chen, M., Kusalik, A., Xu, W., 2020. Predictive analysis methods for human microbiome data with application to Parkinson’s disease. *PLOS ONE* 15(8), e0237779 (pp. 1-20). [https://doi.org/10.1371/journal.pone.0237779](https://doi.org/10.1371/journal.pone.0237779)
`r ritem()` Feng, C., Li, L., Sadeghpour, A., 2020. A comparison of residual diagnosis tools for diagnosing regression models for count data. *BMC Medical Research Methodology* 20, Article 175 (pp. 1-11). [https://doi.org/10.1186/s12874-020-01055-2](https://doi.org/10.1186/s12874-020-01055-2)
`r ritem()` Jiang, L., Greenwood, C.M.T., Yao, W., Li, L., 2020. Bayesian Hyper-LASSO Classification for Feature Selection with Application to Endometrial Cancer RNA-seq Data. *Scientific Reports* 10, Article 9747 (pp. 1-12). [https://doi.org/10.1038/s41598-020-66466-z](https://doi.org/10.1038/s41598-020-66466-z)

**2018-2019**

`r ritem()` Shi, J., Yan, Y., Links, M.G., Li, L., Dillon, J.-A.R., Horsch, M., Kusalik, A., 2019. Antimicrobial resistance genetic factor identification from whole-genome sequence data using deep feature selection. *BMC Bioinformatics* 20, Article 535 (pp. 1-14). [https://doi.org/10.1186/s12859-019-3054-4](https://doi.org/10.1186/s12859-019-3054-4)

**2017-2018**

`r ritem()` Essien, S. K., Feng, C., Sun, W., Farag, M., Li, L., Gao, Y., 2018. Sleep duration and sleep disturbances in association with falls among the middle-aged and older adults in China: a population-based nationwide study. *BMC Geriatrics* 18, Article 196 (pp. 1-9). [https://doi.org/10.1186/s12877-018-0889-x](https://doi.org/10.1186/s12877-018-0889-x)
`r ritem()` Li, L., Yao, W., 2018. Fully Bayesian Logistic Regression with Hyper-Lasso Priors for High-dimensional Feature Selection. *Journal of Statistical Computation and Simulation*, 88(14), 2827–2851. [https://doi.org/10.1080/00949655.2018.1490418](https://doi.org/10.1080/00949655.2018.1490418)

**2016-2017**

`r ritem()` Jin, L., McQuillan, I., Li, L., 2017. Computational Identification of Harmful Mutation Regions to the Activity of Transposable Elements. *BMC Genomics* 18, Article 862 (pp. 1-10). [https://doi.org/10.1186/s12864-017-4227-z](https://doi.org/10.1186/s12864-017-4227-z)
`r ritem()` Li, L., Feng, C.X., Qiu, S., 2017. Estimating Cross-validatory Predictive P-values with Integrated Importance Sampling for Disease Mapping Models. *Statistics in Medicine*, 36(14), 2220–2236. [https://doi.org/10.1002/sim.7278](https://doi.org/10.1002/sim.7278)
`r ritem()` Feng, C. X., Rostami, M., Li, L., 2017. Impact of Misspecified Residual Correlation Structure on the Parameter Estimates in a Shared Spatial Frailty Model. *Journal of Statistical Computation and Simulation*, 87(12), 2384–2410. [https://doi.org/10.1080/00949655.2017.1332196](https://doi.org/10.1080/00949655.2017.1332196)

**2015-2016**

`r ritem()` Li, L., Qiu, S., Zhang, B., Feng, C.X., 2016. Approximating Cross-validatory Predictive Evaluation in Bayesian Latent Variables Models with Integrated IS and WAIC. *Statistics and Computing*, 26(4), 881–897. [https://doi.org/10.1007/s11222-015-9577-2](https://doi.org/10.1007/s11222-015-9577-2)

**2013-2014**

`r ritem()` Yao, W., Li, L., 2014. A New Regression Model: Modal Linear Regression. *Scandinavian Journal of Statistics*, 41(3), 656–671. [https://doi.org/10.1111/sjos.12054](https://doi.org/10.1111/sjos.12054)
`r ritem()` Yao, W., Li, L., 2014. Bayesian Mixture Labeling by Minimizing Deviance of Classification Probabilities to Reference Labels. *Journal of Statistical Computation and Simulation*, 84(2), 310–323.

**2011-2012**

`r ritem()` Khan, S. A., Rana, M., Li, L., Dubin, J. A., 2012. A Comparative Case Study to Monitor and Understand Atmospheric CFC Decline with the Spatial-Longitudinal Bent-Cable Model. *International Journal of Statistics and Probability*, 1(2), 56–68.
`r ritem()` Li, L., 2012. Bias-corrected Hierarchical Bayesian Classification with a Selected Subset of High-dimensional Features. *Journal of American Statistical Association*, 107(497), 120–134. [https://doi.org/10.1198/JASA.2011.AP10446](https://doi.org/10.1198/JASA.2011.AP10446)
`r ritem()` Sajobi, T.T., Lix, L. M., Dansu, B. M., Laverty, W., Li, L., 2012. Robust Descriptive Discriminant Analysis for Repeated Measures Data. *Computational Statistics & Data Analysis*, 56(9), 2782–2794. [https://doi.org/10.1016/j.csda.2012.02.029](https://doi.org/10.1016/j.csda.2012.02.029)

**2010-2011**

`r ritem()` Sajobi, T. T., Lix, L., Li, L., Laverty, W., 2011. Discriminant Analysis for Repeated Measures Data: Effects of Mean and Covariance Misspecification on Bias and Error in Discriminant Function Coefficients. *Journal of Modern Applied Statistical Methods*, 10(2), 571–582. [https://doi.org/10.22237/jmasm/1320120840](https://doi.org/10.22237/jmasm/1320120840)

**2009-2010**

`r ritem()` Li, L., 2010. Are Bayesian Inferences Weak for Wasserman’s Example? *Communications in Statistics – Simulation and Computation*, 39(4), 655–667. [https://doi.org/10.1080/03610910903576540](https://doi.org/10.1080/03610910903576540)

**2007-2008**

`r ritem()` Li, L., Zhang, J., Neal, R.M., 2008. A method for avoiding bias from features selection with application to naive Bayes classification models. *Bayesian Analysis*, 3(1), 171–196. [https://doi.org/10.1214/08-BA307](https://doi.org/10.1214/08-BA307)
`r ritem()` Li, L., Neal, R.M., 2008. Compressing Parameters in Bayesian High-order Models with Application to Logistic Sequence Models. *Bayesian Analysis*, 3(4), 793–822. [https://doi.org/10.1214/08-BA330](https://doi.org/10.1214/08-BA330)

## 13. REFERRED CONFERENCE PUBLICATIONS

`r ritem(3)` Yin, W., Li, L., Wu, F.-X., 2021. A Graph Attention Neural Network for Diagnosing ASD with fMRI Data, in: 2021 IEEE International Conference on Bioinformatics and Biomedicine (BIBM). pp. 1131–1136. [https://doi.org/10.1109/BIBM52615.2021.9669849](https://doi.org/10.1109/BIBM52615.2021.9669849)
`r ritem()` Dagasso, G., Yan, Y., Wang, L., Li, L., Kutcher, R., Zhang, W., Jin, L., 2020. Comprehensive-GWAS: a pipeline for genome-wide association studies utilizing cross-validation to assess the predictivity of genetic variations, in: 2020 IEEE International Conference on Bioinformatics and Biomedicine (BIBM). Presented at the 2020 IEEE International Conference on Bioinformatics and Biomedicine (BIBM), pp. 1361–1367. [https://doi.org/10.1109/BIBM49941.2020.9313355](https://doi.org/10.1109/BIBM49941.2020.9313355)
`r ritem()` Jin, L., McQuillan, I., and Li, L., 2016, Computational Identification of Regions that Influence Activity of Transposable Elements in the Human Genome. Proceeding of 2016 IEEE International Conference on Bioinformatics and Biomedicine, pp. 592-599.

## 14. PRESENTATIONS

### 14.1 Invited Presentations

**2025-2026**

`r ritem(33)` Z-residuals for Checking Bayesian Models. Presented at: University of Calgary, Calgary, AB, Canada; July 28, 2025
`r ritem()` Sparse Learning for Assessing the Association Between Gut Microbiome and Parkinson’s Disease. Presented at: The 3rd JCSDS, Hangzhou, China, July 13, 2025.

**2024-2025**

`r ritem()` Z-residuals for Checking Bayesian Models. Presented at: International Conference on Statistics and Data Science, Vancouver, BC, Canada; June 24, 2025
`r ritem()` Z-residuals for Checking Bayesian Hurdle Models. Presented at: EcoStat 2024; July 17, 2024; Beijing, China.

**2023-2024**

`r ritem()` Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Proportional Hazards Models with Shared Frailty, ICSA Canada Chapter Symp., June 9, 2024, Niagara Falls, Canada
`r ritem()` Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Proportional Hazards Models with Shared Frailty, Annual Meeting of SSC, St John’s, Canada, June 2, 2024
`r ritem()` Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Proportional Hazards Models with Shared Frailty, Dept. Seminar, Texas State University, USA, March 8, 2024
`r ritem()` Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Proportional Hazards Models with Shared Frailty, Dept. Seminar, Sun Yat-sen University, China, Jan. 4, 2024

**2022-2023**

`r ritem()` Cross-validatory Residual Diagnostics for Bayesian Spatial Models, Annual Meeting of SSC, Ottawa, May 29, 2023
`r ritem()` Model Diagnostics for Censored Regression via Randomized Survival Probabilities, the 5th ICSA Canada Symposium, 9 July 2022, Banff, AB, Canada
`r ritem()` Model Diagnostics for Censored Regression via Randomized Survival Probabilities, 17 Aug. 2022, Statistics Conference in Genomics, Pharmaceutical Science, and Health Data Science, University of Victoria, Victoria, BC, Canada

**2021-2022**

`r ritem()` Randomized quantile residuals for diagnosing zero-inflated generalized linear mixed models with applications to microbiome count data, SSC Annual Meeting (virtual), May 2022.
`r ritem()` Model Diagnostics for Censored Regression via Randomized Survival Probabilities, The 6th Canadian Conference in Applied Statistics, Hosted by Concordia University (virtual), 16 July 2021.

**2019-2020**

`r ritem()` Estimating Cross-validatory Predictive P-values with Integrated Importance Sampling for Disease Mapping Models, Aug. 2019, the 4th ICSA-Canada Symposium held at Queen’s University.

**2018-2019**

`r ritem()` Feature Selection Bias in Assessing the Predictivity of SNPs for Alzheimer's Disease, June 2019, Seminar talk, University of Manitoba, Canada

**2017-2018**

`r ritem()` Randomized Quantile Residuals for Checking GLMM with Application to Zero-inflated Microbiome Data, June 2018, Annual Meeting of Statistical Society of Canada, McGill University, Canada.
`r ritem()` Fully Bayesian Classification with Heavy-tailed Priors for Selection in High-Dimensional Features with Grouping Structure, Aug. 2017, the 3rd ICSA-Canada Symposium held at Vancouver.

**2016-2017**

`r ritem()` Randomized Quantile Residuals: an Omnibus Model Diagnostic Tool with Unified Reference Distribution, June 2017, Seminar talk, School of Mathematical Sciences, Xiamen University, China.
`r ritem()` Fully Bayesian Classification with Heavy-tailed Priors for Selection in High-Dimensional Features with Grouping Structure, June 2017, Seminar talk, School of Mathematical Sciences, Xiamen University, China.
`r ritem()` Randomized Quantile Residuals: an Omnibus Model Diagnostic Tool with Unified Reference Distribution, June 2017, Seminar talk, Department of Biostatistics, Southern Medical University, Guangzhou, China.
`r ritem()` Estimating Cross-validatory Predictive P-values with Integrated Importance Sampling for Disease Mapping Models, June 2017, Annual Meeting of Statistical Society of Canada, University of Manitoba, Canada.
`r ritem()` Fully Bayesian Classification with Heavy-tailed Priors for Selection in High-Dimensional Features with Grouping Structure, Dec., 2016, Wuhan University, China.

**2015-2016**

`r ritem()` Cross-validatory Model Comparison and Divergent Regions Detection using iIS for Disease Mapping, Seminar of Dept of Math & Stat, University of Calgary, April 2016, Calgary, AB.
`r ritem()` Cross-validatory Model Comparison and Divergent Regions Detection using iIS for Disease Mapping, Seminar of Dept of Math & Stat, University of Alberta, Edmonton, AB.
`r ritem()` Cross-validatory Model Comparison and Divergent Regions Detection using iIS for Disease Mapping, Seminar of Department of Statistics, University of Manitoba, Jan. 2016, Winnipeg, MB.
`r ritem()` Bias-corrected Hierarchical Bayesian Classification with a Selected Subset of High-dimensional Features, ICSA Canada Chapter Annual Meeting, University of Calgary, Aug. 2015, Calgary, AB.

**2014-2015**

`r ritem()` Approximating Cross-validatory Predictive Evaluation in Bayesian Latent Variables Models with Integrated IS and WAIC, Dec. 2014, Tongji University, Shanghai, China.
`r ritem()` An Introduction to Microarray Data. Workshop on “Statistical Issues in Biomarker and Drug Co-development”, Nov. 2014, Fields Institute, Toronto, ON, Canada.

**2013-2014**

`r ritem()` Approximating Cross-validatory Predictive Evaluation in Bayesian Latent Variables Models with Integrated IS and WAIC. Statistics Seminar, April, Kansas State University, Manhattan, Kansas, USA.

**2011-2012**

`r ritem()` High-dimensional Feature Selection Using Hierarchical Bayesian Logistic Regression with Heavy-tailed Priors. CRM-ISM-GERAD Colloque de Statistique, April, McGill University, Montreal, Quebec, Canada.

**2010-2011**

`r ritem()` High-dimensional Classification using Hierarchical Bayesian Polychotomous Logistic Regression Models. Colloquia talk, Jan., The University of Western Ontario, London, ON, Canada.
`r ritem()` High-dimensional Classification using Hierarchical Bayesian Polychotomous Logistic Regression Models. Colloquia talk, Sept., Penn State University, University Park, PA, USA.

**2007-2008**

`r ritem()` Avoiding Bias from Feature Selection. CRISM 'workshop on Bayesian Analysis of High-dimensional Data, April, University of Warwick, Coventry, UK.

### 14.2 Contributed Presentations

**2013-2014**

`r ritem(10)` Approximating Cross-validatory Predictive Evaluation in Bayesian Latent Variables Models with Integrated IS and WAIC. Annual Meeting of Statistical Society of Canada, May 27, 2014, Toronto, ON, Canada.

**2010-2011**

`r ritem()` High-dimensional Classification using Hierarchical Bayesian Polychotomous Logistic Regression Models. The 8th ICSA International Conference, Dec. 20, 2010, Guangzhou, China.

**2009-2010**

`r ritem()` Sajobi, T., Lix, L., Laverty, W., and Li, L., 2010. Discriminant Analysis for Repeated Measures Data: Effects of Covariance Structure on Bias and Error in Discriminant Function Coefficients. Annual Meeting of Statistical Society of Canada, May 24, 2010, Quebec City, QC, Canada.
`r ritem()` Are Bayesian Inferences Weak for Wasserman’s Example? Annual Meeting of Statistical Society of Canada, May 25, 2010, Quebec City, QC, Canada.

**2008-2009**

`r ritem()` Calibrating Predictions Based on a Selected Subset of Features from Bayesian Gaussian Classification Models. Annual meeting of Statistical Society of Canada, January, Vancouver, BC, Canada.
`r ritem()` Calibrating Predictions Based on a Selected Subset of Features from Bayesian Gaussian Classification Models. Bayesian Biostatistics Conference, January, Houston, TX, USA.

**2007-2008**

`r ritem()` Compressing Parameters in Bayesian High-order Models. Annual Meeting of Statistical Society of Canada, May, Ottawa, ON, Canada.

**2006-2007**

`r ritem()` Compressing Parameters in Bayesian Models with High-order Interactions. The 3rd Monte Carlo Workshop, Harvard University, May, Cambridge, MA, USA.

**2005-2006**

`r ritem()` Avoiding Bias from Feature Selection in Regression and Classification Models. Joint Statistical Meeting, August, Seattle, WA, USA.
`r ritem()` Analysis of Obstructive Sleep Apnea Data with Bayesian Neural Network. Annual Meeting of Statistical Society of Canada, June, London, ON, Canada.

## 15. REPORTS AND OTHER OUTPUTS

### 15.1 Software Released Publicly

`r ritem(11)` Wu, T. and Li, L., 2026. `Zresidual`: Computing and Diagnosing Gaussian-like Residuals. [[Github]](https://tiw150.github.io/Zresidual/index.html) [[Demo]](https://tiw150.github.io/Zresidual_demo.html). Version 0.1-0 on Github (2026).
`r ritem()` Li, L., 2026. R Functions for Computing Z-residuals for `survreg` and `coxph` Objects. [[URL]](https://longhaisk.github.io/software/NRSP/index.html).
`r ritem()` Li, L., et al., 2021. Real-time estimates of $R_t$ for Covid-19 in Canada. [[URL]](https://longhaisk.github.io/CanadaCovidRt/). 
`r ritem()` Li, L. and Liu, S., 2019--2026. `HTLR`: Bayesian Logistic Regression with Hyper-LASSO priors. DOI: 10.32614/CRAN.package.HTLR. [[CRAN]](https://cran.r-project.org/web/packages/HTLR/index.html) [[Github]](https://longhaisk.github.io/HTLR) [[URL]](https://longhaisk.github.io/software/BLRHL/index.html). Version 0.4 (2019), version 0.4-1 (2019), version 0.4-2 (2020), version 0.4-3 (2020), version 0.4-4 (2022), version 1.0 (2026).
`r ritem()` Li, L., 2011--2026. `BCBCSF`: Bias-corrected Bayesian Classification with Selected Features. DOI: 10.32614/CRAN.package.BCBCSF.  [[CRAN]](https://cran.r-project.org/web/packages/BCBCSF/index.html) [[URL]](https://longhaisk.github.io/software/BCBCSF/index.html). Version 0.0-0 (2011), version 0.0-1 (2011), version 0.0-2 (2012), version 1.0-0 (2013), version 1.0-1 (2015), updated to version 1.0-2 (2026).
`r ritem()` Li, L., 2018. `HTLR`: Bayesian Logistic Regression with Hyper-LASSO priors. [[URL]](https://longhaisk.github.io/software/BLRHL/index.html). Pre-CRAN version (2018).
`r ritem()` Li, L., 2016. `iIS`: R code for computing predictive p-values in disease mapping models. [[URL]](https://longhaisk.github.io/software/dmpvalues/dmpvalues-larynx.R). 
`r ritem()` Li, L., 2008. `gibbs.met`: Naive Gibbs Sampling with Metropolis Steps. [[CRAN]](https://cran.r-project.org/web/packages/gibbs.met/index.html) [[URL]](https://longhaisk.github.io/software/gibbs.met/release.html).
`r ritem()` Li, L., 2008. `BPHO`: Bayesian Prediction with High-order Interactions. [[CRAN]](https://cran.r-project.org/web/packages/BPHO/index.html) [[URL]](https://longhaisk.github.io/software/BPHO/release.html).
`r ritem()` Li, L., 2007. `predmixcor`: Classification rule based on Bayesian mixture models with feature selection bias corrected. [[CRAN]](https://cran.r-project.org/web/packages/predmixcor/index.html) [[URL]](https://longhaisk.github.io/software/predmixcor/release.html). 
`r ritem()` Li, L., 2007. `predbayescor`: Classification Rule Based on Bayesian Naive Bayes Models with Features Selection Bias Corrected. [[CRAN]](https://cran.r-project.org/web/packages/predbayescor/index.html) [[URL]](https://longhaisk.github.io/software/predbayescor/release.html). 

### 15.2 Technical Reports

`r ritem(12)` Wu, T., Feng, C. and Li, L., 2023. Cross-validatory Z-Residual for Diagnosing Shared Frailty Models. [https://doi.org/10.48550/arXiv.2303.09616](https://doi.org/10.48550/arXiv.2303.09616). 32 pages, 14 figures.
`r ritem()` Wu, T., Li, L. and Feng, C., 2023. Z-residual diagnostics for detecting misspecification of the functional form of covariates for shared frailty models. [https://doi.org/10.48550/arXiv.2302.09106](https://doi.org/10.48550/arXiv.2302.09106). 21 pages, 7 figures.
`r ritem()` Li, L., Wu, T. and Feng, C., 2019. Model diagnostics for censored regression via randomized survival probabilities. [https://doi.org/10.48550/arXiv.1911.00198](https://doi.org/10.48550/arXiv.1911.00198). 12 pages. (Journal-ref: Statistics in Medicine, 2021, 40(6), 1482-1497).
`r ritem()` Feng, C., Sadeghpour, A. and Li, L., 2017. Randomized Predictive P-values: A Versatile Model Diagnostic Tool with Unified Reference Distribution. [https://doi.org/10.48550/arXiv.1708.08527](https://doi.org/10.48550/arXiv.1708.08527). 26 pages. (Journal-ref: BMC Medical Research Methodology, 2020, 20(175)).
`r ritem()` Jiang, L., Li, L. and Yao, W., 2016. Fully Bayesian Classification with Heavy-tailed Priors for Selection in High-dimensional Features with Grouping Structure. [https://doi.org/10.48550/arXiv.1607.00098](https://doi.org/10.48550/arXiv.1607.00098). 31 pages. (Journal-ref: Sci Rep, 2020, 10(9747)).
`r ritem()` Li, L., Feng, C.X. and Qiu, S., 2016. Estimating Cross-validatory Predictive P-values with Integrated Importance Sampling for Disease Mapping Models. [https://doi.org/10.48550/arXiv.1603.07668](https://doi.org/10.48550/arXiv.1603.07668). 18 pages. (Journal-ref: Statistics in Medicine, 2017, 36(14), 2220-2236).
`r ritem()` Li, L. and Yao, W., 2014. Fully Bayesian Logistic Regression with Hyper-Lasso Priors for High-dimensional Feature Selection. [https://doi.org/10.48550/arXiv.1405.3319](https://doi.org/10.48550/arXiv.1405.3319). 33 pages. (Journal-ref: Journal of Statistical Computation and Simulation, 2018, 88(14), 2827-2851).
`r ritem()` Li, L., Qiu, S., Zhang, B. and Feng, C.X., 2014. Approximating Cross-validatory Predictive Evaluation in Bayesian Latent Variables Models with Integrated IS and WAIC. [https://doi.org/10.48550/arXiv.1404.2918](https://doi.org/10.48550/arXiv.1404.2918). 38 pages. (Journal-ref: Statistics and Computing, 2016, 26(4), 881-897).
`r ritem()` Li, L. and Yao, W., 2013. High-dimensional Feature Selection Using Hierarchical Bayesian Logistic Regression with Heavy-tailed Priors. [https://doi.org/10.48550/arXiv.1308.4690](https://doi.org/10.48550/arXiv.1308.4690). (Earlier version of arXiv:1405.3319).
`r ritem()` Li, L. and Neal, R.M., 2007. A Method for Compressing Parameters in Bayesian Models with Application to Logistic Sequence Prediction Models. [https://doi.org/10.48550/arXiv.0711.4983](https://doi.org/10.48550/arXiv.0711.4983). 29 pages. (Journal-ref: Bayesian Analysis, 2008, 3(4), 793-822).
`r ritem()` Li, L., 2007. Bayesian Classification and Regression with High Dimensional Features. [https://doi.org/10.48550/arXiv.0709.2936](https://doi.org/10.48550/arXiv.0709.2936). PhD Thesis Submitted to University of Toronto, 129 pages.
`r ritem()` Li, L., Zhang, J. and Neal, R.M., 2007. A method for avoiding bias from features selection with application to naive Bayes classification models. Technical Report No 0705, Department of Statistics, University of Toronto.

## 17. RESEARCH FUNDING HISTORY

**2025-2026**

`r ritem(17)` **NSERC Individual Discovery Grant (No. 2026-07053)** – *Prediction-based Methods for Statistical Learning and Inference in Biosciences and Epidemiology*, $185,000 (37K per year), 2026-2031, PI.
`r ritem()` **CANSSI Collaborative Research Team Projects** – *Statistical Methodologies and Computational Tools to Identify Microbial Correlates of Canadian Bee Gut Health*, Project 29, 2025-2028, Co-PI.

**2021-2022**

`r ritem()` **MITACS Accelerate Grant** – *Geospatial Artificial Intelligence Algorithms for Automating Manual Observation Associated with Wheat Production*, $280,000, 2021-2025, PI.

**2020-2021**

`r ritem()` **MITACS Accelerate Grant** – *Develop a web based geospatial artificial intelligence framework to track, visualize, analyze, model, and predict infectious disease spread in real-time*, $105,000, 2020-2021, PI.

**2019-2020**

`r ritem()` **NSERC Individual Discovery Grant** – *Predictive Methods for Analyzing High-throughput and Spatial-temporal Data*, $140,000 (20K per year), 2019-2026, PI.

**2017-2018**

`r ritem()` **The Western Canadian Universities Collaborative Project Seed Funding** – *Genome-wide diet-gene interaction analysis for risk of psychiatric comorbidity in inflammatory bowel disease*, $20,000, 2017-2019, Co-PI.

**2016-2017**

`r ritem()` **Canada First Research Excellence Fund (CFREF)** - *Designing Crops for Global Food Security, Genotype & Environment to Phenotype*, $756,918, 2016-2019, Co-PI.
`r ritem()` **MITACS Accelerate Internship** – *Applications of Neural Network Curve Fitting Methods for Least-squares Monte Carlo Simulations in Financial Risk Management*, $15,000, 2016, PI.

**2014-2015**

`r ritem()` **NSERC Individual Discovery Grant** – *Bayesian Methods for High-Dimensional and Correlated Data*, $70,000, 2014-2019, PI.

**2011-2012**

`r ritem()` **NSERC Individual Discovery Grant ECR Supplement** – *Efficient Bayesian Analysis for Complex Models*, $5,000/year, 2011-2014, PI.

**2009-2010**

`r ritem()` **NSERC Individual Discovery Grant** – *Efficient Bayesian Analysis for Complex Models*, $80,000, 2009-2014, PI.
`r ritem()` **CFI Leaders Opportunity Funds** – *A Computer Cluster for Research on Efficient Bayesian Statistical Methods*, $160,000, 2009, PI.

**2008-2009**

`r ritem()` **MITACS Accelerate Internship** – *Clustering Analysis for Detecting the Types of Vehicles*, $15,000, 2008, Co-PI.
`r ritem()` **University of Saskatchewan President's Award**, $5,000, 2008, PI.
`r ritem()` **College of Graduate Studies and Research at the University of Saskatchewan Award**, $15,000, 2008, PI.
`r ritem()` **College of Arts and Science at the University of Saskatchewan – Supplemental start-up operating grant**, $15,000, 2008, PI.

**2007-2008**

`r ritem()` **University of Saskatchewan – Start-up operating grant**, $5,000, 2007, PI.

## 18. PRACTICE OF PROFESSIONAL SKILLS

**2025-2026**

`r ritem(68)` Refereeing for *Journal of Statistical Computation and Simulation*, June, 2026
`r ritem()` Refereeing for *Journal of Statistical Computation and Simulation*, April, 2026
`r ritem()` Refereeing for *Journal of the Royal Statistical Society: Series C*, April 2026
`r ritem()` Refereeing for *Bioinformatics*, March 2026
`r ritem()` Refereeing for *Journal of Computational and Graphical Statistics*, March 2026
`r ritem()` External Referee for a tenure and promotion of SFU, Dec. 2025
`r ritem()` External Examiner for the doctoral thesis by Xiaoqing Zhang at University of Regina, Dec. 8, 2025
`r ritem()` Refereeing for *Journal of Statistical Computation and Simulation*, Dec. 2025
`r ritem()` Refereeing for *Journal of Computational and Graphical Statistics*, Sept. 2025
`r ritem()` External Examiner for the doctoral thesis by Na Zhang at University of Alberta, August 28, 2025
`r ritem()` Refereeing for *Journal of the Royal Statistical Society: Series C*, August 2025
`r ritem()` Refereeing for *Journal of Applied Statistics*, August 2025

**2024-2025**

`r ritem()` Organizing an invited Session for the 7th Symposium of ICSA Canada Chapter, McGill University, August 2026
`r ritem()` Organizing an invited Session for 2025 SSC Annual Meeting, Saskatoon, SK, Canada, June 2025

**2023-2024**

`r ritem()` External Examiner for the doctoral thesis by Yuping Yang at SFU, June 25, 2024
`r ritem()` Refereeing for *Journal of Computational and Graphical Statistics*, Sept. 2024
`r ritem()` Refereeing for *Journal of Applied Statistics*, Jan. 2024
`r ritem()` Review an MITACS Accelerate Grant, Dec. 2023

**2022-2023**

`r ritem()` Refereeing for *Statistical Methods in Medical Research*, April 2023
`r ritem()` Refereeing for *Statistical Methods in Medical Research*, Jan. 2023
`r ritem()` Refereeing for *Journal of Computational and Graphical Statistics*, Jan. 2023
`r ritem()` Refereeing for *Statistical Methods in Medical Research*, Aug. 2022
`r ritem()` External Examiner for the M.Sc. thesis by Xiangling Ji, University of Victoria, 27 July 2022
`r ritem()` Refereeing for *Canadian Journal of Statistics*, July 2022

**2021-2022**

`r ritem()` Organizer of an invited session for ICSA Canada Symposium 2022, Banff, AB, Canada, July 2022
`r ritem()` External Examiner for one NSERC IDG application
`r ritem()` External Examiner for another NSERC IDG application
`r ritem()` External Examiner for one MITACS Accelerate grant application
`r ritem()` Reviewer for a Canada Research Chair Position
`r ritem()` External M.Sc. Thesis Examiner for Zhongyuan Zhang, University of Toronto
`r ritem()` Refereeing for *Statistical Methods in Medical Research*
`r ritem()` Refereeing for *Journal of Statistical Computation and Simulation*
`r ritem()` Refereeing for *BMC Cancer*
`r ritem()` Refereeing for *Journal of Computational and Graphical Statistics*
`r ritem()` Refereeing for *Canadian Journal of Statistics*
`r ritem()` Refereeing for *IEEE Transactions on Neural Networks and Learning Systems*

**2020-2021**

`r ritem()` Grant refereeing for an application to MITACS Accelerate, May 2021
`r ritem()` External Examination for a PhD thesis of University of Montreal, May 2021
`r ritem()` Grant refereeing for an application to NSERC IDG, Jan. 2021
`r ritem()` Refereeing for *Statistics in Medicine*
`r ritem()` Refereeing for *Computational Statistics and Data Analysis*
`r ritem()` Refereeing for *Frontiers in Genetics*
`r ritem()` Refereeing for *Statistical Methods for Medical Research*
`r ritem()` Refereeing for *Journal of Statistical Computation and Simulation*
`r ritem()` Refereeing for *BMC Cancer*

**2019-2020**

`r ritem()` External doctoral thesis examiner for Shijia Wang, Simon Fraser University
`r ritem()` External doctoral thesis examiner for Kexin Luo, Western University
`r ritem()` Grant Refereeing for an application to MITACS
`r ritem()` Grant Refereeing for an application to NSERC IDG
`r ritem()` Refereeing for *Computational Statistics and Data Analysis*
`r ritem()` Refereeing for *Frontiers in Genetics*
`r ritem()` Refereeing for *Communications in Statistics - Simulation and Computation*

**2017-2018**

`r ritem()` Referee a NSERC discovery grant application
`r ritem()` Refereeing for *Canadian Journal of Statistics*
`r ritem()` Refereeing for *Journal of Royal Statistical Society (C)*

**2016-2017**

`r ritem()` Referee two applications for MITACS Accelerate Grant
`r ritem()` Refereeing for *Statistics in Medicine*
`r ritem()` Refereeing for *Statistics and Computing*
`r ritem()` Refereeing for *PLOS ONE*

**2015-2016**

`r ritem()` Organizer, Invited session “Recent Advances in Statistical Inference Methods in Regression Models for Complex and Big Data”, China Statistics Conference, June 2016, Qingdao, China
`r ritem()` Refereeing application for NSERC individual discovery grant (2015)

**2013-2014**

`r ritem()` Reviewing and revision services for an SHRF Establishment Grant application (Prof. Kelly Penz), funded 2013
`r ritem()` Refereeing for *Biometrika*
`r ritem()` Refereeing for *Statistics In Medicine*
`r ritem()` Refereeing for *Statistical Papers*
`r ritem()` Refereeing for *Computational Statistics*
`r ritem()` Refereeing for *Statistica Sinica*

**2011-2012**

`r ritem()` Refereeing application for NSERC individual discovery grant (2011)

## 19. ADMINISTRATIVE SERVICE

### 19.1 University Committees

**2025-2026**

`r ritem(10)` Chair, Collaborative Biostatistics Program, University of Saskatchewan.
`r ritem()` USASK NSERC Discovery Grant (DG), Internal Reviewer.

**2020-2021**

`r ritem()` USASK NSERC Discovery Grant (DG), Internal Reviewer.

**2019-2020**

`r ritem()` Chair, Collaborative Biostatistics Program, University of Saskatchewan.

**2018-2019**

`r ritem()` Member, Academic Programming Committee, University of Saskatchewan.

**2017-2018**

`r ritem()` Member, Academic Programming Committee, University of Saskatchewan.

**2016-2017**

`r ritem()` Member, Academic Programming Committee, University of Saskatchewan.
`r ritem()` Member, University of Saskatchewan Bioinformatics Program Committee.

**2015-2016**

`r ritem()` Member, University of Saskatchewan Bioinformatics Program Committee.

**2013-2014**

`r ritem()` Dean's Designate for the Ph.D. Defense of Rui Zhang, Department of Veterinary Microbiology, June 12, 2014.

### 19.2 College and Departmental Committees

**2025-2026**

`r ritem(45)` Member, Budgeting and Planning Committee, Dept of Math & Stat
`r ritem()` Co-Chair, Undergraduate Committee (Statistics), Dept of Math & Stat
`r ritem()` Member, Graduate Program Committee in Statistics, Dept of Math & Stat

**2024-2025**

`r ritem()` Member, Graduate Program Committee in Statistics, Dept of Math & Stat
`r ritem()` Search subcommittee for a faculty position in statistics

**2023-2024**

`r ritem()` Member, Graduate Program Committee in Statistics, Dept of Math & Stat
`r ritem()` Member, Sub Search Committee, 4-Year Lecturer Position

**2022-2023**

`r ritem()` Statistics Advisor (credit transferring for the whole university)
`r ritem()` Member, Department Promotion (Associate) Committee (1-Case), Dept of Math & Stat
`r ritem()` Member, Department Renewals and Tenure Committee (1-Case: Tenure), Dept of Math & Stat
`r ritem()` Member, Department Promotion (Full) Committee (1-Case), Dept of Math & Stat

**2021-2022**

`r ritem()` Statistics Advisor (credit transferring for the whole university)
`r ritem()` Member, Department Renewals and Tenure Committee (1-Case), Dept of Math & Stat
`r ritem()` Member, Department Promotion (Full) Committee (1-Case), Dept of Math & Stat
`r ritem()` Search subcommittee for a 4-year term lecturer position (two rounds of searching, Jan 2022--June 2022)

**2019-2020**

`r ritem()` Member, Tenure Committee (1 case), Dept of Math & Stat
`r ritem()` Member, Promotion Committee (1 case), Dept of Math & Stat
`r ritem()` Member, Renewal of Probation Committee (1 case), Dept of Math & Stat
`r ritem()` Member, Graduate Committee, Dept of Math & Stat
`r ritem()` Organizer, Team discussion towards renovating undergraduate statistics program, Dept of Math & Stat

**2018-2019**

`r ritem()` Committee Member, Data Science Boot Camp, University of Saskatchewan (June 10–21, 2019)
`r ritem()` Member, Curriculum Renewal Committee, Dept of Math & Stat, Term 1
`r ritem()` Member, Undergraduate Committee, Dept of Math & Stat, Term 1

**2017-2018**

`r ritem()` Member, Salary Review Committee, Dept of Math & Stat, University of Saskatchewan
`r ritem()` Member, Search Committee, Dept of Math & Stat, University of Saskatchewan

**2016-2017**

`r ritem()` Member, Sub Search Committee for a joint position in ``data science/big data'', College of Arts and Science, University of Saskatchewan.
`r ritem()` Member, Graduate Committee, Dept of Math & Stat
`r ritem()` Member, Sub Search Committee, APA position, Dept of Math & Stat, University of Saskatchewan
`r ritem()` Member, Sub Search Committee, 4 lecturer positions, Dept of Math & Stat, University of Saskatchewan
`r ritem()` Organizer, Statistics and Probability Alumni Networking Day (Nov. 2016)
`r ritem()` Organizer, Qualifying Exams for Trisha Lawrence (Nov. 2016)

**2015-2016**

`r ritem()` Member, Graduate Committee, Dept of Math & Stat
`r ritem()` Organizer, Student Seminar Day, Dept of Math & Stat (May 2016)
`r ritem()` Team leader, submission of U of S courses for accreditation by the Statistical Society of Canada (May 2016)
`r ritem()` Organizer, Qualifying Exams for Trisha Lawrence (May 2016)

**2014-2015**

`r ritem()` Member, Academic Program Committee, College of Arts and Science.
`r ritem()` Member, Graduate Committee, Dept of Math & Stat
`r ritem()` Organizer, Seminar Series, Dept of Math & Stat

**2012-2013**

`r ritem()` Member, Curriculum Renewal Committee, Dept of Math & Stat
`r ritem()` Member, Budget Planning Committee, Dept of Math & Stat
`r ritem()` Member, Colloquium Committee, Dept of Math & Stat

**2011-2012**

`r ritem()` Member, Salary Review Committee, Dept of Math & Stat, University of Saskatchewan
`r ritem()` Organizer, Seminar Series, Dept of Math & Stat
`r ritem()` Member, Colloquium Committee, Dept of Math & Stat

**2009-2010**

`r ritem()` Member, Sub Search Committee, Dept of Math & Stat

## 20. PROFESSIONAL OR ASSOCIATION OFFICES AND COMMITTEE ACTIVITY

**2024-2025**

`r ritem(9)` Member of NSERC Discovery Grant EG 1508 Committee
`r ritem()` Local Organizing Committee, 2025 Annual Meeting of the Statistical Society of Canada held at the U of S

**2023-2024**

`r ritem()` Member of NSERC Discovery Grant EG 1508 Committee

**2022-2023**

`r ritem()` Member of NSERC Discovery Grant EG 1508 Committee
`r ritem()` Co-editor for a special issue "Prediction Methods for Rare Diseases or Outcomes" in the journal *BMC Medical Research Methodology*

**2021-2022**

`r ritem()` Member, CANSSI-SK Health Research Collaborating Center. Participate Substantially in organizing a semester-long seminar series

**2019-2020**

`r ritem()` Program Committee Member for the 4th ICSA-Canada Symposium, Queens University (Aug. 2019)

**2017-2018**

`r ritem()` Co-chair of the scientific program, the 3rd ICSA Canada Chapter Symposium held in Vancouver (Aug. 2017)

**2016-2017**

`r ritem()` Judge for case study competition, Annual Meeting of Statistical Society of Canada (June 2017)

