---
title: "Analysis of Multiple Linear Regression Models"
author: "Longhai Li"
date: "September 8, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    highlight: tango
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup, include=TRUE, echo=FALSE}
library(knitr)
library(rmarkdown)
# This chunk sets up the global options for the document.
knitr::opts_chunk$set(
	fig.height = 6,
	fig.width = 9,
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	comment = "#",
	include = TRUE
)

```

# Fit of the multiple linear regression

This section analyzes the wire bond strength dataset.

## Data Loading and Visualization

**Note:** You must change the file paths in the `read.csv()` functions below to match the location of the files on your computer for example "C:\Users\<YourUsername\>\Documents" on windows.

```{r load-bond-data}

# Read data. Change the path as necessary.
# Example: bond.data <- read.csv("data/wire-bond.csv")
bond.data <- read.csv("wire-bond.csv")

# This will now be automatically rendered as a paged table
bond.data
```

```{r dummy.data, include=FALSE, eval=FALSE}
# Use set.seed() to make sure you get the exact same random data every time
set.seed(42)

# Define the number of data points to create
n <- 150

# Create the bond.data data frame
bond.data <- data.frame(
  # Generate 'length' as random numbers between 10 and 50
  length = runif(n, 10, 50),
  # Generate 'height' as random numbers between 5 and 25
  height = runif(n, 5, 25)
)

# Create 'strength' based on a linear relationship with the other variables
# strength = (some_value * length) + (another_value * height) + random_error
bond.data$strength <- (2.5 * bond.data$length) + 
                      (4.1 * bond.data$height) + 
                      rnorm(n, mean = 0, sd = 20) # Add random noise

# Display the first few rows of the new data frame
head(bond.data)
```

**2D Visualization**

```{r load-bond-data2}

# Create initial scatter plots
par(mfrow = c(1, 2))
plot(bond.data$length, bond.data$strength,
     xlab = "Wire Length", ylab = "Pull strength")
plot(bond.data$height, bond.data$strength,
     xlab = "Die height", ylab = "Pull strength")
```

**3D Visualize**

```{r}
# 1. Install and load the required package

library(scatterplot3d)

par (mfrow = c(1,1))
# 2. Create the 3D scatterplot and store its object
s3d <- with(bond.data, scatterplot3d(
  x = length,
  y = height,
  z = strength,
  pch = 19,
  color = "steelblue",
  main = "3D Scatterplot: Strength vs. Length and Height",
  xlab = "Length",
  ylab = "Height",
  zlab = "Strength",
  angle = 60
))

# 3. Fit a linear model
fit <- lm(strength ~ length + height, data = bond.data)

# 4. Add the regression plane to the plot
s3d$plane3d(fit, lty.box = "solid")

```

**Dynamic 3D Visualization of 3D Data**

```{r}
# 1. Load necessary libraries
library(plotly)
library(dplyr)

# NOTE: This script assumes you have a data frame named 'bond.data'
# with columns 'length', 'height', and 'strength' already loaded.

# 2. Create the initial scatter plot object
fig <- plot_ly(
  data = bond.data, 
  x = ~length, 
  y = ~height, 
  z = ~strength
) %>%
  add_markers(
    marker = list(
      color = ~strength,
      colorscale = 'Viridis',
      showscale = TRUE,
      colorbar = list(title = 'Strength')
    ),
    name = 'Data Points'
  ) %>%
  layout(
    title = "Interactive 3D Scatter Plot of Bond Strength",
    scene = list(
      xaxis = list(title = 'Length'),
      yaxis = list(title = 'Height'),
      zaxis = list(title = 'Strength')
    )
  )

# 3. Display the scatter plot
# In an interactive session (like RStudio), this line will render the plot.
fig
```

**Add Fitted Surface**

```{r}
# 4. Fit a linear model to the data
fit <- lm(strength ~ length + height, data = bond.data)

# 5. Prepare data for the regression plane surface
resolution <- 20
axis_x <- seq(min(bond.data$length), max(bond.data$length), length.out = resolution)
axis_y <- seq(min(bond.data$height), max(bond.data$height), length.out = resolution)

z_matrix <- outer(
  axis_x,
  axis_y,
  function(len, ht) {
    predict(fit, newdata = data.frame(length = len, height = ht))
  }
)

# 6. Add the surface to the EXISTING plot object and display the result
fig <- fig %>%
  add_surface(
    x = ~axis_x,
    y = ~axis_y,
    z = ~z_matrix,
    colorscale = 'Blues',
    opacity = 0.7,
    showscale = FALSE,
    name = 'Regression Plane'
  )

# 7. Display the final, updated plot
fig
```

## Model Fitting and Summary

We fit a multiple linear regression model with `strength` as the response variable and `length` and `height` as predictors.

```{r fit-bond-model}
# Fit the linear model
fit <- lm(strength ~ length + height, data = bond.data)

# Display the classic summary output (this is not a data frame, so it's unaffected)
summary(fit)
```

The summary provides the ANOVA F-test for overall significance, $R^2$, adjusted $R^2$, and t-tests for individual coefficients.

## Confidence Intervals and Model Components

We can extract confidence intervals for the regression coefficients, as well as fitted values and residuals.

```{r bond-model-components}
# Confidence intervals (matrix output, will be auto-paged)
confint(fit)

# Fitted values and ordinary residuals (matrix output, will be auto-paged)
pred <- fitted.values(fit)
e <- resid(fit)
data.frame(y = bond.data$strength, y.hat = pred, e = e)

# Extract covariance matrix (matrix output, will be auto-paged)
cov.mat <- vcov(fit)
cov.mat

# Standard errors (vector output, needs to be a data.frame to be paged)
data.frame(std.error = sqrt(diag(cov.mat)))
```

# Partial F-test and t-test

This section uses data on the weight, height, and age of children to demonstrate partial F-tests.

```{r setup-child-data}
# Data: Weight, height and age of children
wgt <- c(64, 71, 53, 67, 55, 58, 77, 57, 56, 51, 76, 68)
hgt <- c(57, 59, 49, 62, 51, 50, 55, 48, 42, 42, 61, 57)
age <- c(8, 10, 6, 11, 8, 7, 10, 9, 10, 6, 12, 9)
child.data <- data.frame(wgt, hgt, age)
```

## Problem 1: Simple Linear Regression

```{r partial-f-test-1}
fit11 <- lm(wgt ~ hgt, data = child.data)
summary(fit11)
```

## Problem 2: Adding a Predictor

Here we test if adding `age` significantly improves the model that already contains `hgt`.

```{r partial-f-test-2}
fit21 <- lm(wgt ~ hgt + age, data = child.data)
summary(fit21)

# Compare the nested models (anova output is a data frame, will be auto-paged)
fit22 <- lm(wgt ~ hgt, data = child.data)
anova(fit22, fit21)
```

## Problem 3: Adding a Polynomial Term

We test if adding a quadratic term for `age` improves the model with `hgt` and `age`.

```{r partial-f-test-3}
fit31 <- lm(wgt ~ hgt + age + I(age^2), data = child.data)
summary(fit31)

fit32 <- lm(wgt ~ hgt + age, data = child.data)
anova(fit32, fit31)
```

## Problem 4: Testing Multiple Coefficients

We test if the `age` and `age^2` terms are jointly significant.

```{r partial-f-test-4}
fit41 <- lm(wgt ~ hgt + age + I(age^2), data = child.data)
fit42 <- lm(wgt ~ hgt, data = child.data)
anova(fit42, fit41)
```

# Mean Response and Prediction Intervals

This section demonstrates how to calculate confidence and prediction intervals for the wire bond strength data.

## Confidence Interval for Mean Response

Construct a 95% CI on the mean pull strength for a wire bond with wire length = 8 and die height = 275.

```{r confidence-interval}
# We use the model 'fit' from the first section (predict output is a matrix)
predict(fit, newdata = data.frame(length = 8, height = 275),
        interval = "confidence", level = 0.95)
```

## Prediction Interval for a New Observation

Construct a 95% PI on the pull strength for a new wire bond with wire length = 8 and die height = 275.

```{r prediction-interval}
predict(fit, newdata = data.frame(length = 8, height = 275),
        interval = "prediction", level = 0.95)
```

# Model Diagnostics

This section covers various methods for diagnosing the model fit using residuals and other measures.

## Residual Calculations

We can calculate hat values, and various types of residuals. They are combined here into a single data frame which will be auto-paged.

```{r calculate-residuals}
residuals_df <- data.frame(
  hat_values = hatvalues(fit),
  ordinary_resid = resid(fit),
  standardized_resid = resid(fit) / sigma(fit),
  studentized_internal = rstandard(fit),
  studentized_external = rstudent(fit)
)
residuals_df
```

## Residual Plots

A common practice is to analyze plots of the studentized residuals to check for non-linearity, non-constant variance, and non-normality.

```{r plot-residuals}
# Residual analysis using internally studentized residuals
n <- nrow(bond.data)
r <- rstandard(fit) 
y.hat <- fitted.values(fit)

par(mfrow = c(2, 3))
qqnorm(r, main = "Normal Q-Q Plot")
qqline(r)
plot(y.hat, r, xlab = "Fitted values", ylab = "Studentized Residuals")
abline(h = 0)
plot(1:n, r, xlab = "Observation Number", ylab = "Studentized Residuals")
abline(h = 0)
plot(bond.data$length, r, xlab = "Wire Length", ylab = "Studentized Residuals")
abline(h = 0)
plot(bond.data$height, r, xlab = "Die Height", ylab = "Studentized Residuals")
abline(h = 0)
```

# Influential Observations

We can identify influential data points using metrics like DFFITS, DFBETAS, and Cook's D.

```{r influential-obs-calcs}
# Find DFFITS, DFBETAS and Cook's D for the wire bond strength data.
influence_df <- data.frame(dffits = dffits(fit), cook.D = cooks.distance(fit), dfbetas(fit))
influence_df
```

## Plotting with the `olsrr` Package

The `olsrr` package provides convenient functions for plotting these influence diagnostics. Note: You only need to run `install.packages("olsrr")` once.

```{r plot-influential-obs, fig.height=12}
# install.packages("olsrr") # Run this once if you don't have the package
library(olsrr)

# Plot for Cook's D
ols_plot_cooksd_chart(fit)

# Plot for DFFITS
ols_plot_dffits(fit)

# Plot for DFBETAS
ols_plot_dfbetas(fit)
```

# Polynomial Regression

This example uses data on the cost and production lot size of airplane sidewall panels.

```{r poly-regression-setup}
y <- c(1.81, 1.70, 1.65, 1.55, 1.48, 1.40, 1.30, 1.26, 1.24, 1.21, 1.20, 1.18)
x <- c(20, 25, 30, 35, 40, 50, 60, 65, 70, 75, 80, 90)
fit_poly <- lm(y ~ x + I(x^2))
summary(fit_poly)
```

The plot shows the fitted quadratic curve.

```{r poly-regression-plot}
plot(x, y, xlab = "Lot size, x", ylab = "Average cost per unit, y")
lines(x, predict(fit_poly, newdata = data.frame(x = x)), type = "l")
```

We can use a partial F-test to see if the quadratic term significantly improves the model.

```{r poly-regression-ftest}
fit1 <- lm(y ~ x)
anova(fit1, fit_poly)
```

# Indicator Variables

This section uses the SBP (Systolic Blood Pressure) dataset to model the effect of a categorical variable (`sex`).

```{r indicator-setup}
# Note: Update this path to your local file location
sbpdata <- read.csv("sbpdata.csv")

# Fit the full model with an interaction term
fit.full <- lm(sbp ~ age + sex + age:sex, data = sbpdata)
summary(fit.full)
```

## Hypothesis Tests for Indicator Models

We can perform various hypothesis tests to see if the lines are coincident, parallel, or have equal intercepts.

```{r indicator-hyp-tests}
# Test of Coincidence. H0: beta2=beta3=0
fit.coin <- lm(sbp ~ age, data = sbpdata)
anova(fit.coin, fit.full)

# Test of Parallelism. H0: beta3=0
fit.para <- lm(sbp ~ age + sex, data = sbpdata)
anova(fit.para, fit.full)

# Test for Equal Intercepts. H0: beta2=0 (assuming parallel lines)
fit1.inter <- lm(sbp ~ age + sex, data = sbpdata)
fit2.inter <- lm(sbp ~ age, data = sbpdata)
anova(fit2.inter, fit1.inter)
```

## Plotting the Fitted Model

The plot below shows the separate regression lines for males and females.

```{r indicator-plot}
# Define color for males (sex=1) and females (sex=0).
colors <- ifelse(sbpdata$sex == 1, "black", "gray")

# Scatter plot
plot(sbpdata$age, sbpdata$sbp, xlab = "Age", ylab = "SBP", col = colors)

# Add the fitted lines for males and females
# The design matrix is data.frame(1, age, 1, age) for sex = 1,
# and data.frame(1, age, 0, 0) for sex = 0.
lines(sbpdata$age, cbind(1, sbpdata$age, 1, sbpdata$age) %*% coef(fit.full), col = "black")
lines(sbpdata$age, cbind(1, sbpdata$age, 0, 0) %*% coef(fit.full), col = "gray")

# Add legend
legend("topleft", legend = c("sex = 1 (male)", "sex = 0 (female)"), 
       lty = c(1, 1), col = c("black", "gray"))
```

# Model Building

This section uses the `olsrr` package and wine quality data to demonstrate automated model selection techniques.

```{r model-building-setup}
library(olsrr)
# Note: Update this path to your local file location
wine <- read.csv("wine.csv")

# A dot in the lm function means: use all other variables as predictors.
model <- lm(quality ~ ., data = wine)
```

## All Possible Regression

This method evaluates every possible combination of predictors. The output is a data frame and will be auto-paged.

```{r model-building-all-possible}
ols_step_best_subset(model)
```

## Automated Stepwise Procedures

We can also use backward, forward, or stepwise selection to find a good model. These functions also return data frames.

```{r model-building-stepwise}
# Backward Elimination (alpha_out = 0.1)
ols_step_backward_p(model, p_val = 0.1)

# Forward Selection (alpha_in = 0.1)
ols_step_forward_p(model, p_val = 0.1)

# Stepwise Regression (alpha_in = 0.1, alpha_out = 0.1)
ols_step_both_p(model, p_enter = 0.1, p_remove = 0.1)
```

# Multicollinearity

This section explores the issue of multicollinearity, where predictors are highly correlated.

## A Simple Example

```{r multicollinearity-simple}
y <- c(19, 20, 37, 39, 36, 38)
x1 <- c(4, 4, 7, 7, 7.1, 7.1)
x2 <- c(16, 16, 49, 49, 50.4, 50.4)
cor(data.frame(x1, x2))

# Full model
fit_multi <- lm(y ~ x1 + x2)
summary(fit_multi)

# Simple model
fit1_multi <- lm(y ~ x1)
summary(fit1_multi)
```

## VIFs in the Wine Quality Data

We can calculate the Variance Inflation Factor (VIF) for each predictor to diagnose multicollinearity. A VIF \> 10 is often considered problematic.

```{r multicollinearity-wine}
wine.x <- wine[, -ncol(wine)] # Assuming quality is the last column
cor(wine.x)

# Find VIF's using the olsrr package (returns a data frame).
ols_vif_tol(model)
```

```         
```
