---
title: 'STAT 812: Computational Statistics'
author: "Longhai Li"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: no
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
  word_document:
    toc: yes
  pdf_document:
    toc: yes
    fig_height: 6
    fig_width: 10
    number_sections: yes
subtitle: Computer Arithmetic
editor_options:
  chunk_output_type: console
---
```{r}
options(digits=22) ## display values in 22 digits
```

# Integer
An integer $u$ is represented by  a vector of binary numbers $(d_0,\ldots,d_{31})$ as follows:
$$
u=\sum_{i=0}^{31} d_i2^i - 2^{31}.
$$
The larguest integer is 
$$u^{max}=\sum_{i=0}^{31} 1\times 2^i - 2^{31}=2^{31}-1.$$
The smallest integer is 
$$u^{min}= - 2^{31}\approx -(2^{31}-1).$$

Checking the largest and smallest integers
```{r}
as.integer(2^31)

as.integer(2^31 - 1)
as.integer(2^31 - 1) + as.integer(1) 

as.integer(-(2^31 - 1))
as.integer(- 2^31 )
as.integer(- 2^31 -1 )
```

# Floating point numbers

A floating point number $x$ is reprented by a vector of binary numbers $(s,d_0,\ldots,d_{t-1},e_1,\ldots, e_{k-1})$ as follows:
$$
x = (-1)^s \sum_{i=0}^{t-1}d_i2^{-i}\times 2^{\sum_{i=0}^{k-1}e_i2^i-2^k}.
$$
Typically, in double precision (64bits), $t= 52$ and $k=11$. The largest positive floating point number is roughly $2^{1024}$ and the smallest positive floating point number is roughly $2^{-1074}$. The numbers greater than $2^{1024}$ are overflow, called infinity. The numbers smaller than  $2^{-1074}$ are underlfow, treated as 0. 

```{r}
# checking the largest double precision floating-point number: 
2^1023

2^1023+1

2^1024
#smallest double precision floating-point number: > 2^(-1073)

2^(-1074)

2^(-1075)
```


# Rounding error 

## Computing sum: avoiding computing large + small
    
```{r}

### demo of roundoff error: avoid large +- small


# a number f0
f0 <- 1 + 2^(-40)

f0

f1 <- 1
i <- 1
while(i <= 2^12){     
    f1 <- f1 + 2^(-52)   
    i <- i + 1   
}

f1


#A method that will give your wrong answer

f2 <- 1
i <- 1
while(i <= 2^13){     
    f2 <- f2 + 2^(-53)   
    i <- i + 1
}


f2

# this difference is not due to number display in R:
f0==f1

f0==f2


#R built-in function 'sum'

f3 <- sum (c(1, rep (2^(-53),2^13)))

f3

f1 == f3

f4 <- sum (c(1, rep (2^(-54),2^14)))

f4

f0 == f4

# a possible solution is to sort the numbers before adding

f6s <- 0
i <- 1
while(i <= 2^14){     
    f6s <- f6s + 2^(-54)   
    i <- i + 1
}
print (f6s)
f6 <- f6s + 1; f6
```
Let's get a rough sense of mantissa in different bases:

```{r}

#base= exp(1) = 2.71
log(2^(-52))

1+ exp (-35)
1+ exp (-37)

#based 10
log(2^(-52),base=10)

```

## Computing exp (x): avoiding computing large - large

```{r}

fexp <- function(x, debug=FALSE) 
{
    i <- 0
    expx <- 1
    u <- 1
    while(abs(u)>1e-20*abs(expx)) {
        i <- i+1
        u <- u*x/i
        expx <- expx+u
        if (debug){
            cat(sprintf ("Step %d: adding %f, exp(%f) = %f\n", i, u, x, expx))
        }
    }
    expx
}

matrix(c(exp(10),fexp(10)))

matrix(c(exp(20),fexp(20)))

matrix(c(exp(60),fexp(60)))

matrix(c(exp(-1),fexp(-1)))

matrix(c(exp(-10),fexp(-10)))

matrix(c(exp(-20),fexp(-20)))

matrix(c(exp(-50),fexp(-50)))

# See what's happening
fexp (-50, debug = TRUE)

## using another expression to avoid computing large - large
fexp2 <- function(x) {
    xa <- abs(x)
    i <- 0
    expx <- 1
    u <- 1
    while(u>1e-20*expx) {
        i <- i+1
        u <- u*xa/i
        expx <- expx+u
    }
    if (x >= 0) expx else 1/expx
}






matrix(c(exp(-10),fexp2(-10)))

matrix(c(exp(-20),fexp2(-20)))

matrix(c(exp(-50),fexp2(-50)))
```
## Computing variance: avoiding computing large - large

```{r}

var1 <- function(x) 
{ n <- length(x)
( sum(x^2) - sum(x)^2/n ) / (n-1)
}

var2 <- function(x)
{
    n <- length(x)
    
    sum((x-mean(x))^2) / (n-1)
}

x <- c(1,2,3)

c(var1(x),var2(x),var(x))


x <- c(1,2,3) + 1e10

c(var1(x),var2(x),var(x))


```

# Overflow and underflow 
```{r}

## demo of overflow problem: avoid Inf/Inf, 0/0
p1 <- function(theta)
{
    exp(theta)/(1+exp(theta))
}

p2 <- function(theta)
{
    1/(1+exp(-theta))
}


theta <- 2000

p1(theta)

p2(theta)

theta <- -2000

p1(theta)

p2(theta)


```
# Use logarithm to handle overflow and underflow numbers

```{r}

log_x <- 800

x <- exp(log_x)

x

log_y <- 805

y <- exp(log_y)

y

x/y

log_xovery <- log_x- log_y

log_xovery

exp(log_xovery)


#looking for log of sum of numbers in logarithm

log_sum_exp <- function(log_x)
{
    max_log_x <- max(log_x)
    
    max_log_x + log( sum(exp(log_x - max_log_x) ))
}

log_x <- c(2000,2010,2030)

exp(log_x)

log_sum_exp(log_x)

log_x <- - c(2000,2010,2030)

exp(log_x)

log_sum_exp(log_x)

log_minus_exp <- function(log_x,log_y)
    
{  if(log_x < log_y)
    stop("The first argument is bigger than the second")
    
    log_x + log(1-exp(log_y-log_x))
    
}

log_minus_exp(2020,2000)


exp(2000) - exp(1999)

```

