---
title: "Factorial Design and Design"
author:
  - name: "Longhai Li"
    affiliation: "University of Saskatchewan"
    url: "https://longhaisk.github.io/"
date: today
editor: source
execute:
  echo: true
  warning: false
  message: false
  cache: false
  fig-width: 9
  fig-height: 6
editor_options: 
  chunk_output_type: console
---

## Battery Design Experiment

"This analysis explores data from a reliability engineering study designed to select the optimal power source for industrial sensors used in harsh environments (e.g., downhole drilling or autoclaves). The experiment investigates the Service Life (in hours) of the batteries based on two factors: 

Here is the modified and formatted text, ready for your R Markdown file:

* **Factor 1: Battery Material** (3 levels)
    * `1`: Standard Li-ion (Consumer Grade)
    * `2`: Industrial Li-MnO$_2$ (Lithium Manganese Dioxide)
    * `3`: High-Temp Li-SOCl$_2$ (Lithium Thionyl Chloride)

* **Factor 2: Ambient Temperature** (3 levels)
    * $15^\circ\text{C}$ (Cold Start)
    * $70^\circ\text{C}$ (Operating Temp)
    * $125^\circ\text{C}$ (Extreme Heat)
    
The objective is to characterize the reliability of these power sources across the full thermal range. A critical specific aim is to investigate the presence of an interaction effect: to determine if the 'High-Temp' chemistry successfully maintains its voltage stability at 125°C, a point where standard consumer-grade batteries are expected to suffer catastrophic capacity loss.

## Data Setup and Preparation

First, we organize the raw data into a structured `data.frame`. This is a best practice in R that makes the data easier to manage and the code more readable. We create columns for the response variable `life` and the two factors, `material` and `temperature`, ensuring they are treated as categorical variables (factors) for the analysis.

```{r}
#| label: setup-data

## Response variable: battery life
life <- c(130,155,74,180,  34,40,80,75,   20,70,82,58,
          150,188,159,126, 136,122,106,115, 25,70,58,45,
          138,110,168,160, 174,120,150,139, 96,104,82,60)

## Create the data frame
battery_df <- data.frame(
  life = life,
  material = factor(rep(1:3, each = 12)),
  temperature = factor(rep(rep(c(15, 70, 125), each = 4), 3))
)

## Preview the data
battery_df
```

## Exploratory Data Analysis and Visualization

Before fitting a formal model, we visualize the data to get an intuition for the relationships between the factors and the response.

## Boxplots of Main Effects

Boxplots are excellent for examining the distribution of battery life for each level of our factors independently. This gives us a preliminary look at the **main effects**—the individual impact of material type and temperature.

```{r}
#| label: plot-boxplots
#| fig-cap: "Distribution of Battery Life by Material and Temperature."
#| layout-ncol: 2

library(ggplot2)

## Boxplot for Material Type
ggplot(battery_df, aes(x = material, y = life, fill = material)) +
  geom_boxplot() +
  labs(title = "Battery Life by Material Type", x = "Material Type", y = "Life (hours)") +
  theme_minimal() +
  theme(legend.position = "none")

## Boxplot for Temperature
ggplot(battery_df, aes(x = temperature, y = life, fill = temperature)) +
  geom_boxplot() +
  labs(title = "Battery Life by Temperature", x = "Temperature (°C)", y = "Life (hours)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Interaction Plot

The most crucial plot for a factorial experiment is the **interaction plot**. It displays the mean battery life for each combination of material and temperature. If the lines are parallel, it suggests there is no interaction. If the lines are not parallel (i.e., they cross or diverge), it indicates that the effect of temperature on battery life is different for each material type, signaling a likely interaction.

```{r}
#| label: plot-interaction
#| fig-cap: "Interaction between Material Type and Temperature."

ggplot(battery_df, aes(x = temperature, y = life, group = material, color = material)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  labs(
    title = "Interaction Plot: Material Type and Temperature",
    x = "Temperature (°C)",
    y = "Average Battery Life (hours)",
    color = "Material Type"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


The interaction plot reveals a strong **interaction effect** between Material Type and Temperature, meaning the performance profile is distinct for each chemistry:

* **Material 1 (Standard Li-ion):**
    * This material is suitable **only for cool environments**. While it provides adequate life at $15^\circ\text{C}$ (~135 hours), it suffers a catastrophic drop in performance at $70^\circ\text{C}$ and stays consistently poor at $125^\circ\text{C}$ (~55 hours). It is not viable for high-temperature operations.

* **Material 2 (Industrial Li-MnO$_2$):**
    * This is the **superior choice for cold starts**, offering the highest lifespan of all materials at $15^\circ\text{C}$ (~155 hours). However, it degrades linearly and severely as temperature rises, ultimately becoming the worst performer at extreme heat ($125^\circ\text{C}$), dropping to ~50 hours.

* **Material 3 (High-Temp Li-SOCl$_2$):**
    * This material demonstrates **exceptional thermal stability**. Unlike the others, its performance actually peaks at the operating temperature of $70^\circ\text{C}$. Crucially, it is the **only viable option** for extreme heat ($125^\circ\text{C}$), maintaining a usable lifespan of ~85 hours when others have failed.

**Summary:**
There is no single "best" material overall. The optimal choice depends entirely on the environment: **Material 2** for cold applications and **Material 3** for any environment exceeding room temperature.



## Model Fitting and Analysis of Variance (ANOVA)

We now fit a linear model to formally test the significance of the main effects and the interaction term. The model `life ~ material * temperature` is shorthand for `life ~ material + temperature + material:temperature`. We use a sum-to-zero contrast (`contr.sum`) for balanced interpretation of the effects. The **ANOVA table** will tell us if the variation caused by our factors is statistically significant compared to the random variation in the data.

```{r}
#| label: fit-model

## Fit the full factorial model
battery_fit <- lm(life ~ material * temperature, 
                  data = battery_df,
                  contrasts = list(material = contr.sum, temperature = contr.sum))

summary(battery_fit)

## Generate the ANOVA table
anova(battery_fit)
```

The ANOVA table shows very small p-values (`Pr(>F)`) for `material`, `temperature`, and, most importantly, the `material:temperature` interaction. This confirms our visual inspection: all effects are statistically significant. **Because the interaction is significant, our interpretation should focus on the interaction itself rather than the main effects in isolation.**

## Model Adequacy Checks

The validity of our ANOVA results depends on the model's residuals meeting certain assumptions (normality, constant variance, independence). We check these with diagnostic plots.

```{r}
#| label: plot-diagnostics
#| fig-cap: "Diagnostic plots for the battery life model."
#| layout-ncol: 2

## Extract standardized residuals and fitted values
battery_fit_diag <- data.frame(
  residuals = rstandard(battery_fit),
  fitted = fitted.values(battery_fit)
)

## Normal Q-Q Plot
p1 <- ggplot(battery_fit_diag, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot", x = "Theoretical Quantiles", y = "Standardized Residuals") +
  theme_minimal()

## Residuals vs. Fitted Plot
p2 <- ggplot(battery_fit_diag, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted Values", x = "Fitted Values", y = "Standardized Residuals") +
  theme_minimal()

p1 
p2
```

The Normal Q-Q plot shows the points falling roughly along the line, suggesting the normality assumption is met. The Residuals vs. Fitted plot shows a random scatter of points around the zero line, indicating that the variance is reasonably constant. The model assumptions appear to be satisfied.

## Post-Hoc Analysis: Pairwise Comparisons

Since the interaction is significant, we must compare the means of the nine specific treatment combinations (3 materials × 3 temperatures). Simply comparing the average effect of Material 1 vs. Material 2 would be misleading, as that difference depends on the temperature.

## Tukey's HSD Test

**Tukey's Honest Significant Difference (HSD)** test is a post-hoc test that compares all possible pairs of means while controlling the family-wise error rate. We apply it to an `aov` model object. The output for the `material:temperature` interaction shows which specific combinations are significantly different from one another.

```{r}
#| label: test-tukey

## Fit the model using aov() for Tukey's test
battery_aov <- aov(life ~ material * temperature, data = battery_df)

## Perform Tukey's HSD test
TukeyHSD(battery_aov)
```

## Fisher's LSD Method

The **Fisher's Least Significant Difference (LSD)** method is another option for pairwise comparisons. To test the interaction means, we must specify both factors in the `trt` argument.

```{r}
#| label: test-lsd

library(agricolae)

## Perform LSD test on the interaction term
lsd_results <- LSD.test(battery_aov, trt = c("material", "temperature"),
                        p.adj = "none", group = FALSE)

## Print the comparison table
print(lsd_results$comparison)
```

The results from both Tukey's HSD and Fisher's LSD provide detailed p-values for comparing pairs of treatment combinations, allowing us to make specific conclusions, such as "at 125°C, Material 3 has a significantly longer life than Materials 1 and 2."