---
filters: 
  - ../resources/coursetable.lua
engine: knitr
metadata-files:
  - ../_website.yaml 
format: 
  html: default
---

```{r}
#| label: setup-ritem
#| include: false
#| cache: true

renv::load("~/Github/longhaiSK.github.io/")
.current_item_count <- 0

ritem <- function(n = NULL) {
  if (!is.null(n)) {
    .current_item_count <<- n
  }
  
  num <- .current_item_count
  .current_item_count <<- .current_item_count - 1
  
  # Fetch the pandoc target format safely
  out_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
  
  # The \n\n forces Pandoc to start a brand new paragraph for EVERY item
  if (knitr::is_latex_output()) {
    out <- sprintf("\n\n\\hangindent=2.5em \\hangafter=1 \\makebox[2.5em][l]{%d.}", num)
    return(knitr::asis_output(out))
  } else if (knitr::is_html_output()) {
    out <- sprintf("\n\n<span class='ritem-num'>[%d]</span>", num)
    return(knitr::asis_output(out))
  } else if (!is.null(out_format) && out_format == "typst") {
    # Emit a raw inline "N. " at the start of a new paragraph.
    # In the .typ output this line begins with e.g. "31. ", which Typst
    # parses as an explicitly-numbered enum item. The following text
    # becomes the item body -> native hanging indent on wrapped lines.
    out <- sprintf("\n\n`%d. `{=typst}", num)
    return(knitr::asis_output(out))
  } else {
    return(paste0("\n\n[", num, "]"))
  }
}
```

```{=typst}
#align(right)[#text(size: 14pt, weight: "bold")[Form 1]]
#v(1em)
#grid(
  columns: (1fr, 1fr),
  row-gutter: 1em,
  [*Name:* #underline[Longhai Li]], [*Date:* #underline[June 30, 2026]],
  [*College:* #underline[Arts & Science]], [*Department:* #underline[Mathematics & Statistics]]
)
#v(0.5em)
#line(length: 100%)
#v(1em)
#align(center)[
  #strong[
    INFORMATION FOR UPDATE OF CURRICULUM VITAE -- NEW ITEMS \
    (ITEMS NOT PREVIOUSLY REPORTED) \
    The cut off date for items to be reported is JUNE 30.
  ]
]
```

## 9. TEACHING ACTIVITIES

### 9.1 Scheduled Instructional Activity

 
::: {.course-tables}

**2025-2026**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|-----|--|-----------|--|--|--|--|
| STAT 851 | T2 | Linear Statistical Models | LEC | 4 | 39 | 156 |
| STAT 443 | T2 | Linear Models | LEC | 3 | 39 | 117 |
| STAT 442 | T2 | Statistical Inference | LEC | 3 | 39 | 117 |
| STAT 850 | T2 | Mathematical Statistics and Inference | LEC | 1 | 39 | 39 |
| STAT 845 | T1 | Statistical Methods for Research | LEC | 10 | 39 | 390 |
| MATH 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |

:::

### 9.4 Teaching Materials

**2025-2026**

`r ritem(3)` STAT 442/850: a textbook entitled "Theory of Statistical Inference" in PDF ($\ge$ 200 pages) and HTML.
`r ritem()`  STAT 845: a textbook entitled "Statistical Inference and Learning Methods for Research" in PDF ($\ge$ 200 pages) and HTML
`r ritem()`  STAT 443/851: a textbook entitled "Theory of Linear Model" in PDF ($\ge$ 200 pages) and HTML.

### 9.5 Other Teaching-Related Activities

**2025-2026**

`r ritem(1)` Peer evaluation for Matthew Schmirler, Mar. 2026.

## 10. SUPERVISION AND ADVISORY ACTIVITIES

### 10.1 Undergraduate Student Supervision

**2025-2026**

`r ritem(2)` George Chen, B.Sc., Computer Science, Simon Fraser University, June 1, 2025 -- Aug. 31, 2025, Supervised.
`r ritem()` Shruti Kaur, B.Sc., Computer Science and Statistics, May -- Aug., 2025, Supervised.

### 10.2 Graduate Student Supervision

`r ritem(2)` Jing Wang, Ph.D., Biostatistics, School of Public Health, Co-supervised with Prof. Li Xing, 2023--2026 (Transferred to other supervisor)
`r ritem()` Dananji Egodage, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Cindy Feng, 2023--2025 (Defended: Aug. 30, 2025)

### 10.3 Graduate Theses Supervised

**2025-2026**

`r ritem(1)` Dananji Egodage, 2025, Component-wise Z-residuals for Checking Bayesian Hurdle Models, M.Sc., defended on Aug. 30, 2025.

   

### 10.4 Supervision of Post-Doctoral Fellows and Research Associates

`r ritem(2)` Tingxuan Wu, University of Saskatchewan, Postdoc Research Associate,  Jan. 2025 – July 2026.
`r ritem()` Ming Ming Zhang, MITACS Postdoc, 2022-2025.

### 10.6 Thesis Committee Memberships

```{r}
#| label: tbl-thesis-committee
#| tbl-cap: "Thesis Committee Memberships"
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(gt)

# Create the dataframe using tribble for easy row-by-row data entry
committee_data <- tribble(
  ~`#`, ~NAME, ~DEG, ~PROGRAM, ~`TIME FRAME`, ~ROLE,
  31, "Tiansui Wu", "M.Sc.", "Biostatistics", "2025--Present", "Chair",
  30, "Prabhawi Kahatapitiye", "Ph.D.", "Statistics", "2025--Present", "Member",
  29, "Rasel Kabir", "Ph.D.", "Biostatistics", "2025--Present", "Chair",
  28, "Mohammad Toranjsimin", "M.Sc.", "Biostatistics", "2023--2025 (Def. 09/2025)", "Member",
  27, "Lina Li", "Ph.D.", "Biostatistics", "2023--Present", "Member",
  25, "Mangladeep Bhullar", "Ph.D.", "Physics", "2023--2025 (Def. 11/2025)", "Cognate",
  22, "Han Wang", "Ph.D.", "Sociology", "2020--Present", "Cognate"
)

# Render the gt table
committee_data |> 
  gt() |> 
  tab_options(
    table.width = pct(100),
    
    table.layout = "fixed",
    table.font.size = px(12), 
    container.overflow.x = TRUE,
    # --- remove all lines ---
    table.border.top.style = "hidden",
    table.border.bottom.style = "hidden",
    column_labels.border.top.style = "hidden",
    column_labels.border.bottom.style = "hidden",
    table_body.hlines.style = "none",
    table_body.vlines.style = "none",
    table_body.border.top.style = "hidden",
    table_body.border.bottom.style = "hidden",
    stub.border.style = "hidden"
  ) |> 
  cols_width(
    `#` ~ pct(5),
    NAME ~ pct(25),
    DEG ~ pct(5),
    PROGRAM ~ pct(20),
    `TIME FRAME` ~ pct(30),
    ROLE ~ pct(10)
  ) |>  
  opt_row_striping() |> 
  # --- gray header background + bold ---
  tab_style(
    style = list(
      cell_fill(color = "#d9d9d9"),
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  ) |> 
  cols_align(
    align = "center",
    columns = c(`#`, DEG, ROLE)
  ) 
```

## 12. PAPERS IN REFERRED JOURNALS

**2025-2026**
 
`r ritem(2)` Wu, T., Li, L., and Feng, C., $Z$-residuals Diagnostics for Cox Proportional Hazards Models: Distinguishing Functional Form Misspecification from Nonproportional Hazards, with an Application to Biliary Cirrhosis Survival Times, *Canadian Journal of Statistics*, Accepted on June 5, 2026.
`r ritem()` Nolan, J., Su, C., Li, L., 2025. Evaluating Railroad Duopoly Behavior: A Market Level Analysis. *Review of Network Economics* 24, 87–111. [https://doi.org/10.1515/rne-2025-0034](https://doi.org/10.1515/rne-2025-0034)
                   

  

## 14. PRESENTATIONS

### 14.1 Invited Presentations

**2025-2026**

`r ritem(2)` Z-residuals for Checking Bayesian Models. Presented at: University of Calgary, Calgary, AB, Canada; July 28, 2025
`r ritem()` Sparse Learning for Assessing the Association Between Gut Microbiome and Parkinson’s Disease. Presented at: The 3rd JCSDS, Hangzhou, China, July 13, 2025.

## 15. REPORTS AND OTHER OUTPUTS

### 15.1 Software Released Publicly

`r ritem(4)` Wu, T. and Li, L., 2026. `Zresidual`: Computing and Diagnosing Gaussian-like Residuals. [[Github]](https://tiw150.github.io/Zresidual/index.html) [[Demo]](https://tiw150.github.io/Zresidual_demo.html). Version 0.1-0 on Github (2026).
`r ritem()` Li, L., 2026. R Functions for Computing Z-residuals for `survreg` and `coxph` Objects. [[URL]](https://longhaisk.github.io/software/NRSP/index.html).
`r ritem()` Li, L. and Liu, S., 2019--2026. `HTLR`: Bayesian Logistic Regression with Hyper-LASSO priors. DOI: 10.32614/CRAN.package.HTLR. [[CRAN]](https://cran.r-project.org/web/packages/HTLR/index.html) [[Github]](https://longhaisk.github.io/HTLR) [[URL]](https://longhaisk.github.io/software/BLRHL/index.html). Version 0.4 (2019), version 0.4-1 (2019), version 0.4-2 (2020), version 0.4-3 (2020), version 0.4-4 (2022), version 1.0 (2026).
`r ritem()` Li, L., 2011--2026. `BCBCSF`: Bias-corrected Bayesian Classification with Selected Features. DOI: 10.32614/CRAN.package.BCBCSF.  [[CRAN]](https://cran.r-project.org/web/packages/BCBCSF/index.html) [[URL]](https://longhaisk.github.io/software/BCBCSF/index.html). Version 0.0-0 (2011), version 0.0-1 (2011), version 0.0-2 (2012), version 1.0-0 (2013), version 1.0-1 (2015), updated to version 1.0-2 (2026).

## 17. RESEARCH FUNDING HISTORY

**2025-2026**

`r ritem(2)` **NSERC Individual Discovery Grant (No. 2026-07053)** – *Prediction-based Methods for Statistical Learning and Inference in Biosciences and Epidemiology*, $185,000 (37K per year), 2026-2031, PI.
`r ritem()` **CANSSI Collaborative Research Team Projects** – *Statistical Methodologies and Computational Tools to Identify Microbial Correlates of Canadian Bee Gut Health*, Project 29, 2025-2028, Co-PI.

## 18. PRACTICE OF PROFESSIONAL SKILLS

**2025-2026**

`r ritem(12)` Refereeing for *Journal of Statistical Computation and Simulation*, June, 2026
`r ritem()` Refereeing for *Journal of Statistical Computation and Simulation*, April, 2026
`r ritem()` Refereeing for *Journal of the Royal Statistical Society: Series C*, April 2026
`r ritem()` Refereeing for *Bioinformatics*, March 2026
`r ritem()` Refereeing for *Journal of Computational and Graphical Statistics*, March 2026
`r ritem()` External Referee for a tenure and promotion of SFU, Dec. 2025
`r ritem()` External Examiner for the doctoral thesis by Xiaoqing Zhang at University of Regina, Dec. 8, 2025
`r ritem()` Refereeing for *Journal of Statistical Computation and Simulation*, Dec. 2025
`r ritem()` Refereeing for *Journal of Computational and Graphical Statistics*, Sept. 2025
`r ritem()` External Examiner for the doctoral thesis by Na Zhang at University of Alberta, August 28, 2025
`r ritem()` Refereeing for *Journal of the Royal Statistical Society: Series C*, August 2025
`r ritem()` Refereeing for *Journal of Applied Statistics*, August 2025

## 19. ADMINISTRATIVE SERVICE

### 19.1 University Committees

**2025-2026**

`r ritem(2)` Chair, Collaborative Biostatistics Program, University of Saskatchewan.
`r ritem()` USASK NSERC Discovery Grant (DG), Internal Reviewer.

### 19.2 College and Departmental Committees

**2025-2026**

`r ritem(3)` Member, Budgeting and Planning Committee, Dept of Math & Stat
`r ritem()` Co-Chair, Undergraduate Committee (Statistics), Dept of Math & Stat
`r ritem()` Member, Graduate Program Committee in Statistics, Dept of Math & Stat

```{=typst}
#colbreak()

#align(right)[#text(size: 14pt, weight: "bold")[Form 2]]
#v(1em)
#grid(
  columns: (1fr, 1fr),
  row-gutter: 1em,
  [*Name:* #underline[Longhai Li]], [*Date:* #underline[June 30, 2026]],
  [*College:* #underline[Arts & Science]], [*Department:* #underline[Mathematics & Statistics]]
)
#v(0.5em)
#line(length: 100%)
#v(1em)
#align(center)[
  #strong[
    INFORMATION FOR UPDATE OF CURRICULUM VITAE \
    (REVISION OF ITEMS PREVIOUSLY REPORTED AND CONSIDERED) \
    #v(0.5em)
    The cut off date for items to be reported is JUNE 30.
  ]
]
#v(2em)
#align(center)[*Nothing to report for timeframe: 2025-2026*]
```

