---
filters: 
  - ../resources/coursetable.lua
engine: knitr
metadata-files:
  - ../_website.yaml 
format: 
  html: default
---

```{r}
#| label: setup-ritem
#| include: false
#| cache: false

renv::load("~/Github/longhaiSK.github.io/")
.current_item_count <- 0

ritem <- function(n = NULL) {
  if (!is.null(n)) {
    .current_item_count <<- n
  }
  
  num <- .current_item_count
  .current_item_count <<- .current_item_count - 1
  
  # Fetch the pandoc target format safely
  out_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
  
  # The \n\n forces Pandoc to start a brand new paragraph for EVERY item
  if (knitr::is_latex_output()) {
    out <- sprintf("\n\n\\hangindent=2.5em \\hangafter=1 \\makebox[2.5em][l]{%d.}", num)
    return(knitr::asis_output(out))
  } else if (knitr::is_html_output()) {
    out <- sprintf("\n\n<span class='ritem-num'>[%d]</span>", num)
    return(knitr::asis_output(out))
  } else if (!is.null(out_format) && out_format == "typst") {
    # Emit a raw inline "N. " at the start of a new paragraph.
    # In the .typ output this line begins with e.g. "31. ", which Typst
    # parses as an explicitly-numbered enum item. The following text
    # becomes the item body -> native hanging indent on wrapped lines.
    out <- sprintf("\n\n`%d. `{=typst}", num)
    return(knitr::asis_output(out))
  } else {
    return(paste0("\n\n[", num, "]"))
  }
}
```

```{=typst}
#align(right)[#text(size: 14pt, weight: "bold")[Form 1]]
#v(1em)
#grid(
  columns: (1fr, 1fr),
  row-gutter: 1em,
  [*Name:* #underline[Longhai Li]], [*Date:* #underline[June 30, 2024]],
  [*College:* #underline[Arts & Science]], [*Department:* #underline[Mathematics & Statistics]]
)
#v(0.5em)
#line(length: 100%)
#v(1em)
#align(center)[
  #strong[
    INFORMATION FOR UPDATE OF CURRICULUM VITAE -- NEW ITEMS \
    (ITEMS NOT PREVIOUSLY REPORTED) \
    The cut off date for items to be reported is JUNE 30.
  ]
]
```

## 9. TEACHING ACTIVITIES

### 9.1 Scheduled Instructional Activity

 
::: {.course-tables}

**2023-2024**

| COURSE | TERM | TITLE | TYPE | ENRL | YIH | YCSH |
|-----|--|-----------|--|--|--|--|
| STAT 245 | T1 | Introduction to Statistical Methods | LEC | 91 | 39 | 3549 |
| STAT 812 | T1 | Computational Statistics | LEC | 5 | 39 | 195 |
| STAT 420 | T1 | Topics in Computational Statistics | LEC | 1 | 39 | 39 |
| STAT 443 | T2 | Linear Statistical Models | LEC | 4 | 39 | 156 |
| STAT 851 | T2 | Linear Models | LEC | 4 | 39 | 156 |
| BIOS 996 | T1T2 | Research Supervision (Ph.D.) | RES | 1 | | |
| BIOS 996 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |
| MATH 994 | T1T2 | Research Supervision (M.Sc.) | RES | 1 | | |

:::

### 9.5 Other Teaching-Related Activities

**2023-2024**

`r ritem(1)` Data Science Bootcamp, a case study, June 19, 2023.

## 10. SUPERVISION AND ADVISORY ACTIVITIES

### 10.1 Undergraduate Student Supervision

**2023-2024**

`r ritem(1)` George Chen, B.Sc., Computer Science, Simon Fraser University, May 7, 2023 -- Aug. 25, 2023, Supervised.

### 10.2 Graduate Student Supervision

`r ritem(3)` Jing Wang, Ph.D., Biostatistics, School of Public Health, Co-supervised with Prof. Li Xing, 2023--2026 (Transferred to other supervisor)
`r ritem()` Dananji Egodage, M.Sc., Statistics, Math & Stat, Co-supervised with Prof. Cindy Feng, 2023--2025 (Defended: Aug. 30, 2025)
`r ritem()` Wuqian Effie Gao, M.Sc., Biostatistics, School of Public Health, Co-supervised with Prof. Cindy Feng, 2022--2024 (Defended: Aug. 30, 2024)

### 10.3 Graduate Theses Supervised

**2023-2024**

`r ritem(1)` Tingxuan Wu, 2023, Residual Diagnostics and Statistical Inference for Shared Frailty Models, Ph.D., defended on May 24, 2023.

   

### 10.4 Supervision of Post-Doctoral Fellows and Research Associates

`r ritem(2)` Tingxuan Wu, University of Saskatchewan, Postdoc Fellow, June 2023 – April 2024.
`r ritem()` Ming Ming Zhang, MITACS Postdoc, 2022-2025.

### 10.6 Thesis Committee Memberships

```{r}
#| label: tbl-thesis-committee
#| tbl-cap: "Thesis Committee Memberships"
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(gt)

# Create the dataframe using tribble for easy row-by-row data entry
committee_data <- tribble(
  ~`#`, ~NAME, ~DEG, ~PROGRAM, ~`TIME FRAME`, ~ROLE,
  28, "Mohammad Toranjsimin", "M.Sc.", "Biostatistics", "2023--2025 (Def. 09/2025)", "Member",
  27, "Lina Li", "Ph.D.", "Biostatistics", "2023--Present", "Member",
  26, "Kyle Gardiner", "M.Sc.", "Statistics", "2023--2024 (Def. 09/2024)", "Member",
  25, "Mangladeep Bhullar", "Ph.D.", "Physics", "2023--2025 (Def. 11/2025)", "Cognate",
  24, "Hammed Jimoh", "M.Sc.", "Statistics", "2022--2023", "Member",
  22, "Han Wang", "Ph.D.", "Sociology", "2020--Present", "Cognate"
)

# Render the gt table
committee_data |> 
  gt() |> 
  tab_options(
    table.width = pct(100),
    
    table.layout = "fixed",
    table.font.size = px(12), 
    container.overflow.x = TRUE,
    # --- remove all lines ---
    table.border.top.style = "hidden",
    table.border.bottom.style = "hidden",
    column_labels.border.top.style = "hidden",
    column_labels.border.bottom.style = "hidden",
    table_body.hlines.style = "none",
    table_body.vlines.style = "none",
    table_body.border.top.style = "hidden",
    table_body.border.bottom.style = "hidden",
    stub.border.style = "hidden"
  ) |> 
  cols_width(
    `#` ~ pct(5),
    NAME ~ pct(25),
    DEG ~ pct(5),
    PROGRAM ~ pct(20),
    `TIME FRAME` ~ pct(30),
    ROLE ~ pct(10)
  ) |>  
  opt_row_striping() |> 
  # --- gray header background + bold ---
  tab_style(
    style = list(
      cell_fill(color = "#d9d9d9"),
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  ) |> 
  cols_align(
    align = "center",
    columns = c(`#`, DEG, ROLE)
  ) 
```

## 12. PAPERS IN REFERRED JOURNALS

 
                   

  
**2023-2024**

`r ritem(1)` Feng, C., Li, L., Xu, C., 2023. Advancements in predicting and modeling rare event outcomes for enhanced decision-making. *BMC Medical Research Methodology* 23, Article 243 (pp. 1-3). [https://doi.org/10.1186/s12874-023-02060-x](https://doi.org/10.1186/s12874-023-02060-x)

## 14. PRESENTATIONS

### 14.1 Invited Presentations

**2023-2024**

`r ritem(4)` Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Proportional Hazards Models with Shared Frailty, ICSA Canada Chapter Symp., June 9, 2024, Niagara Falls, Canada
`r ritem()` Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Proportional Hazards Models with Shared Frailty, Annual Meeting of SSC, St John’s, Canada, June 2, 2024
`r ritem()` Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Proportional Hazards Models with Shared Frailty, Dept. Seminar, Texas State University, USA, March 8, 2024
`r ritem()` Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Proportional Hazards Models with Shared Frailty, Dept. Seminar, Sun Yat-sen University, China, Jan. 4, 2024

## 15. REPORTS AND OTHER OUTPUTS

### 15.1 Software Released Publicly

`r ritem(2)` Li, L. and Liu, S., 2019--2026. `HTLR`: Bayesian Logistic Regression with Hyper-LASSO priors. DOI: 10.32614/CRAN.package.HTLR. [[CRAN]](https://cran.r-project.org/web/packages/HTLR/index.html) [[Github]](https://longhaisk.github.io/HTLR) [[URL]](https://longhaisk.github.io/software/BLRHL/index.html). Version 0.4 (2019), version 0.4-1 (2019), version 0.4-2 (2020), version 0.4-3 (2020), version 0.4-4 (2022), version 1.0 (2026).
`r ritem()` Li, L., 2011--2026. `BCBCSF`: Bias-corrected Bayesian Classification with Selected Features. DOI: 10.32614/CRAN.package.BCBCSF.  [[CRAN]](https://cran.r-project.org/web/packages/BCBCSF/index.html) [[URL]](https://longhaisk.github.io/software/BCBCSF/index.html). Version 0.0-0 (2011), version 0.0-1 (2011), version 0.0-2 (2012), version 1.0-0 (2013), version 1.0-1 (2015), updated to version 1.0-2 (2026).

### 15.2 Technical Reports

`r ritem(2)` Wu, T., Feng, C. and Li, L., 2023. Cross-validatory Z-Residual for Diagnosing Shared Frailty Models. [https://doi.org/10.48550/arXiv.2303.09616](https://doi.org/10.48550/arXiv.2303.09616). 32 pages, 14 figures.
`r ritem()` Wu, T., Li, L. and Feng, C., 2023. Z-residual diagnostics for detecting misspecification of the functional form of covariates for shared frailty models. [https://doi.org/10.48550/arXiv.2302.09106](https://doi.org/10.48550/arXiv.2302.09106). 21 pages, 7 figures.

## 18. PRACTICE OF PROFESSIONAL SKILLS

**2023-2024**

`r ritem(4)` External Examiner for the doctoral thesis by Yuping Yang at SFU, June 25, 2024
`r ritem()` Refereeing for *Journal of Computational and Graphical Statistics*, Sept. 2024
`r ritem()` Refereeing for *Journal of Applied Statistics*, Jan. 2024
`r ritem()` Review an MITACS Accelerate Grant, Dec. 2023

## 19. ADMINISTRATIVE SERVICE

### 19.2 College and Departmental Committees

**2023-2024**

`r ritem(2)` Member, Graduate Program Committee in Statistics, Dept of Math & Stat
`r ritem()` Member, Sub Search Committee, 4-Year Lecturer Position

## 20. PROFESSIONAL OR ASSOCIATION OFFICES AND COMMITTEE ACTIVITY

**2023-2024**

`r ritem(1)` Member of NSERC Discovery Grant EG 1508 Committee

```{=typst}
#colbreak()

#align(right)[#text(size: 14pt, weight: "bold")[Form 2]]
#v(1em)
#grid(
  columns: (1fr, 1fr),
  row-gutter: 1em,
  [*Name:* #underline[Longhai Li]], [*Date:* #underline[June 30, 2024]],
  [*College:* #underline[Arts & Science]], [*Department:* #underline[Mathematics & Statistics]]
)
#v(0.5em)
#line(length: 100%)
#v(1em)
#align(center)[
  #strong[
    INFORMATION FOR UPDATE OF CURRICULUM VITAE \
    (REVISION OF ITEMS PREVIOUSLY REPORTED AND CONSIDERED) \
    #v(0.5em)
    The cut off date for items to be reported is JUNE 30.
  ]
]
#v(2em)
#align(center)[*Nothing to report for timeframe: 2023-2024*]
```

