---
title: "One-factor Design"
author:
  - name: "Longhai Li"
    affiliation: "University of Saskatchewan"
    url: "https://longhaisk.github.io/"
date: today
editor: source
format: 
  html
execute:
  echo: true
  warning: false
  message: false
  cache: false
  fig-width: 9
  fig-height: 6
editor_options: 
  chunk_output_type: console
---

## Completely Randomized Design
### Plasma Etching Experiment

This section analyzes data from a **Completely Randomized Design (CRD)**. In a CRD, experimental units (in this case, the silicon wafers being etched) are assigned to treatments (the RF Power levels) completely at random. The primary goal is to determine if changing the RF Power level has a statistically significant effect on the mean etch rate.

#### Data and Visualization

We begin by loading the data into a single, tidy `data.frame`. The response variable, `rate`, contains all the etch rate observations. The predictor variable, `power`, is a **factor**, which is R's way of representing a categorical variable. This tells R to treat the different power levels as distinct groups.

```{r}
#| label: setup-crd
#| echo: true

## Define the data vectors
rate <- c(575, 542, 530, 539, 570, 565, 593, 590, 579, 610,
          600, 651, 610, 637, 629, 725, 700, 715, 685, 710)
power_levels <- c(160, 180, 200, 220)

## Create the data frame
etching_df <- data.frame(
  rate = rate,
  power = factor(rep(power_levels, each = 5))
)

## Display the first few rows
etching_df
```

**Grouped Boxplots**

```{r}
boxplot(rate~power, data=etching_df)

```
**Using ggplot to visualize grouped data**
```{r}
#| label: index-plot-rate-power-segments
#| fig-cap: "Index Plot of Etch Rate with Group-Specific Mean Lines"

library(ggplot2)
library(dplyr) # Using dplyr for easier data manipulation

## Calculate group means and their start/end indices
mean_rates <- etching_df %>%
  mutate(obs_index = row_number()) %>%
  group_by(power) %>%
  summarise(
    mean_rate = mean(rate),
    x_start = min(obs_index) - 0.5,
    x_end = max(obs_index) + 0.5
  )

ggplot(etching_df, aes(x = 1:nrow(etching_df), y = rate, color = power)) +
  geom_point(size = 3, alpha = 0.7) + # Plot individual data points
  geom_segment(
    data = mean_rates, 
    aes(x = x_start, xend = x_end, y = mean_rate, yend = mean_rate),
    linetype = "dashed", 
    size = 1.2
  ) + # Add line segments for group means
  labs(
    title = "Etch Rate Observations by RF Power Level",
    x = "Observation Index",
    y = "Etch Rate",
    color = "RF Power (W)"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

#### Model Fitting with Sum-to-Zero Constraint

We fit a linear model using the `lm()` function to perform an **Analysis of Variance (ANOVA)**. The model is specified as `rate ~ power`, and we now include the `data = etching_df` argument.

To get interpretable estimates for the treatment effects ($\tau_i$), we use a **sum-to-zero constraint** (`contr.sum`), which forces the sum of the treatment effects to be zero ($\sum \tau_i = 0$).

```{r}
#| label: fit-sum-to-zero

fit <- lm(rate ~ power, data = etching_df, contrasts = list(power = contr.sum))
cat ("Model Matrix:\n")
model.matrix(fit)
summary.fit <- summary(fit)
cat ("Summary of lm fitting results:\n")

summary.fit
```

#### Point Estimation of Parameters

The output of the model provides estimates for the overall mean ($\hat{\mu}$) and the treatment effects for the first k-1 levels ($\hat{\tau}_1, \hat{\tau}_2, \hat{\tau}_3$).

* $\hat{\mu}$ (the Intercept) is the estimate of the grand mean etch rate across all power levels.
* $\hat{\tau}_i$ is the estimated effect of the i-th power level, representing how much that level's mean deviates from the grand mean.

Using the sum-to-zero constraint, we can manually calculate the effect for the final level, $\hat{\tau}_4$.

```{r}
#| label: point-estimation

## Extract coefficients
est <- coef(fit)
tau4.hat <- -sum(est[-1])
taui.hat <- c(est[-1], tau4.hat)
print("Estimated Treatment Effects (tau_i):")
print(taui.hat)

## Estimates of treatment means (mu_i)
mu.hat <- est[1]
mui.hat <- mu.hat + taui.hat
print("Estimated Treatment Means (mu_i):")
print(mui.hat)
```

#### ANOVA Table

The ANOVA table partitions the total variation into variation **between** treatment groups (`power`) and variation **within** treatment groups (random error). The **p-value** (Pr(>F)) indicates if the treatment has a significant effect.

```{r}
#| label: anova-crd

anova(fit)
```

#### 95% Confidence Intervals for Treatment Means

A **confidence interval** provides a range of plausible values for the true mean etch rate at each power level.

```{r}
#| label: confidence-intervals

## Number of replicates
n <- 5 
## Extract sqrt(MSE) and error df
sqrt.MSE <- summary.fit$sigma
DF <- fit$df.residual
## Find t-value
t.value <- qt(0.975, DF)
## Calculate CIs
CI.lower <- mui.hat - t.value * sqrt.MSE / sqrt(n)
CI.upper <- mui.hat + t.value * sqrt.MSE / sqrt(n)

## Display CIs
data.frame(Power_Level = power_levels, Mean = mui.hat, Lower_CI = CI.lower, Upper_CI = CI.upper)
```

**Alternatively, one can use a model without intercept**
```{r}
fit_nointercpt <- lm(rate ~ 0+power, data = etching_df)
summary(fit_nointercpt)
confint(fit_nointercpt)
```

#### Comparison with Default "Treatment" Contrast

Fitting the model without specifying contrasts uses R's default ("treatment" contrast), which sets $\tau_1 = 0$. The fundamental results (ANOVA, treatment means) remain unchanged.

```{r}
#| label: fit-default-contrast

fit1 <- lm(rate ~ power, data = etching_df)
summary(fit1)
```

#### Pairwise Comparisons

Since our ANOVA result was significant, we perform **post-hoc tests** to determine exactly which pairs of power levels have different means.

##### Tukey's HSD Test

**Tukey's Honest Significant Difference (HSD)** controls the **family-wise error rate**, adjusting p-values to account for multiple comparisons.

```{r}
#| label: tukey-test

fit.aov <- aov(rate ~ power, data = etching_df)
TukeyHSD(fit.aov)
```

##### Fisher's LSD Test

The **Fisher's Least Significant Difference (LSD)** test does not control the family-wise error rate but is more powerful.

```{r}
#| label: fisher-lsd

with(etching_df, pairwise.t.test(rate, power, p.adj = "none"))
```

#### Checking Model Assumptions

The validity of our ANOVA results depends on three key assumptions about the model's residuals. We use diagnostic plots to check them.

```{r}
#| label: residuals-setup

r <- rstudent(fit)
fitted <- fitted.values(fit)
```

##### Normality of Residuals

A **Normal Q-Q plot** is used to check if the residuals are normally distributed. The points should fall closely along the straight diagonal line.

```{r}
#| label: qq-plot
#| fig-cap: "Normal Q-Q plot of standardized residuals."

qqnorm(r)
qqline(r)
```

##### Independence of Residuals

A plot of **residuals versus run order** helps check for independence. We look for random scatter around the zero line.

```{r}
#| label: residuals-vs-order
#| fig-cap: "Standardized residuals vs. run order."

plot(r, ylab = "Standardized residuals", xlab = "Run order",
     main = "Plot of residuals vs. run order")
abline(h = 0)
```

##### Constant Variance (Homoscedasticity)

A plot of **residuals versus fitted values** helps check for constant variance. The spread of residuals should be roughly constant across all fitted values.

```{r}
#| label: residuals-vs-fitted
#| fig-cap: "Standardized residuals vs. fitted values."

plot(fitted, r, ylab = "Standardized residuals", 
     xlab = "Fitted values", main = "Plot of residuals vs. fitted values")
abline(h = 0)
```

## Unbalanced Designs with unequal Sample Sizes

The ANOVA framework also handles **unbalanced designs**. We again start by creating a data frame.

```{r}
#| label: unequal-samples

## Create the data frame
bricks_df <- data.frame(
  density = c(21.8, 21.9, 21.7, 21.6, 21.7,
              21.7, 21.4, 21.5, 21.4,
              21.9, 21.8, 21.8, 21.6, 21.5,
              21.9, 21.7, 21.8, 21.4),
  temperature = factor(c(rep(100, 5), rep(125, 4), rep(150, 5), rep(175, 4)))
)

bricks_df

## Fit the model and get the ANOVA table
fit2 <- lm(density ~ temperature, data = bricks_df)
summary(fit2)
anova(fit2)
```

In this case, the large p-value (0.133) indicates that there is no statistically significant evidence that firing temperature affects brick density.

## Randomized Complete Block Design
### Vascular Graft Experiment 

This section analyzes a **Randomized Complete Block Design (RCBD)**, used to control for a known source of variability (here, "batches of resin," treated as **blocks**).

#### Data and Visulization

We structure the data in a `data.frame` to identify the response, treatment (`pressure`), and block (`batch`) for each observation.

```{r}
#| label: setup-rcbd

## Define data vectors
strength <- c(90.3, 89.2, 98.2, 93.9, 87.4, 97.9,
              92.5, 89.5, 90.6, 94.7, 87.0, 95.8,
              85.5, 90.8, 89.6, 86.2, 88.0, 93.4,
              82.5, 89.5, 85.6, 87.4, 78.9, 90.7)
pressure_levels <- rep(c(8500, 8700, 8900, 9100), each = 6)
batch_levels <- rep(1:6, 4)

## Create the data frame
graft_df <- data.frame(
  strength = strength,
  pressure = factor(pressure_levels),
  batch = factor(batch_levels)
)


graft_df
```

**Visualize the Block and Treatment Effects**

```{r}
par (mfrow = c(1,2))
#boxplot
plot(strength ~ batch, data=graft_df , main = "Block")
plot(strength ~ pressure, data=graft_df , main = "Pressure")
```


**Interaction Plots**



```{r}
#| label: plot-interaction
#| fig-cap: "Interaction between Material Type and Temperature."

ggplot(graft_df, aes(x = pressure, y = strength, group = batch, color = batch)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  labs(
    title = "Interaction Plot: Batch and Pressure",
    x = "Pressue",
    y = "Strength",
    color = "Batch"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


#### Model Fitting and ANOVA

The model `strength ~ pressure + batch` partitions the total variance into treatment, block, and error components. Our primary interest is in the significance of the `pressure` factor.

```{r}
#| label: fit-rcbd

rcbd.fit1 <- aov(strength ~ pressure + batch, data = graft_df)
anova(rcbd.fit1)
```

The small p-value for `pressure` (0.0019) provides strong evidence that extrusion pressure significantly affects graft strength after accounting for batch differences.

#### Model Adequacy Checks

The assumptions for an RCBD are the same as for a CRD. We perform the same diagnostic checks.

```{r}
#| label: plot-diagnostics-rcbd
#| layout-ncol: 2
#| fig-cap: 
#|  - "Normal Q-Q Plot"
#|  - "Residuals vs. Fitted Values"
#|  - "Residuals vs. Treatment (Pressure)"
#|  - "Residuals vs. Block (Batch)"

rcbd.r1 <- rstudent(rcbd.fit1)
rcbd.fitted1 <- fitted.values(rcbd.fit1)

qqnorm(rcbd.r1, main = "Normal Q-Q Plot")
qqline(rcbd.r1)

plot(rcbd.fitted1, rcbd.r1, ylab = "Standardized residuals", 
     xlab = "Fitted values", main = "Residuals vs. Fitted")
abline(h = 0)

plot(graft_df$pressure, rcbd.r1, ylab = "Standardized residuals", 
     xlab = "Extrusion pressure", main = "Residuals vs. Treatment")
abline(h = 0)

plot(graft_df$batch, rcbd.r1, ylab = "Standardized residuals", 
     xlab = "Batches of raw material", main = "Residuals vs. Block")
abline(h = 0)
```

#### Pairwise Comparisons

Again, since the treatment factor (`pressure`) is significant, we perform post-hoc tests.

##### Tukey's HSD Test

Tukey's HSD compares all pairs of treatment levels while controlling the family-wise error rate.

```{r}
#| label: tukey-rcbd

TukeyHSD(rcbd.fit1, which = "pressure")
```

##### Fisher's LSD Test

The `LSD.test()` function from the `agricolae` package correctly handles the error structure of an RCBD.

```{r}
#| label: lsd-rcbd
#| message: false

## install.packages("agricolae")
library(agricolae)

out <- LSD.test(rcbd.fit1, trt = "pressure", p.adj = "none", group = FALSE)
print(out$comparison)
```