---
title: "Unequal Probability Sampling"
author: "Longhai Li"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: no
    toc_depth: 2
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
    df_print: paged
  word_document:
    toc: yes
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
  pdf_document: 
    toc: yes
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---


\section{Used Functions or Packages}
# Functions and Packages
```{r}
## ydata --- observations of the variable of interest
## xdata --- observations of the auxilliary variable
## N --- population size

## the output is the ratio of ybarU/xbarU
srs_ratio_est <- function (ydata, xdata, N = Inf)
{	
  n <- length (xdata)
  xbar <- mean (xdata)
  ybar <- mean (ydata)
  B_hat <- ybar / xbar
  d <- ydata - B_hat * xdata
  var_d <- sum (d^2) / (n - 1)
  sd_B_hat <- sqrt ((1 - n/N) * var_d / n) / xbar
  mem <- qt (0.975, df = n - 1) * sd_B_hat
  output <- c (B_hat, sd_B_hat, B_hat - mem, B_hat + mem )
  
  
  names (output) <- c("Est.", "S.E.", "ci.low", "ci.upp" )
  output
}

## ydata --- observations of the variable of interest
## xdata --- observations of the auxilliary variable
## N --- population size
## xbarU --- population mean of auxilliary variable

## the output is the estimate mean or total (est.total=TRUE)
srs_reg_est <- function (ydata, xdata, xbarU, N = Inf, est.total = FALSE)
{
  n <- length (ydata)
  lmfit <- lm (ydata ~ xdata)
  Bhat <- lmfit$coefficients
  efit <- lmfit$residuals
  SSe <- sum (efit^2) / (n - 2) 
  yhat_reg <- Bhat[1] + Bhat[2] * xbarU
  se_yhat_reg <- sqrt ((1-n/N) * SSe / n)
  mem <- qt (0.975, df = n - 2) * se_yhat_reg
  output <- c(yhat_reg, se_yhat_reg, yhat_reg - mem, yhat_reg + mem)
  
  if (est.total) {
      if(!is.finite(N)) stop("N must be finite for estimating population total" )
      output <- output * N
  }

  names (output) <- c("Est.", "S.E.", "ci.low", "ci.upp" )
  output
}

## sdata --- a vector of original survey data
## N --- population size
## to find total, multiply N to the estimate returned by this function

srs_mean_est <- function (sdata, N = Inf)
{
	n <- length (sdata)
	ybar <- mean (sdata)
	se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
	mem <- qt (0.975, df = n - 1) * se.ybar
	c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}

# total estimation for datasets collected with UPS with Replacement
upswr_total_est <- function (total, psi)
{	
  srs_mean_est (total/psi, N = Inf)
}
# ratio estimation for datasets collected with UPS with Replacement
upswr_ratio_est <- function (total, M, psi)
{
  srs_ratio_est (total/psi, M/psi, N = Inf)
}

```

# Analysis of a dataset (student study hours) collected by UPSWR
```{r}

psi <- c(24, 100, 100, 76, 44) / 647
Mi <- c(24, 100, 100, 76, 44)
hrs <- c(75, 203, 203, 191, 168)

plot(hrs~Mi)
summary (lm(hrs~Mi))

## total hrs spent by all students

upswr_total_est (total = hrs, psi = psi)

## mean hrs spent by each student
upswr_ratio_est (total = hrs, M = Mi, psi = psi)

```

# Analysis of a dataset (statepop.csv) collected with one-stage UPSWR

The file "statepop.csv" is a dataset from an unequal-probability sample of 100 counties in the United States. Counties were chosen using the cumulative-size method from the listings in the County and City Data Book (U.S. Census Bureau, 1994) with probabilities proportional to their populations. The total population for all counties is $M_0 = \sum_{i=1} M_i = 255,077,536$. Sampling was done with replacement, so very large counties occur multiple times in the sample. For example, Los Angeles County, with the largest population in the United States, occurs four times.

```{r}
statepop <- read.csv ("data/statepop.csv")
statepop

M0 <- 255077536 ## pop of USA in 1992
popn <- statepop$popn ## pop per county in 1992
psi <- popn / M0
phys <- statepop$phys

plot (popn, phys, log="xy")

summary(lm (phys~psi))
# estimate the total number of physicians in USA
upswr_total_est (phys, psi)

# estimate number of physicians per capita
upswr_ratio_est (phys, popn, psi)
```

**Compare with an estimate based on an SRS sample (Assignment 3 Question 3)**

```{r}
# read srs sample of size 100
counties <- read.csv ("data/counties.csv", na = "-99")
counties
plot(physician~totpop, data=counties, log="xy")

summary (lm(physician~totpop, data=counties))
totpop <-  255077536
N <- 3141 # total number of counties
# naive estimate 
srs_mean_est (counties$physician, N = 3141) * 3141

# ratio estimate
srs_ratio_est (counties$physician, counties$totpop, N = 3141) * totpop

# regression estimate
srs_reg_est (counties$physician, counties$totpop, totpop/N, N = N) * N 

# From the textbook by Lohr, the true value is: 
total_physician <- 532638
```

# Analysis of a dataset (studenthrs.csv) collected with two-stage UPSWR
```{r}
studenthrs <- read.table ("data/studentshrs.txt", header = T)


## add sampling probability to the data
studenthrs$cpsi <- studenthrs$class_size/647
studenthrs
## data --- original data frame for holding data
## sid --- variable recording sample id of psu, note that, 
## sid must be different when the same psu is surveyed multiple times
## csize --- variable recording cluster (psu) population size (not sample size)
## cpik --- sampling probability for each cluster
## yvar --- variable of interest
## N --- total number of clusters (psus)

cluster_upswr_ratio_est <- function (data, sid, csize, cpsi, yvar,show.cluster.summary=TRUE)
{
    clust <- data[, sid]
    ydata <- data [, yvar]  
    
    ## cluster-wise summary
    ybari <- tapply (ydata, clust, mean)
    Mi <- tapply (data [,csize], clust, function (x) x[1])
    psi <- tapply (data [,cpsi], clust, function (x) x[1])
    ## the same as total in cluster if Mi = mi
    t_hat_cls <- ybari * Mi 
    
    if(show.cluster.summary){
        cat ("Cluster level Summary:\n")
        print(data.frame(Mi,psi,ybari, ti=t_hat_cls, ti_psi=t_hat_cls/psi,Mi_psi=Mi/psi))
    }
    
    ## apply ratio estimate to t_hat_cls and Mi
    
    srs_ratio_est (t_hat_cls/psi, Mi/psi)
}

cluster_upswr_ratio_est (studenthrs, "sid" , "class_size", "cpsi", "hrs")
```


# Simulation Studies for UPSWR using agpop data


```{r}
# Read population data
agpop <- read.csv ("data/agpop.csv", na = "-99")
## removing the counties with NA in either acres92 or acres87
agpop <- subset(agpop, subset = !is.na(agpop$acres92) & !is.na(agpop$acres87))

N <- nrow (agpop); N
total_acres92<- sum (agpop$acres92)
total_acres87 <- sum (agpop$acres87)

```

## Using acres87 to construct sampling probabilities
```{r cache=TRUE}
plot (agpop$acres87, agpop$acres92)
summary(lm(agpop$acres92~agpop$acres87))
acres87 <- agpop[, "acres87"]
est_srs300 <- est_ups300_acres87 <- est_srs300_reg_acres87 <- rep (0, 2000)
for (i in 1:2000)
{
  ## srs sample
  srs300 <- sample (1:N, size = 300)
  ## survey
  acres92_srs300 <- agpop[srs300, "acres92"]
  acres87_srs300 <- agpop[srs300, "acres87"]
  ## estimate
  est_srs300 [i] <-  (srs_mean_est (acres92_srs300, N = N) * N) [1] 
  est_srs300_reg_acres87[i] <- srs_reg_est(
      acres92_srs300, acres87_srs300, total_acres87/N, N = N, est.total = TRUE )[1]
  
  ## upswr sample
  ups300 <- sample (1:N, size = 300, prob = acres87, replace = TRUE)

  ## survey (obtaining measurements on acres92)
  acres92_ups300 <- agpop[ups300, "acres92"]

  ## estimate
  psi <- (acres87 / sum (acres87)) [ups300]
  est_ups300_acres87 [i] <-  upswr_total_est (acres92_ups300, psi) [1] 

}
```


```{r fig.height=15}
par (mfrow = c(2,1))
boxplot (data.frame(est_srs300, est_srs300_reg_acres87, est_ups300_acres87))
abline(h=total_acres92, col="red")
boxplot (data.frame(est_srs300_reg_acres87, est_ups300_acres87))

abline(h=total_acres92, col="red")
```


## Using largef87 to construct sampling probabilities

```{r cache=TRUE}
plot (agpop$largef92, agpop$acres92)
largef87 <- agpop[, "largef87"]
summary(lm(agpop$acres92~agpop$largef87))

est_ups300_largef87 <- rep (0, 2000)

for (i in 1:2000)
{

    ## ups sample
    ups300 <- sample (1:N, size = 300, prob = largef87, replace = TRUE)
    ## survey
    acres92_ups300 <- agpop[ups300, "acres92"]
    
    ## estimate
    psi <- (largef87 / sum (largef87)) [ups300]
    est_ups300_largef87 [i] <-  upswr_total_est (acres92_ups300, psi) [1] 
    
}

boxplot (data.frame(est_srs300,est_srs300_reg_acres87, est_ups300_acres87, est_ups300_largef87))
abline(h=total_acres92, col="red")

```

# Analysis of a dataset (student study hours) collected by UPSWO (without replacement)

```{r}

## data --- original data frame for holding data
## cname --- variable recording cluster (psu) identity
## csize --- variable recording cluster (psu) population size (not sample size)
## cpik --- sampling probability for each cluster
## yvar --- variable of interest
## N --- total number of clusters (psus)

cluster_upswo_ratio_est <- function (data, cname, csize, cpik, yvar)
{
  clust <- data[,cname]
  ydata <- data [, yvar]  
  
  ## cluster-wise summary
  ybari <- tapply (ydata, clust, mean)
  Mi <- tapply (data [,csize], clust, function (x) x[1])
  pik <- tapply (data [,cpik], clust, function (x) x[1])
  ## the same as total in cluster if Mi = mi
  t_hat_cls <- ybari * Mi 
  
  ## apply ratio estimate to t_hat_cls and Mi
  srs_ratio_est (t_hat_cls/pik, Mi/pik)
}

classpps <- read.csv ("data/classpps.csv")

classpps

cluster_upswo_ratio_est (classpps, "class", "clssize", "clssize", "hours")


```

