---
title: "Data Analysis and Simulation for Simple Random Sampling"
author: "Longhai Li"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: no
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
    df_print: paged
  word_document:
    toc: yes
---
```{r}
library(latex2exp)
```


# Analysis of agsrs.csv Data
## Step by step calculation without using a function 
```{r}
## read survey data
agsrs <- read.csv ("data/agsrs.csv")
head(agsrs)

## extract the variable of interest
sdata <- agsrs$acres92
N <- 3078

## do calculation
n <- length (sdata)
ybar <- mean (sdata)
se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
mem <- qt (0.975, df = n - 1) * se.ybar
## return estimate vector for pop mean
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)

## return estimate vector for pop total
c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem) * N
```

## Write a function for repeated use
### A function for doing data analysis for srs sample
```{r }
# 
# sdata -- a vector of sampling survey data
# N -- population size 
# to find total, multiply N to the estimate returned by this function
srs_mean_est <- function (sdata, N)
{
	n <- length (sdata)
	ybar <- mean (sdata)
	se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
	mem <- qt (0.975, df = n - 1) * se.ybar
	c (ybar = ybar, se = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}
```

### Apply srs_mean_est to agsrs.csv data
*Import Data*
```{r}
agsrs <- read.csv ("data/agsrs.csv")
```
*Estimating the mean of acre92*
```{r}
srs_mean_est (agsrs[,"acres92"], N = 3078)
```

*Estimating the total of acre92*
```{r}
srs_mean_est (agsrs[,"acres92"], N = 3078) * 3078
```

*Estimating the proportion of counties with fewer than 200K acres for farming in 1992*
```{r}
acres92.is.fewer.200k <- as.numeric (agsrs[,"acres92"] < 200000)
head(acres92.is.fewer.200k)
srs_mean_est (acres92.is.fewer.200k, N = 3078)
```

*Estimating the total number of counties with fewer than 200K acres for farming in 1992*
```{r}
srs_mean_est (acres92.is.fewer.200k, N = 3078) * 3078
```

## Comparing with true value

```{r}
agpop <- read.csv ("data/agpop.csv", na = "-99")
#true mean
mean (agpop[, "acres92"], na.rm = T)
# true total
sum (agpop[, "acres92"], na.rm = T)

# true proportion of counties with less than 200K acres for farming
mean (agpop[, "acres92"] < 200000, na.rm = T)
# true number of counties with less than 200K acres for farming
sum (agpop[, "acres92"] < 200000, na.rm = T)

```

# A Simulation Demonstration of SRS Inference
```{r }
# read population data 
agpop <- read.csv ("data/agpop.csv")
# remove those counties with na
agpop <- subset( agpop, acres92 != -99)
```



## True Values
```{r }
# sample size
n <- 300
# population size
N <- nrow (agpop); N

# true value of population mean
ybarU <- mean (agpop[,"acres92"]); ybarU
# true value of deviation of sample mean
true.se.ybar <- sqrt (1- n/N) * sd (agpop[,"acres92"]) / sqrt (n); true.se.ybar
```

## One SRS sampling

```{r }
## 
# srs sampling
srs <- sample (1:N,n)
head(agpop [srs, ])
# get data of variable "acres92"
sdata <- agpop [srs, "acres92"]
# analysis
srs_mean_est (sdata, N)
```

## Repeating SRS sampling 5000 times
```{r fig.height=10}
nres <- 5000 # number of repeated sampling
simulation.results <- matrix (0, nres, 4) # matrix recording repeated results
colnames(simulation.results) <- c( "Est.",   "S.E.",   "ci.low", "ci.upp")

for (i in 1:nres)
{
    srs <- sample (N, n)
	sdata <- agpop [srs, "acres92"]
	simulation.results [i,] <- srs_mean_est (sdata, N)
}

head(simulation.results)
# look at the distribution of sample mean
par (mfrow= c(2,2))
hist (agpop$acres92,main = "Population Distribution of acre92")
hist (simulation.results[,1], main = "Sampling Distribution of Sample Mean for acre92")
abline (v = ybarU, col = "red")
qqnorm (simulation.results[,1], main="QQ plot of Sample Mean"); qqline(simulation.results[,1])
boxplot (simulation.results[,1], main = "Boxplot of Sample Mean")
abline (h = ybarU, col = "red")

mean (simulation.results[,1])
ybarU
sd (simulation.results [,1])
true.se.ybar
```


## Empirical Coverage Rate of CIs

```{r fig.height=10}
simulation.results <- cbind (simulation.results, (simulation.results[,3] < ybarU) * (ybarU < simulation.results[,4] ))
colnames(simulation.results)[5] <- "Covered?"
head(simulation.results)

library("plotrix") 
par(mfrow=c(1,1))
plotCI(x=1:100,
       y=simulation.results[1:100,1],
       li = simulation.results[1:100,3],
       ui = simulation.results[1:100,4],
       col = 2-simulation.results[,5])
abline(h=ybarU, col = "blue")

# Empirical coverage rate 
mean (simulation.results[,"Covered?"])

```



