---
title: "STAT 348 Sampling Techniques (Univ. of Saskatchewan, 2026-09)"
metadata-files: 
  - ../../../_website.yaml
---

## Description

Theory and applications of sampling from finite populations. Includes: simple random sampling, stratified random sampling, cluster sampling, systematic sampling, probability proportionate to size sampling, and the difference, ratio and regression methods of estimation.

## Prerequisites

* STAT 242, or STAT 245, or STAT 246

## Instructor
* [Longhai Li](https://longhaisk.github.io), Professor
* Department of Mathematics and Statistics, University of Saskatchewan
* Email: longhai.li@usask.ca.

## Times and Places
* **Lectures:** MWF 12:30 - 1:20, **Arts Building 208**
* **Office Hours:** TBA with Students

## Textbook and Course Materials

Textbooks are not required, but you are advised to have one of the following two books:

* **Recommended Text 1:** *Sampling: Design and Analysis*, 2nd Edition, by Sharon L. Lohr (Brooks/Cole). We will roughly cover materials in Chapters 1-6.
* **Recommended Text 2:** *Elementary Survey Sampling*, 7th Edition, by Richard L. Scheaffer, William Mendenhall III, R. Lyman Ott, Kenneth G. Gerow (ISBN-13: 978-0-8400-5361-9). We will roughly cover materials in Chapters 1-9.

* **Lecture Notes:** Available from a shared OneDrive folder with the link given on Canvas (the password is also released on the Canvas page). Assignment questions, solutions, and R code for demonstration are all in this folder.
* **R Code and Spreadsheet Demonstration:** [https://longhaisk.github.io/teaching/stat348_rdemo/stat348.html](https://longhaisk.github.io/teaching/stat348_rdemo/)

## Computing

We will use RStudio and R for this course. 

* **Personal Computer:** Download R, RStudio, Positron, VS-code to your local machine.
* **USASK vlab:** If you don't have a personal computer, you can use the USask remote desktop, the browser-based vlab ([https://vlab.usask.ca/](https://vlab.usask.ca/)), 

* **Posit Cloud:** ([https://posit.cloud/](https://posit.cloud/)).
* **Google Colab:** You can also run R in the cloud using Google Colaboratory. To open a notebook with R pre-configured, use this direct link: [https://colab.research.google.com/#create=true&language=r](https://colab.research.google.com/#create=true&language=r). Alternatively, you can create a new notebook in Colab and change the runtime type to R (Runtime > Change runtime type > R).

## Tentative Schedule / List of Topics

```{r}
#| echo: false
#| message: false
#| warning: false
#| out-width: "100%"

library(gt)
library(dplyr)
library(tibble)

# 1. SETUP DATES (Estimated for Fall 2024)
start_date <- as.Date("2026-09-02")
break_date <- as.Date("2026-11-09")
end_date   <- as.Date("2026-12-07")

# Helper function to find the Monday of the week for any given date
get_monday <- function(d) { d - as.numeric(format(d, "%u")) + 1 }

first_monday <- get_monday(start_date)
break_monday <- get_monday(break_date)
last_monday  <- get_monday(end_date)

# Generate all Mondays for the term
term_dates <- seq(from = first_monday, to = last_monday, by = "week")

# 2. DEFINE COURSE TOPICS & TASKS (IN ORDER)
content_list <- tribble(
  ~Topic,                                                                 ~Base_Task,
  "1 Introduction to Sampling Techniques, R and R Markdown",              "",
  "1 Introduction to Sampling Techniques, R and R Markdown",              "",
  "2 Simple Random Sampling",                                             "",
  "2 Simple Random Sampling",                                             "",
  "3 Stratified Sampling",                                                "**Assignment 1 due**",
  "3 Stratified Sampling",                                                "",
  "4 Ratio and Regression Estimate",                                      "",
  "4 Ratio and Regression Estimate",                                      "**Midterm**",
  "4 Ratio and Regression Estimate",                                  "",
  "5 Cluster Sampling",                                  "**Assignment 2 due**",
  "5 Cluster Sampling",                                       "",
  "6 Unequal probability sampling",                                       "",
  "6 Unequal probability sampling",                                       "",
  "Review",                                                               "**Assignment 3 due**"
)

# 3. BUILD THE SCHEDULE DYNAMICALLY
schedule_list <- list()
acad_counter <- 1 # Tracks which topic we are on

for (i in seq_along(term_dates)) {
  curr_date <- term_dates[i]
  
  if (curr_date == break_monday) {
    # --- Reading Week ---
    schedule_list[[i]] <- data.frame(
      Date_Val  = curr_date,
      Date      = format(curr_date, "%b %d"),
      Acad_Week = "N/A",
      Topic     = "—",
      Task      = "**Fall Break – No classes**",
      stringsAsFactors = FALSE
    )
  } else {
    # --- Regular Academic Week ---
    curr_topic <- ifelse(acad_counter <= nrow(content_list), content_list$Topic[acad_counter], "TBD")
    curr_task  <- ifelse(acad_counter <= nrow(content_list), content_list$Base_Task[acad_counter], "")
    
    schedule_list[[i]] <- data.frame(
      Date_Val  = curr_date,
      Date      = format(curr_date, "%b %d"),
      Acad_Week = as.character(acad_counter),
      Topic     = curr_topic,
      Task      = curr_task,
      stringsAsFactors = FALSE
    )
    
    # Increment the topic counter
    acad_counter <- acad_counter + 1
  }
}

# Combine into a single dataframe
schedule_data <- do.call(rbind, schedule_list)

# 4. APPEND START & END TASKS
# Find the start week and append the start date note
start_idx <- which(schedule_data$Date_Val == first_monday)
if (length(start_idx) > 0) {
  prefix <- ifelse(schedule_data$Task[start_idx] == "", "", paste0(schedule_data$Task[start_idx], "<br>"))
  schedule_data$Task[start_idx] <- paste0(prefix, "**Course Starts (", format(start_date, "%b %d"), ")**")
}

# Find the end week and append the end date note
end_idx <- which(schedule_data$Date_Val == last_monday)
if (length(end_idx) > 0) {
  prefix <- ifelse(schedule_data$Task[end_idx] == "", "", paste0(schedule_data$Task[end_idx], "<br>"))
  schedule_data$Task[end_idx] <- paste0(prefix, "**Course Ends (", format(end_date, "%b %d"), ")**")
}

# Save index for highlighting the Reading Week
break_idx <- which(schedule_data$Date_Val == break_monday)

# Clean up the temporary internal date column
schedule_data$Date_Val <- NULL

# 5. CREATE THE GT TABLE
gt_tbl <- schedule_data %>%
  gt() %>%
  
  # Format column labels
  cols_label(
    Acad_Week = md("**Acad. Week**"),
    Date      = md("**Date**"),
    Topic     = md("**Topic**"),
    Task      = md("**Remark**")
  ) %>%
  
  # Process markdown in the Task column
  fmt_markdown(columns = c(Task)) %>%
  
  # Alignment
  cols_align(align = "center", columns = c(Acad_Week)) %>%
  cols_align(align = "left", columns = c(Date, Topic, Task)) %>%
  
  # Column widths
  cols_width(
    Acad_Week ~ pct(10),
    Date      ~ pct(10),
    Topic     ~ pct(55),
    Task      ~ pct(25)
  ) %>%
  
  opt_row_striping() %>%
  
  # Table Styling
  tab_options(
    table_body.hlines.style = "solid",
    table_body.vlines.style = "solid",
    table_body.hlines.color = "#E0E0E0",
    
    column_labels.vlines.style = "solid",
    column_labels.border.top.style = "solid",
    column_labels.border.bottom.style = "solid",
    column_labels.background.color = "#F0F0F0",
    
    table.border.top.style = "solid",
    table.border.bottom.style = "solid",
    table.font.size = px(13)
  ) %>%
  
  # Highlight the dynamically found break_date row
  tab_style(
    style = list(
      cell_fill(color = "#d1e7dd"),
      cell_text(weight = "bold", style = "italic")
    ),
    locations = cells_body(rows = break_idx)
  )

gt_tbl

```

:::{.callout-important}
The schedule is only for reference and may change depending on the actual class pace. The exact assignment due dates and the midterm test date are given on Canvas in the "Assignments" section.
:::

## Learning Outcomes

After completing this course, students are expected to grasp the following knowledges and skills:

```{r}
#| echo: false
#| message: false
library(gt)
library(dplyr)

# Constructing the dataframe with updated Sampling Techniques skills
final_guide <- data.frame(
  Topic = c(
    "Principles & Bias",
    "Sampling Schemes",
    "Data Analysis",
    "R & Dissemination"
  ),
  Knowledge = c(
    "Understand the principles and methods used to design sampling schemes.",
    "Understand the characteristics of a well-designed survey, identifying possible sources of bias and measurement errors.",
    "Gain a comparative understanding of different schemes (simple, strata, cluster, unequal probabilities).",
    "Understand how to interpret outputs from datasets collected with different sampling schemes."
  ),
  Skills = c(
    "Design appropriate sampling schemes for finite populations.",
    "Critically evaluate survey designs and diagnose sources of bias.",
    "Analyze datasets collected with different schemes including simple sampling, strata, cluster, and unequal probabilities.",
    "Use R and Rmarkdown to analyze the datasets and disseminate the findings effectively."
  ),
  Percentages = c("20%", "20%", "40%", "20%")
)

outcome_tbl <- final_guide %>%
  gt() %>%
  cols_label(
    Topic = md("**Topic**"),
    Knowledge = md("**Knowledge**"),
    Skills = md("**Skills**"),
    Percentages = md("**Perc**")
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Topic)
  ) %>%
  opt_row_striping() %>%
  tab_options(
    table.width = pct(100),
    column_labels.background.color = "#E0E0E0",
    column_labels.font.weight = "bold",
    table.font.size = px(12),
    data_row.padding = px(5),
    row.striping.background_color = "#F5F5F5",
    table_body.hlines.color = "#B0B0B0", 
    table_body.hlines.width = px(1)
  ) %>%
  cols_width(
    Topic ~ pct(15),
    Knowledge ~ pct(35),
    Skills ~ pct(40),
    Percentages ~ pct(10)
  )

outcome_tbl

```

## Evaluation

### Grading Scheme

**3 Assignments: 3 x 10%, 1 Term Test: 20%, 1 Final Exam: 50%.**

### Assignments and Tests

**Assignment questions are released in the one-drive folder**. You will submit your solutions via Canvas. **If you miss an assignment without proper excuse, the weight will NOT be shifted to the final.** Undergraduate students will be assigned with different assignments and tests.

### Assignments

* I will accept late assignments only for three (3) days beyond the due date. The penalty for your delay is 10 percentage points per day of lateness from the value of the assignment (including weekends). **Extensions are only granted in rare instances (notably as a result of family or medical emergencies) and upon receipt of adequate documentation/proof.**
* Answer the questions in the order they appear in the assignment. Neatness is important.
* Solutions to problems are to be included. Hence, simple answers without work will receive few (or no!) marks.
* Most problems in statistics have a “real-life” basis. Hence, solutions should include not only numerical solutions but also a statement as to what the numbers say about the problem.
* The work handed in must not be an exact duplicate of others.
* Submitting Assignments: The assignment can be typed and/or handwritten. Save your assignment as **one PDF file** (for handwritten assignments, feel free to take a picture/scan of your work and save it as one PDF file). Upload the **PDF file** as an assignment submission in Canvas.
* More details will be provided ahead of each assignment.
* Due Date: See Course Schedule.

### Midterm

* The midterm is given in class period.
* Midterms must be written on the dates scheduled. Students must do midterms completely on their own. More details (including syllabus) will be provided ahead of each midterm.
* Type: Short-answer questions, problem-solving, open-book.
* Calculator: A scientific calculator is allowed.
* Make-up exam will not be given. If you miss an exam for a legitimate reason (e.g., illness, emergency) and notify me within 48 hours of the scheduled exam, the weight of the missed exam will be transferred to the final exam.

### Final Exam

* Scheduling: Final examinations may be scheduled at any time during the examination period; students should therefore avoid making prior travel, employment, or other commitments for this period. If a student is unable to write an exam through no fault of their own for medical or other valid reasons, documentation must be provided and an opportunity to write the missed exam may be given. Students are encouraged to review all examination policies and procedures: [http://students.usask.ca/academics/exams.php](http://students.usask.ca/academics/exams.php).
* The final exam will cover material of the entire course. More details will be provided ahead of the exam.
* Length: 3-hour in-person exam.
* Type: Short-answer questions, problem-solving, open-book.

### Criteria That Must Be Met to Pass

The **final exam is a required component of the course**. Students must complete the final exam in order to be eligible to receive a passing grade in this class.

## Attendance Expectation

Attendance is highly correlated with student performance. While a syllabus and suggested readings are provided, it is not an adequate substitute for attending class. Your **attendance is highly recommended** but not required, and you will not be graded on your attendance.

## Recording of the Course

Recording of the lectures will only be allowed in certain circumstances. Please see the instructor for information on how to receive approval. In general, there will be no videos available for in-person lectures. Therefore, **attendance is strongly recommended**.

## Use of Generative AI and Electronic Devices

* AI for Learning vs. Assessment. Students are free (and **encouraged**) to use Generative AI tools as a study aid to understand course concepts, debug code, or explain complex theorems. However, **all submitted work for assignments must be your own.** You must write your own solutions. Directly copying text, derivations, or code from an AI tool and submitting it as your own may receive a **severe penalty** (up to receiving a 0% on the assignment).
* Electronic Devices During Tests. All term tests and the final exam are **Open Book**, meaning you may bring printed notes, textbooks, and lecture slides.
* **No Electronic Devices:** You are **NOT allowed** to use laptops, tablets, smartwatches, or any other electronic devices during the exam.
* **Phone Exception:** You are permitted to bring a smartphone, but it must remain stowed away during the writing period. It may **only** be used at the very end of the exam for the specific purpose of taking photos of your answer sheets for submission (if required). Using the phone for any other reason during the exam will be treated as academic misconduct.

