---
engine: knitr
format: 
  html: default
  pdf: default
---

# Hypothesis Testing in Linear Models

## Testing Reduced Model vs Full Model (Partial F-test)

Consider the general linear model partitioned as follows:

$$
y = X\beta + e = (X_{1}, X_{2}) \begin{pmatrix} \beta_{1} \\ \beta_{2} \end{pmatrix} + e = X_1 \beta_1 + X_2 \beta_2 + e
$$

where $X$ is an $n \times (k+1)$ matrix, $X_1$ corresponds to the first set of predictors, and $X_2$ corresponds to the remaining $h$ predictors . We assume $e \sim N(0, \sigma^2 I)$.

We are often interested in testing the hypothesis that the second set of coefficients is zero:

$$
H_0: \beta_2 = 0
$$

This leads to a comparison between two models:

1.  **Full Model (FM)**: The maintained hypothesis where $\mu \in C(X) = C([X_1, X_2])$ .

    
    $$
    y = X_1 \beta_1 + X_2 \beta_2 + e
    $$
    
2.  **Reduced Model (RM)**: The model under $H_0$ where $\mu \in C(X_1)$.

    
    $$
    y = X_1 \beta_1 + e^*
    $$

The testing problem is to test $H_0: \mu \in C(X_1)$ versus $H_1: \mu \in C(X)$ but $\mu \notin C(X_1)$ .

### Geometric Interpretation

Since $C(X_1) \subset C(X)$, if the Reduced Model is true, the Full Model must also be true. Under $H_0$, the least squares estimates of the mean $\mu$ from both models, $P_{C(X_1)}y$ and $P_{C(X)}y$, estimate the same quantity.

This suggests that the difference vector :

$$
P_{C(X)}y - P_{C(X_1)}y = (P_{C(X)} - P_{C(X_1)})y = \hat{y} - \hat{y}_0
$$

should be "small" under $H_0$. The matrix $P_{C(X)} - P_{C(X_1)}$ is the projection matrix onto $C(X_1)^{\perp} \cap C(X)$, which is the orthogonal complement of $C(X_1)$ with respect to $C(X)$.

```{tikz}
%| label: fig-projection-geometry
%| fig-cap: "Geometric interpretation of the Full vs Reduced Model. The vector y is projected onto the Full Model space C(X) to get y_hat, and onto the Reduced Model space C(X1) to get y_hat0."
%| echo: false
%| fig-align: "center"
%| out-extra: 'style="width: 70% !important;"'

\begin{tikzpicture}[scale=1.2]
  % Define coordinates
  \coordinate (O) at (0,0);
  \coordinate (X1) at (4,0.5); % Reduced model space vector
  \coordinate (X2) at (1.5,3); % Another vector in Full model space
  \coordinate (Y) at (1.5,4.5); % Observation y

  % Draw plane C(X)
  \fill[blue!10] (0,0) -- (5,0.625) -- (6.5,3.625) -- (1.5,3) -- cycle;
  \node at (5.5,3) [blue] {$C(X)$ (Full Model)};

  % Draw line C(X1)
  \draw[thick, blue] (O) -- (5,0.625);
  \node at (4.5,0.3) [blue] {$C(X_1)$ (Reduced Model)};

  % Projections
  \coordinate (Yhat) at (2.5,2.2); % Projection onto C(X)
  \coordinate (Yhat0) at (2.0, 0.25); % Projection onto C(X1) approx

  % Draw vectors
  \draw[->, thick] (O) -- (Y) node[above left] {$y$};
  \draw[->, thick, red] (O) -- (Yhat) node[below right] {$\hat{y} = P_{C(X)}y$};
  \draw[->, thick, red] (O) -- (Yhat0) node[below] {$\hat{y}_0 = P_{C(X_1)}y$};

  % Difference vectors
  \draw[dashed] (Y) -- (Yhat);
  \draw[dashed] (Yhat) -- (Yhat0);
  \draw[->, thick, green!60!black] (Yhat0) -- (Yhat) node[midway, left] {$\hat{y} - \hat{y}_0$};
  \draw[dashed] (Y) -- (Yhat0);

  % Right angles
  \draw (Yhat) -- ++(0.2,0.1) -- ++(-0.1,0.2); % roughly perp to plane
  
\end{tikzpicture}

```

The squared length of this difference is used as a test statistic:

$$
\text{SSH} = ||\hat{y} - \hat{y}_0||^2 = y^T (P_{C(X)} - P_{C(X_1)}) y
$$

### Distributional Properties

To derive a test, we analyze the distribution of the sum of squares.

::: {#thm-quadratic-independence}

### Distribution of Quadratic Forms
Suppose $y \sim N(X\beta, \sigma^2 I)$ where $X$ is $n \times (k+1)$ of full rank, $X\beta = X_1\beta_1 + X_2\beta_2$, and $X_2$ is $n \times h$. Let $\hat{y} = P_{C(X)}y$, $\hat{y}_0 = P_{C(X_1)}y$, and $\mu_0 = P_{C(X_1)}\mu$. Then:

1.  $\frac{1}{\sigma^2} ||y - \hat{y}||^2 = \frac{1}{\sigma^2} y^T (I - P_{C(X)}) y \sim \chi^2(n - k - 1)$.

    
2.  $\frac{1}{\sigma^2} ||\hat{y} - \hat{y}_0||^2 = \frac{1}{\sigma^2} y^T (P_{C(X)} - P_{C(X_1)}) y \sim \chi^2(h, \lambda_1)$.

    
    where the non-centrality parameter is $\lambda_1 = \frac{1}{\sigma^2} ||(P_{C(X)} - P_{C(X_1)})\mu||^2 = \frac{1}{\sigma^2} ||\mu - \mu_0||^2$.
    
3.  The two quadratic forms are independent.

:::

Under $H_0$, $\mu \in C(X_1)$, so $\mu = \mu_0$ and $\lambda_1 = 0$. Thus, the numerator sum of squares follows a central $\chi^2$ distribution.

### The F-Test

Based on the independence and distribution of the quadratic forms, we construct the F-statistic.

::: {#thm-f-test-reduced}

### F-Test for Reduced Model
Under the conditions of @thm-quadratic-independence, the statistic

$$
F = \frac{||\hat{y} - \hat{y}_0||^2 / h}{||y - \hat{y}||^2 / (n - k - 1)} = \frac{\text{SSH}/h}{\text{SSE}_{\text{FM}}/(n - k - 1)}
$$

follows a non-central F-distribution $F(h, n - k - 1, \lambda_1)$. Under $H_0: \beta_2 = 0$, it follows a central F-distribution $F(h, n - k - 1)$ .

:::

Using the Pythagorean Theorem, we can express the numerator in terms of Sum of Squares for Error (SSE) or Regression (SSR):

$$
||\hat{y} - \hat{y}_0||^2 = SSE_{\text{RM}} - \text{SSE}_{\text{FM}}
$$

$$
||\hat{y} - \hat{y}_0||^2 = \text{SSR}_{\text{FM}} - \text{SSR}_{\text{RM}}
$$

This quantity is often denoted as $\text{SS}(\beta_2 | \beta_1)$, the "extra" regression sum of squares due to $\beta_2$ after accounting for $\beta_1$.

The ANOVA table for this comparison is constructed as follows:

| Source | SS | MS | F |
| :--- | :--- | :--- | :--- |
| Hypothesis | $\text{SSR}_{\text{FM}} - \text{SSR}_{\text{RM}}$ | $\text{SSH}/h$ | $\frac{\text{SSH}/h}{\text{MSE}_{\text{FM}}}$ |
| Error | $\text{SSE}_{\text{FM}}$ | $\text{SSE}_{\text{FM}}/(n-k-1)$ | |
| Total | $\text{SST}$ | | |

## Overall Regression Test

A special case of the test $H_0: \beta_2 = 0$ is when $\beta_1$ contains only the intercept $\beta_0$, and $\beta_2$ contains all slope coefficients. The model is:

$$
y = \beta_0 j_n + X_2 \beta_2 + e
$$

The hypothesis $H_0: \beta_2 = 0$ is equivalent to $H_0: \beta_1 = \beta_2 = \dots = \beta_k = 0$. In this case:

1.  The reduced model estimates $\hat{y}_0 = \bar{y}j_n$.
2.  The degrees of freedom $h = k$.
3.  The numerator becomes $\text{SSR}/k \equiv \text{MSR}$.

::: {#thm-overall-regression}

### Overall Regression F-Statistic
The test statistic for overall regression is given by:

$$
F = \frac{\text{SSR}/k}{\text{SSE}_{\text{FM}}/(n - k - 1)} = \frac{\text{MSR}}{\text{MSE}}
$$

Under $H_0$, $F \sim F(k, n - k - 1)$.

:::

This statistic can be expressed in terms of the coefficient of determination $R^2$:

$$
F = \frac{R^2 / k}{(1 - R^2) / (n - k - 1)}
$$

## The General Linear Hypothesis

We can generalize these tests to the hypothesis $H_0: C\beta = t$, where $C$ is a $q \times (k+1)$ matrix of rank $q \le k+1$.

Common examples include :

* Testing a subset of coefficients: $C = (0, I_h)$.
* Overall regression: $C = (0, I_k)$.
* Equality of coefficients (e.g., $\beta_1 = \beta_2$).

### Test Statistic for $C\beta = 0$

The test compares $C\hat{\beta}$ to its null value 0 using a squared statistical distance. Since $\hat{\beta} \sim N(\beta, \sigma^2(X^T X)^{-1})$, it follows that $C\hat{\beta} \sim N(C\beta, \sigma^2 C(X^T X)^{-1} C^T)$ .

::: {#thm-general-linear-hypothesis}

### F-Test for General Linear Hypothesis
If $y \sim N(X\beta, \sigma^2 I)$ and $C$ has rank $q$, then under $H_0: C\beta = 0$ :

$$
F = \frac{(C\hat{\beta})^T \{C(X^T X)^{-1} C^T\}^{-1} C\hat{\beta} / q}{\text{SSE}_{\text{FM}} / (n - k - 1)} \sim F(q, n - k - 1)
$$

If $H_0$ is false, it follows a non-central F distribution with parameter $\lambda = \frac{(C\beta)^T [C(X^T X)^{-1} C^T]^{-1} C\beta}{\sigma^2}$ .

:::

### Nested Models Interpretation

The general linear hypothesis test is fundamentally a test of nested models.

::: {#thm-nested-models}

### General Linear Hypothesis as Nested Models
The F-test for the general linear hypothesis $H_0: C\beta = 0$ is equivalent to a full-and-reduced model test. Specifically, testing $H_0$ is equivalent to testing whether the mean vector $\mu$ lies in a subspace $V_0 \subset C(X)$. The test statistic satisfies:

$$
\text{SSH} = (C\hat{\beta})^T \{C(X^T X)^{-1} C^T\}^{-1} C\hat{\beta} = ||P_{C(Z)}y||^2
$$

where $C(Z)$ is the orthogonal complement of the reduced model space $V_0$ with respect to the full model space $C(X)$.

:::

::: {.proof}

Under the null hypothesis $H_0: C\beta = 0$, we have a constrained model. We observe that:

$$
C\beta = 0 \implies C(X^T X)^{-1} X^T (X\beta) = 0
$$

Let $Z^T = C(X^T X)^{-1} X^T$. Then the condition becomes $Z^T \mu = 0$ (since $\mu = X\beta$). This implies that under $H_0$, $\mu$ must be orthogonal to the column space of $Z$, denoted $C(Z)$ .

Since $\mu$ must also lie in the full model space $C(X)$, under $H_0$, $\mu$ belongs to the intersection:

$$
V_0 = C(Z)^{\perp} \cap C(X)
$$

This $V_0$ represents the subspace for the Reduced Model (RM), while $C(X)$ is the subspace for the Full Model (FM). The hypotheses correspond to nested models since $V_0 \subset C(X)$.

The numerator sum of squares for comparing these models is the squared length of the projection of $y$ onto the difference space. Since $V_0$ is the orthogonal complement of $C(Z)$ relative to $C(X)$, the difference space is exactly $C(Z)$. Note that $Z = X(X^T X)^{-1} C^T$, and since $C$ has rank $q$, $Z$ has rank $q$.

The sum of squares for the hypothesis is:
$$
\text{SSH} = ||(P_{C(X)} - P_{V_0})y||^2 = ||P_{C(Z)}y||^2
$$

Expanding the projection matrix $P_{C(Z)} = Z(Z^T Z)^{-1} Z^T$:

$$
y^T P_{C(Z)} y = y^T Z(Z^T Z)^{-1} Z^T y
$$

Substituting $Z = X(X^T X)^{-1} C^T$:

1.  **Term $Z^T y$**:

    $$
    Z^T y = C(X^T X)^{-1} X^T y = C\hat{\beta}
    $$

2.  **Term $Z^T Z$**:

    $$
    Z^T Z = C(X^T X)^{-1} X^T X (X^T X)^{-1} C^T = C(X^T X)^{-1} C^T
    $$

Thus, the quadratic form becomes:

$$
y^T P_{C(Z)} y = (C\hat{\beta})^T [C(X^T X)^{-1} C^T]^{-1} (C\hat{\beta})
$$

This confirms that the Wald-type statistic derived for the General Linear Hypothesis is algebraically identical to the difference in sums of squares between the implied full and reduced models.

:::

### Extension to $C\beta = t$

For the non-homogeneous case $H_0: C\beta = t$ ($t \neq 0$), the statistic is modified to center the numerator around $t$:

$$
F = \frac{(C\hat{\beta} - t)^T [C(X^T X)^{-1} C^T]^{-1} (C\hat{\beta} - t) / q}{\text{SSE}_{\text{FM}} / (n - k - 1)}
$$

:::: {#exm-chemical-reaction}

### Chemical Reaction Data

Consider an experiment designed to optimize the yield of a chemical reaction. The explanatory variables are:

* $x_1$: Temperature ($^\circ C$)
* $x_2$: Concentration of a reagent (%)
* $x_3$: Time of reaction (hours)

The response variable $y_1$ is the percent of unchanged starting material. The data is provided below :

| $y_1$ | $x_1$ | $x_2$ | $x_3$ |
| :--- | :--- | :--- | :--- |
| 41.5 | 162 | 23 | 3 |
| 33.8 | 162 | 23 | 8 |
| 27.7 | 162 | 30 | 5 |
| 21.7 | 162 | 30 | 8 |
| 19.9 | 172 | 25 | 5 |
| 15.0 | 172 | 25 | 8 |
| 12.2 | 172 | 30 | 5 |
| 4.3 | 172 | 30 | 8 |
| 19.3 | 167 | 27.5 | 6.5 |
| 6.4 | 177 | 27.5 | 6.5 |
| 37.6 | 157 | 27.5 | 6.5 |
| 18.0 | 167 | 32.5 | 6.5 |
| 26.3 | 167 | 22.5 | 6.5 |
| 9.9 | 167 | 27.5 | 9.5 |
| 25.0 | 167 | 27.5 | 3.5 |
| 14.1 | 177 | 20 | 6.5 |
| 15.2 | 177 | 20 | 6.5 |
| 15.9 | 160 | 34 | 7.5 |
| 19.6 | 160 | 34 | 7.5 |

We fit the full model:
$$
y_1 = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \epsilon
$$

Suppose we wish to test the hypothesis that the effect of temperature and concentration are equal (scaled by 2) and related to time as follows: $H_0: 2\beta_1 = 2\beta_2 = \beta_3$. This can be broken down into two independent linear constraints:

1.  $2\beta_1 - 2\beta_2 = 0$
2.  $2\beta_2 - \beta_3 = 0$

We formulate this as a general linear hypothesis $C\beta = 0$, where $\beta = (\beta_0, \beta_1, \beta_2, \beta_3)^T$ and $C$ is a $2 \times 4$ matrix :

$$
C = \begin{pmatrix} 0 & 2 & -2 & 0 \\ 0 & 0 & 2 & -1 \end{pmatrix}
$$

Using the data, the estimated coefficients $\hat{\beta}$ yield $C\hat{\beta} = \begin{pmatrix} -0.1214 \\ -0.6118 \end{pmatrix}$. The variance term involving the design matrix is calculated as:

$$
C(X^T X)^{-1} C^T = \begin{pmatrix} 0.003366 & -0.006943 \\ -0.006943 & 0.044974 \end{pmatrix}
$$

The test statistic is computed using the quadratic form:

$$
F = \frac{(C\hat{\beta})^T [C(X^T X)^{-1} C^T]^{-1} (C\hat{\beta}) / 2}{s^2} = \frac{28.62301 / 2}{5.3449} = 2.6776
$$

where $s^2 = \text{MSE}_{\text{FM}} = 5.3449$. This statistic follows an $F(2, 15)$ distribution. The corresponding p-value is $0.101$, suggesting that we fail to reject the null hypothesis at the $\alpha = 0.05$ level.

**Alternative Approach (Full vs. Reduced Model):**
The null hypothesis implies relationships between the parameters that allow us to rewrite the model. If $2\beta_1 = \beta_3$ and $2\beta_2 = \beta_3$, then $\beta_1 = 0.5\beta_3$ and $\beta_2 = 0.5\beta_3$. Substituting these into the full model:

$$
y_1 = \beta_0 + 0.5\beta_3 x_1 + 0.5\beta_3 x_2 + \beta_3 x_3 + \epsilon
$$
$$
y_1 = \beta_0 + \beta_3 (0.5 x_1 + 0.5 x_2 + x_3) + \epsilon
$$

This reduced model has only two parameters: an intercept $\beta_0$ and a slope $\beta_3$ for the constructed variable $z = 0.5x_1 + 0.5x_2 + x_3$. We could then calculate $F$ using the difference in SSE between the full model and this reduced model .


**Method 1: General Linear Hypothesis (Manual Matrix & car Package)**

This approach tests the hypothesis $H_0: C\beta = 0$. We calculate the Sum of Squares for the Hypothesis ($SSH$) using the general linear hypothesis formula and then structure the results into a standard ANOVA table format.
**Method 1: General Linear Hypothesis (Manual Matrix & car Package)**

This approach tests the hypothesis $H_0: C\beta = 0$. We calculate the Sum of Squares for the Hypothesis ($SSH$) using the general linear hypothesis formula and then structure the results into a standard ANOVA table format.

```{r}
#| echo: true
#| label: method-1-manual-and-car
#| message: false
library(knitr)
library(car)

# Load Data 
chemical_data <- data.frame(
  y1 = c(41.5, 33.8, 27.7, 21.7, 19.9, 15.0, 12.2, 4.3, 19.3, 6.4, 
         37.6, 18.0, 26.3, 9.9, 25.0, 14.1, 15.2, 15.9, 19.6),
  x1 = c(162, 162, 162, 162, 172, 172, 172, 172, 167, 177, 
         157, 167, 167, 167, 167, 177, 177, 160, 160),
  x2 = c(23, 23, 30, 30, 25, 25, 30, 30, 27.5, 27.5, 
         27.5, 32.5, 22.5, 27.5, 27.5, 20, 20, 34, 34),
  x3 = c(3, 8, 5, 8, 5, 8, 5, 8, 6.5, 6.5, 
         6.5, 6.5, 6.5, 9.5, 3.5, 6.5, 6.5, 7.5, 7.5)
)

# Fit Full Model
full_model <- lm(y1 ~ x1 + x2 + x3, data = chemical_data)
beta_hat <- coef(full_model)
XtX_inv  <- summary(full_model)$cov.unscaled
SSE_FM   <- sum(residuals(full_model)^2)
n <- nrow(chemical_data)
k <- 3

# --- Manual Matrix Implementation ---
C <- matrix(c(0, 2, -2,  0,
              0, 0,  2, -1), nrow = 2, byrow = TRUE)
q <- nrow(C) 

Cb <- C %*% beta_hat
middle_term <- solve(C %*% XtX_inv %*% t(C))
SSH <- as.numeric(t(Cb) %*% middle_term %*% Cb)

# Constructing ANOVA components
df_error <- n - k - 1
MSE <- SSE_FM / df_error
F_stat <- (SSH / q) / MSE
p_val <- pf(F_stat, q, df_error, lower.tail = FALSE)

# Display as an ANOVA table
anova_manual <- data.frame(
  Df = c(q, df_error),
  `Sum Sq` = c(SSH, SSE_FM),
  `Mean Sq` = c(SSH/q, MSE),
  `F value` = c(F_stat, NA),
  `Pr(>F)` = c(p_val, NA),
  row.names = c("Hypothesis", "Residuals")
)

print(as.table(as.matrix(anova_manual)))

# --- Verification with car package ---
hypotheses <- c("2*x1 - 2*x2 = 0", "2*x2 - x3 = 0")
print(linearHypothesis(full_model, hypotheses))
```

---

**Method 2: Nested Models Approach**

In this method, we analyze the change in error when moving from the Full Model to a Reduced Model. The plot illustrates $SSE$ against the number of parameters $p$. Labels are positioned to the right of each data point for clarity.



```{r}
#| echo: true
#| label: method-2-nested
#| message: false
library(ggplot2)

# Create the reduced variable z based on constraints
chemical_data$z <- 0.5 * chemical_data$x1 + 0.5 * chemical_data$x2 + chemical_data$x3

# Fit Models
reduced_model <- lm(y1 ~ z, data = chemical_data)

# Data for Plotting
n_obs <- nrow(chemical_data)
p_full <- length(coef(full_model))
p_red  <- length(coef(reduced_model))
p_sat  <- n_obs

viz_data <- data.frame(
  p = c(p_sat, p_full, p_red),
  sse = c(0, sum(residuals(full_model)^2), sum(residuals(reduced_model)^2)),
  Model = c("Saturated", "Full Model", "Reduced Model")
)

ggplot(viz_data, aes(x = p, y = sse)) +
  geom_line(color = "gray80", linetype = "dashed") +
  geom_point(aes(color = Model), size = 4) +
  # Use nudge_x and hjust=0 to place text on the right
  geom_text(aes(label = Model), hjust = 0, nudge_x = 0.5, size = 3.5) +
  labs(
    title = "SSE vs. Number of Parameters (p)",
    x = "Number of Parameters (p)",
    y = "Sum of Squares Error (SSE)"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Full Model" = "red", 
                                "Reduced Model" = "blue", 
                                "Saturated" = "darkgreen")) +
  # Expand x-axis slightly to make room for text on the right
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.3))) +
  ylim(0, max(viz_data$sse) * 1.1)

# Display raw text output from anova for nested models
anova(reduced_model, full_model)
```
::::

## Specific Tests on Coefficients

Tests for individual coefficients or linear combinations are special cases of the general linear hypothesis where $q=1$ .

### Test for $a^T \beta = 0$

To test a linear combination $a^T \beta = 0$, the F-statistic is:

$$
F = \frac{(a^T \hat{\beta})^2}{s^2 a^T (X^T X)^{-1} a} \sim F(1, n - k - 1)
$$

This is equivalent to the t-test:

$$
t = \frac{a^T \hat{\beta}}{s \sqrt{a^T (X^T X)^{-1} a}} \sim t(n - k - 1)
$$

### Test for $\beta_j = 0$

For a single coefficient $\beta_j$, we set $a$ to extract the $j$-th element. The test statistic becomes:

$$
t = \frac{\hat{\beta}_j}{s \sqrt{g_{jj}}} = \frac{\hat{\beta}_j}{\text{SE}(\hat{\beta}_j)}
$$

where $g_{jj}$ is the $j$-th diagonal element of $(X^T X)^{-1}$.

## Confidence and Prediction Intervals

### Confidence Region for $\beta$

A $100(1-\alpha)\%$ confidence region for the vector $\beta$ is the set of all vectors satisfying :

$$
(\hat{\beta} - \beta)^T X^T X (\hat{\beta} - \beta) \le (k + 1) s^2 F_{1-\alpha}(k + 1, n - k - 1)
$$

This region forms an ellipsoid centered at $\hat{\beta}$.

### Confidence Interval for $a^T \beta$ and $E(y_0)$

A $100(1-\alpha)\%$ confidence interval for a linear combination $a^T \beta$ is given by:

$$
a^T \hat{\beta} \pm t_{1-\alpha/2}(n - k - 1) s \sqrt{a^T (X^T X)^{-1} a}
$$

To estimate the mean response $E(y_0) = x_0^T \beta$ for a given $x_0$, we set $a = x_0$ :

$$
x_0^T \hat{\beta} \pm t_{1-\alpha/2}(n - k - 1) s \sqrt{x_0^T (X^T X)^{-1} x_0}
$$

### Prediction Interval for $y_0$

When predicting a new random observation $y_0 = x_0^T \beta + e_0$, we must account for both the variance of the estimator and the variance of the new error term.

$$
\text{Var}(y_0 - \hat{y}_0) = \text{Var}(e_0) + \text{Var}(x_0^T \hat{\beta}) = \sigma^2 (1 + x_0^T (X^T X)^{-1} x_0)
$$

The $100(1-\alpha)\%$ prediction interval is :

$$
\hat{y}_0 \pm t_{1-\alpha/2}(n - k - 1) s \sqrt{1 + x_0^T (X^T X)^{-1} x_0}
$$