---
engine: knitr
format: 
  html: default
  pdf: default
---

# Estimation and Inference with Non-full-rank Models

## Non-full-rank Models and Parameter Non-identifiability

**The One-way Model**

Consider the balanced one-way layout model for $y_{ij}$ a response on the $j^{th}$ unit in the $i^{th}$ treatment group. Suppose that there are $a$ treatments and $n$ units in the $i^{th}$ treatment group.

The **cell-means model** for this situation is:

$$
y_{ij} = \mu_i + e_{ij}, \quad i=1,...,a, \quad j=1,...,n
$$

where the $e_{ij}$ are i.i.d. $N(0,\sigma^2)$.

An alternative, but equivalent, linear model is the **effects model** for the one-way layout:

$$
y_{ij} = \mu + \alpha_i + e_{ij}, \quad i=1,...,a, \quad j=1,...,n
$$

with the same assumptions on the errors.

The cell means model can be written in vector notation as:

$$
y = \mu_1 x_1 + \mu_2 x_2 + \dots + \mu_a x_a + e, \quad e \sim N(0, \sigma^2 I)
$$

and the effects model can be written as:

$$
y = \mu j_N + \alpha_1 x_1 + \alpha_2 x_2 + \dots + \alpha_a x_a + e, \quad e \sim N(0, \sigma^2 I)
$$

where $x_i$ is an indicator for treatment $i$, and $N=an$ is the total sample size.


The effects model has the same model matrix as the cell-means model, but with one extra column, a column of ones, in the first position.

Notice that $\sum_i x_i = j_N$. Therefore, the columns of the model matrix for the effects model are linearly dependent.

Let $X_1$ denote the model matrix in the cell-means model, and $X_2 = (j_N, X_1)$ denote the model matrix in the effects model. Note that $\text{Col}(X_1) = \text{Col}(X_2)$.




Without constraints, the parameters in the effects model are not uniquely identified. $\mu$ is in no sense the grand mean; it is just an arbitrary baseline value.
However, subject to the constraint $\sum_i \alpha_i = 0$, the parameters have specific interpretations:

1.  **Grand Mean ($\mu$)**: The grand mean response across all treatments.
2.  **Treatment Effect ($\alpha_i$)**: The deviation from the grand mean.

Adding the constraint $\sum_i \alpha_i = 0$ essentially reparameterizes the overparameterized (non-full rank) effects model to a just-parameterized (full rank) model equivalent to the cell means model.


::: {#def-equivalent-models}

#### Equivalent Linear Models
In general, two linear models $y = X_1 \beta_1 + e_1$ and $y = X_2 \beta_2 + e_2$ with the same assumptions on $e_1$ and $e_2$ are equivalent linear models if $\text{Col}(X_1) = \text{Col}(X_2)$.

:::


**Strategies for Non-Full Rank Models**

When faced with a non-full rank model, we have three ways to proceed:

1.  **Reparameterize to a full rank model** (e.g., cell-means model).
2.  **Add constraints (side-conditions)** to the model parameters (e.g., $\sum \alpha_i = 0$).
3.  **Analyze the model as a non-full rank model**, limiting inference to **estimable** functions.

## Least Square Estimation of $\beta$ and $\mu=X\beta$

Even if $X$ is not of full rank, the least-squares criterion is still reasonable and leads to the normal equation:

$$
X^T X \hat{\beta} = X^T y
$$

::: {#thm-consistency-normal-eqn}

#### Consistency of Normal Equations
For $X$ an $n \times p$ matrix of rank $k < p \le n$, the normal equations form a consistent system of equations because $\text{Col}(X^T X) = \text{Col}(X^T)$.

:::

Since the system is consistent, it has a (non-unique) solution given by:

$$
\hat{\beta} = (X^T X)^- X^T y
$$

where $(X^T X)^-$ is some (any) generalized inverse of $X^T X$.

::: {#thm-projection-uniqueness}

#### Uniqueness of Projection
$\hat{\beta} = (X^T X)^- X^T y$ is a solution to $(X^T X)\beta = X^T y$.
Then $\hat{y} = X\hat{\beta} = X(X^T X)^- X^T y$ is the projection of $y$ onto $\text{Col}(X)$.

Since the projection onto a subspace is unique, $\hat{y}$ is unique regardless of the choice of generalized inverse.

:::

### Distribution of $\hat{\beta}$ and $s^2$

In the model $y = X\beta + e$, where $E(e)=0$, $\text{Var}(e)=\sigma^2 I$, and $X$ is $n \times p$ of rank $k \le p \le n$:

1.  **Unbiasedness**: $E(s^2) = \sigma^2$.
2.  **Uniqueness**: $s^2$ is invariant to the choice of $\hat{\beta}$ (i.e., to the choice of $(X^T X)^-$).

$$
s^2 = \frac{SSE}{n-k} = \frac{||y - X\hat{\beta}||^2}{n-k}
$$

where $k = \text{rank}(X)$.

::: {#thm-distributions-non-full-rank}

#### Distributions in Non-Full Rank Models
In the normal-errors model $y \sim N(X\beta, \sigma^2 I)$:

1.  For any given choice of $(X^T X)^-$:

    
    $$
    \hat{\beta} \sim N_p [(X^T X)^- X^T X \beta, \sigma^2 (X^T X)^- X^T X \{(X^T X)^-\}^T]
    $$

2.  $(n-k)s^2 / \sigma^2 \sim \chi^2(n-k)$ and $SSE/\sigma^2 \sim \chi^2(n-k)$.
3.  For any given choice of $(X^T X)^-$, $\hat{\beta}$ and $s^2$ are independent.

:::

## Sum Squares and F-test

Suppose $y \sim N_n(X\beta, \sigma^2 I)$ where $X$ is a matrix with rank $k+1$. Let $X = [X_1, X_2]$ where $\text{rank}(X) = k+1$ and assume we are testing a reduced model involving $X_1$.

Let:

* $\hat{y} = P_{\text{Col}(X)} y$ (Full Model Projection)
* $\hat{y}_0 = P_{\text{Col}(X_1)} y$ (Reduced Model Projection)

::: {#thm-partitioning-chi-square}

#### Partitioning of Sum of Squares

1.  **Full Model Residual Sum of Squares**:

    
    $$
    \frac{1}{\sigma^2} ||y - \hat{y}||^2 = \frac{1}{\sigma^2} y^T (I - P_{\text{Col}(X)}) y \sim \chi^2(n - k - 1)
    $$
    
    

2.  **Difference in Sum of Squares (Hypothesis)**:

    
    $$
    \frac{1}{\sigma^2} ||\hat{y} - \hat{y}_0||^2 = \frac{1}{\sigma^2} y^T (P_{\text{Col}(X)} - P_{\text{Col}(X_1)}) y \sim \chi^2(h, \lambda_1)
    $$
    
    where $h$ is the difference in rank (degrees of freedom). The non-centrality parameter is:
    
    $$
    \lambda_1 = \frac{1}{2\sigma^2} ||(P_{\text{Col}(X)} - P_{\text{Col}(X_1)}) \mu||^2 = \frac{1}{2\sigma^2} ||\mu - \mu_0||^2
    $$
    
    

3.  **Independence**: The two quadratic forms above are independent.

:::

::: {#thm-f-test-statistic}

#### F-Statistic for Nested Models
Under the conditions of the previous theorem:

$$
F = \frac{||\hat{y} - \hat{y}_0||^2 / h}{s^2} = \frac{y^T (P_{\text{Col}(X)} - P_{\text{Col}(X_1)}) y / h}{y^T (I - P_{\text{Col}(X)}) y / (n - k - 1)}
$$

$$
\sim \begin{cases} F(h, n - k - 1), & \text{under } H_0 \\ F(h, n - k - 1, \lambda_1), & \text{under } H_1 \end{cases}
$$


:::

### Examples


:::{#exm-anova-non-full-rank}

**Numerical Example: One-way ANOVA with Non-Full Rank Matrix**

We generate a balanced dataset with $a=3$ groups and $n=4$ observations per group. We then explicitly construct the non-full-rank design matrix $X = [j_N, x_1, x_2, x_3]$ and compute projections. We add a column for $\hat{y} - \hat{y}_0$ to explicitly show the vector whose squared length equals the Between Group Sum of Squares.

```{r}
#| echo: true
#| label: anova-ginv-example
#| message: false
library(MASS)  # For ginv()
library(knitr) # For kable()

# 1. Data Generation (3 Groups, 4 obs per group)
group_means_true <- c(10, 20, 15)
n_per_group <- 4
groups <- factor(rep(1:3, each = n_per_group))
# Adding some random noise to exact means for realism
set.seed(123)
y <- c(rnorm(n_per_group, group_means_true[1], 2),
       rnorm(n_per_group, group_means_true[2], 2),
       rnorm(n_per_group, group_means_true[3], 2))
n <- length(y)

# 2. Construct Non-full-rank Design Matrix X
# j_N is the column of ones
j_N <- rep(1, n)
# Indicators for all 3 groups
X_ind <- model.matrix(~ groups - 1) 
colnames(X_ind) <- c("x_1", "x_2", "x_3")
# Full non-full-rank matrix
X <- cbind(Intercept = j_N, X_ind)

# 3. Compute Projections
# Projection onto j_N (Grand Mean) - Reduced Model
P0 <- j_N %*% t(j_N) / as.numeric(t(j_N) %*% j_N)
y_grand_mean <- P0 %*% y

# Projection onto Group Means (Cell Means Model)
X_ind_inv <- ginv(t(X_ind) %*% X_ind)
P_groups <- X_ind %*% X_ind_inv %*% t(X_ind)
y_group_means <- P_groups %*% y

# Projection onto [j_N, X] (Full Column Space)
XtX <- t(X) %*% X
X_inv_gen <- ginv(XtX)
P_X <- X %*% X_inv_gen %*% t(X)
y_proj_X <- P_X %*% y

# Difference Vector (Hypothesis Space)
y_diff <- y_proj_X - y_grand_mean

# Residuals (Error)
e <- y - y_proj_X

# Combine into a data frame
results_df <- data.frame(
  y = y,
  j_N = j_N,
  x1 = X_ind[,1],
  x2 = X_ind[,2],
  x3 = X_ind[,3],
  Proj_Grand = as.vector(y_grand_mean),
  Proj_FullX = as.vector(y_proj_X),
  Diff = as.vector(y_diff),          # Added Column
  Residuals = as.vector(e)
)

# Calculate Sum of Squares for each column
SS_row <- colSums(results_df^2)
results_table <- rbind(results_df, SS_row)
rownames(results_table)[nrow(results_table)] <- "Sum of Squares"

# 4. Compute F and p-value
# SSH = || Diff ||^2
SSH <- sum(y_diff^2)
# SSE = || Residuals ||^2
SSE <- sum(e^2)

# Degrees of Freedom
rank_X <- 3 
df_hyp <- rank_X - 1
df_err <- n - rank_X

MSH <- SSH / df_hyp
MSE <- SSE / df_err
F_stat <- MSH / MSE
p_val <- pf(F_stat, df_hyp, df_err, lower.tail = FALSE)

# Display Detailed Data Table
kable(results_table, digits = 2, 
      col.names = c("$y$", "$j_N$", "$x_1$", "$x_2$", "$x_3$", 
                    "$\\hat{y}_0$", 
                    "$\\hat{y}$", 
                    "$\\hat{y} - \\hat{y}_0$", # Label for new column
                    "$e$"),
      caption = "Data, Projections, and Difference Vectors with Sum of Squares")

# 5. Output ANOVA Table
anova_tab <- data.frame(
  Source = c("Groups (Model)", "Residuals (Error)", "Total"),
  DF = c(df_hyp, df_err, n - 1),
  SS = c(SSH, SSE, SSH + SSE),
  MS = c(MSH, MSE, NA),
  F = c(F_stat, NA, NA),
  P_value = c(p_val, NA, NA)
)

kable(anova_tab, digits = 4, caption = "ANOVA Table derived from Projections")
```



**Key Observation from the Table**

* Note that the projection onto the group means ($\hat{y}_{groups}$) and the projection onto the overparameterized space ($\hat{y} = P_{[j_N, X]}y$) are identical (Columns 7 and 8). This confirms numerically that adding the linearly dependent intercept column does not change the column space of the model; $\text{Col}(X) = \text{Col}(\text{Indicators})$.

* The Sum of Squares for the new column $\hat{y} - \hat{y}_0$ (Column 8) is exactly equal to the **Between Group Sum of Squares (SSH)** in the ANOVA table. This numerically demonstrates that $SSH = ||\hat{y} - \hat{y}_0||^2$.
:::



:::{#exm-regression-non-full-rank}

**Numerical Example: Regression with Linear Dependency**

We generate a dataset with $n=12$ observations and 4 predictors where $x_4 = 0.1x_1 + 0.5x_2 + x_3$. This makes the design matrix rank-deficient. We use the generalized inverse to compute projections onto the full model space and compare them with projections onto the subspace of just $x_1, x_2, x_3$.

```{r}
#| echo: true
#| label: regression-ginv-example
#| message: false
#| warning: false
library(MASS)  # For ginv()
library(knitr) # For kable()
library(gt)    # For HTML tables

# 1. Data Generation
set.seed(42)
n <- 12
x1 <- rnorm(n, mean=5, sd=2)
x2 <- rnorm(n, mean=10, sd=2)
x3 <- rnorm(n, mean=0, sd=1)
# Linear Dependency
x4 <- 0.1*x1 + 0.5*x2 + x3 

# Generate y
beta_true <- c(10, 2, -1, 3) 
y <- beta_true[1] + beta_true[2]*x1 + beta_true[3]*x2 + beta_true[4]*x3 + rnorm(n, sd=1)

# 2. Construct Design Matrices
j_N <- rep(1, n)
X_full <- cbind(Intercept = j_N, x1, x2, x3, x4)
X_sub  <- cbind(Intercept = j_N, x1, x2, x3)

# 3. Compute Projections

# A. Projection onto Null Model (Grand Mean)
P0 <- j_N %*% t(j_N) / as.numeric(t(j_N) %*% j_N)
y_0 <- P0 %*% y

# B. Projection onto Full Model Space [j_N, x1...x4] using ginv
#    (Rank Deficient)
XtX_full <- t(X_full) %*% X_full
X_inv_full <- ginv(XtX_full)
beta_hat_full <- X_inv_full %*% t(X_full) %*% y
y_hat_full <- X_full %*% beta_hat_full

# C. Projection onto Subspace [j_N, x1, x2, x3]
#    (Full Rank)
XtX_sub <- t(X_sub) %*% X_sub
# Standard inverse since this matrix is full rank
X_inv_sub <- solve(XtX_sub) 
beta_hat_sub <- X_inv_sub %*% t(X_sub) %*% y
y_hat_sub <- X_sub %*% beta_hat_sub

# D. Difference Vector and Residuals
y_diff <- y_hat_full - y_0
e <- y - y_hat_full

# 4. Create Detailed Results Table
results_df <- data.frame(
  y = y,
  x1 = x1,
  x2 = x2,
  x3 = x3,
  x4 = x4,
  y_hat_0 = as.vector(y_0),
  y_hat_sub = as.vector(y_hat_sub),   # New Column
  y_hat_full = as.vector(y_hat_full),
  y_diff = as.vector(y_diff),
  e = as.vector(e)
)

# Calculate Sum of Squares for each column
SS_row <- colSums(results_df^2)
results_table <- rbind(results_df, SS_row)
rownames(results_table)[nrow(results_table)] <- "Sum of Squares"

# Display Table using kable (for standard rendering)
kable(results_table, digits = 2,
      col.names = c("$y$", "$x_1$", "$x_2$", "$x_3$", "$x_4$",
                    "$\\hat{y}_0$", 
                    "$\\hat{y}_{sub}$",      # Projection onto x1-x3
                    "$\\hat{y}_{full}$",     # Projection onto x1-x4
                    "$\\hat{y}_{full} - \\hat{y}_0$", 
                    "$e$"),
      caption = "Comparison of Projections: Full (Rank-Deficient) vs Subspace")

# 5. ANOVA Statistics
SSH <- sum(y_diff^2)
SSE <- sum(e^2)
rank_X <- qr(X_full)$rank 
df_hyp <- rank_X - 1 
df_err <- n - rank_X
MSH <- SSH / df_hyp
MSE <- SSE / df_err
F_stat <- MSH / MSE
p_val <- pf(F_stat, df_hyp, df_err, lower.tail = FALSE)

# 6. Output ANOVA Table with Coloring
anova_df <- data.frame(
  Source = c("Regression (Model)", "Residuals (Error)", "Total"),
  DF = c(df_hyp, df_err, n - 1),
  SS = c(SSH, SSE, SSH + SSE),
  MS = c(MSH, MSE, NA),
  F = c(F_stat, NA, NA),
  P_value = c(p_val, NA, NA)
)

if (knitr::is_html_output()) {
  anova_df |>
    gt() |>
    fmt_number(columns = c(SS, MS, F, P_value), decimals = 4) |>
    tab_style(
      style = cell_text(color = "red", weight = "bold"),
      locations = cells_body(columns = DF, rows = Source == "Regression (Model)")
    ) |>
    tab_header(title = "ANOVA Table")
} else {
  anova_df$DF <- as.character(anova_df$DF)
  anova_df$DF[1] <- paste0("\\textcolor{red}{", anova_df$DF[1], "}")
  kable(anova_df, digits = 4, escape = FALSE, caption = "ANOVA Table")
}
```

**Key Observation**

Notice that the columns $\hat{y}_{sub}$ (projection onto $x_1, x_2, x_3$) and $\hat{y}_{full}$ (projection onto $x_1, \dots, x_4$) are identical. This confirms that adding the linearly dependent predictor $x_4$ does not change the column space of the design matrix or the fitted values, even though it makes individual coefficients non-unique.
:::


:::{#exm-lm-rank-deficient}

**Fitting Rank-Deficient Models with `lm()`**

In R, the standard `lm()` function automatically detects singularities. It uses QR decomposition with pivoting to handle rank deficiency. If a column is found to be linearly dependent on previous columns, its coefficient is set to `NA`.

We use the same dataset as before, where $x_4 = 0.1x_1 + 0.5x_2 + x_3$.

```{r}
#| echo: true
#| label: lm-singularity-example
#| message: false
#| warning: false

# 1. Data Generation (Same as previous)
set.seed(42)
n <- 12
x1 <- rnorm(n, mean=5, sd=2)
x2 <- rnorm(n, mean=10, sd=2)
x3 <- rnorm(n, mean=0, sd=1)
# Linear Dependency
x4 <- 0.1*x1 + 0.5*x2 + x3 
# Generate y
beta_true <- c(10, 2, -1, 3) 
y <- beta_true[1] + beta_true[2]*x1 + beta_true[3]*x2 + beta_true[4]*x3 + rnorm(n, sd=1)

# Combine into data frame
df <- data.frame(y=y, x1=x1, x2=x2, x3=x3, x4=x4)

# 2. Fit model using lm() with tryCatch
# Note: lm() usually just drops coefs, but we use tryCatch as requested
lm_fit <- tryCatch({
  model <- lm(y ~ x1 + x2 + x3 + x4, data = df)
  print("Model fitted successfully (with potential NAs).")
  model
}, error = function(e) {
  print(paste("Error in lm:", e$message))
  return(NULL)
})

# 3. Examine Coefficients
# Expecting one coefficient to be NA because of the singularity
print(coef(lm_fit))

# 4. Verify Fitted Values
# Even though coefficients are not unique (NA appears), fitted values ARE unique
y_hat_lm <- predict(lm_fit)

# Compare with previous ginv result (re-calculating locally for comparison)
X <- model.matrix(lm_fit)
library(MASS)
beta_ginv <- ginv(t(X) %*% X) %*% t(X) %*% y
y_hat_ginv <- X %*% beta_ginv

# Check difference
max_diff <- max(abs(y_hat_lm - y_hat_ginv))
print(paste("Maximum difference between lm() and ginv() fitted values:", max_diff))
```

**Observation**

The `lm()` output shows `NA` for `x4` (or whichever variable is pivoted out last). This confirms that while the individual parameters are not identifiable (infinite solutions exist), the `lm()` function picks one specific solution (setting one coef to 0) to allow for prediction. The fitted values $\hat{y}$ are identical to those obtained via the generalized inverse.
:::

## Inference for Estimable Parameters

::: {#def-estimability}

#### Estimability
Let $\lambda = (\lambda_1, ..., \lambda_p)^T$ be a vector of constants. The parameter $\lambda^T \beta = \sum_j \lambda_j \beta_j$ is said to be estimable if there exists a vector $a$ in $R^n$ such that:

$$
E(a^T y) = \lambda^T \beta
$$

for all $\beta \in \mathcal{R}^p$.

:::

This condition is equivalent to $\lambda^T \beta$ being estimable if and only if there exists $a$ such that $X^T a = \lambda$ (i.e., iff $\lambda$ lies in the row space of $X$).

::: {#thm-estimability-conditions}

#### Equivalent Conditions for Estimability
In the model $y = X\beta + e$, the linear function $\lambda^T \beta$ is estimable if and only if any one of the following equivalent conditions holds:

1.  $\lambda^T$ is a linear combination of the rows of $X$; that is, there exists a vector $a$ such that $a^T X = \lambda^T$.
2.  $\lambda^T$ is a linear combination of the rows of $X^T X$, or $\lambda$ is a linear combination of the columns of $X^T X$; that is, there exists a vector $r$ such that $X^T X r = \lambda$.
3.  $\lambda$ is such that $X^T X (X^T X)^- \lambda = \lambda$.
4.  $\lambda \in \text{Row}(X) = \text{Col}(X^TX)$

:::

::: {#thm-number-estimable-functions}

#### Number of Estimable Functions
In the non-full-rank model $y = X\beta + e$, the number of linearly independent estimable functions of $\beta$ is the rank of $X$.

:::

**Properties of Estimators**

Let $\lambda^T \beta$ be an estimable function. Let $\hat{\beta}$ be any solution to the normal equations. Then the estimator $\lambda^T \hat{\beta}$ has the following properties:

1.  **Unbiasedness**: $E(\lambda^T \hat{\beta}) = \lambda^T \beta$.
2.  **Uniqueness**: $\lambda^T \hat{\beta}$ is invariant to the choice of $\hat{\beta}$ (to the choice of generalized inverse).

::: {#thm-variance-covariance}

#### Variance and Covariance

1.  **Variance**: The variance of $\lambda^T \hat{\beta}$ is unique and is given by:

    
    $$
    \text{Var}(\lambda^T \hat{\beta}) = \sigma^2 \lambda^T (X^T X)^- \lambda
    $$

2.  **Covariance**: For two estimable functions $\lambda_1^T \beta$ and $\lambda_2^T \beta$:

    
    $$
    \text{Cov}(\lambda_1^T \hat{\beta}, \lambda_2^T \hat{\beta}) = \sigma^2 \lambda_1^T (X^T X)^- \lambda_2
    $$

:::

::: {#thm-gauss-markov-non-full}

#### Gauss-Markov in Non-Full Rank Case
If $\lambda^T \beta$ is estimable in the spherical errors non-full rank linear model, then $\lambda^T \hat{\beta}$ is its Best Linear Unbiased Estimator (BLUE).

:::

::: {#def-testable-hypothesis}

#### Testable Hypothesis
A hypothesis such as $H_0: \beta_1 = ... = \beta_q$ is said to be testable if there exists a set of linearly independent estimable functions such that $H_0$ is true if and only if these functions equal zero.

:::

## Re-parametrization and Side Conditions

**Re-parametrization**

The idea is to transform from the vector of non-estimable parameters $\beta$ to a vector of linearly independent estimable functions of $\beta$: $\gamma = U\beta$.
The new full-rank model is $y = Z\gamma + e$, where $Z\gamma = X\beta$.

To find $Z$, we solve $Z U = X$. One solution is $Z = X U^T (U U^T)^{-1}$.

**Side Conditions**

::: {#thm-side-conditions}

#### Unique Solution with Side Conditions
If $y = X\beta + e$ where $X$ is $n \times p$ of rank $k$, and if $T$ is a $(p-k) \times p$ matrix of rank $p-k$ such that $T\beta$ is a set of non-estimable functions, then there is a unique vector $\hat{\beta}$ that satisfies both $X^T X \hat{\beta} = X^T y$ and $T\hat{\beta} = 0$.

:::

The solution is given by:

$$
\hat{\beta} = (X^T X + T^T T)^{-1} X^T y
$$

