---
engine: knitr
format: 
  html: default
  pdf: default
---

# Estimation and Inference with Non-full-rank Models

## Non-full-rank Models and Parameter Non-identifiability

**The One-way Model**

Consider the balanced one-way layout model for $y_{ij}$ a response on the $j^{th}$ unit in the $i^{th}$ treatment group. Suppose that there are $a$ treatments and $n$ units in the $i^{th}$ treatment group.

The **cell-means model** for this situation is:

$$
y_{ij} = \mu_i + e_{ij}, \quad i=1,...,a, \quad j=1,...,n
$$

where the $e_{ij}$ are i.i.d. $N(0,\sigma^2)$.

An alternative, but equivalent, linear model is the **effects model** for the one-way layout:

$$
y_{ij} = \mu + \alpha_i + e_{ij}, \quad i=1,...,a, \quad j=1,...,n
$$

with the same assumptions on the errors.

The cell means model can be written in vector notation as:

$$
y = \mu_1 x_1 + \mu_2 x_2 + \dots + \mu_a x_a + e, \quad e \sim N(0, \sigma^2 I)
$$

and the effects model can be written as:

$$
y = \mu j_N + \alpha_1 x_1 + \alpha_2 x_2 + \dots + \alpha_a x_a + e, \quad e \sim N(0, \sigma^2 I)
$$

where $x_i$ is an indicator for treatment $i$, and $N=an$ is the total sample size.


The effects model has the same model matrix as the cell-means model, but with one extra column, a column of ones, in the first position.

Notice that $\sum_i x_i = j_N$. Therefore, the columns of the model matrix for the effects model are linearly dependent.

Let $X_1$ denote the model matrix in the cell-means model, and $X_2 = (j_N, X_1)$ denote the model matrix in the effects model. Note that $\text{Col}(X_1) = \text{Col}(X_2)$.




Without constraints, the parameters in the effects model are not uniquely identified. $\mu$ is in no sense the grand mean; it is just an arbitrary baseline value.
However, subject to the constraint $\sum_i \alpha_i = 0$, the parameters have specific interpretations:

1.  **Grand Mean ($\mu$)**: The grand mean response across all treatments.
2.  **Treatment Effect ($\alpha_i$)**: The deviation from the grand mean.

Adding the constraint $\sum_i \alpha_i = 0$ essentially reparameterizes the overparameterized (non-full rank) effects model to a just-parameterized (full rank) model equivalent to the cell means model.


::: {#def-equivalent-models}

#### Equivalent Linear Models
In general, two linear models $y = X_1 \beta_1 + e_1$ and $y = X_2 \beta_2 + e_2$ with the same assumptions on $e_1$ and $e_2$ are equivalent linear models if $\text{Col}(X_1) = \text{Col}(X_2)$.

:::


**Strategies for Non-Full Rank Models**

When faced with a non-full rank model, we have three ways to proceed:

1.  **Reparameterize to a full rank model** (e.g., cell-means model).
2.  **Add constraints (side-conditions)** to the model parameters (e.g., $\sum \alpha_i = 0$).
3.  **Analyze the model as a non-full rank model**, limiting inference to **estimable** functions.

## Least Square Estimation of $\beta$ and $\mu=X\beta$

Even if $X$ is not of full rank, the least-squares criterion is still reasonable and leads to the normal equation:

$$
X^T X \hat{\beta} = X^T y
$$

::: {#thm-consistency-normal-eqn}

#### Consistency of Normal Equations
For $X$ an $n \times p$ matrix of rank $k < p \le n$, the normal equations form a consistent system of equations because $\text{Col}(X^T X) = \text{Col}(X^T)$.

:::

Since the system is consistent, it has a (non-unique) solution given by:

$$
\hat{\beta} = (X^T X)^- X^T y
$$

where $(X^T X)^-$ is some (any) generalized inverse of $X^T X$.

::: {#thm-projection-uniqueness-nonfulrank}

#### Uniqueness of Projection
$\hat{\beta} = (X^T X)^- X^T y$ is a solution to $(X^T X)\beta = X^T y$.
Then $\hat{y} = X\hat{\beta} = X(X^T X)^- X^T y$ is the projection of $y$ onto $\text{Col}(X)$.

Since the projection onto a subspace is unique, $\hat{y}$ is unique regardless of the choice of generalized inverse.

:::

### Distribution of $\hat{\beta}$ and $s^2$

In the model $y = X\beta + e$, where $E(e)=0$, $\text{Var}(e)=\sigma^2 I$, and $X$ is $n \times p$ of rank $k \le p \le n$:

1.  **Unbiasedness**: $E(s^2) = \sigma^2$.
2.  **Uniqueness**: $s^2$ is invariant to the choice of $\hat{\beta}$ (i.e., to the choice of $(X^T X)^-$).

$$
s^2 = \frac{SSE}{n-k} = \frac{||y - X\hat{\beta}||^2}{n-k}
$$

where $k = \text{rank}(X)$.

::: {#thm-distributions-non-full-rank}

#### Distributions in Non-Full Rank Models
In the normal-errors model $y \sim N(X\beta, \sigma^2 I)$:

1.  For any given choice of $(X^T X)^-$:

    
    $$
    \hat{\beta} \sim N_p [(X^T X)^- X^T X \beta, \sigma^2 (X^T X)^- X^T X \{(X^T X)^-\}^T]
    $$

2.  $(n-k)s^2 / \sigma^2 \sim \chi^2(n-k)$ and $SSE/\sigma^2 \sim \chi^2(n-k)$.
3.  For any given choice of $(X^T X)^-$, $\hat{\beta}$ and $s^2$ are independent.

:::

## Sum Squares and F-test

Suppose $y \sim N_n(X\beta, \sigma^2 I)$ where $X$ is a matrix with rank $k+1$. Let $X = [X_1, X_2]$ where $\text{rank}(X) = k+1$ and assume we are testing a reduced model involving $X_1$.

Let:

* $\hat{y} = P_{\text{Col}(X)} y$ (Full Model Projection)
* $\hat{y}_0 = P_{\text{Col}(X_1)} y$ (Reduced Model Projection)

::: {#thm-partitioning-chi-square}

#### Partitioning of Sum of Squares

1.  **Full Model Residual Sum of Squares**:

    
    $$
    \frac{1}{\sigma^2} ||y - \hat{y}||^2 = \frac{1}{\sigma^2} y^T (I - P_{\text{Col}(X)}) y \sim \chi^2(n - k - 1)
    $$
    
    

2.  **Difference in Sum of Squares (Hypothesis)**:

    
    $$
    \frac{1}{\sigma^2} ||\hat{y} - \hat{y}_0||^2 = \frac{1}{\sigma^2} y^T (P_{\text{Col}(X)} - P_{\text{Col}(X_1)}) y \sim \chi^2(h, \lambda_1)
    $$
    
    where $h$ is the difference in rank (degrees of freedom). The non-centrality parameter is:
    
    $$
    \lambda_1 = \frac{1}{2\sigma^2} ||(P_{\text{Col}(X)} - P_{\text{Col}(X_1)}) \mu||^2 = \frac{1}{2\sigma^2} ||\mu - \mu_0||^2
    $$
    
    

3.  **Independence**: The two quadratic forms above are independent.

:::

::: {#thm-f-test-statistic}

#### F-Statistic for Nested Models
Under the conditions of the previous theorem:

$$
F = \frac{||\hat{y} - \hat{y}_0||^2 / h}{s^2} = \frac{y^T (P_{\text{Col}(X)} - P_{\text{Col}(X_1)}) y / h}{y^T (I - P_{\text{Col}(X)}) y / (n - k - 1)}
$$

$$
\sim \begin{cases} F(h, n - k - 1), & \text{under } H_0 \\ F(h, n - k - 1, \lambda_1), & \text{under } H_1 \end{cases}
$$


:::

### Examples

#### Numerical Example: One-way ANOVA with Non-Full Rank Matrix


:::{#exm-anova-non-full-rank}


We generate a balanced dataset with $a=3$ groups and $n=4$ observations per group. We then explicitly construct the non-full-rank design matrix $X = [j_N, x_1, x_2, x_3]$ and compute projections. We add a column for $\hat{y} - \hat{y}_0$ to explicitly show the vector whose squared length equals the Between Group Sum of Squares.

```{r}
#| echo: true
#| label: anova-ginv-example-corrected
#| message: false
library(MASS)  # For ginv()
library(knitr) # For kable()

# 1. Data Generation (3 Groups, 4 obs per group)
group_means_true <- c(10, 20, 15)
n_per_group <- 4
groups <- factor(rep(1:3, each = n_per_group))

# Adding some random noise
set.seed(123)
y <- c(rnorm(n_per_group, group_means_true[1], 2),
       rnorm(n_per_group, group_means_true[2], 2),
       rnorm(n_per_group, group_means_true[3], 2))
n <- length(y)

# 2. Construct Non-full-rank Design Matrix X
j_N <- rep(1, n)
X_ind <- model.matrix(~ groups - 1) 
colnames(X_ind) <- c("x_1", "x_2", "x_3")
X <- cbind(Intercept = j_N, X_ind)

# 3. Compute Projections
# A. Projection onto j_N (Grand Mean)
P0 <- j_N %*% t(j_N) / as.numeric(t(j_N) %*% j_N)
y_grand_mean <- P0 %*% y

# B. Projection onto [j_N, X] (Full Column Space)
XtX <- t(X) %*% X
X_inv_gen <- ginv(XtX)
P_X <- X %*% X_inv_gen %*% t(X)
y_proj_X <- P_X %*% y

# C. Difference Vector and Residuals
y_diff <- y_proj_X - y_grand_mean
e <- y - y_proj_X

# --- Compute Arithmetic Group Means for Verification ---
group_means_vec <- ave(y, groups)

# Combine into a data frame
results_df <- data.frame(
  y = y,
  j_N = j_N,
  x1 = X_ind[,1],
  x2 = X_ind[,2],
  x3 = X_ind[,3],
  Proj_Grand = as.vector(y_grand_mean),
  Proj_FullX = as.vector(y_proj_X),
  Group_Means = group_means_vec,
  Diff = as.vector(y_diff),          
  Residuals = as.vector(e)
)

# Calculate Sum of Squares for each column
SS_row <- colSums(results_df^2)
results_table <- rbind(results_df, SS_row)
rownames(results_table)[nrow(results_table)] <- "Sum of Squares"

# Display Detailed Data Table
kable(results_table, digits = 2, 
      col.names = c("$y$", "$j_N$", "$x_1$", "$x_2$", "$x_3$", 
                    "$\\hat{y}_0$", 
                    "$\\hat{y}$", 
                    "$\\bar{y}_{i\\cdot}$", # <--- Verification Header
                    "$\\hat{y} - \\hat{y}_0$", 
                    "$e$"),
      caption = "Data, Projections, and Verification of Group Means")

# 4. Compute F and p-value
SSH <- sum(y_diff^2)
SSE <- sum(e^2)
SST <- SSH + SSE  # Total Sum of Squares

rank_X <- 3 
df_hyp <- rank_X - 1
df_err <- n - rank_X
df_tot <- n - 1

MSH <- SSH / df_hyp
MSE <- SSE / df_err
MST <- SST / df_tot # Mean Square Total

F_stat <- MSH / MSE
p_val <- pf(F_stat, df_hyp, df_err, lower.tail = FALSE)

# 5. Adjusted R-squared
R2_adj <- 1 - (MSE / MST)

# 6. Output Extended ANOVA Table
anova_tab <- data.frame(
  Source = c("Groups (Model)", "Residuals (Error)", "Total"),
  DF = c(df_hyp, df_err, df_tot),
  SS = c(SSH, SSE, SST),
  MS = c(MSH, MSE, MST),
  F = c(F_stat, NA, NA),
  P_value = c(p_val, NA, NA),
  
  # Custom Variance Estimation Column
  Sigma_Hat_Sq = c(MST - MSE,  # Group Row: MST - MSE
                   MSE,        # Error Row: MSE
                   MST),       # Total Row: MST
  
  R_sq_adj = c(R2_adj, NA, NA)
)

kable(anova_tab, digits = 4, 
      col.names = c("Source", "DF", "SS", "MS", "F", "P-value", 
                    "$\\hat{\\sigma}^2$", "$R_a^2$"),
      caption = "ANOVA Table with Variance Estimates and Adjusted R-squared",
      escape = FALSE)
```



**Key Observation from the Table**

* Note that the projection onto the group means ($\hat{y}_{groups}$) and the projection onto the overparameterized space ($\hat{y} = P_{[j_N, X]}y$) are identical (Columns 7 and 8). This confirms numerically that adding the linearly dependent intercept column does not change the column space of the model; $\text{Col}(X) = \text{Col}(\text{Indicators})$.

* The Sum of Squares for the new column $\hat{y} - \hat{y}_0$ (Column 8) is exactly equal to the **Between Group Sum of Squares (SSH)** in the ANOVA table. This numerically demonstrates that $SSH = ||\hat{y} - \hat{y}_0||^2$.

:::


#### Numerical Example: Regression with Linear Dependency


::::{#exm-regression-non-full-rank}


We generate a dataset with $n=12$ observations and 4 predictors where $x_4 = 0.1x_1 + 0.5x_2 + x_3$. This makes the design matrix rank-deficient. We use the generalized inverse to compute projections onto the full model space and compare them with projections onto the subspace of just $x_1, x_2, x_3$.

1. Fitting Models and F-test
```{r}
#| echo: true
#| label: regression-ginv-example-extended-latex-gt
#| message: false
#| warning: false
library(MASS)  # For ginv()
library(knitr) # For kable()
library(gt)    # For HTML tables

# 1. Data Generation
set.seed(42)
n <- 12
x1 <- rnorm(n, mean=5, sd=2)
x2 <- rnorm(n, mean=10, sd=2)
x3 <- rnorm(n, mean=0, sd=1)
x4 <- 0.1*x1 + 0.5*x2 + 0.1*x3 
beta_true <- c(10, 1, -1, 1) 
y <- beta_true[1] + beta_true[2]*x1 + beta_true[3]*x2 + beta_true[4]*x3 + rnorm(n, sd=3)

# 2. Projections (Condensed for brevity, same as before)
j_N <- rep(1, n)
X_full <- cbind(Intercept = j_N, x1, x2, x3, x4)
X_sub  <- cbind(Intercept = j_N, x1, x2, x3)
P0 <- j_N %*% t(j_N) / as.numeric(t(j_N) %*% j_N)
y_0 <- P0 %*% y

# B. Projection onto Full Model Space (Rank Deficient)
XtX_full <- t(X_full) %*% X_full
X_inv_full <- ginv(XtX_full)
beta_hat_full <- X_inv_full %*% t(X_full) %*% y
y_hat_full <- X_full %*% beta_hat_full

# C. Projection onto Subspace (Full Rank)
XtX_sub <- t(X_sub) %*% X_sub
X_inv_sub <- solve(XtX_sub) 
beta_hat_sub <- X_inv_sub %*% t(X_sub) %*% y
y_hat_sub <- X_sub %*% beta_hat_sub

# D. Difference Vector and Residuals
y_diff <- y_hat_full - y_0
e <- y - y_hat_full

# 4. Create Detailed Results Table
results_df <- data.frame(
y = y,
x1 = x1,
x2 = x2,
x3 = x3,
x4 = x4,
y_hat_0 = as.vector(y_0),
y_hat_sub = as.vector(y_hat_sub),   # New Column
y_hat_full = as.vector(y_hat_full),
y_diff = as.vector(y_diff),
e = as.vector(e)
)

# Calculate Sum of Squares for each column
SS_row <- colSums(results_df^2)
results_table <- rbind(results_df, SS_row)
rownames(results_table)[nrow(results_table)] <- "Sum of Squares"

# Display Table using kable (for standard rendering)
kable(results_table, digits = 2,
      col.names = c("$y$", "$x_1$", "$x_2$", "$x_3$", "$x_4$",
                  "$\\hat{y}_0$", 
                  "$\\hat{y}_{sub}$",      # Projection onto x1-x3
                  "$\\hat{y}_{full}$",     # Projection onto x1-x4
                  "$\\hat{y}_{full} - \\hat{y}_0$", 
                  "$e$"),
      caption = "Comparison of Projections: Full (Rank-Deficient) vs Subspace")

# 5. ANOVA Statistics
SSH <- sum(y_diff^2)
SSE <- sum(e^2)
rank_X <- qr(X_full)$rank 
df_hyp <- rank_X - 1 
df_err <- n - rank_X
df_tol <- df_hyp + df_err
MSH <- SSH / df_hyp
MSE <- SSE / df_err
F_stat <- MSH / MSE
p_val <- pf(F_stat, df_hyp, df_err, lower.tail = FALSE)
SST <- SSH + SSE 
MST <- SST/df_tol
R2_adj <- 1-MSE/MST

# 4. Create ANOVA Data Frame
anova_df <- data.frame(
  Source = c("Regression (Model)", "Residuals (Error)", "Total"),
  DF = c(df_hyp, df_err, df_tot),
  SS = c(SSH, SSE, SST),
  MS = c(MSH, MSE, MST),
  F = c(F_stat, NA, NA),
  P_value = c(p_val, NA, NA),
  Sigma_Hat_Sq = c(MST - MSE, MSE, MST),
  R_sq_adj = c(R2_adj, NA, NA)
)

# 5. Output with LaTeX Headers in gt
if (knitr::is_html_output()) {
  anova_df |>
    gt() |>
    # Use md() to enable LaTeX parsing in headers
    cols_label(
      Source = "Source",
      DF = "DF",
      SS = "SS",
      MS = "MS",
      F = "F",
      P_value = "P-value",
      Sigma_Hat_Sq = md("$\\hat{\\sigma}^2$"),
      R_sq_adj = md("$R_a^2$")
    ) |>
    fmt_number(
      columns = c(SS, MS, F, P_value, Sigma_Hat_Sq, R_sq_adj), 
      decimals = 4
    ) |>
    sub_missing(columns = everything(), missing_text = "") |>
    tab_style(
      style = cell_text(color = "red", weight = "bold"),
      locations = cells_body(columns = DF, rows = Source == "Regression (Model)")
    ) |>
    tab_header(title = "ANOVA Table")
} else {
  # Standard LaTeX Output for PDF
  anova_df$DF <- as.character(anova_df$DF)
  anova_df$DF[1] <- paste0("\\textcolor{red}{", anova_df$DF[1], "}")
  
  kable(anova_df, digits = 4, escape = FALSE, 
        col.names = c("Source", "DF", "SS", "MS", "F", "P-value", 
                      "$\\hat{\\sigma}^2$", "$R_a^2$"),
        caption = "ANOVA Table")
}
```

   **Key Observation**

   Notice that the columns $\hat{y}_{sub}$ (projection onto $x_1, x_2, x_3$) and $\hat{y}_{full}$ (projection onto $x_1, \dots, x_4$) are identical. This confirms that adding the linearly dependent predictor $x_4$ does not change the column space of the design matrix or the fitted values, even though it makes individual coefficients non-unique.


2. Model Performance and Inference for $\rho^2$

   In a rank-deficient model where $r = \text{rank}(\mathbf{X})$, we define the following metrics based on the projection onto the model space $\mathcal{C}(\mathbf{X})$.

   **Mathematical Formulas**

   The **Coefficient of Determination** ($R^2$) and its adjusted version ($R^2_a$) are defined as:

   $$R^2 = \frac{SS_{Reg}}{SS_{Total}} = \frac{\| \hat{\mathbf{y}}_{full} - \bar{\mathbf{y}} \|^2}{\| \mathbf{y} - \bar{\mathbf{y}} \|^2}$$

   $$R^2_a = 1 - (1 - R^2) \frac{n - 1}{n - r}$$

   For the **Confidence Interval of $\rho^2$**, we utilize the non-central $F$-distribution. The non-centrality parameter $\lambda$ is related to the population squared multiple correlation $\rho^2$ by:

   $$\lambda = n \frac{\rho^2}{1 - \rho^2} \iff \rho^2 = \frac{\lambda}{\lambda + n}$$

   We find the confidence limits $[\lambda_{L}, \lambda_{U}]$ by inverting the non-central $F$ cumulative distribution function such that:
   $$P(F \le F_{obs} \mid \lambda_{U}) = \alpha/2 \quad \text{and} \quad P(F \le F_{obs} \mid \lambda_{L}) = 1 - \alpha/2$$



   **R Implementation**

   ```{r}
   #| label: non-full-rank-predictivity
   #| echo: true

   # 7. R-Squared and Adjusted R-Squared
   SS_Total <- sum((y - mean(y))^2)
   r_sq <- SSH / SS_Total

   # Using the rank computed earlier (rank_X = 4)
   adj_r_sq <- 1 - ((1 - r_sq) * (n - 1) / (n - rank_X))

   # 8. Confidence Interval for rho^2 (95%)
   ci_rho2_F <- function(F_stat, df1, df2, n, conf_level = 0.95) {
  
      # Helper function to find the non-centrality parameter (lambda)
      get_ncp <- function(f_val, df1, df2, prob) {
         # If observed F is less than the critical value, the lower bound is 0
         if (f_val <= qf(prob, df1, df2)) return(0)
         
         # Objective function: find lambda such that P(F < f_val | lambda) = prob
         obj <- function(lambda) pf(f_val, df1, df2, ncp = lambda) - prob
         
         # Attempt to solve using uniroot
         result <- try(uniroot(obj, interval = c(0, 10000))$root, silent = TRUE)
         
         if (inherits(result, "try-error")) return(NA) else return(result)
      }
      
      # Calculate alpha tail probabilities
      alpha <- 1 - conf_level
      prob_lower <- 1 - (alpha / 2) # Upper quantile for lower limit of lambda
      prob_upper <- alpha / 2       # Lower quantile for upper limit of lambda
      
      # Calculate Non-centrality Parameters (Lambdas)
      # Note: The "Upper" probability yields the "Lower" lambda bound and vice versa
      lambda_low  <- get_ncp(F_stat, df1, df2, prob_lower)
      lambda_high <- get_ncp(F_stat, df1, df2, prob_upper)
      
      # Convert Lambda to Rho-squared
      # Formula: rho^2 = lambda / (lambda + n)
      rho2_low  <- lambda_low / (lambda_low + n)
      rho2_high <- lambda_high / (lambda_high + n)
      
      return(c(lower = rho2_low, upper = rho2_high))
   }
 
   ci_rho2 <- ci_rho2_F(F_stat, df_hyp, df_err, n)
   rho2_low  <- ci_rho2[1]
   rho2_high <- ci_rho2[2]

   # 9. Summary Table
   metrics_df <- data.frame(
   Metric = c("R-Squared", "Adjusted R-Squared", "95% CI for $\\rho^2$"),
   Value = c(
      round(r_sq, 4), 
      round(adj_r_sq, 4), 
      paste0("[", round(rho2_low, 4), ", ", round(rho2_high, 4), "]")
   )
   )

   kable(metrics_df, 
         caption = "Model Fit Metrics and Population Correlation Interval")
   ```




::::

:::{#exm-lm-anova-combined}

### Rank-Deficient Model Fitting and ANOVA Analysis

This example demonstrates the end-to-end process of fitting a model with linear dependencies using R's `lm()` function and generating a formatted ANOVA table. We explicitly handle the potential for fitting errors and highlight the reduced degrees of freedom resulting from the singularity.

1. **Data Generation and Model Fitting** 

   We first generate a dataset where $x_4$ is a perfect linear combination of $x_1, x_2$, and $x_3$. We then attempt to fit the full model using `lm()` within a `tryCatch` block to ensure robustness.

   ```{r}
   #| echo: true
   #| label: lm-anova-combined
   #| message: false
   #| warning: false
   library(MASS)  # For ginv()
   library(knitr) # For kable()
   library(gt)    # For HTML tables

   # 1. Generate Data with linear dependency: x4 = 0.1*x1 + 0.5*x2 + x3
   set.seed(42)
   n <- 12
   x1 <- rnorm(n, mean=5, sd=2)
   x2 <- rnorm(n, mean=10, sd=2)
   x3 <- rnorm(n, mean=0, sd=1)
   x4 <- 0.1*x1 + 0.5*x2 + x3 
   y <- 10 + 2*x1 - 1*x2 + 3*x3 + rnorm(n, sd=1)
   df <- data.frame(y=y, x1=x1, x2=x2, x3=x3, x4=x4)

   # 2. Fit model using lm() with tryCatch
   lm_fit <- tryCatch({
     model <- lm(y ~ x1 + x2 + x3 + x4, data = df)
     model
   }, error = function(e) {
     message("Error in fitting: ", e$message)
     return(NULL)
   })

   # 3. Analyze Coefficients and ANOVA
   # Note: lm() sets the coefficient of x4 to NA due to the singularity
   print("Estimated Coefficients:")
   print(coef(lm_fit))

   anova_results <- anova(lm_fit)
   anova_df <- as.data.frame(anova_results)
   anova_df$Term <- rownames(anova_results)
   anova_df <- anova_df[, c("Term", "Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)")]

   # 4. Combined Formatting and Output
   if (knitr::is_html_output()) {
     # HTML Output with gt highlighting
     anova_df |>
       gt() |>
       fmt_number(columns = c("Sum Sq", "Mean Sq", "F value", "Pr(>F)"), decimals = 4) |>
       sub_missing(columns = everything(), missing_text = "") |>
       tab_style(
         style = cell_text(color = "red", weight = "bold"),
         locations = cells_body(columns = Df, rows = Term %in% c("x1", "x2", "x3")) 
       ) |>
       tab_header(title = "ANOVA Table")
       
   } else {
     # LaTeX Output with color highlighting
     anova_df$Df <- as.character(anova_df$Df)
     # Identify rows for model components and color them
     model_rows <- which(anova_df$Term %in% c("x1", "x2", "x3"))
     anova_df$Df[model_rows] <- paste0("\\textcolor{red}{", anova_df$Df[model_rows], "}")
     
     kable(anova_df, digits = 4, escape = FALSE, 
           caption = "Sequential ANOVA Table")
   }
   ```

2. **Interpretation of Results** 

   The `lm()` function automatically identifies that $x_4$ is aliased with the other predictors and assigns it `NA`. Consequently, in the ANOVA table, the variation is partitioned among the first three predictors, and $x_4$ contributes no additional degrees of freedom or sum of squares. This demonstrates that for non-full rank models, inference is limited to the estimable part of the parameter space.

:::

## Inference for Estimable Parameters

### Estimability

::: {#def-estimability}

#### Estimability
Let $\lambda = (\lambda_1, ..., \lambda_p)^T$ be a vector of constants. The parameter $\lambda^T \beta = \sum_j \lambda_j \beta_j$ is said to be estimable if there exists a vector $a$ in $R^n$ such that:

$$
E(a^T y) = \lambda^T \beta
$$

for all $\beta \in \mathcal{R}^p$.

:::

This condition is equivalent to $\lambda^T \beta$ being estimable if and only if there exists $a$ such that $X^T a = \lambda$ (i.e., iff $\lambda$ lies in the row space of $X$).

::: {#thm-estimability-conditions}

#### Equivalent Conditions for Estimability
In the model $y = X\beta + e$, the linear function $\lambda^T \beta$ is estimable if and only if any one of the following equivalent conditions holds:

1.  $\lambda^T$ is a linear combination of the rows of $X$; that is, there exists a vector $a$ such that $a^T X = \lambda^T$.
2.  $\lambda^T$ is a linear combination of the rows of $X^T X$, or $\lambda$ is a linear combination of the columns of $X^T X$; that is, there exists a vector $r$ such that $X^T X r = \lambda$.
3.  $\lambda$ is such that $X^T X (X^T X)^- \lambda = \lambda$.
4.  $\lambda \in \text{Row}(X) = \text{Col}(X^TX)$

:::

::: {.proof}
We prove the equivalence of the conditions for estimability as follows:

1. **Condition (1) $\iff$ Definition of Estimability** 

   By definition, $\lambda^T \beta$ is estimable if there exists a vector $a$ such that $E(a^T y) = \lambda^T \beta$ for all $\beta$. Since $E(a^T y) = a^T E(y) = a^T X \beta$, the requirement $a^T X \beta = \lambda^T \beta$ must hold for all $\beta \in \mathbb{R}^p$. This is true if and only if $a^T X = \lambda^T$, which means $\lambda$ is a linear combination of the rows of $X$.

2. **Condition (1) $\iff$ Condition (4)** 

   Condition (1) states there exists $a$ such that $X^T a = \lambda$. This is the literal definition of $\lambda$ being in the column space of $X^T$, which is the row space of $X$, denoted $\text{Col}(X^T) = \text{Row}(X)$. Furthermore, a fundamental result in matrix algebra states that $\text{Col}(X^T) = \text{Col}(X^T X)$. Thus, $\lambda \in \text{Row}(X) \iff \lambda \in \text{Col}(X^T X)$.

3. **Condition (4) $\iff$ Condition (2)** 

   If $\lambda \in \text{Col}(X^T X)$, then by the definition of column space, there must exist a vector $r$ such that $X^T X r = \lambda$. Taking the transpose gives $r^T X^T X = \lambda^T$, showing $\lambda^T$ is a linear combination of the rows of $X^T X$.

4. **Condition (2) $\iff$ Condition (3)** 

   This follows from the properties of generalized inverses. A system $X^T X r = \lambda$ is consistent if and only if $X^T X (X^T X)^- \lambda = \lambda$ for any generalized inverse $(X^T X)^-$. If such an $r$ exists, then:
   
   $$
   X^T X (X^T X)^- (X^T X r) = X^T X r = \lambda
   $$
   
   using the defining property $AAA^- A = A$.

:::

### Example: Checking Estimability in One-Way ANOVA

:::{#exm-estimability-comprehensive}

### Checking Estimability in One-Way ANOVA

We use the balanced one-way ANOVA effects model matrix $X$ ($a=3$ groups, $n=4$ observations per group) to verify different linear functions $\lambda^T \beta$. In this model, the parameter vector is $\beta = (\mu, \alpha_1, \alpha_2, \alpha_3)^T$.

1. **Defining the Vectors and Explicit Matrix Form** 

   
   First, we observe the explicit form of the $X^T X$ matrix. For a balanced one-way layout with $n=4$ per group, the matrix is as follows :

   $$ 
   X^T X = \begin{pmatrix} 12 & 4 & 4 & 4 \\ 4 & 4 & 0 & 0 \\ 4 & 0 & 4 & 0 \\ 4 & 0 & 0 & 4 \end{pmatrix} 
   $$
   
   We examine the following functions of the parameters:

   * $\lambda_1 = (1, 1, 0, 0)^T$: The cell mean for group 1 ($\mu + \alpha_1$).
   * $\lambda_2 = (0, 1, 0, 0)^T$: The individual treatment effect $\alpha_1$.
   * $\lambda_3 = (0, 1, -1, 0)^T$: The contrast between group 1 and group 2 ($\alpha_1 - \alpha_2$).
   * $\lambda_4 = (1, 0, 0, 1)^T$: The cell mean for group 3 ($\mu + \alpha_3$).
   * $\lambda_5 = (12, 4, 4, 4)^T$: This corresponds to $12\mu + 4\sum \alpha_i$, which is the first column of $X^TX$.

2. **Numerical Consistency Check (Condition 3)** 

   A linear function $\lambda^T\beta$ is estimable if and only if $X^T X (X^T X)^- \lambda = \lambda$. We perform this check for all vectors using the Moore-Penrose generalized inverse.

   ```{r}
   #| echo: true
   #| label: comprehensive-estimability-check
   library(MASS)
   library(knitr)

   # Setup X'X for balanced one-way layout (n=4 per group) 
   XtX <- matrix(c(
     12,  4,  4,  4,
      4,  4,  0,  0,
      4,  0,  4,  0,
      4,  0,  0,  4
   ), nrow = 4, byrow = TRUE)
   
   XtX_inv <- ginv(XtX)

   # Define lambda vectors
   L1 <- c(1, 1, 0, 0)      # mu + alpha1 
   L2 <- c(0, 1, 0, 0)      # alpha1      
   L3 <- c(0, 1, -1, 0)     # alpha1-alpha2 
   L4 <- c(1, 0, 0, 1)      # mu + alpha3 
   L5 <- c(12, 4, 4, 4)     # 12mu + 4(alpha1 + alpha2 + alpha3)

   # Compute (X'X)(X'X)^- * lambda for each
   check_L1 <- XtX %*% XtX_inv %*% L1
   check_L2 <- XtX %*% XtX_inv %*% L2
   check_L3 <- XtX %*% XtX_inv %*% L3
   check_L4 <- XtX %*% XtX_inv %*% L4
   check_L5 <- XtX %*% XtX_inv %*% L5

   # Combine results for display
   final_comp <- data.frame(
     Param = c("mu", "alpha1", "alpha2", "alpha3"),
     L1_Target = L1, L1_Res = as.vector(check_L1),
     L2_Target = L2, L2_Res = as.vector(check_L2),
     L3_Target = L3, L3_Res = as.vector(check_L3),
     L4_Target = L4, L4_Res = as.vector(check_L4),
     L5_Target = L5, L5_Res = as.vector(check_L5)
   )

   kable(final_comp, digits = 3,
         col.names = c("Param", "$\\lambda_1$", "$L_1$", 
                       "$\\lambda_2$", "$L_2$", 
                       "$\\lambda_3$", "$L_3$", 
                       "$\\lambda_4$", "$L_4$", 
                       "$\\lambda_5$", "$L_5$"),
         caption = "Estimability Check for Multiple Parameter Functions")
   ```

3. **Summary of Results** 

   * **Estimable Functions**: $\lambda_1, \lambda_3, \lambda_4$, and $\lambda_5$ all satisfy the condition because their result columns ($L_i$) match their target columns ($\lambda_i$). These represent cell means, contrasts, and linear combinations of the rows of $X^TX$.
   * **Non-Estimable Function**: $\lambda_2$ (individual $\alpha_1$) fails the condition as the target and result do not match.
   * **The First Column ($\lambda_5$)**: This vector is estimable because it is explicitly the first row/column of $X^TX$. Since any linear combination of rows of $X^TX$ is estimable, $\lambda_5^T \beta = 12\mu + 4\alpha_1 + 4\alpha_2 + 4\alpha_3$ is estimable. Note that this is equal to $\sum_{i,j} E(y_{ij})$, the sum of all expected cell means.

4. **Theoretical Conclusion** 

   In non-full-rank models, we cannot estimate individual parameters that are not uniquely identified by the data. However, linear functions that lie in the row space of $X$ (or the column space of $X^TX$)—such as cell means or treatment contrasts—are invariant to the choice of solution $\hat{\beta}$ and remain uniquely estimable.

:::

::: {#thm-number-estimable-functions-nonfullrank}

#### Number of Estimable Functions
In the non-full-rank model $y = X\beta + e$, the number of linearly independent estimable functions of $\beta$ is the rank of $X$.

:::

### Properties of Estimators for Estimabile Parameters

Let $\lambda^T \beta$ be an estimable function. Let $\hat{\beta}$ be any solution to the normal equations. Then the estimator $\lambda^T \hat{\beta}$ has the following properties:

1.  **Unbiasedness**: $E(\lambda^T \hat{\beta}) = \lambda^T \beta$.
2.  **Uniqueness**: $\lambda^T \hat{\beta}$ is invariant to the choice of $\hat{\beta}$ (to the choice of generalized inverse).

::: {#thm-variance-covariance}

#### Variance and Covariance

1.  **Variance**: The variance of $\lambda^T \hat{\beta}$ is unique and is given by:

    
    $$
    \text{Var}(\lambda^T \hat{\beta}) = \sigma^2 \lambda^T (X^T X)^- \lambda
    $$
    

2.  **Covariance**: For two estimable functions $\lambda_1^T \beta$ and $\lambda_2^T \beta$:

    
    $$
    \text{Cov}(\lambda_1^T \hat{\beta}, \lambda_2^T \hat{\beta}) = \sigma^2 \lambda_1^T (X^T X)^- \lambda_2
    $$
    

:::


### Properties of Estimators for Estimabile Parameters

Let $\lambda^T \beta$ be an estimable function. Let $\hat{\beta}$ be any solution to the normal equations. Then the estimator $\lambda^T \hat{\beta}$ has the following properties:

1.  **Unbiasedness**: $E(\lambda^T \hat{\beta}) = \lambda^T \beta$.
2.  **Uniqueness**: $\lambda^T \hat{\beta}$ is invariant to the choice of $\hat{\beta}$ (to the choice of generalized inverse).

::: {#thm-variance-covariance-nonfullrank}

#### Variance and Covariance

1.  **Variance**: The variance of $\lambda^T \hat{\beta}$ is unique and is given by:

    
    $$
    \text{Var}(\lambda^T \hat{\beta}) = \sigma^2 \lambda^T (X^T X)^- \lambda
    $$

2.  **Covariance**: For two estimable functions $\lambda_1^T \beta$ and $\lambda_2^T \beta$:

    
    $$
    \text{Cov}(\lambda_1^T \hat{\beta}, \lambda_2^T \hat{\beta}) = \sigma^2 \lambda_1^T (X^T X)^- \lambda_2
    $$

:::

::: {#thm-gauss-markov-non-full}

#### Gauss-Markov in Non-Full Rank Case
If $\lambda^T \beta$ is estimable in the spherical errors non-full rank linear model, then $\lambda^T \hat{\beta}$ is its Best Linear Unbiased Estimator (BLUE).

:::

### Testable Hypotheses in Non-Full Rank Models

In non-full rank models, we cannot test hypotheses about individual parameters that are not uniquely identified by the data. Instead, we must restrict our inference to **testable hypotheses**, which are formulated using estimable functions.

::: {#def-testable-hypothesis}

#### Formal Description of a Testable Hypothesis
A linear hypothesis $H_0: C\beta = 0$ is said to be **testable** if and only if every row of the $m \times p$ matrix $C$ is an estimable function of $\beta$. 

This implies that there exists a matrix $A$ such that $C = AX$. Physically, this means $H_0$ can be expressed entirely in terms of the expected values of the observations.

:::



### The General Linear Hypothesis F-Test

If $y \sim N_n(X\beta, \sigma^2 I)$, where $X$ is $n \times p$ with $\text{rank}(X) = k < p$, and $C\beta$ is a set of $m$ linearly independent estimable functions (where $m \le k$), the following results hold:

1.  **Distribution of the Estimator** 

    The least-squares estimator $C\hat{\beta}$ follows a multivariate normal distribution:
    
    $$C\hat{\beta} \sim N_m\left[C\beta, \sigma^2 C(X^T X)^- C^T\right]$$ 

2.  **The F-Statistic** 

    The test statistic for $H_0: C\beta = 0$ is defined as:
    
    $$F = \frac{SSH / m}{SSE / (n - k)}$$ 
    
    where:

    * $SSH = (C\hat{\beta})^T [C(X^T X)^- C^T]^{-1} C\hat{\beta}$ is the Hypothesis Sum of Squares.
    * $SSE = y^T [I - X(X^T X)^- X^T] y$ is the Error Sum of Squares.
    * $k$ is the **rank** of the matrix $X$, not the number of columns $p$.

3.  **Sampling Distribution** 

    The statistic $F$ follows an F-distribution:
    
    $$F \sim F(m, n - k)$$ 
    
    Under the alternative hypothesis $H_1: C\beta eq 0$, the statistic follows a non-central F-distribution $F(m, n - k, \lambda)$, with non-centrality parameter:
    
    $$\lambda = \frac{(C\beta)^T [C(X^T X)^- C^T]^{-1} C\beta}{2\sigma^2}$$ 



**Summary of Degrees of Freedom**

| Component | Degrees of Freedom |
| :--- | :--- |
| **Numerator (Model/Hypothesis)** | $m$ (Number of linearly independent estimable constraints)  |
| **Denominator (Error)** | $n - k$ (Total observations minus **rank** of $X$)  |


