---
title: "Data Analysis and Simulation for Stratified Random Sampling"
author: "Longhai Li"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: no
    toc_depth: 2
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
  word_document:
    toc: yes
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
  pdf_document:
    toc: yes
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
editor_options:
  chunk_output_type: inline
---


# Functions and packages for Analyzing Data
```{r}
library(latex2exp)
library(sampling)
library(plyr)
## this function finds statistical estimates given a dataset with sampling weight
# stratdata --- data.frame containing stratified sample
# y --- name of variable for which we want to estiamte population mean
# stratum --- name of variable that will be used as stratum variable
# weight --- name of variable indicating sampling weight
# note: from weights we can find Nh (see the code for formula)
str_mean_estimate_data <- function (stratdata, y, stratum, weight, post=FALSE)
{
    ## compute stratum-wise data
    n <- nrow (stratdata)
    sh <- tapply (stratdata[, y], stratdata[,stratum], sd)
    ybarh <- tapply (stratdata[, y], stratdata[,stratum], mean)
    ## find population stratum size using sampling weight included in the data set
    Nh <- tapply (stratdata[, weight], stratdata[,stratum], sum)
    if (!post)    
        nh <- tapply (1:nrow(stratdata), stratdata[,stratum], length)
    
    else 
        nh <- n * Nh/sum (Nh)
    
    ## find mean estimates
    N <- sum (Nh)
    Pi_h <- Nh/N
    ybar <- sum(ybarh * Pi_h)
    seybar <- sqrt(sum((1-nh/Nh)*Pi_h^2*sh^2/nh))
    mem <- 1.96 * seybar
    c(Est. = ybar, S.E. = seybar, ci.low = ybar - mem, ci.upp = ybar + mem)
    
}


## to find total, multiply N to the estimate returned by this function
## for poststratification, use nh = n * Nh/N
str_mean_estimate <- function (ybarh, sh, nh, Nh)
{
    N <- sum (Nh)
    Pi_h <- Nh/N
    ybar <- sum(ybarh * Pi_h)
    seybar <- sqrt(sum((1-nh/Nh)*Pi_h^2*sh^2/nh))
    mem <- 1.96 * seybar
    c(Est. = ybar, S.E. = seybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}



## sdata --- a vector of original survey data
## N --- population size
## to find total, multiply N to the estimate returned by this function

srs_mean_est <- function (sdata, N = Inf)
{
	n <- length (sdata)
	ybar <- mean (sdata)
	se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
	mem <- qt (0.975, df = n - 1) * se.ybar
	c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}

```




# Analysis of "agstrat.csv" data
## Importing agstrat.csv data
```{r}
agstrat <- read.csv("data/agstrat.csv")
head (agstrat) 
```

## Spreadsheet calculation
### Summarizing acre92 in each stratum

```{r}
nh <- tapply (1:nrow(agstrat), agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
# create a vector with external information
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)

data.frame (Nh, nh, ybarh, sh)
```

### Estimates
```{r}
N <- sum (Nh)
pi_h <- Nh/N
ybar <- sum(ybarh * pi_h)
data.frame (Nh, nh=nh,pi_h, ybarh, sh=sh, "var component"=(1-nh/Nh)*pi_h^2*sh^2/nh)
```


```{r}
seybar <- sqrt(sum((1-nh/Nh)*pi_h^2*sh^2/nh))
mem <- 1.96 * seybar
c(Est. = ybar, S.E. = seybar, ci.low = ybar - mem, ci.upp = ybar + mem)
```

## Using the function "str_mean_estimate"
```{r}
nh <- tapply (1:nrow(agstrat), agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
# create a vector with external information
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)

## strata mean estimate
str_mean_estimate (ybarh, sh, nh, Nh)
```


```{r}
## population total estimate 
str_mean_estimate (ybarh, sh, nh, Nh) * sum (Nh)
```
## Using the function "str_mean_estimate_data"
```{r}
## In the fucntion "str_mean_estimate_data", we can find the stratum size with the variable "weights":
tapply (agstrat$weight, agstrat$region, sum)
```


```{r}
## if the dataset contains a variable "weight"
str_mean_estimate_data (agstrat, y="acres92", stratum="region", weight="weight")
```


```{r}
## estimating the mean of the number of small farms in 1992
str_mean_estimate_data (agstrat, y="smallf92", stratum="region", weight="weight")

```

## Comparing with SRS estimate
```{r}
agsrs <- read.csv ("data/agsrs.csv")
srs_mean_est(agsrs[, "acres92"], N=3078)

## the ratio of estimated variance
(16379.87/18898.43)^2

```
```{r}
## percentage of reduction of variance of str estimates from that of SRS estimate
1- (16379.87/18898.43)^2
```

# Allocation of stratum sample size

## Analyzing seals.csv collected with stratified sampling
Lydersen and Ryg (1991) used stratification techniques to estimate ringed seal pop- ulations in Svalbard fjords. The 200 km2 study area was divided into three zones: Zone 1, outer Sassenfjorden, was covered with relatively new ice during the study period in March, 1990, and had little snow cover; Zone 3, Tempelfjorden, had a stable ice cover throughout the year; Zone 2, inner Sassenfjorden, was intermediate between the stable Zone 3 and the unstable Zone 1. Ringed seals need good ice to establish territories with breathing holes, and snow cover enables females to dig out birth lairs. Thus, it was thought that the three zones would have different seal densities. The investigators took a stratified random sample of 20% of the 200 1-km2 areas. 
```{r}
###### ########

## load data
seals <- read.csv("data/seals.csv")
head (seals)
```


```{r}
# survey data summary in each stratum
nh <- as.vector(table(seals[,"zone"]))
sh <- tapply (seals[, "holes"], seals[,"zone"], sd)
ybarh <- tapply (seals[, "holes"], seals[,"zone"], mean)
Nh <- c(68,84,48)
data.frame (Nh,nh, ybarh, sh)
```


```{r}
str_mean_estimate (ybarh, sh, nh, Nh)
```

## Neyman allocation of stratum sample size

```{r}
## assuming the cost of surving each zone is the same
Ch <- rep (1, 3)
Sh <- sh ## estimate Sh with sh
NhShDCh <- Nh * Sh / sqrt (Ch)

## the optimal allocation scheme for estimating the population total or means holes:
Lh <- NhShDCh / sum (NhShDCh)
data.frame(Ch,Nh, Sh,NhShDCh, Lh)

```

# Simulation to Study the Efficiency of Stratified Sampling with Different Allocation

## Using "Region" to Stratify
### Read Population Data
```{r}
agpop <- read.csv ("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na
N <- nrow(agpop)
```


### Define Stratum Variable
```{r}
agpop$stratum <- agpop$region
## reorder agpop for the ease of using strata of "sampling" (very important)
agpop <- agpop[order (agpop$stratum), ]

```

Look at variance decomposition
```{r}
boxplot (acres92 ~ stratum, data = agpop)
anova(lm (agpop$acres92~agpop$stratum))
```
Look at the $R^2$ of predicting "acres92" with "region".
```{r}
summary(lm (agpop$acres92~agpop$stratum))$r.squared
```

### Stratified Sampling with P2S allocation 
```{r}
# doing one stratified sampling
Nh <- tapply (1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum (Nh)*300)
data.frame(Nh,nh)
nh

## doing stratified sampling
strsample <- strata (agpop, "stratum", size = nh,  method = "srswor")
head(strsample)
## checking sampling results
table (strsample [,1]) 

# collecting data on sampled counties
agstrat <- agpop [strsample$ID_unit, ]
agstrat$weight <- 1/strsample$Prob

str_mean_estimate_data (agstrat, "acres92", "stratum", "weight")

```
Estimate with spreadsheet method for a double checking
```{r}
## double checking the estimates
nh <- tapply (1:nrow(agstrat), agstrat[,"region"], length)
sh <- tapply (agstrat[, "acres92"], agstrat[,"region"], sd)
ybarh <- tapply (agstrat[, "acres92"], agstrat[,"region"], mean)
# create a vector with external information
Nh <- tapply (1:nrow(agpop), agpop[,"region"], length)
## strata mean estimate
str_mean_estimate (ybarh, sh, nh, Nh)
```

### Repeat stratified sampling with P2S allocation 2000 times
```{r cache=TRUE}

Nh <- tapply (1:nrow (agpop), agpop$stratum, length) 
nh <- round(Nh/sum (Nh)*300) ## make sure the order matches Nh

nres <- 2000
str_p2s_simulated <- matrix (0, nres, 4)
for (i in 1:nres)
{
    ## doing stratified sampling
    strsample <- strata (agpop, "stratum", size = nh,  method = "srswor")
    agstrat <- agpop [strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    ## analyziing data
    str_p2s_simulated[i,] <- str_mean_estimate_data (agstrat, "acres92", "stratum", "weight")
    
}
```


### Repeat stratified sampling with optimal allocation 2000 times
```{r}
Nh <- tapply (1:nrow (agpop), agpop$stratum, length) 

Sh <- tapply (agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum (Nh*Sh) * 300);nh_opt
data.frame(Nh,Sh,nh_opt)
```


```{r cache=TRUE}
nres <- 2000
str_neyman_simulated <- matrix (0, nres, 4)
for (i in 1:nres)
{
    ## doing stratified sampling
    strsample <- strata (agpop, "stratum", size = nh_opt,  method = "srswor")
    # collecting data on sampled counties
    agstrat <- agpop [strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    
    str_neyman_simulated[i,] <- str_mean_estimate_data (agstrat, "acres92", "stratum", "weight")
    
}
```

### Repeat simple random sampling  2000 times
```{r}

nres <- 2000
srs_simulated <- matrix (0, nres, 4)
for (i in 1:nres)
{
    ## doing stratified sampling
    srs <- sample (sum(Nh), sum (nh))
    # collecting data on sampled counties
    agsrs <- agpop [srs, ]
    srs_simulated[i,] <- srs_mean_est (agsrs[, "acres92"], N= sum(Nh))
    
}

```
### Compare the efficiency of different methods

```{r}
sim_results_str_region <- data.frame("SRS"=srs_simulated[,1], 
                      "Prop2size"=str_p2s_simulated[,1], 
                      "Neyman"=str_neyman_simulated[,1])
boxplot (sim_results_str_region)
abline (h = mean (agpop$acres92), col = "red")

sapply (sim_results_str_region, mean) -> sim_means
sapply (sim_results_str_region, var) -> sim_var

sim_var/sim_var[1] -> sim_var_relative

#Percentage of reduction of variances of estimates compared to SRS

1-sim_var/sim_var[1] -> sim_var_reduction

cbind("Mean"=sim_means, "Variance"=sim_var, "Relative Variance"= sim_var_relative, "Percentage of Variance Reduction"=sim_var_reduction)

```

## Using "acres82" to Define Strata
### Read Population Data
```{r}
agpop <- read.csv ("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na
N <- nrow(agpop)
```

### Define Stratum Variable with Quantiles of "acres82"
```{r}
plot (acres92~acres82, data = agpop)
summary (lm(acres92~acres82, data = agpop))
agpop$stratum <- cut(agpop$acres82,
                     breaks = quantile (agpop$acres82, 
                                        probs = c(0,0.25,0.75,0.95,1)),
                     include.lowest =  T)
agpop$stratum <- mapvalues(agpop$stratum, 
                          from = (sort(unique (agpop$stratum))), 
                          to= paste0("acres82",c("Q1", "Q2", "Q3", "Q4")))
agpop <- agpop[order(agpop$stratum), ]
```

Look at variance decomposition
```{r}
boxplot (acres92 ~ stratum, data = agpop)
anova(lm (agpop$acres92~agpop$stratum))
```
Look at the $R^2$ of predicting "acres92" with "acres82" quantile.
```{r}
summary(lm (agpop$acres92~agpop$stratum))$r.squared
```

### Stratified sampling with P2S allocation
```{r}
# doing one stratified sampling
Nh <- tapply (1:nrow(agpop), agpop$stratum, length) 
nh <- round(Nh/sum (Nh)*300)
nh

## doing stratified sampling
strsample <- strata (agpop, "stratum", size = nh,  method = "srswor")
head(strsample)
## checking sampling results
table (strsample [,1]) 

# collecting data on sampled counties
agstrat <- agpop [strsample$ID_unit, ]
agstrat$weight <- 1/strsample$Prob

str_mean_estimate_data (agstrat, "acres92", "stratum", "weight")

```


### Repeat stratified sampling with P2S allocation 2000 times
```{r cache=TRUE}

Nh <- tapply (1:nrow (agpop), agpop$stratum, length) 
nh <- round(Nh/sum (Nh)*300) ## make sure the order matches Nh
data.frame(Nh, nh)
nres <- 2000
str_p2s_simulated <- matrix (0, nres, 4)
for (i in 1:nres)
{
    ## doing stratified sampling
    strsample <- strata (agpop, "stratum", size = nh,  method = "srswor")
    agstrat <- agpop [strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    ## analyziing data
    str_p2s_simulated[i,] <- str_mean_estimate_data (agstrat, "acres92", "stratum", "weight")
    
}
```


### Repeat stratified sampling with optimal allocation 2000 times
```{r}
Nh <- tapply (1:nrow (agpop), agpop$stratum, length) 
# find order of stratum
# unique (agpop$stratum)
Sh <- tapply (agpop$acres92, agpop$stratum, sd)
nh_opt <- round((Nh*Sh)/sum (Nh*Sh) * 300);nh_opt
data.frame(Nh,Sh,nh_opt)
```
```{r cache=TRUE}
nres <- 2000
str_neyman_simulated <- matrix (0, nres, 4)
for (i in 1:nres)
{
    ## doing stratified sampling
    strsample <- strata (agpop, "stratum", size = nh_opt,  method = "srswor")
    # collecting data on sampled counties
    agstrat <- agpop [strsample$ID_unit, ]
    agstrat$weight <- 1/strsample$Prob
    
    str_neyman_simulated[i,] <- str_mean_estimate_data (agstrat, "acres92", "stratum", "weight")
    
}
```

### Repeat simple random sampling  2000 times
```{r}

nres <- 2000
srs_simulated <- matrix (0, nres, 4)
for (i in 1:nres)
{
    ## doing stratified sampling
    srs <- sample (sum(Nh), sum (nh))
    # collecting data on sampled counties
    agsrs <- agpop [srs, ]
    srs_simulated[i,] <- srs_mean_est (agsrs[, "acres92"], N= sum(Nh))
    
}

```
### Compare the efficiency of different methods

```{r}
sim_results_str_acres82 <- data.frame("SRS"=srs_simulated[,1], 
                      "Prop2size"=str_p2s_simulated[,1], 
                      "Neyman"=str_neyman_simulated[,1])
boxplot (sim_results_str_acres82)
abline (h = mean (agpop$acres92), col = "red")

sapply (sim_results_str_acres82, mean) -> sim_means
sapply (sim_results_str_acres82, var) -> sim_var

sim_var/sim_var[1] -> sim_var_relative

#Percentage of reduction of variances of estimates compared to SRS

1-sim_var/sim_var[1] -> sim_var_reduction

cbind("Mean"=sim_means, "Variance"=sim_var, "Relative Variance"= sim_var_relative, "Percentage of Variance Reduction"=sim_var_reduction)

```


<!-- # Functions in the file estimates.r -->
<!-- ```{r comment=""} -->
<!-- cat(paste0(readLines("estimates.r"), collapse="\n")) -->
<!-- ``` -->

