---
title: "Decision Theory"
format: html
---

## Formulation of Decision Theory

In decision theory, we formalize the process of making decisions under uncertainty using the following components:

1.  **Parameter Space ($\Theta$):**
    The set of all possible states of nature or values that the parameter can take.
    $\theta \in \Theta$ (e.g., mean, variance).

2.  **Sample Space ($\mathcal{X}$):**
    The space where the data $X$ lies.
    Example: $X = (X_1, X_2, \dots, X_n)$ where $X_i \in \mathbb{R}$. So $\mathcal{X} \in \mathbb{R}^n$.

3.  **Family of Probability Distributions:**
    $\{P_\theta(x) : \theta \in \Theta\}$.
    This describes how likely we are to see the data $X$ given a specific parameter $\theta$.

    * If $X$ is continuous: $P_\theta(x) = f(x, \theta)$ (Probability Density Function).
    * If $X$ is discrete: $P_\theta(x) = f(x, \theta)$ (Probability Mass Function).

4.  **Action Space ($\mathcal{A}$):**
    The set of all actions or decisions available to the experimenter.

5.  **Loss Function:**
    $L: \Theta \times \mathcal{A} \rightarrow \mathbb{R}$.
    $L(\theta, a)$ specifies the loss incurred if the true parameter is $\theta$ and we take action $a$. Generally, $L(\theta, a) \ge 0$.

## Decision Rules and Risk Functions

### Decision Rule
A decision rule is a function $d: \mathcal{X} \rightarrow \mathcal{A}$. It dictates the action $d(x)$ we take when we observe data $x$.

### Risk Function
The risk function is the expected loss for a given decision rule $d$ as a function of the parameter $\theta$.

$$R(\theta, d) = E_\theta[L(\theta, d(X))]$$

* Continuous Case: $R(\theta, d) = \int_{\mathcal{X}} L(\theta, d(x)) f(x, \theta) dx$
* Discrete Case: $R(\theta, d) = \sum_{x \in \mathcal{X}} L(\theta, d(x)) P_\theta(x)$

## Examples of Decision Problems

This section applies the definitions above to common statistical problems.

### Example 1: Hypothesis Testing
We want to test $H_0$ vs $H_1$.

* **Action Space:** $\mathcal{A} = \{0, 1\}$ where $0$ means "Accept $H_0$" and $1$ means "Reject $H_0$".
* **Loss Function (0-1 Loss):**
    $$
    L(\theta, a) =
    \begin{cases}
    0 & \text{if decision is correct} \\
    1 & \text{if decision is wrong}
    \end{cases}
    $$
    Specifically:
    * If $\theta \in H_0$ and $a=1$ (Type I Error), Loss = 1.
    * If $\theta \in H_1$ and $a=0$ (Type II Error), Loss = 1.
* **Decision Rule:** A common rule is $d(x) = 1$ if $\bar{x} > c$, else $0$.
* **Risk Function:**
    * If $\theta \in H_0$: $R(\theta, d) = 1 \cdot P_\theta(d(X)=1) + 0 = P(\text{Type I Error})$.
    * If $\theta \in H_1$: $R(\theta, d) = 1 \cdot P_\theta(d(X)=0) + 0 = P(\text{Type II Error})$.

### Example 2: Point Estimation
We want to estimate a parameter $\theta$.

* **Action Space:** $\mathcal{A} = \Theta$. An action is an estimate of $\theta$.
* **Loss Function:**
    * *Squared Error:* $L(\theta, a) = (\theta - a)^2$
    * *Absolute Error:* $L(\theta, a) = |\theta - a|$
* **Decision Rule:** A common rule is $d(x) = \bar{x}$ (Sample mean).
* **Risk Function (for Squared Error):**
    $$R(\theta, d) = E_\theta[(\bar{x} - \theta)^2] = \text{MSE} = \text{Var}(\bar{x}) + \text{Bias}^2$$

### Example 3: Interval Estimation
We want to estimate a range for the parameter.

* **Action Space:** $\mathcal{A} = \{(l, u) : l \in \mathbb{R}, u \in \mathbb{R}, l \le u\}$.

## Example: The Duchess and the Emerald Necklace

**Scenario:**
You are the Duchess of Omnium. You have two necklaces: a priceless **Real** one and a valueless **Imitation**. They are indistinguishable to you. One is in the **Left Drawer (Box 1)**, the other is in the **Right Drawer (Box 2)**.

**The Data (Great Aunt):**
You consult your Great Aunt. She inspects the Left Drawer first, then the Right.
* If the **Real** necklace is in the **Left** ($\theta_1$): She identifies it correctly. (Infallible).
* If the **Real** necklace is in the **Right** ($\theta_2$): She sees the fake first, gets confused, and guesses randomly ($50/50$).
## Formulation of the Decision Problem

### Parameter Space ($\Theta$)

The set of possible states of nature regarding the location of the real necklace:

* $\theta_1$: The **Real** necklace is in the **Left** Drawer.
* $\theta_2$: The **Real** necklace is in the **Right** Drawer.

### Action Space ($\mathcal{A}$)

The set of possible actions available to the Duchess:

* $a_1$: Wear the **Left** necklace.
* $a_2$: Wear the **Right** necklace.

### The Data ($X$)

The data consists of the Great Aunt's judgment after inspecting the necklaces:

* $X \in \{1, 2\}$
* $X=1$: Aunt says "Left is Real".
* $X=2$: Aunt says "Right is Real".

### Probability of Data Given Parameter ($P_\theta(X)$)

The probability of the Aunt's advice changes depending on the true state of nature ($\theta$).

| State | $P(X=1 \mid \theta)$ | $P(X=2 \mid \theta)$ |
| :--- | :---: | :---: |
| **$\theta_1$ (Real Left)** | 1 | 0 |
| **$\theta_2$ (Real Right)** | 0.5 | 0.5 |

### Loss Function ($L$)

The loss is defined as 0 for a correct choice and 1 (representing £1M) for an incorrect choice.

| State \ Action | $a_1$ (Wear Left) | $a_2$ (Wear Right) |
| :--- | :---: | :---: |
| $\theta_1$ (Real Left) | **0** | **1** |
| $\theta_2$ (Real Right) | **1** | **0** |

### Decision Rules

There are four possible deterministic decision rules ($d: \mathcal{X} \rightarrow \mathcal{A}$).

| Rule | Description | Action if $X=1$ (Left) | Action if $X=2$ (Right) |
| :--- | :--- | :---: | :---: |
| **$d_1$** | Always Left | $a_1$ | $a_1$ |
| **$d_2$** | Always Right | $a_2$ | $a_2$ |
| **$d_3$** | Follow Aunt | $a_1$ | $a_2$ |
| **$d_4$** | Do Opposite | $a_2$ | $a_1$ |

### Risk Calculation Tables

For each decision rule, the table below calculates the risk by summing the expected loss across possible outcomes ($X$) for both states of nature.

#### **Rule $d_1$ (Always Left)**

| State | Component | Outcome $X=1$ | Outcome $X=2$ | Risk (Row Sum) |
| :--- | :--- | :---: | :---: | :---: |
| **$\theta_1$** | Action $d_1(X)$ | $a_1$ | $a_1$ | |
| | Loss $L(\theta_1, d)$ | 0 | 0 | |
| | Prob $P(X \mid \theta_1)$ | 1 | 0 | |
| | **Product ($L \times P$)** | **0** | **0** | **Risk = 0** |
| **$\theta_2$** | Action $d_1(X)$ | $a_1$ | $a_1$ | |
| | Loss $L(\theta_2, d)$ | 1 | 1 | |
| | Prob $P(X \mid \theta_2)$ | 0.5 | 0.5 | |
| | **Product ($L \times P$)** | **0.5** | **0.5** | **Risk = 1** |

#### **Rule $d_2$ (Always Right)**

| State | Component | Outcome $X=1$ | Outcome $X=2$ | Risk (Row Sum) |
| :--- | :--- | :---: | :---: | :---: |
| **$\theta_1$** | Action $d_2(X)$ | $a_2$ | $a_2$ | |
| | Loss $L(\theta_1, d)$ | 1 | 1 | |
| | Prob $P(X \mid \theta_1)$ | 1 | 0 | |
| | **Product ($L \times P$)** | **1** | **0** | **Risk = 1** |
| **$\theta_2$** | Action $d_2(X)$ | $a_2$ | $a_2$ | |
| | Loss $L(\theta_2, d)$ | 0 | 0 | |
| | Prob $P(X \mid \theta_2)$ | 0.5 | 0.5 | |
| | **Product ($L \times P$)** | **0** | **0** | **Risk = 0** |

#### **Rule $d_3$ (Follow Aunt)**

| State | Component | Outcome $X=1$ | Outcome $X=2$ | Risk (Row Sum) |
| :--- | :--- | :---: | :---: | :---: |
| **$\theta_1$** | Action $d_3(X)$ | $a_1$ | $a_2$ | |
| | Loss $L(\theta_1, d)$ | 0 | 1 | |
| | Prob $P(X \mid \theta_1)$ | 1 | 0 | |
| | **Product ($L \times P$)** | **0** | **0** | **Risk = 0** |
| **$\theta_2$** | Action $d_3(X)$ | $a_1$ | $a_2$ | |
| | Loss $L(\theta_2, d)$ | 1 | 0 | |
| | Prob $P(X \mid \theta_2)$ | 0.5 | 0.5 | |
| | **Product ($L \times P$)** | **0.5** | **0** | **Risk = 0.5** |

#### **Rule $d_4$ (Do Opposite)**

| State | Component | Outcome $X=1$ | Outcome $X=2$ | Risk (Row Sum) |
| :--- | :--- | :---: | :---: | :---: |
| **$\theta_1$** | Action $d_4(X)$ | $a_2$ | $a_1$ | |
| | Loss $L(\theta_1, d)$ | 1 | 0 | |
| | Prob $P(X \mid \theta_1)$ | 1 | 0 | |
| | **Product ($L \times P$)** | **1** | **0** | **Risk = 1** |
| **$\theta_2$** | Action $d_4(X)$ | $a_2$ | $a_1$ | |
| | Loss $L(\theta_2, d)$ | 0 | 1 | |
| | Prob $P(X \mid \theta_2)$ | 0.5 | 0.5 | |
| | **Product ($L \times P$)** | **0** | **0.5** | **Risk = 0.5** |
## Comparing Decision Rules

We typically do not have a single rule $d$ that is better than all other rules for all $\theta$.

* $d$ strictly dominates $d'$ if $R(\theta, d) \le R(\theta, d')$ for all $\theta$, with strict inequality for at least one $\theta$.
* **Admissibility:** A decision rule $d$ is admissible if it is not dominated by any other rule. If it is dominated, it is *inadmissible*.

### Minimax Principle
A rule $d$ is Minimax if it minimizes the maximum possible risk.

$$\sup_{\theta} R(\theta, d) \le \sup_{\theta} R(\theta, d') \quad \text{for all } d' \in \mathcal{D}$$

Visualize the risk functions of two rules, $d_1$ and $d_2$. $d_2$ might have a higher risk in some areas but a lower "peak" risk, making it Minimax.

```{r}
#| label: fig-minimax
#| echo: false
#| fig-cap: "Illustration of Minimax: Rule d2 has a lower maximum risk than d1, making it the Minimax rule, even though d1 is better for some values of theta."
library(ggplot2)

theta <- seq(-3, 3, length.out = 100)
## d1 has high variance/peak
risk_d1 <- 1.5 * exp(-theta^2 / 2) 
## d2 is flatter (lower max risk)
risk_d2 <- 0.8 * exp(-theta^2 / 5) + 0.1

df <- data.frame(
  theta = rep(theta, 2),
  Risk = c(risk_d1, risk_d2),
  Rule = rep(c("d1", "d2 (Minimax)"), each = 100)
)

ggplot(df, aes(x = theta, y = Risk, color = Rule)) +
  geom_line(size = 1.2) +
  theme_minimal() +
  labs(title = "Comparison of Risk Functions", y = "Risk R(theta, d)", x = "Parameter Theta")
```

## Bayes Decision Rules

We specify a **prior distribution** $\pi(\theta)$ on the parameter space $\Theta$.

### Bayes Risk
The Bayes risk of a rule $d$ with respect to prior $\pi$ is the weighted average of the frequentist risk:

$$r(\pi, d) = E_\pi [R(\theta, d)] = \int_\Theta R(\theta, d) \pi(\theta) d\theta$$

### Bayes Rule Definition
A decision rule $d_\pi$ is a Bayes rule with respect to $\pi$ if it minimizes the Bayes risk:

$$r(\pi, d_\pi) = \inf_{d' \in \mathcal{D}} r(\pi, d')$$

## Randomized Decision Rules & Risk Sets

### Randomized Rules
A randomized decision rule chooses between deterministic rules $d_1, d_2, \dots$ with probabilities $p_1, p_2, \dots$.
The risk of a randomized rule $d^* = \sum p_i d_i$ is the linear combination of their risks:
$$R(\theta, d^*) = \sum p_i R(\theta, d_i)$$

### Geometric Interpretation (Finite Parameter Space)
If $\Theta = \{\theta_1, \theta_2, \dots, \theta_k\}$ is finite, we can plot the risk vector $(R(\theta_1, d), \dots, R(\theta_k, d))$ in $\mathbb{R}^k$.

**Risk Set ($S$):** The set of all possible risk vectors.
Lemma: The Risk Set $S$ is **convex**. This is because we can form randomized rules that lie on the line segment connecting any two deterministic rules.

```{r}
#| label: fig-riskset
#| echo: false
#| fig-cap: "The Risk Set S. Points d1 and d2 are deterministic rules. The line connecting them represents randomized rules. The Minimax rule lies on the line R1=R2 (if accessible). The Bayes rule is found by bringing a tangent line with slope -pi1/pi2 toward the origin."
library(ggplot2)

## Define a polygon representing the risk set
risk_polygon <- data.frame(
  R1 = c(1, 4, 4, 2, 1),
  R2 = c(4, 4, 1, 1, 4)
)

ggplot() +
  ## Draw the convex risk set
  geom_polygon(data = risk_polygon, aes(x = R1, y = R2), fill = "lightblue", alpha = 0.5, color = "black") +
  ## Add 45 degree line for minimax
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  annotate("text", x = 3.5, y = 3.8, label = "R1 = R2 (Minimax Line)", color = "red") +
  coord_fixed(xlim = c(0, 5), ylim = c(0, 5)) +
  labs(title = "Geometric Representation of Risk Set", x = "Risk R(theta_1)", y = "Risk R(theta_2)") +
  theme_minimal()
```

## Theorems relating Minimax and Bayes

**Theorem 2.1:**
If a sequence of Bayes rules $\delta_n$ (w.r.t priors $\pi_n$) has Bayes risk converging to $C$, and $R(\theta, \delta_0) \le C$ for all $\theta$, then $\delta_0$ is Minimax.

**Theorem 2.2 (Equalizer Rule):**
An "Extended Bayes" rule that is an **equalizer** rule (constant risk across $\theta$) must be Minimax.
*Proof sketch:* If it were not minimax, there would exist a rule with strictly lower maximum risk. This would imply it has lower Bayes risk for the specific prior, leading to a contradiction with the assumption that the original rule was Bayes.

**Theorem 2.3**

Assume that the parameter space $\Theta = \{\theta_1, \dots, \theta_t\}$ is finite, and that the prior $\pi(\theta)$ gives positive probability to each $\theta_i$. Then, a Bayes rule with respect to $\pi$ is admissible.

**Theorem 2.4**

If a Bayes rule is unique, it is admissible.

**Theorem 2.5**

Let $\Theta$ be a subset of the real line.
1.  Assume that the risk functions $R(\theta, d)$ are continuous in $\theta$ for all decision rules $d$.
2.  Suppose that for any $\theta$ and any $\epsilon > 0$, the interval $(\theta - \epsilon, \theta + \epsilon)$ has positive probability under the prior $\pi(\theta)$.

Then, a Bayes rule with respect to $\pi$ is admissible.