---
title: "Brief Introduction of Neural Networks"
date: "2026-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Define the packages you need
required_pkgs <- c("ggraph", "tidygraph", "knitr")

# Find any that are missing
missing_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]

# Install missing packages if needed
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs)
}

# Load them into the session
library(ggraph)
library(tidygraph)
library(knitr)
```



## Non-linear Poisson Regression

In this architecture, we extend the standard Poisson GLM by introducing a hidden layer with $m$ units. This allows the model to learn non-linear combinations of the inputs before predicting the rate parameter $\lambda$.

### Mathematical Description

The model assumes the response variable $y$ follows a Poisson distribution: $y \sim \text{Poisson}(\lambda)$.

#### **The Forward Pass**
1.  **Hidden Layer (Linear):** $\eta_k = \sum_{i=1}^{p} \beta_{ki} x_i + b_{1,k}$
2.  **Activation:** $a_k = \exp(\eta_k)$
3.  **Output Layer:** $\lambda = \sum_{k=1}^{m} w_k a_k + b_2$
4.  **Poisson Loss:** $L = \lambda - y \ln(\lambda)$

```{r poisson_forward_graph, fig.width=5, fig.height=4, out.width="100%",echo=FALSE}
library(ggraph)
library(tidygraph)

# 1. Define Nodes (Input p=3, Hidden m=3, Total Nodes = 11)
nodes_fwd <- data.frame(
  id = 1:11,
  label = c("italic(x)[1]", "italic(x)[2]", "italic(x)[3]",
            "eta[1]", "eta[2]", "eta[3]", 
            "italic(a)[1]", "italic(a)[2]", "italic(a)[3]",
            "lambda", "italic(L)"),
  x = c(1, 1, 1, 5, 5, 5, 9, 9, 9, 13, 17),
  y = c(3, 2, 1, 3, 2, 1, 3, 2, 1, 2, 2),
  fill_color = c(rep("#F0F8FF", 3), rep("#FFFACD", 3), rep("#E6E6FA", 3), "#E0FFF0", "#FFE4E1")
)

# 2. Define Edges (Full connectivity between x and eta)
edges_fwd <- data.frame(
  from = c(1,1,1, 2,2,2, 3,3,3,  4,5,6,  7,8,9,  10),
  to   = c(4,5,6, 4,5,6, 4,5,6,  7,8,9,  10,10,10, 11),
  label = c("beta[11]", "beta[21]", "beta[31]", 
            "beta[12]", "beta[22]", "beta[32]",
            "beta[13]", "beta[23]", "beta[33]",
            "exp", "exp", "exp",
            "italic(w)[1]", "italic(w)[2]", "italic(w)[3]",
            "Poisson")
)

net_fwd <- tbl_graph(nodes = nodes_fwd, edges = edges_fwd)

ggraph(net_fwd, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(label = label), 
                 arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                 start_cap = circle(4, 'mm'), end_cap = circle(4, 'mm'),   
                 label_parse = TRUE, label_pos = 0.6, vjust = -0.7,         
                 label_size = 3, color = "gray70", edge_width = 0.5) +
  geom_node_point(aes(fill = I(fill_color)), shape = 21, size = 11, color = "#4575B4", stroke = 1.2) + 
  geom_node_text(aes(label = label), parse = TRUE, size = 3) +
  theme_void() + 
  scale_y_continuous(expand = expansion(mult = 0.4)) +
  labs(caption = "Forward Pass: Flow of data with 3 symmetric input nodes.")
```

---

### Computational Gradient Graph

In the gradient network, we compute the partial derivatives. The gradient for the hidden weight $\beta_{ki}$ is:
$$\frac{\partial L}{\partial \beta_{ki}} = x_i \cdot \exp(\eta_k) \cdot w_k \cdot \frac{\lambda - y}{\lambda}$$

```{r poisson_nonlinear_graph, fig.width=5, fig.height=4, out.width="100%",echo=FALSE}
# Using the same nodes for symmetry
edges_back <- data.frame(
  from = c(1,1,1, 2,2,2, 3,3,3,  4,5,6,  7,8,9,  10),
  to   = c(4,5,6, 4,5,6, 4,5,6,  7,8,9,  10,10,10, 11),
  label = c("italic(x)[1]", "italic(x)[1]", "italic(x)[1]", 
            "italic(x)[2]", "italic(x)[2]", "italic(x)[2]",
            "italic(x)[3]", "italic(x)[3]", "italic(x)[3]",
            "exp(eta[1])", "exp(eta[2])", "exp(eta[3])",
            "italic(w)[1]", "italic(w)[2]", "italic(w)[3]",
            "frac(lambda - italic(y), lambda)")
)

net_back <- tbl_graph(nodes = nodes_fwd, edges = edges_back)

ggraph(net_back, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(label = label), 
                 arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                 start_cap = circle(4, 'mm'), end_cap = circle(4, 'mm'),   
                 label_parse = TRUE, label_pos = 0.3, vjust = -0.7,         
                 label_size = 3, color = "gray70", edge_width = 0.5) +
  geom_node_point(aes(fill = I(fill_color)), shape = 21, size = 11, color = "#4575B4", stroke = 1.2) + 
  geom_node_text(aes(label = label), parse = TRUE, size = 3) +
  theme_void() + 
  scale_y_continuous(expand = expansion(mult = 0.4)) +
  labs(caption = "Backward Pass: Local gradients shown on symmetric edges.")
```
```{r}
library(ggplot2)

# 1. Setup Parameters
set.seed(123)
n <- 400
x <- seq(-2, 2, length.out = n)

# Layer 1: Increased absolute beta for steeper curves
# We use opposing signs to create a "V" or "M" shape
beta <- c(3.0, -1.0, 0) 

# Layer 1 Biases: Spread these out so the 'activation' happens 
# at different points on the x-axis
b1   <- c(-1, 4.0, 0)

# Layer 2: Output weights (w) and Bias (b2)
# Using a negative weight for one unit creates a 'dip' or valley
w    <- c(2.0, 1.0, 0)*0.1 
b2   <- -2

# 2. Forward Pass Simulation
eta1 <- beta[1] * x + b1[1]
eta2 <- beta[2] * x + b1[2]
eta3 <- beta[3] * x + b1[3]

a1 <- exp(eta1)
a2 <- exp(eta2)
a3 <- exp(eta3)

# Compute final rate lambda
# The combination of high positive and negative weights creates 
# severe non-linearity
lambda <- pmax(w[1]*a1 + w[2]*a2 + w[3]*a3 + b2, 0.1)

# 3. Generate Poisson Responses
y <- rpois(n, lambda = lambda)

# 4. Visualization
df <- data.frame(x = x, lambda = lambda, y = y)

ggplot(df, aes(x = x)) +
  geom_point(aes(y = y), alpha = 0.3, color = "gray40") +
  geom_line(aes(y = lambda), color = "#D6604D", linewidth = 1.2) +
  theme_minimal() +
  labs(
    title = "Severe Non-linear Poisson Regression Simulation",
    subtitle = expression(lambda(x) == sum(w[k] * exp(beta[k]*x + b[1][k]), k==1, 3) + b[2]),
    x = "Predictor (x)",
    y = "Count / Rate"
  )
```





## Neural Network: Forward and Backward Computational Graphs


### The Forward Pass: Computing $L$ from $x$

In the forward pass, information flows from the inputs to the loss. The edges represent the **parameters** (weights) and **functions** that transform the data.

#### Mathematical Description
For a single observation:

1. **Layer 1:** $\eta = \mathbf{B}x + b_1 \rightarrow h = \tanh(\eta)$

2. **Layer 2:** $\zeta = \mathbf{A}h + b_2 \rightarrow a = \tanh(\zeta)$

3. **Output:** $\mu = w^T a + b_3$

4. **Loss:** $L = \frac{(y - \mu)^2}{2\sigma^2}$ (Gaussian Negative Log-Likelihood)

```{r forward_graph, fig.width=5, fig.height=4, out.width="100%",echo=FALSE}
nodes <- data.frame(
  id = 1:12,
  label = c("italic(x)[1]", "italic(x)[2]", "eta[1]", "eta[2]", "italic(h)[1]", "italic(h)[2]", 
            "zeta[1]", "zeta[2]", "italic(a)[1]", "italic(a)[2]", "mu", "italic(L)"),
  x = c(1, 1, 5, 5, 9, 9, 13, 13, 17, 17, 21, 25), 
  y = c(2.0, 1.0, 2.0, 1.0, 2.0, 1.0, 2.0, 1.0, 2.0, 1.0, 1.5, 1.5),
  fill_color = c(rep("#F0F8FF", 2), rep("#FFFACD", 2), rep("#E6E6FA", 2), 
                 rep("#FFE4B5", 2), rep("#FFDAB9", 2), "#E0FFF0", "#FFE4E1")
)

edges_fwd <- data.frame(
  from = c(1, 1, 2, 2, 3, 4, 5, 5, 6, 6, 7, 8, 9, 10, 11),
  to   = c(3, 4, 3, 4, 5, 6, 7, 8, 7, 8, 9, 10, 11, 11, 12),
  label = c("beta[11]", "beta[21]", "beta[12]", "beta[22]",
            "tanh", "tanh",
            "alpha[11]", "alpha[21]", "alpha[12]", "alpha[22]",
            "tanh", "tanh",
            "italic(w)[1]", "italic(w)[2]",
            "Gaussian")
)

net_fwd <- tbl_graph(nodes = nodes, edges = edges_fwd)

ggraph(net_fwd, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(label = label), arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                 start_cap = circle(3.5, 'mm'), end_cap = circle(3.5, 'mm'),   
                 label_parse = TRUE, label_pos = 0.4, vjust = -0.8,         
                 label_size = 3.5, color = "gray70", edge_width = 0.6) +
  geom_node_point(aes(fill = I(fill_color)), shape = 21, size = 11, color = "#4575B4", stroke = 1.2) + 
  geom_node_text(aes(label = label), parse = TRUE, size = 3) +
  theme_void() + 
  scale_y_continuous(expand = expansion(mult = 0.5)) +
  labs(title = "Forward Pass: Flow of Values and Parameters")
```

---

### The Backward Pass: Gradient Network

In the gradient network, edges represent the **local derivatives**. The gradient of any node is the sum of the products of all paths leading to the loss.

#### The "Sum of Paths" Rule
To find $\frac{\partial L}{\partial h_j}$, we sum the branches going into the next layer:
$$\frac{\partial L}{\partial h_j} = \sum_{k} \left( \frac{\partial \zeta_k}{\partial h_j} \cdot \frac{\partial L}{\partial \zeta_k} \right)$$

```{r backward_graph,  fig.width=6, fig.height=5, out.width="100%", echo=FALSE}
edges_back <- data.frame(
  from = c(1, 1, 2, 2, 3, 4, 5, 5, 6, 6, 7, 8, 9, 10, 11),
  to   = c(3, 4, 3, 4, 5, 6, 7, 8, 7, 8, 9, 10, 11, 11, 12),
  label = c("italic(x)[1]", "italic(x)[1]", "italic(x)[2]", "italic(x)[2]",
            "1 - italic(h)[1]^2", "1 - italic(h)[2]^2",
            "italic(h)[1]", "italic(h)[1]", "italic(h)[2]", "italic(h)[2]",
            "1 - italic(a)[1]^2", "1 - italic(a)[2]^2",
            "italic(a)[1]", "italic(a)[2]",
            "frac(mu - italic(y), sigma^2)")
)

net_back <- tbl_graph(nodes = nodes, edges = edges_back)

ggraph(net_back, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(label = label), arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                 start_cap = circle(3.5, 'mm'), end_cap = circle(3.5, 'mm'),   
                 label_parse = TRUE, label_pos = 0.4, vjust = -0.8,         
                 label_size = 3.5, color = "gray70", edge_width = 0.6) +
  geom_node_point(aes(fill = I(fill_color)), shape = 21, size = 11, color = "#4575B4", stroke = 1.2) + 
  geom_node_text(aes(label = label), parse = TRUE, size = 3) +
  theme_void() + 
  scale_y_continuous(expand = expansion(mult = 0.5)) +
  labs(title = "Backward Pass: Flow of Local Gradients")

```


