---
title: "Likelihood-based Hypothesis Testing"
engine: knitr
format: 
  html: default
  pdf: default
---

## Formulation for Hypothesis Testing

### Hypothesis

We formulate the problem of hypothesis testing as deciding between two competing claims about a parameter $\theta$:

$$
H_0: \theta \in \Theta_0 \quad \text{(Null Hypothesis)}
$$

$$
H_1: \theta \in \Theta_1 \quad \text{(Alternative Hypothesis)}
$$

::: {#def-simple-composite}
### Simple and Composite Hypotheses
A hypothesis is called **simple** if it specifies a single value for the parameter (e.g., $\Theta_0$ contains only one point). It is called **composite** if it specifies more than one value.
:::

::: {#exm-normal-hypotheses}
### Normal Mean Test
Let $X_1, \dots, X_n \overset{iid}{\sim} N(\mu, \sigma^2)$.

* If $\sigma^2$ is known, $H_0: \mu = \mu_0$ is a simple hypothesis.
* If $\sigma^2$ is unknown, $H_0: \mu = \mu_0$ is a composite hypothesis (since $\sigma^2$ can vary).
:::

### Test Functions

A test is defined by a **critical region** $C_\alpha$ such that we reject $H_0$ if the data $x \in C_\alpha$. Equivalently, we can define a **test function** $\phi(x)$ representing the probability of rejecting $H_0$ given data $x$.

* A **non-randomized** test is given as follows:

$$
\phi(x) = I(x \in C_\alpha) = \begin{cases} 1 & \text{if } x \in C_\alpha \text{ (Reject } H_0 \text{)} \\ 0 & \text{otherwise} \end{cases}
$$

* A **randomized** test, $\phi(x)$ can take values in $[0, 1]$, which can be expressed typically as follows:


   $$
   \phi(x) = \begin{cases} 
   1 & \text{if } x \in C_1 \\
   \gamma & \text{if } x \in C_* \\
   0 & \text{otherwise}
   \end{cases}
   $$

   where:

   * $C_1$ is the region where we strictly reject $H_0$.
   * $C_*$ is the boundary region (often where $T(x) = k$) where we reject $H_0$ with probability $\gamma$.

* More generally, $\phi(x)$ is just a function of $x$ with values in $[0,1]$, which represents the probability that we will reject $H_0$. 


::: {#exm-randomized-test}
### Randomized Test for Binomial
Let $X \sim \text{Bin}(n=10, \theta)$. Consider testing $H_0: \theta = 1/2$ vs $H_1: \theta > 1/2$ with target size $\alpha = 0.05$.

Suppose we choose a critical region $X \ge k$.

* If $k=9$, $P(X \ge 9 | \theta=0.5) \approx 0.0107$.
* If $k=8$, $P(X \ge 8 | \theta=0.5) \approx 0.0547$.

Since we cannot achieve exactly 0.05 with a non-randomized test (the survival function jumps over 0.05), we must use a randomized test function.


The randomized test is defined as:

$$
\phi(x) = \begin{cases} 
1 & \text{if } x \in C_1 \text{ (i.e., } x \ge 9) \\
\gamma & \text{if } x \in C_* \text{ (i.e., } x = 8) \\
0 & \text{otherwise}
\end{cases}
$$

From the figure, we see that $\alpha = 0.05$ lies between $P(X \ge 9)$ and $P(X \ge 8)$. We always reject the "tail" where probabilities are strictly less than $\alpha$ (here $x \ge 9$). At the boundary $x=8$, we cannot reject with probability 1 (which would give total size 0.0547), nor with probability 0 (which would give total size 0.0107).

We choose $\gamma$ to bridge this gap:

$$
\begin{aligned}
\alpha &= P(X \ge 9) + \gamma \cdot P(X = 8) \\
0.05 &= 0.01074 + \gamma \cdot (P(X \ge 8) - P(X \ge 9)) \\
0.05 &= 0.01074 + \gamma \cdot (0.05469 - 0.01074)
\end{aligned}
$$

Solving for $\gamma$:

$$
\gamma = \frac{0.05 - 0.01074}{0.04395} \approx \frac{39}{44} \approx 0.89
$$
:::


```{tikz }
%| label: fig-binomial-survival-zoom
%| fig-cap: "Survival Function P(X >= k) with randomization detail"
%| echo: false
%| fig-align: "center"
%| out-extra: 'style="width: 90% !important;"'
%| engine.opts:
%|   extra.preamble: "\\usetikzlibrary{patterns, spy}"

\begin{tikzpicture}[
    scale=1.2,
    spy using outlines={circle, magnification=5, size=3.5cm, connect spies, thick, gray}
]

% --- Definitions ---
\def\ysc{5} % Y-scale factor

% --- Axes ---
\draw[->] (-0.5,0) -- (11.5,0) node[right] {$k$};
\draw[->] (0,0) -- (0,\ysc+0.5) node[above] {$P(X \ge k)$};

% --- Ticks ---
\foreach \x in {0,1,...,10} \draw (\x,0.1) -- (\x,-0.1) node[below] {\footnotesize \x};
\draw (0.1, \ysc) -- (-0.1, \ysc) node[left] {\footnotesize 1.0};
\draw (0.1, 0.5*\ysc) -- (-0.1, 0.5*\ysc) node[left] {\footnotesize 0.5};

% --- Data Points (Binomial 10, 0.5 Survival) ---
% P(>=8) = 0.0547, P(>=9) = 0.0107
\coordinate (P0) at (0, 1.0*\ysc);
\coordinate (P1) at (1, 0.999*\ysc);
\coordinate (P2) at (2, 0.989*\ysc);
\coordinate (P3) at (3, 0.945*\ysc);
\coordinate (P4) at (4, 0.828*\ysc);
\coordinate (P5) at (5, 0.623*\ysc);
\coordinate (P6) at (6, 0.377*\ysc);
\coordinate (P7) at (7, 0.172*\ysc);
\coordinate (P8) at (8, 0.0547*\ysc);
\coordinate (P9) at (9, 0.0107*\ysc);
\coordinate (P10) at (10, 0.001*\ysc);

% --- Drawing Sticks (Vertical Lines) ---
% We use very small points (0.4pt) so they don't look huge in the zoom
\foreach \i in {0,...,10} {
    \draw[blue, thick] (\i,0) -- (P\i);
    \fill[blue] (P\i) circle (0.4pt); 
}

% --- Alpha Threshold Line ---
\draw[red, dashed, thick] (-0.5, 0.05*\ysc) -- (11, 0.05*\ysc);
\node[red, left] at (-0.5, 0.05*\ysc) {$\alpha=0.05$};

% --- Zoom / Spy Region ---
% We spy on the gap between k=8 and k=9 around the alpha line
% Spy point: x=8.5, y=0.03*5 (approx middle of gap)
\coordinate (FocusPoint) at (8.2, 0.035*\ysc);
\coordinate (ZoomLocation) at (7, 0.5*\ysc); 

\spy [black, magnification=6] on (FocusPoint) in node at (ZoomLocation);

% --- Annotations (Outside spy to avoid double text) ---
\node[right, font=\footnotesize, align=left] at (6.5, 0.85*\ysc) {
    \textbf{Randomization Zone}\\
    Zoom shows $k=8$ and $k=9$.\\
    $\alpha$ passes between them.
};

\end{tikzpicture}
```





### Size

::: {#def-test-size}
### Size of a Test
The **size** of a test $\phi(x)$ is the maximum probability of rejecting the null hypothesis when it is true:

$$
\text{Size}(\phi) = \sup_{\theta \in \Theta_0} W_\phi(\theta) = \sup_{\theta \in \Theta_0} E_\theta[\phi(X)]
$$
:::


### Power 
We distinguish between the power function varying over parameters and the power metric of a specific test.

1. **Power Function ($W_\phi(\theta)$)**
The probability of rejecting $H_0$ as a function of the parameter $\theta$:

$$
W_\phi(\theta) = E_\theta[\phi(X)]
$$

2. **Power of the Test ($\text{Power}(\phi)$)**
In the context of a specific alternative hypothesis (e.g., $H_1: \theta = \theta_1$), we define the power as a scalar functional of $\phi$:

$$
\text{Power}(\phi) = E_{\theta_1}[\phi(X)]
$$

Ideally, we want:

* $W_\phi(\theta) \le \text{Size}(\phi)$ for all $\theta \in \Theta_0$ (Control Type I error).
* $\text{Power}(\phi)$ to be as large as possible (Maximize sensitivity to $H_1$).
  

```{r}
#| label: fig-test-function-size-power
#| fig-cap: "Illustration of a Test Function $\\phi(x)$ (Pink) relative to Size ($H_0$, Blue) and Power ($H_1$, Red)."
#| fig-width: 7
#| fig-height: 5

library(ggplot2)

# 1. Define Parameters
mu0 <- 0
mu1 <- 3
sigma <- 1
c_val <- 1.5      # Critical value
gamma_val <- 0.5  # Randomization constant

# 2. Scaling Constants
max_dens <- dnorm(mu0, mean = mu0, sd = sigma)
y_limit <- max_dens * 1.1
phi_scale <- 0.1 * y_limit

# 3. Define the Test Function phi(x) (Single Test)
# Step function: 0 -> 1 at c_val
df_phi <- data.frame(
  x_start = c(-3, c_val),
  x_end   = c(c_val, 6),
  y_start = c(0, phi_scale),
  y_end   = c(0, phi_scale)
)

ggplot() +
  # --- Layer 1: Densities (Solid Lines, No Fill) ---
  
  # H0: Normal(0, 1) - Blue (Cool)
  stat_function(fun = dnorm, args = list(mean = mu0, sd = sigma), 
                color = "blue", size = 0.8) +
  
  # H1: Normal(3, 1) - Red (Hot)
  stat_function(fun = dnorm, args = list(mean = mu1, sd = sigma), 
                color = "red", size = 0.8) +

  # --- Layer 2: Test Function phi(x) (Thick Pink Line) ---
  
  # The horizontal segments
  geom_segment(data = df_phi, 
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end), 
               color = "deeppink", size = 2.5, alpha = 0.5) +
  
  # The vertical threshold line
  geom_segment(aes(x = c_val, xend = c_val, y = 0, yend = phi_scale), 
               linetype = "dotted", color = "deeppink", size = 0.8) +
  
  # Points at discontinuity
  geom_point(aes(x = c_val, y = gamma_val * phi_scale), 
             color = "deeppink", size = 3) + 
  geom_point(aes(x = c_val, y = 0), size = 3, shape = 21, fill = "white", color = "deeppink") +
  geom_point(aes(x = c_val, y = phi_scale), size = 3, shape = 21, fill = "white", color = "deeppink") +

  # --- Layer 3: Annotations ---
  
  # Gamma label
  annotate("text", x = c_val + 0.2, y = gamma_val * phi_scale, 
           label = expression(gamma), hjust = 0, fontface = "bold", color = "deeppink") +
  
  # H0 / H1 Labels
  annotate("text", x = mu0, y = max_dens * 0.9, 
           label = expression(H[0]), color = "blue", size = 5) +
  annotate("text", x = mu1, y = max_dens * 0.9, 
           label = expression(H[1]), color = "red", size = 5) +
  
  # Critical Value Label
  annotate("text", x = c_val, y = -0.01, label = "c", vjust = 1) +

  # --- Layer 4: Scales ---
  scale_y_continuous(
    name = "Density f(x)",
    limits = c(-0.02, y_limit),
    expand = c(0, 0),
    
    # Secondary Axis for phi
    sec.axis = sec_axis(~ . / phi_scale, 
                        name = expression(phi(x)),
                        breaks = c(0, 1))
  ) +
  scale_x_continuous(name = "Observation x", limits = c(-3, 6)) +
  
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(angle = 90, vjust = 0.5),
    panel.grid.minor = element_blank()
  )
```


## The Neyman-Pearson Lemma

Consider testing a simple null against a simple alternative:
$H_0: \theta = \theta_0$ vs $H_1: \theta = \theta_1$.

We define the **Likelihood Ratio** $\Lambda(x)$ as:

$$
\Lambda(x) = \frac{f_1(x)}{f_0(x)} = \frac{f(x; \theta_1)}{f(x; \theta_0)}
$$

::: {#def-lrt}
### Likelihood Ratio Test (LRT)
A test $\phi(x)$ is a Likelihood Ratio Test if it has the form:

$$
\phi_{\text{LRT}}(x) = \begin{cases} 
1 & \text{if } \Lambda(x) > k \\
\gamma(x) & \text{if } \Lambda(x) = k \\
0 & \text{if } \Lambda(x) < k
\end{cases}
$$

where $k \ge 0$ is a constant and $0 \le \gamma(x) \le 1$.
:::



### Neyman-Pearson Lemma
::: {#thm-neyman-pearson}
### Neyman-Pearson Lemma
a) **Optimality:** For any $k$ and $\gamma(x)$, the LRT $\phi_0(x)$ defined above has maximum power among all tests whose size is less than or equal to the size of $\phi_0(x)$.

a) **Existence:** Given $\alpha \in (0, 1)$, there exist constants $k$ and $\gamma_0$ such that the LRT defined by this $k$ and $\gamma(x) = \gamma_0$ has size exactly $\alpha$.

c) **Uniqueness:** If a test $\phi$ has size $\alpha$ and is of maximum power among all tests of size $\alpha$, then $\phi$ is necessarily an LRT, except possibly on a set of measure zero under $H_0$ and $H_1$.
:::

### A Derivation with The Lagrange Multiplier Approach

To make the optimality of the Likelihood Ratio Test (LRT) intuitive, we can frame the search for the best test function $\phi(x)$ as a constrained optimization problem.

We want to maximize the power of the test:

$$
\text{Power}(\phi) = \int \phi(x) f_1(x) dx
$$

subject to the constraint on the size of the test $\alpha$:

$$
\text{Size}(\phi) = \int \phi(x) f_0(x) dx = \alpha
$$

Using the method of Lagrange multipliers, we define the objective function $L$ with a multiplier $k$:

$$
L(\phi, k) = \int \phi(x) f_1(x) dx - k \left( \int \phi(x) f_0(x) dx - \alpha \right)
$$

Rearranging the terms inside the integral, we get:

$$
L(\phi, k) = \int \phi(x) [f_1(x) - k f_0(x)] dx + k\alpha
$$

To maximize $L$ with respect to $\phi(x)$, we look at the integrand. Since $0 \le \phi(x) \le 1$, we should choose $\phi(x)$ to be as large as possible whenever its coefficient is positive, and as small as possible whenever its coefficient is negative:

* If $f_1(x) - k f_0(x) > 0$, set $\phi(x) = 1$.
* If $f_1(x) - k f_0(x) < 0$, set $\phi(x) = 0$.
* If $f_1(x) - k f_0(x) = 0$, the value of $\phi(x)$ does not affect the integral (this is where $\gamma$ comes in).

This decision rule is equivalent to:

$$
\phi(x) = 
\begin{cases} 
1 & \text{if } \frac{f_1(x)}{f_0(x)} > k \\
0 & \text{if } \frac{f_1(x)}{f_0(x)} < k
\end{cases}
$$

This is precisely the form of the Likelihood Ratio Test. The "shadow price" or Lagrange multiplier $k$ represents the critical threshold that balances the gain in power against the cost of increasing the Type I error.

### Proof of NP Lemma

::: {.proof}

**Proof of (a) Optimality:**
Let $\phi_{\text{LRT}}$ be the LRT with size $\alpha$, and $\phi$ be any other test with size $\le \alpha$.
Define the function $U(x)$ as the difference in test functions weighted by the linear combination of densities:

$$
U(x) = (\phi_{}(x) - \phi(x))(f_1(x) - k f_0(x))
$$

We analyze the sign of $U(x)$ by looking at the sign of its two factors in three cases:

* If $f_1(x) - k f_0(x) > 0$ (implies $\Lambda(x) > k$). Since $\phi_{\text{LRT}}(x) = 1$ and $\phi(x) \le 1$, we have:
    
    $$
    \begin{aligned}
    \phi_{\text{LRT}}(x) - \phi(x) &\ge 0 \\
    U(x) = (\phi_{\text{LRT}}(x) - \phi(x))(f_1(x) - k f_0(x)) &\ge 0
    \end{aligned}
    $$

* If $f_1(x) - k f_0(x) < 0$ (implies $\Lambda(x) < k$). Since $\phi_{\text{LRT}}(x) = 0$ and $\phi(x) \ge 0$, we have:

    $$
    \begin{aligned}
    \phi_{\text{LRT}}(x) - \phi(x) &\le 0 \\
    U(x) = (\phi_{\text{LRT}}(x) - \phi(x))(f_1(x) - k f_0(x)) &\ge 0
    \end{aligned}
    $$

* If $f_1(x) - k f_0(x) = 0$.
    The product is zero regardless of the test functions.
    
    $$
    U(x) = 0
    $$

Combining these cases, we conclude that the product is non-negative for all $x$:

$$
U(x) = (\phi_{\text{LRT}}(x) - \phi(x))(f_1(x) - k f_0(x)) \ge 0
$$

Therefore, integrating $U(x)$ over the entire domain:

$$
\int U(x) dx = \int (\phi_{\text{LRT}}(x) - \phi(x))(f_1(x) - k f_0(x)) dx \ge 0
$$

Expanding the integral:

$$
\int \phi_{\text{LRT}}(x) f_1(x) \, dx - \int \phi(x) f_1(x) \, dx - k \left( \int \phi_{\text{LRT}}(x) f_0(x) \, dx - \int \phi(x) f_0(x) \, dx \right) \ge 0
$$
Converting to expectations:

$$
E_{\theta_1}[\phi_{\text{LRT}}] - E_{\theta_1}[\phi] \ge k (E_{\theta_0}[\phi_{\text{LRT}}] - E_{\theta_0}[\phi])  
$$

Since $E_{\theta_0}[\phi_{\text{LRT}}] =\text{Size}(\phi_{\text{LRT}})= \alpha$ and we require that $E_{\theta_0}[\phi] = \text{Size}(\phi)\le \alpha$,  
$$
E_{\theta_0}[\phi_{\text{LRT}}] - E_{\theta_0}[\phi] \ge 0
$$

Thereore, given that $k \ge 0$:

$$
\text{Power}(\phi_{\text{LRT}}) \ge \text{Power}(\phi)
$$

**Proof of (b) Existence:**

Let $G(k) = P_{\theta_0}(\Lambda(X) \le k)$. $G(k)$ is the cumulative distribution function of the random variable $\Lambda(X)$, so it is non-decreasing.
We seek $k_0$ such that $1 - G(k_0) \approx \alpha$.
Because of discrete jumps, we might not hit $\alpha$ exactly.
We choose $k_0$ such that:

$$
P_{\theta_0}(\Lambda(X) > k_0) \le \alpha \le P_{\theta_0}(\Lambda(X) \ge k_0)
$$

Set $\gamma_0 = \frac{\alpha - P_{\theta_0}(\Lambda(X) > k_0)}{P_{\theta_0}(\Lambda(X) = k_0)}$.


**Proof of (c) Uniqueness**

Let $\phi_{\text{LRT}}$ be the LRT of size $\alpha$. Suppose there exists another test $\phi$ that is also Most Powerful (MP) with size $\le \alpha$. We wish to show that $\phi(x) = \phi_{\text{LRT}}(x)$ for almost all $x$ where $f_1(x) \ne k f_0(x)$.

As established in the optimality proof, the function:
$$
U(x) = (\phi_{\text{LRT}}(x) - \phi(x))(f_1(x) - k f_0(x))
$$
is non-negative for all $x$. Since both tests are MP, they have the same power: $E_{\theta_1}[\phi_{\text{LRT}}] = E_{\theta_1}[\phi]$. 

From the integral of $U(x)$, we have:
$$
0 \le \int U(x) dx = (E_{\theta_1}[\phi_{\text{LRT}}] - E_{\theta_1}[\phi]) - k(E_{\theta_0}[\phi_{\text{LRT}}] - E_{\theta_0}[\phi])
$$

Substituting the equality of power:
$$
0 \le -k(\alpha - E_{\theta_0}[\phi])
$$

Since $k > 0$ and $E_{\theta_0}[\phi] \le \alpha$, the term $-k(\alpha - E_{\theta_0}[\phi])$ is $\le 0$. The only way for the integral of a non-negative function $U(x)$ to be $\le 0$ is if the integral is exactly zero:
$$
\int (\phi_{\text{LRT}}(x) - \phi(x))(f_1(x) - k f_0(x)) \, dx = 0
$$

For the integral of a non-negative function to be zero, the integrand must be zero almost everywhere:
$$
(\phi_{\text{LRT}}(x) - \phi(x))(f_1(x) - k f_0(x)) = 0 \quad \text{a.e.}
$$

This implies that for any $x$ where $f_1(x) - k f_0(x) \ne 0$, we must have:
$$
\phi_{\text{LRT}}(x) - \phi(x) = 0 \implies \phi(x) = \phi_{\text{LRT}}(x)
$$

Thus, the test is unique except possibly on the boundary set $\{x : f_1(x) = k f_0(x)\}$. If $P_{\theta_0}(\Lambda(X) = k) = 0$ (as in continuous distributions like the Normal), the MP test is unique almost everywhere.

:::

```{r}
#| label: fig-NPlemma-proof
#| fig-cap: "Visualizing the NP Lemma. The thick, transparent pink line is $\\phi_{\text{LRT}}$. The thin solid black line is $\\phi$. Overlap is visible as a black line inside pink."
#| fig-width: 7
#| fig-height: 5

library(ggplot2)

# 1. Define Parameters
mu0 <- 0
mu1 <- 3
sigma <- 1
c_lrt <- 2
gamma_val <- 0.5

# Scaling constants
max_dens <- dnorm(mu0, mean = mu0, sd = sigma)
y_limit <- max_dens * 1.1
phi_scale <- 0.1 * y_limit

# 2. Define Test Functions Data
df_lrt <- data.frame(
  x_start = c(-3, c_lrt),
  x_end   = c(c_lrt, 6),
  y_val   = c(0, phi_scale),
  Test    = "phi[LRT]" 
)

df_other <- data.frame(
  x_start = c(-3, 0.5, 1, 3),
  x_end   = c(0.5, 1, 3,6),
  y_val   = c(0, phi_scale, 0, 0),
  Test    = "phi"
)

ggplot() +
  # --- Layer 1: Densities (Solid Lines, No Fill) ---
  stat_function(fun = dnorm, args = list(mean = mu0, sd = sigma), 
                color = "blue", size = 0.8) +
  stat_function(fun = dnorm, args = list(mean = mu1, sd = sigma), 
                color = "red", size = 0.8) +

  # --- Layer 2: Test Functions (Segments) ---
  
  # 2a. LRT Segments (Thick, Transparent Pink)
  geom_segment(data = df_lrt,
               aes(x = x_start, xend = x_end, 
                   y = y_val, yend = y_val, 
                   color = Test, linetype = Test, size = Test),
               alpha = 0.4) + 
  
  # Vertical connector for LRT
  geom_segment(aes(x = c_lrt, xend = c_lrt, y = 0, yend = phi_scale),
               color = "deeppink", linetype = "dotted", size = 0.5, alpha = 0.6) +

  # 2b. Other Test Segments (Thin Black Opaque)
  geom_segment(data = df_other,
               aes(x = x_start, xend = x_end, 
                   y = y_val, yend = y_val, 
                   color = Test, linetype = Test, size = Test)) +
  
  # Vertical connectors for Other Test
  geom_segment(aes(x = 1, xend = 1, y = 0, yend = phi_scale),
               color = "black", linetype = "dotted", size = 0.5) +
  geom_segment(aes(x = 3, xend = 3, y = 0, yend = phi_scale),
               color = "black", linetype = "dotted", size = 0.5) +

  # --- Layer 3: Annotations ---
  # LRT Gamma Point (Pink)
  geom_point(aes(x = c_lrt, y = gamma_val * phi_scale), 
             color = "deeppink", size = 3) +
  
  # Density Labels
  annotate("text", x = mu0, y = max_dens * 0.9, 
           label = expression(H[0]), color = "blue", size = 5) +
  annotate("text", x = mu1, y = max_dens * 0.9, 
           label = expression(H[1]), color = "red", size = 5) +

  # --- Layer 4: Scales and Legend ---
  scale_y_continuous(
    name = "Density f(x)", limits = c(-0.02, y_limit), expand = c(0, 0),
    sec.axis = sec_axis(~ . / phi_scale, name = expression(phi(x)), breaks = c(0, 1))
  ) +
  scale_x_continuous(name = "Observation x", limits = c(-3, 6)) +
  
  # Manual Legend Controls (FIXED using named vectors for labels)
  scale_color_manual(name = "Test Function",
                     values = c("phi[LRT]" = "deeppink", "phi" = "black"),
                     labels = c("phi[LRT]" = expression(phi[LRT]), "phi" = expression(phi))) +
  scale_linetype_manual(name = "Test Function",
                        values = c("phi[LRT]" = "solid", "phi" = "solid"),
                        labels = c("phi[LRT]" = expression(phi[LRT]), "phi" = expression(phi))) +
  scale_size_manual(name = "Test Function",
                    values = c("phi[LRT]" = 4,   # Extra Thick
                               "phi" = 1), # Thin
                    labels = c("phi[LRT]" = expression(phi[LRT]), "phi" = expression(phi))) +
  
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    axis.title.y.right = element_text(angle = 90, vjust = 0.5),
    panel.grid.minor = element_blank()
  )
  
```


## Uniformly Most Powerful (UMP) Tests

When the alternative hypothesis is composite ($H_1: \theta \in \Theta_1$), we seek a test that is "best" for *all* $\theta \in \Theta_1$.

::: {#def-ump}
### Uniformly Most Powerful Test

A test $\phi_0(x)$ of size $\alpha$ is **Uniformly Most Powerful (UMP)** if:

1.  $E_{\theta}[\phi_0(X)] \le \alpha$ for all $\theta \in \Theta_0$.
2.  For any other test $\phi(x)$ satisfying (1), $E_{\theta}[\phi_0(X)] \ge E_{\theta}[\phi(X)]$ for all $\theta \in \Theta_1$.
   
:::

```{tikz }
%| label: fig-ump-power
%| fig-cap: "Diagram of UMP Tests"
%| echo: false
%| fig-align: "center"
%| out-extra: 'style="width: 80% !important;"'

\begin{tikzpicture}
\draw[->] (0,0) -- (6,0) node[right] {$\theta$};
\draw[->] (0,0) -- (0,4) node[above] {Power $W(\theta)$};

\draw[dashed] (3,0) -- (3,4);
\node at (3,-0.3) {$\theta_0$};
\node at (1.5, -0.3) {$\Theta_0$};
\node at (4.5, -0.3) {$\Theta_1$};

\draw[thick, blue] (0,0.5) .. controls (2,0.6) and (3,1.5) .. (6,3.5) node[right] {$\phi_0$ (UMP)};
\draw[thick, red] (0,0.8) .. controls (2,0.9) and (3,1.5) .. (6,2.5) node[right] {$\phi$};

\draw[dotted] (0,1.5) -- (3,1.5);
\node at (-0.3, 1.5) {$\alpha$};

\end{tikzpicture}

```

## Monotone Likelihood Ratio (MLR)

::: {#def-mlr}
### Monotone Likelihood Ratio
A family of densities $\{f(x; \theta)\}$ has a **Monotone Likelihood Ratio (MLR)** with respect to a statistic $T(x)$ if for any $\theta_1 > \theta_0$, the ratio:

$$
\frac{f(x; \theta_1)}{f(x; \theta_0)}
$$

is a non-decreasing function of $T(x)$.
:::

Common examples include the one-parameter Exponential Family:
$f(x; \theta) = h(x) c(\theta) \exp\{w(\theta) T(x)\}$.
If $w(\theta)$ is increasing, the family has MLR w.r.t $T(x)$.


:::: {#exm-mlr-gamma}

Let $X_1, \dots, X_n \overset{iid}{\sim} \text{Exp}(\theta)$ with pdf $f(x) = \frac{1}{\theta} e^{-x/\theta}$.
Test $$H_0: \theta = \theta_0 \text{ vs } H_1: \theta > \theta_0.$$


The Likelihood Ratio for $\theta_1 > \theta_0$ is:

$$
\frac{L(\theta_1)}{L(\theta_0)} = \frac{\theta_1^{-n} e^{-\sum x_i / \theta_1}}{\theta_0^{-n} e^{-\sum x_i / \theta_0}} = \left(\frac{\theta_0}{\theta_1}\right)^n \exp \left\{ \left( \frac{1}{\theta_0} - \frac{1}{\theta_1} \right) \sum x_i \right\}
$$

Since $\theta_1 > \theta_0$, the term $(\frac{1}{\theta_0} - \frac{1}{\theta_1})$ is positive. Thus, $\Lambda(x)$ is an increasing function of the sum $T (x) = \sum x_i$.

Rejecting for large $\Lambda(x)$ is equivalent to rejecting for $T(x)=\sum x_i > C$.

Under $H_0$, $X_i \sim \text{Exp}(\theta_0)$, which is equivalent to $\text{Gamma}(1, \theta_0)$. By the reproductive property of the Gamma distribution:

$$T = \sum_{i=1}^n X_i \sim \text{Gamma}(n, \theta_0)$$

Alternatively, using the relationship with the Chi-square distribution:

$$\frac{2T}{\theta_0} \sim \chi^2_{2n}$$

To find $C$ for a significance level $\alpha$, we set $P(T > C | \theta_0) = \alpha$. Using the $\chi^2$ transformation:

$$P\left(\frac{2T}{\theta_0} > \frac{2C}{\theta_0}\right) = \alpha \implies \frac{2C}{\theta_0} = \chi^2_{2n, \alpha}$$

Thus, the critical value is:

$$C = \frac{\theta_0}{2} \chi^2_{2n, \alpha}$$

where $\chi^2_{2n, \alpha}$ is the upper-$\alpha$ quantile of a Chi-square distribution with $2n$ degrees of freedom. 

::: {.callout-important}
We note that the value $C$ does not depend on $\theta_1$.
:::

::::



### Karlin-Rubin Theorem 

::: {.theorem #thm-chebyshev-covariance}
#### Chebyshev's Association Inequality

Let $X$ be a random variable, and let $f(x)$ and $g(x)$ be two functions that are both non-decreasing (or both non-increasing). Then:
$$
E[f(X)g(X)] \ge E[f(X)] \cdot E[g(X)]
$$
Equivalently, the covariance is non-negative: $\text{Cov}(f(X), g(X)) \ge 0$.
:::

::: {.proof}
Let $Y$ be an independent copy of $X$ (i.e., $X$ and $Y$ are i.i.d.). Consider the quantity:
$$
\Delta = (f(X) - f(Y))(g(X) - g(Y))
$$

Since $f$ and $g$ are both non-decreasing (or both non-increasing), the terms $(f(X) - f(Y))$ and $(g(X) - g(Y))$ always share the same sign. Thus, their product is always non-negative:
$$
\Delta \ge 0
$$

Taking the expectation:
$$
E[(f(X) - f(Y))(g(X) - g(Y))] \ge 0
$$

Expanding the product and using linearity of expectation:
$$
E[f(X)g(X)] - E[f(X)g(Y)] - E[f(Y)g(X)] + E[f(Y)g(Y)] \ge 0
$$

Since $X$ and $Y$ are i.i.d.:

1.  $E[f(Y)g(Y)] = E[f(X)g(X)]$
2.  $E[f(X)g(Y)] = E[f(X)]E[g(Y)] = E[f(X)]E[g(X)]$ (Independence)
3.  $E[f(Y)g(X)] = E[f(Y)]E[g(X)] = E[f(X)]E[g(X)]$ (Independence)

Substituting these back yields:
$$
2E[f(X)g(X)] - 2E[f(X)]E[g(X)] \ge 0
$$

Dividing by 2 proves the inequality:
$$
E[f(X)g(X)] \ge E[f(X)]E[g(X)]
$$
:::


::: {#thm-karlin-rubin}
#### Karlin-Rubin Theorem 
Suppose $X$ has a distribution from a family with MLR with respect to $T(X)$, and the distribution of $T(X)$ is continuous.
Consider testing $H_0: \theta \le \theta_0$ vs $H_1: \theta > \theta_0$.

The test:

$$
\phi(x) = \begin{cases} 
1 & \text{if } T(x) > t_0 \\
0 & \text{if } T(x) \le t_0
\end{cases}
$$

where $t_0$ is determined by $P_{\theta_0}(T(X) > t_0) = \alpha$, is the UMP size $\alpha$ test.
:::

::: {.proof}

**The Test:**
Define the test $\phi_0(x)$ as:

$$
\phi_0(x) = \begin{cases} 
1 & \text{if } T(x) > t_0 \\
0 & \text{if } T(x) \le t_0
\end{cases}
$$

where $t_0$ is determined such that the power at the boundary is $\alpha$, i.e., $W_{\phi_0}(\theta_0) = \alpha$.

1. **Monotonicity of the Power Function**

   We first establish that $W_{\phi_0}(\theta)$ is non-decreasing. Let $\theta_1 > \theta_0$.
   Define the test indicator function $h(t) = \mathbb{I}(t > t_0)$ and the likelihood ratio $\Lambda(t) = \frac{f_{\theta_1}(t)}{f_{\theta_0}(t)}$.
   
   Because of the Monotone Likelihood Ratio (MLR) property, $\Lambda(t)$ is a non-decreasing function of $t$. The indicator function $h(t)$ is clearly non-decreasing.
   
   The power at $\theta_1$ can be written as an integral involving the density under $\theta_0$:
   $$
   W_{\phi_0}(\theta_1) = \int_{-\infty}^{\infty} h(t) f_{\theta_1}(t) \, dt = \int_{-\infty}^{\infty} h(t) \frac{f_{\theta_1}(t)}{f_{\theta_0}(t)} f_{\theta_0}(t) \, dt = \int_{-\infty}^{\infty} h(t) \Lambda(t) f_{\theta_0}(t) \, dt
   $$
   
   This integral is the expectation $E_{\theta_0}[h(T) \Lambda(T)]$. By Chebyshev's Association Inequality (Covariance Inequality), since both $h(t)$ and $\Lambda(t)$ are non-decreasing, the expectation of their product is at least the product of their expectations:
   $$
   \int_{-\infty}^{\infty} h(t) \Lambda(t) f_{\theta_0}(t) \, dt \ge \left( \int_{-\infty}^{\infty} h(t) f_{\theta_0}(t) \, dt \right) \left( \int_{-\infty}^{\infty} \Lambda(t) f_{\theta_0}(t) \, dt \right)
   $$
   
   We evaluate the two integrals on the right-hand side:
   1. The first term is the power at $\theta_0$: $\int h(t) f_{\theta_0}(t) \, dt = W_{\phi_0}(\theta_0)$.
   2. The second term integrates the likelihood ratio: $\int \frac{f_{\theta_1}(t)}{f_{\theta_0}(t)} f_{\theta_0}(t) \, dt = \int f_{\theta_1}(t) \, dt = 1$.
   
   Substituting these back, we get:
   $$
   W_{\phi_0}(\theta_1) \ge W_{\phi_0}(\theta_0) \cdot 1
   $$
   Thus, $W_{\phi_0}(\theta)$ is non-decreasing.

2. **Size Control**

   For the composite null $H_0: \theta \le \theta_0$, we require the size to be at most $\alpha$. Since $W_{\phi_0}(\theta)$ is non-decreasing (from Step 1) and we explicitly set $W_{\phi_0}(\theta_0) = \alpha$:
    
   $$
   W_{\phi_0}(\theta) \le W_{\phi_0}(\theta_0) = \alpha \quad \text{for all } \theta \le \theta_0
   $$
    
   This confirms $\phi_0$ is a valid level-$\alpha$ test.

3. **Uniformly Most Powerful (UMP) via Neyman-Pearson Lemma**

   Let $\phi'(x)$ be any other valid test of size $\alpha$ for $H_0$. This implies $W_{\phi'}(\theta_0) \le \alpha$.
    
   Consider any specific alternative $\theta_1 > \theta_0$. Because the family has MLR, the likelihood ratio $\Lambda(x)$ is increasing in $T(x)$. Therefore, the test $\phi_0$ (which rejects for large $T$) is identified by the Neyman-Pearson Lemma as the Most Powerful (MP) test for the simple hypothesis $\theta_0$ vs $\theta_1$.
    
   Comparing the power of $\phi_0$ and $\phi'$ at this specific $\theta_1$:
   $$
   W_{\phi_0}(\theta_1) \ge W_{\phi'}(\theta_1)
   $$
   Since $\theta_1$ was arbitrary, this inequality holds for all $\theta > \theta_0$. Thus, $\phi_0$ is the UMP test.

:::

```{r }
#| label: fig-monotone-power-MLR
#| echo: false
#| fig-cap: "Visualizing Monotonicity of Power Function $W(\\theta)$ for MLR Distributions"

# Parameters
mu0 <- 0   # Corresponds to theta_0
mu1 <- 2   # Corresponds to theta_1 (where theta_1 > theta_0)
sigma <- 1.5
t0 <- 1.0

# Grid
x <- seq(-4, 6, length.out = 500)
f0 <- dnorm(x, mean = mu0, sd = sigma)
f1 <- dnorm(x, mean = mu1, sd = sigma)

# Calculate exact densities at threshold t0 for polygon closing
y0_t0 <- dnorm(t0, mean = mu0, sd = sigma)
y1_t0 <- dnorm(t0, mean = mu1, sd = sigma)

# Plotting
plot(x, f0, type = "l", col = "blue", lwd = 2, 
     ylab = "Density f(t)", xlab = "T(x)", 
     main = expression(paste("Monotonicity of Power: ", W(theta[1]) > W(theta[0]))))
lines(x, f1, col = "red", lwd = 2)

# Define the domain for the tails (t > t0)
idx_tail <- x > t0
x_poly <- c(t0, t0, x[idx_tail], max(x))

# Shading W(theta_1) - The red tail
# Corresponds to the Alternative theta_1
y_poly1 <- c(0, y1_t0, f1[idx_tail], 0)
polygon(x_poly, y_poly1, 
        col = rgb(1, 0, 0, 0.2), border = NA)

# Shading W(theta_0) - The blue tail
# Corresponds to the Null theta_0
y_poly0 <- c(0, y0_t0, f0[idx_tail], 0)
polygon(x_poly, y_poly0, 
        col = rgb(0, 0, 1, 0.2), border = NA)

# Add Threshold Line
abline(v = t0, col = "black", lwd = 2, lty = 2)

# Legend
legend("topleft", 
       legend = c(expression(f[theta[0]] ~ "(Null)"), 
                  expression(f[theta[1]] ~ "(Alternative)"), 
                  expression(W(theta[0]) == P(T > t[0] ~ "|" ~ theta[0])), 
                  expression(W(theta[1]) == P(T > t[0] ~ "|" ~ theta[1]))),
       col = c("blue", "red", rgb(0, 0, 1, 0.2), rgb(1, 0, 0, 0.2)), 
       lwd = c(2, 2, 10, 10), bty = "n")

text(t0, max(f0)*0.8, expression(t[0]), pos = 4, col = "black", cex = 1.2)

# Annotations pointing to the areas
text(3, 0.05, expression(W(theta[0])), col = "blue", pos = 3)
text(3.5, 0.15, expression(W(theta[1])), col = "red", pos = 3)
```

::: {#exm-ump-gamma}
### UMP Test for Exponential/Gamma
Let $X_1, \dots, X_n \overset{iid}{\sim} \text{Exp}(\theta)$ with pdf $f(x) = \frac{1}{\theta} e^{-x/\theta}$.
Test $H_0: \theta = \theta_0$ vs $H_1: \theta > \theta_0$.
The sum $T = \sum X_i$ is a sufficient statistic, and $T \sim \text{Gamma}(n, \theta)$.

The Likelihood Ratio for $\theta_1 > \theta_0$ is:

$$
\frac{L(\theta_1)}{L(\theta_0)} = \frac{\theta_1^{-n} e^{-\sum x_i / \theta_1}}{\theta_0^{-n} e^{-\sum x_i / \theta_0}} = \left(\frac{\theta_0}{\theta_1}\right)^n \exp \left\{ \left( \frac{1}{\theta_0} - \frac{1}{\theta_1} \right) \sum x_i \right\}
$$

Since $\theta_1 > \theta_0$, the term $(\frac{1}{\theta_0} - \frac{1}{\theta_1})$ is positive. Thus, $\Lambda(x)$ is an increasing function of $\sum x_i$.

Rejecting for large $\Lambda(x)$ is equivalent to rejecting for $\sum x_i > C$.


$T \sim \text{Gamma}(n, \theta)$

This test form does not depend on the specific $\theta_1$, so it is UMP for all $\theta > \theta_0$.

:::


### Non-Existence of UMP for Two-Sided Hypotheses

For testing a point null hypothesis $H_0: \theta = \theta_0$ against a two-sided alternative $H_1: \theta \neq \theta_0$ in a family with a monotone likelihood ratio (e.g., Normal, Exponential), a Uniformly Most Powerful (UMP) test generally **does not exist**. The non-existence proof relies on the uniqueness of the Most Powerful (MP) test derived from the Neyman-Pearson Lemma:

1.  **Conflict of Optimal Regions:**
    Consider a specific alternative $\theta_1 > \theta_0$. By the Neyman-Pearson Lemma, the MP test $\phi_1$ rejects $H_0$ for large values of the sufficient statistic $T(\mathbf{X}) > k_1$.
    Conversely, consider an alternative $\theta_2 < \theta_0$. The MP test $\phi_2$ rejects $H_0$ for small values of the statistic $T(\mathbf{X}) < k_2$.

2.  **Failure of Uniformity:**
    A UMP test $\phi^*$ would need to be the MP test for *every* $\theta \in H_1$.
    * For $\phi^*$ to be most powerful against $\theta_1$, it must be equivalent to $\phi_1$ (rejecting in the right tail).
    * For $\phi^*$ to be most powerful against $\theta_2$, it must be equivalent to $\phi_2$ (rejecting in the left tail).

3.  **Biased Power Function:**
    The MP test $\phi_1$ (Right-Sided) has a power function that drops below the size $\alpha$ for values $\theta < \theta_0$. Therefore, it cannot be the most powerful test for $\theta_2$, as there exists a valid test (e.g., $\phi_2$) with power strictly greater than $\alpha$ at $\theta_2$.
    
Since no single critical region can simultaneously maximize power for both $\theta > \theta_0$ and $\theta < \theta_0$, no UMP test exists. We typically restrict our search to **Unbiased** tests (UMPU) to resolve this.



```{r}
#| label: fig-two-powers-normal
#| echo: false
#| fig-height: 4
#| fig-width: 6
#| fig-cap: Illustration of Non-existence of UMP Tests for Two-sided Tests
theta <- seq(-3, 3, length.out=200)
power_right <- 1 - pnorm(1.645 - theta) 
power_left <- pnorm(-1.645 - theta)

plot(theta, power_right, type='l', col='red', lwd=2, 
   ylab="Power W(theta)", xlab=expression(theta),
   main="Power Functions of One-Sided Tests")
lines(theta, power_left, col='blue', lwd=2)
abline(h=0.05, lty=2, col="gray")
abline(v=0, lty=3)
legend("top", legend=c(expression(W[right](theta)), expression(W[left](theta))), 
      col=c("red", "blue"), lwd=2, bty="n")
```

## LRT, Score, and Wald's Tests

### Likelihood Ratio Tests

The Likelihood Ratio Test (LRT) provides a unified approach for constructing hypothesis tests for parametric models. Let $\mathbf{Y} = (Y_1, \dots, Y_n)$ be a random sample with joint probability density function $L(\boldsymbol{\theta}; \mathbf{Y})$, where $\boldsymbol{\theta} \in \Theta$. We consider the general hypothesis testing problem:
$$H_0: \boldsymbol{\theta} \in \Theta_0 \quad \text{vs} \quad H_1: \boldsymbol{\theta} \in \Theta \setminus \Theta_0$$

The **Likelihood Ratio Statistic** is defined as the ratio of the maximum likelihood achievable under the null hypothesis to the maximum likelihood achievable under the full parameter space:

$$\Lambda(\mathbf{Y}) = \frac{\sup_{\boldsymbol{\theta} \in \Theta_0} L(\boldsymbol{\theta}; \mathbf{Y})}{\sup_{\boldsymbol{\theta} \in \Theta} L(\boldsymbol{\theta}; \mathbf{Y})} = \frac{L(\hat{\boldsymbol{\theta}}^{(0)})}{L(\hat{\boldsymbol{\theta}}^{(1)})}$$

where $\hat{\boldsymbol{\theta}}^{(0)}$ is the restricted MLE and $\hat{\boldsymbol{\theta}}^{(1)}$ is the unrestricted (global) MLE. By definition, the likelihood ratio is bounded by $0 \le \Lambda(\mathbf{Y}) \le 1$. The Likelihood Ratio Test (LRT) is defined by the decision function:

$$
\phi(\mathbf{Y}) = \begin{cases} 
1 & \text{if } \Lambda(\mathbf{Y}) \le c \quad (\text{Reject } H_0) \\
0 & \text{if } \Lambda(\mathbf{Y}) > c \quad (\text{Fail to reject } H_0)
\end{cases}
$$

where $c \in (0, 1)$ is a critical value determined by the significance level $\alpha$. This rejection region $\{\mathbf{Y} : \Lambda(\mathbf{Y}) \le c\}$ corresponds to cases where the restricted likelihood (under $H_0$) is significantly smaller than the unrestricted likelihood.

### Geometry of Likelihood

Before deriving the asymptotic distribution of the LRT, we establish the geometric framework. We define the inner product on the parameter space induced by the Fisher Information $\mathcal{I}(\boldsymbol{\theta}_0)$ as $\langle \mathbf{u}, \mathbf{v} \rangle_{\mathcal{I}} = \mathbf{u}^T \mathcal{I}(\boldsymbol{\theta}_0) \mathbf{v}$.

::: {#def-orthogonality}

#### Definition: Orthogonal Projection under Fisher Information

Let $\Theta_0$ be a linear subspace of the parameter space. The **orthogonal projection** of a point $\hat{\boldsymbol{\theta}}^{(1)}$ onto $\Theta_0$ with respect to the inner product $\langle \cdot, \cdot \rangle_{\mathcal{I}}$ is the unique vector $\tilde{\boldsymbol{\theta}}^{(0)} \in \Theta_0$ such that the residual is orthogonal to the subspace:
$$ \langle \hat{\boldsymbol{\theta}}^{(1)} - \tilde{\boldsymbol{\theta}}^{(0)}, \mathbf{v} \rangle_{\mathcal{I}} = 0 \quad \text{for all } \mathbf{v} \in \Theta_0 $$

**Consequence (Pythagorean Identity):**
As a direct consequence of this definition, for any $\boldsymbol{\theta}_0 \in \Theta_0$, the vector $(\tilde{\boldsymbol{\theta}}^{(0)} - \boldsymbol{\theta}_0)$ lies in $\Theta_0$ and is therefore orthogonal to the residual $(\hat{\boldsymbol{\theta}}^{(1)} - \tilde{\boldsymbol{\theta}}^{(0)})$. This implies the Pythagorean relationship holds exactly:
$$\|\hat{\boldsymbol{\theta}}^{(1)} - \boldsymbol{\theta}_0\|_{\mathcal{I}}^2 = \|\hat{\boldsymbol{\theta}}^{(1)} - \tilde{\boldsymbol{\theta}}^{(0)}\|_{\mathcal{I}}^2 + \|\tilde{\boldsymbol{\theta}}^{(0)} - \boldsymbol{\theta}_0\|_{\mathcal{I}}^2$$

:::

### Wilks' Theorem for Testing Reduced Models vs Full Models

The asymptotic distribution of the Likelihood Ratio Test statistic is given by the celebrated result from Wilks (1938). This theorem allows us to construct hypothesis tests without deriving the exact distribution of the likelihood ratio $\Lambda$ for every specific model.

::: {#thm-wilks}

### Wilks' Theorem for Testing Reduced Models vs Full Models

Let $\mathbf{Y} = (Y_1, \dots, Y_n)$ be an i.i.d. sample from a regular family of distributions with parameters $\boldsymbol{\theta} \in \Theta \subseteq \mathbb{R}^{p_1}$. Let $\boldsymbol{\theta}_0$ denote the true parameter value. Consider the test:
$$H_0: \boldsymbol{\theta} \in \Theta_0 \quad \text{vs} \quad H_1: \boldsymbol{\theta} \in \Theta_1$$
where $\Theta_0 \subset \Theta_1$ are smooth parameter spaces with dimensions $p_0 < p_1$. Let $\hat{\boldsymbol{\theta}}^{(1)}$ be the unrestricted MLE under $H_1$, and $\hat{\boldsymbol{\theta}}^{(0)}$ be the restricted MLE under $H_0$.

The likelihood ratio statistic is $D = 2[\ell(\hat{\boldsymbol{\theta}}^{(1)}) - \ell(\hat{\boldsymbol{\theta}}^{(0)})]$. Under $H_0$, as $n \to \infty$:
$$D \xrightarrow{d} \chi^2_{k}$$
where $k = p_1 - p_0$ is the difference in dimensionality between the models.

:::

::: {.proof}

**Geometric Proof via Quadratic Approximation**

We rely on the geometric interpretation of the log-likelihood function as a squared distance metric induced by the Fisher Information matrix $\mathcal{I}(\boldsymbol{\theta}_0)$.

1.  **Quadratic Approximation of the Log-Likelihood**

    We expand the log-likelihood $\ell(\boldsymbol{\theta})$ around the unrestricted MLE $\hat{\boldsymbol{\theta}}^{(1)}$. Since $\nabla \ell(\hat{\boldsymbol{\theta}}^{(1)}) = \mathbf{0}$, the linear term vanishes. Asymptotically, the Hessian converges to $-n\mathcal{I}(\boldsymbol{\theta}_0)$:
    $$\ell(\boldsymbol{\theta}) \approx \ell(\hat{\boldsymbol{\theta}}^{(1)}) - \frac{n}{2} (\hat{\boldsymbol{\theta}}^{(1)} - \boldsymbol{\theta})^T \mathcal{I}(\boldsymbol{\theta}_0) (\hat{\boldsymbol{\theta}}^{(1)} - \boldsymbol{\theta})$$
    $$\ell(\boldsymbol{\theta}) \approx \ell(\hat{\boldsymbol{\theta}}^{(1)}) - \frac{n}{2} \|\hat{\boldsymbol{\theta}}^{(1)} - \boldsymbol{\theta}\|_{\mathcal{I}}^2$$

    This approximation replaces the likelihood surface with a parabola $Q(\boldsymbol{\theta})$.

2.  **Approximation of the Restricted Estimator (Projection)**

    The true restricted MLE $\hat{\boldsymbol{\theta}}^{(0)}$ maximizes the *exact* log-likelihood $\ell(\boldsymbol{\theta})$ over $\Theta_0$. However, for the quadratic approximation, we define the **Approximate Restricted MLE** $\tilde{\boldsymbol{\theta}}^{(0)}$ as the maximizer of the *parabolic* approximation $Q(\boldsymbol{\theta})$ over $\Theta_0$. This is equivalent to minimizing the distance to the unrestricted MLE:
    $$\tilde{\boldsymbol{\theta}}^{(0)} = \arg\min_{\boldsymbol{\theta} \in \Theta_0} \|\hat{\boldsymbol{\theta}}^{(1)} - \boldsymbol{\theta}\|_{\mathcal{I}}^2$$

    Geometrically, $\tilde{\boldsymbol{\theta}}^{(0)}$ is the **orthogonal projection** (as defined in @def-orthogonality) of $\hat{\boldsymbol{\theta}}^{(1)}$ onto the subspace $\Theta_0$ with respect to the metric $\mathcal{I}$.

    **Crucial Justification:** Why can we replace $\hat{\boldsymbol{\theta}}^{(0)}$ with $\tilde{\boldsymbol{\theta}}^{(0)}$?
    According to **Le Cam's Local Asymptotic Normality (LAN)** theory (Le Cam, 1986), the likelihood surface becomes locally quadratic and the parameter space locally linear. Consequently, the difference between the true restricted maximum and the quadratic projection vanishes at a rate faster than the estimation error:
    $$\|\hat{\boldsymbol{\theta}}^{(0)} - \tilde{\boldsymbol{\theta}}^{(0)}\| = o_p(n^{-1/2})$$
    Thus, for the purpose of the asymptotic distribution, we use $\tilde{\boldsymbol{\theta}}^{(0)}$.

3.  **Orthogonal Decomposition**

    Substituting the projection $\tilde{\boldsymbol{\theta}}^{(0)}$ into the test statistic approximation:
    $$D \approx n \|\hat{\boldsymbol{\theta}}^{(1)} - \tilde{\boldsymbol{\theta}}^{(0)}\|_{\mathcal{I}}^2$$
    
    Using the **Pythagorean Identity** from @def-orthogonality with $\boldsymbol{\theta}_0$ as the true parameter (which lies in $\Theta_0$ under the null hypothesis):
    $$n \|\hat{\boldsymbol{\theta}}^{(1)} - \tilde{\boldsymbol{\theta}}^{(0)}\|_{\mathcal{I}}^2 = n \|\hat{\boldsymbol{\theta}}^{(1)} - \boldsymbol{\theta}_0\|_{\mathcal{I}}^2 - n \|\tilde{\boldsymbol{\theta}}^{(0)} - \boldsymbol{\theta}_0\|_{\mathcal{I}}^2$$

4.  **Distributional Conclusion**

    Multiplying by $n$, we identify the terms with Chi-square variables:
    $$\underbrace{D}_{\text{Statistic}} \approx \underbrace{n\|\hat{\boldsymbol{\theta}}^{(1)} - \boldsymbol{\theta}_0\|_{\mathcal{I}}^2}_{\sim \chi^2_{p_1}} - \underbrace{n\|\tilde{\boldsymbol{\theta}}^{(0)} - \boldsymbol{\theta}_0\|_{\mathcal{I}}^2}_{\sim \chi^2_{p_0}}$$

    By Cochran's Theorem, the squared length of the projection of a standard normal vector onto the orthogonal complement of a subspace of dimension $p_0$ is distributed as a Chi-square with degrees of freedom equal to the codimension:
    $$D \xrightarrow{d} \chi^2_{p_1 - p_0}$$

**References:**
Le Cam, L. (1986). *Asymptotic Methods in Statistical Decision Theory*. Springer-Verlag.

:::

::: {#exm-lrt-normal}

**LRT for Normal Mean with Unknown Variance**

Let $X_1, \dots, X_n \overset{i.i.d.}{\sim} N(\mu, \sigma^2)$ where both $\mu$ and $\sigma^2$ are unknown. The parameter vector is $\theta = (\mu, \sigma^2)$. We wish to test:
$$H_0: \mu = \mu_0 \quad \text{vs} \quad H_1: \mu \neq \mu_0$$

1.  **Find the Maximum Likelihood Estimators:** The log-likelihood function is $\ell(\mu, \sigma^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n (x_i - \mu)^2$.

    * **Unrestricted MLEs ($\hat{\theta}$):** $\hat{\mu} = \bar{X}, \quad \hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \bar{X})^2$. 
      The maximized log-likelihood is:
      $$\ell(\hat{\mu}, \hat{\sigma}^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\hat{\sigma}^2) - \frac{n}{2}$$

    * **Restricted MLEs ($\hat{\theta}_0$):** Under $H_0$, $\hat{\mu}_0 = \mu_0, \quad \hat{\sigma}_0^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \mu_0)^2$.
      The restricted maximized log-likelihood is:
      $$\ell(\hat{\mu}_0, \hat{\sigma}_0^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\hat{\sigma}_0^2) - \frac{n}{2}$$

2.  **Define the Test Statistic $D$:** The test statistic $D$ is defined as twice the difference between the unrestricted and restricted log-likelihoods:
    $$D = 2 \left[ \ell(\hat{\mu}, \hat{\sigma}^2) - \ell(\hat{\mu}_0, \hat{\sigma}_0^2) \right]$$
    Substituting the expressions above, the constant terms cancel out:
    $$D = 2 \left[ -\frac{n}{2}\ln(\hat{\sigma}^2) - \left( -\frac{n}{2}\ln(\hat{\sigma}_0^2) \right) \right] = n \ln \left( \frac{\hat{\sigma}_0^2}{\hat{\sigma}^2} \right)$$

3.  **Relate to the t-statistic:** Using the sum of squares decomposition $\hat{\sigma}_0^2 = \hat{\sigma}^2 + (\bar{X} - \mu_0)^2$, we have:
    $$\frac{\hat{\sigma}_0^2}{\hat{\sigma}^2} = 1 + \frac{(\bar{X} - \mu_0)^2}{\hat{\sigma}^2}$$
    Substituting the t-statistic $T = \frac{\bar{X} - \mu_0}{S/\sqrt{n}}$, where $S^2 = \frac{n}{n-1}\hat{\sigma}^2$, we find:
    $$\frac{(\bar{X} - \mu_0)^2}{\hat{\sigma}^2} = \frac{T^2}{n-1}$$
    Thus, $D$ is a monotone increasing function of $T^2$:
    $$D = n \ln \left( 1 + \frac{T^2}{n-1} \right)$$
    Rejecting $H_0$ for large $D$ is equivalent to rejecting for large $|T|$.

4.  **Asymptotic Distribution (Wilks' Theorem):** By Wilks' Theorem, $D \xrightarrow{d} \chi^2_1$. For large $n$, using the approximation $\ln(1+x) \approx x$:
    $$D = n \ln \left( 1 + \frac{T^2}{n-1} \right) \approx n \left( \frac{T^2}{n-1} \right) \approx T^2$$
    
    As $n \to \infty$, $T^2 \xrightarrow{d} \chi^2_1$, confirming the theorem.

:::


### Rao's Score Theorem for Testing Restricted Models

The asymptotic distribution of the Score Test statistic relies on the behavior of the gradient of the log-likelihood (the Score vector) evaluated at the restricted Maximum Likelihood Estimator. This theorem provides a powerful method for hypothesis testing that only requires fitting the model under the null hypothesis.

::: {#thm-score}

### Asymptotic Distribution of the Score Statistic

Let $\mathbf{Y} = (Y_1, \dots, Y_n)$ be an i.i.d. sample from a regular family of distributions with parameters $\boldsymbol{\theta} \in \Theta \subseteq \mathbb{R}^{p}$. Partition the parameter vector as $\boldsymbol{\theta} = (\boldsymbol{\theta}_0^T, \boldsymbol{\theta}_1^T)^T$, where $\boldsymbol{\theta}_0$ is the $q$-dimensional parameter of interest and $\boldsymbol{\theta}_1$ is the $(p-q)$-dimensional nuisance parameter. Consider the test:
$$H_0: \boldsymbol{\theta}_0 = \boldsymbol{\theta}_0^* \quad \text{vs} \quad H_1: \boldsymbol{\theta}_0 \neq \boldsymbol{\theta}_0^*$$

Let $\hat{\boldsymbol{\theta}}_0 = (\boldsymbol{\theta}_0^*, \hat{\boldsymbol{\theta}}_1)$ be the restricted MLE under $H_0$. Let $\mathbf{U}(\boldsymbol{\theta})$ be the total Score vector partitioned as $\mathbf{U} = [\mathbf{U}_0^T, \mathbf{U}_1^T]^T$, and let $\mathcal{I}(\boldsymbol{\theta})$ be the expected Fisher Information matrix with $\mathcal{I}^{00}$ representing the top-left block of its inverse $\mathcal{I}^{-1}$.

The Score statistic is $S = \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)^T \mathcal{I}^{00}(\hat{\boldsymbol{\theta}}_0) \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)$. Under $H_0$, as $n \to \infty$:
$$S \xrightarrow{d} \chi^2_q$$
where $q$ is the dimension of the parameter of interest $\boldsymbol{\theta}_0$.

:::

::: {.proof}

**Algebraic Proof via Orthogonal Projection of the Score**

We rely on the first-order Taylor expansion to observe how estimating the nuisance parameter $\boldsymbol{\theta}_1$ impacts the sampling distribution of the score vector for the parameter of interest $\boldsymbol{\theta}_0$.

1.  **Partitioning the Information Matrix**
    The expected Fisher Information matrix and its inverse are partitioned conformably with $\boldsymbol{\theta} = (\boldsymbol{\theta}_0^T, \boldsymbol{\theta}_1^T)^T$:
    $$\mathcal{I} = \begin{pmatrix} \mathcal{I}_{00} & \mathcal{I}_{01} \\ \mathcal{I}_{10} & \mathcal{I}_{11} \end{pmatrix}, \quad \mathcal{I}^{-1} = \begin{pmatrix} \mathcal{I}^{00} & \mathcal{I}^{01} \\ \mathcal{I}^{10} & \mathcal{I}^{11} \end{pmatrix}$$

2.  **Estimation Error of the Nuisance Parameter**
    Let $\mathbf{U}_0$ and $\mathbf{U}_1$ denote the unobservable true score vectors evaluated at the true parameters $(\boldsymbol{\theta}_0^*, \boldsymbol{\theta}_1)$. By definition, the restricted MLE $\hat{\boldsymbol{\theta}}_1$ forces the nuisance score to zero: $\mathbf{U}_1(\boldsymbol{\theta}_0^*, \hat{\boldsymbol{\theta}}_1) = \mathbf{0}$. 
    
    Taking a first-order Taylor expansion of $\mathbf{U}_1(\boldsymbol{\theta}_0^*, \hat{\boldsymbol{\theta}}_1)$ around the true parameter $\boldsymbol{\theta}_1$:
    $$\mathbf{0} \approx \mathbf{U}_1 - \mathcal{I}_{11} (\hat{\boldsymbol{\theta}}_1 - \boldsymbol{\theta}_1)$$
    Rearranging gives the asymptotic estimation error:
    $$(\hat{\boldsymbol{\theta}}_1 - \boldsymbol{\theta}_1) \approx \mathcal{I}_{11}^{-1} \mathbf{U}_1$$

3.  **The Effective Score (Orthogonal Projection)**
    Next, we expand the evaluated score of interest $\mathbf{U}_0(\boldsymbol{\theta}_0^*, \hat{\boldsymbol{\theta}}_1)$ around the true parameter $\boldsymbol{\theta}_1$:
    $$\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0) \approx \mathbf{U}_0 - \mathcal{I}_{01} (\hat{\boldsymbol{\theta}}_1 - \boldsymbol{\theta}_1)$$
    
    Substituting the estimation error from Step 2:
    $$\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0) \approx \mathbf{U}_0 - \mathcal{I}_{01} \mathcal{I}_{11}^{-1} \mathbf{U}_1$$
    

    This quantity is the **Effective Score**. Geometrically, it represents the residual of $\mathbf{U}_0$ after projecting it orthogonally onto the nuisance score $\mathbf{U}_1$.

4.  **Asymptotic Variance of the Effective Score**
    We verify independence by checking the covariance between the effective score and the nuisance score:
    $$\text{Cov}(\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0), \mathbf{U}_1) \approx \text{Cov}(\mathbf{U}_0, \mathbf{U}_1) - \mathcal{I}_{01} \mathcal{I}_{11}^{-1} \text{Var}(\mathbf{U}_1) = \mathcal{I}_{01} - \mathcal{I}_{01} \mathcal{I}_{11}^{-1} \mathcal{I}_{11} = \mathbf{0}$$
    
    Because they are jointly normal with zero covariance, they are independent. The variance of this orthogonalized score is the Schur complement of $\mathcal{I}_{11}$:
    $$\text{Var}(\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)) = \mathcal{I}_{00} - \mathcal{I}_{01} \mathcal{I}_{11}^{-1} \mathcal{I}_{10}$$
    By the block matrix inversion theorem, this Schur complement is exactly the inverse of the top-left block of $\mathcal{I}^{-1}$:
    $$\text{Var}(\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)) = (\mathcal{I}^{00})^{-1}$$

5.  **Distributional Conclusion**
    Because the evaluated score $\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)$ is asymptotically normal with mean zero and variance $(\mathcal{I}^{00})^{-1}$, standardizing it by its variance creates a quadratic form that follows a Chi-square distribution:
    $$S = \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)^T \mathcal{I}^{00} \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0) \xrightarrow{d} \chi^2_q$$

:::

::: {#exm-score-normal}

**Score Test for Normal Mean with Unknown Variance**

Let $X_1, \dots, X_n \overset{i.i.d.}{\sim} N(\mu, \sigma^2)$ where both $\mu$ and $\sigma^2$ are unknown. The parameter vector is $\boldsymbol{\theta} = (\mu, \sigma^2)^T$. We wish to test:
$$H_0: \mu = \mu_0 \quad \text{vs} \quad H_1: \mu \neq \mu_0$$

1.  **Find the Restricted MLEs:** Under $H_0$, we fix $\mu = \mu_0$. Maximizing the restricted log-likelihood with respect to the nuisance parameter $\sigma^2$ yields:
    $$\hat{\sigma}_0^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \mu_0)^2$$
    The restricted parameter vector is $\hat{\boldsymbol{\theta}}_0 = (\mu_0, \hat{\sigma}_0^2)^T$.

2.  **Evaluate the Full Score Vector:** The full log-likelihood function for the normal sample is:
    $$\ell(\mu, \sigma^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n (X_i - \mu)^2$$
    
    The total Score vector $\mathbf{U}(\boldsymbol{\theta})$ is the gradient with respect to both parameters:
    $$\mathbf{U}(\mu, \sigma^2) = \begin{pmatrix} U_\mu(\mu, \sigma^2) \\ U_{\sigma^2}(\mu, \sigma^2) \end{pmatrix} = \begin{pmatrix} \frac{n(\bar{X} - \mu)}{\sigma^2} \\ -\frac{n}{2\sigma^2} + \frac{1}{2\sigma^4}\sum_{i=1}^n (X_i - \mu)^2 \end{pmatrix}$$
    
    Now we evaluate this full vector at the restricted MLE $\hat{\boldsymbol{\theta}}_0 = (\mu_0, \hat{\sigma}_0^2)^T$:
    $$\mathbf{U}(\hat{\boldsymbol{\theta}}_0) = \begin{pmatrix} \frac{n(\bar{X} - \mu_0)}{\hat{\sigma}_0^2} \\ -\frac{n}{2\hat{\sigma}_0^2} + \frac{1}{2\hat{\sigma}_0^4}\sum_{i=1}^n (X_i - \mu_0)^2 \end{pmatrix}$$
    
    Because $\hat{\sigma}_0^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \mu_0)^2$, the second component perfectly cancels out. By definition, the score for the nuisance parameter evaluated at the restricted MLE is zero:
    $$\mathbf{U}(\hat{\boldsymbol{\theta}}_0) = \begin{pmatrix} \frac{n(\bar{X} - \mu_0)}{\hat{\sigma}_0^2} \\ 0 \end{pmatrix}$$

3.  **Determine the Fisher Information:** The expected Fisher Information matrix for a Normal distribution is diagonal because the mean and variance are orthogonal parameters:
    $$\mathcal{I}(\mu, \sigma^2) = \begin{pmatrix} \frac{n}{\sigma^2} & 0 \\ 0 & \frac{n}{2\sigma^4} \end{pmatrix}$$
    Because the matrix is diagonal, the top-left block of the inverse $\mathcal{I}^{-1}$ is simply the reciprocal of the top-left block of $\mathcal{I}$. Evaluated at $\hat{\boldsymbol{\theta}}_0$:
    $$\mathcal{I}^{\mu\mu}(\hat{\boldsymbol{\theta}}_0) = \frac{\hat{\sigma}_0^2}{n}$$

4.  **Compute the Score Statistic $S$:** Constructing the quadratic form for the parameter of interest:
    $$S = U_\mu(\hat{\boldsymbol{\theta}}_0)^T \mathcal{I}^{\mu\mu}(\hat{\boldsymbol{\theta}}_0) U_\mu(\hat{\boldsymbol{\theta}}_0)$$
    $$S = \left( \frac{n(\bar{X} - \mu_0)}{\hat{\sigma}_0^2} \right) \left( \frac{\hat{\sigma}_0^2}{n} \right) \left( \frac{n(\bar{X} - \mu_0)}{\hat{\sigma}_0^2} \right) = \frac{n(\bar{X} - \mu_0)^2}{\hat{\sigma}_0^2}$$

5.  **Asymptotic Distribution:** By the Score Theorem, since we are testing $q=1$ parameter restriction, the statistic follows a $\chi^2_1$ distribution:
    $$S = \frac{n(\bar{X} - \mu_0)^2}{\hat{\sigma}_0^2} \xrightarrow{d} \chi^2_1$$
    This is highly analogous to the squared t-statistic, but specifically uses the variance estimated under the null hypothesis constraint.

:::

### Wald's Theorem for Testing Parameter Restrictions

The Wald Test evaluates the distance between the unrestricted Maximum Likelihood Estimator (MLE) and the null hypothesis value, standardized by the curvature of the log-likelihood at the unrestricted estimate. Unlike the Score test, the Wald test requires estimating the full, unrestricted model.

::: {#thm-wald}

### Asymptotic Distribution of the Wald Statistic

Let $\mathbf{Y} = (Y_1, \dots, Y_n)$ be an i.i.d. sample from a regular family of distributions with parameters $\boldsymbol{\theta} \in \Theta \subseteq \mathbb{R}^{p}$. Partition the parameter vector as $\boldsymbol{\theta} = (\boldsymbol{\theta}_0^T, \boldsymbol{\theta}_1^T)^T$, where $\boldsymbol{\theta}_0$ is the $q$-dimensional parameter of interest and $\boldsymbol{\theta}_1$ is the $(p-q)$-dimensional nuisance parameter. Consider the test:
$$H_0: \boldsymbol{\theta}_0 = \boldsymbol{\theta}_0^* \quad \text{vs} \quad H_1: \boldsymbol{\theta}_0 \neq \boldsymbol{\theta}_0^*$$

Let $\hat{\boldsymbol{\theta}} = (\hat{\boldsymbol{\theta}}_0^T, \hat{\boldsymbol{\theta}}_1^T)^T$ be the unrestricted MLE under $H_1$. Let $\mathcal{I}(\boldsymbol{\theta})$ be the expected Fisher Information matrix, and let $\mathcal{I}^{00}(\boldsymbol{\theta})$ denote the top-left $q \times q$ block of its inverse $\mathcal{I}^{-1}(\boldsymbol{\theta})$.

The Wald statistic is defined as the quadratic form:
$$W = (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)^T [\mathcal{I}^{00}(\hat{\boldsymbol{\theta}})]^{-1} (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)$$

Under standard regularity conditions, if $H_0$ is true, as $n \to \infty$:
$$W \xrightarrow{d} \chi^2_q$$
where $q$ is the dimension of the parameter of interest $\boldsymbol{\theta}_0$.

:::

::: {.proof}

**Algebraic Proof via Asymptotic Normality of the MLE**

The derivation of the Wald statistic relies directly on the asymptotic sampling distribution of the Maximum Likelihood Estimator.

1.  **Asymptotic Normality of the Full MLE**
    By standard large-sample theory, the unrestricted MLE $\hat{\boldsymbol{\theta}}$ is consistent and asymptotically normal. Its asymptotic variance-covariance matrix is the inverse of the Fisher Information matrix evaluated at the true parameter $\boldsymbol{\theta}$:
    $$\sqrt{n}(\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}) \xrightarrow{d} N(\mathbf{0}, [n^{-1}\mathcal{I}(\boldsymbol{\theta})]^{-1})$$
    Equivalently, for the unscaled estimator:
    $$\hat{\boldsymbol{\theta}} \overset{a}{\sim} N(\boldsymbol{\theta}, \mathcal{I}^{-1}(\boldsymbol{\theta}))$$

2.  **Marginal Distribution of the Parameter of Interest**
    We partition the inverse Fisher Information matrix $\mathcal{I}^{-1}$ conformably with $\boldsymbol{\theta} = (\boldsymbol{\theta}_0^T, \boldsymbol{\theta}_1^T)^T$. The marginal asymptotic distribution of the sub-vector $\hat{\boldsymbol{\theta}}_0$ depends only on the corresponding top-left block of the inverse matrix, $\mathcal{I}^{00}$:
    $$\hat{\boldsymbol{\theta}}_0 \overset{a}{\sim} N(\boldsymbol{\theta}_0, \mathcal{I}^{00}(\boldsymbol{\theta}))$$
    
    Under the null hypothesis $H_0: \boldsymbol{\theta}_0 = \boldsymbol{\theta}_0^*$, this becomes:
    $$\hat{\boldsymbol{\theta}}_0 \overset{a}{\sim} N(\boldsymbol{\theta}_0^*, \mathcal{I}^{00}(\boldsymbol{\theta}_0^*, \boldsymbol{\theta}_1))$$

3.  **Constructing the Quadratic Form**
    
    To create a test statistic, we measure the squared distance from the estimate $\hat{\boldsymbol{\theta}}_0$ to the hypothesized value $\boldsymbol{\theta}_0^*$, standardized by its variance. If a random vector $\mathbf{Z} \sim N(\boldsymbol{\mu}, \mathbf{\Sigma})$, then the quadratic form $(\mathbf{Z} - \boldsymbol{\mu})^T \mathbf{\Sigma}^{-1} (\mathbf{Z} - \boldsymbol{\mu})$ follows a $\chi^2$ distribution with degrees of freedom equal to the dimension of $\mathbf{Z}$.
    
    Applying this to our marginal distribution:
    $$(\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)^T [\mathcal{I}^{00}(\boldsymbol{\theta}_0^*, \boldsymbol{\theta}_1)]^{-1} (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*) \xrightarrow{d} \chi^2_q$$

4.  **Substituting the Consistent Estimator**
    In practice, the true parameters in the variance matrix are unknown. Because the Fisher Information matrix is a continuous function of the parameters, and $\hat{\boldsymbol{\theta}}$ is a consistent estimator of $\boldsymbol{\theta}$, we can invoke Slutsky's Theorem. We substitute the unknown true information with the expected information evaluated at the unrestricted MLE, $\mathcal{I}(\hat{\boldsymbol{\theta}})$, without changing the asymptotic distribution:
    $$W = (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)^T [\mathcal{I}^{00}(\hat{\boldsymbol{\theta}})]^{-1} (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*) \xrightarrow{d} \chi^2_q$$

:::

::: {#exm-wald-normal}

**Wald Test for Normal Mean with Unknown Variance**

Let $X_1, \dots, X_n \overset{i.i.d.}{\sim} N(\mu, \sigma^2)$ where both $\mu$ and $\sigma^2$ are unknown. The parameter vector is $\boldsymbol{\theta} = (\mu, \sigma^2)^T$. We wish to test:
$$H_0: \mu = \mu_0 \quad \text{vs} \quad H_1: \mu \neq \mu_0$$

1.  **Find the Unrestricted MLEs:** Maximizing the full log-likelihood with respect to both parameters yields the unrestricted estimators:
    $$\hat{\mu} = \bar{X}, \quad \hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \bar{X})^2$$
    Our unrestricted parameter estimate is $\hat{\boldsymbol{\theta}} = (\hat{\mu}, \hat{\sigma}^2)^T$.

2.  **Determine the Fisher Information at the MLE:** The expected Fisher Information matrix for a Normal distribution is diagonal:
    $$\mathcal{I}(\mu, \sigma^2) = \begin{pmatrix} \frac{n}{\sigma^2} & 0 \\ 0 & \frac{n}{2\sigma^4} \end{pmatrix}$$
    Because the matrix is diagonal, the top-left block of the inverse $\mathcal{I}^{-1}$ (the asymptotic variance of $\hat{\mu}$) is simply the reciprocal of the top-left block of $\mathcal{I}$. Evaluated at the unrestricted MLE $\hat{\boldsymbol{\theta}}$:
    $$\mathcal{I}^{\mu\mu}(\hat{\boldsymbol{\theta}}) = \frac{\hat{\sigma}^2}{n}$$

3.  **Compute the Wald Statistic $W$:** We construct the quadratic form using the parameter of interest ($\mu$), its null value ($\mu_0$), and its estimated variance:
    $$W = (\hat{\mu} - \mu_0)^T [\mathcal{I}^{\mu\mu}(\hat{\boldsymbol{\theta}})]^{-1} (\hat{\mu} - \mu_0)$$
    Substituting our values:
    $$W = (\bar{X} - \mu_0) \left( \frac{n}{\hat{\sigma}^2} \right) (\bar{X} - \mu_0) = \frac{n(\bar{X} - \mu_0)^2}{\hat{\sigma}^2}$$

4.  **Contrast with the Score Statistic:**
    By Wald's Theorem, this statistic follows a $\chi^2_1$ distribution asymptotically. Note the critical difference from the Score test derived previously. The Wald statistic uses the unrestricted variance estimator $\hat{\sigma}^2$, whereas the Score statistic used the restricted variance estimator $\hat{\sigma}_0^2 = \frac{1}{n}\sum (X_i - \mu_0)^2$. Because $\hat{\sigma}_0^2 \ge \hat{\sigma}^2$, it guarantees that mathematically $W \ge S$ for this model.

:::

### A Comparison of Finite-Sample Distributions of the Three Asymptotic Tests in OLS

#### Exact Finite-Sample Distributions via the F-Statistic

When testing linear restrictions in Ordinary Least Squares (OLS) under the assumption of normal errors, we do not need to rely on asymptotic approximations. The true, exact finite-sample test statistic follows an $F$-distribution. 

Because the Wald ($W$), Likelihood Ratio ($LR$), and Score ($S$) statistics are strictly monotonic algebraic transformations of the exact finite-sample $F$-statistic, we can use the transformation of variables to derive their exact finite-sample Cumulative Distribution Functions (CDFs).

Let us consider testing a linear hypothesis involving $q$ restrictions in an OLS model with $k$ total parameters and $n$ observations. Let $F$ denote the exact finite-sample $F$-statistic, which follows an $F_{q, n-k}$ distribution. Let $G_{q, n-k}(x)$ denote the CDF of this $F$-distribution.

By isolating $F$ in the algebraic definitions of $W$, $LR$, and $S$, we derive their exact finite-sample CDFs:

1.  **Wald Statistic ($W$):**
    $$W = \frac{nq}{n-k} F \implies F = \frac{n-k}{nq} W$$
    $$F_W(w) = P(W \le w) = G_{q, n-k}\left( \frac{n-k}{nq} w \right)$$

2.  **Likelihood Ratio Statistic ($LR$):**
    $$LR = n \ln \left( 1 + \frac{q}{n-k} F \right) \implies F = \frac{n-k}{q} (e^{LR/n} - 1)$$
    $$F_{LR}(l) = P(LR \le l) = G_{q, n-k}\left( \frac{n-k}{q} (e^{l/n} - 1) \right)$$

3.  **Score Statistic ($S$):**
    $$S = \frac{n q F}{n-k + q F} \implies F = \frac{S(n-k)}{q(n - S)}$$
    Because the Score statistic is bounded ($S < n$), its exact CDF for $s < n$ is:
    $$F_S(s) = P(S \le s) = G_{q, n-k}\left( \frac{s(n-k)}{q(n - s)} \right)$$
    For $s \ge n$, $F_S(s) = 1$.



#### Visualizing Finite-Sample Calibration (Size Distortion)

To visualize how badly the asymptotic $\chi^2_q$ approximation distorts the Type I error rates in finite samples, we can plot the true Cumulative Distribution Function (CDF) of the asymptotic p-values. 

If a test is perfectly calibrated, the probability that its p-value is less than a nominal threshold $\alpha$ should be exactly $\alpha$, forming a perfect $y = x$ diagonal line (the Uniform(0,1) CDF). 

The event that the asymptotic p-value is less than $\alpha$ is mathematically equivalent to the test statistic being greater than the $\chi^2_q$ critical value ($c_\alpha$). Therefore, the true rejection probability $P(p < \alpha)$ is exactly $1 - F(c_\alpha)$, where $F$ is the exact finite-sample CDF derived above.

#### Exact Finite-Sample Distribution Calibration Plot

The following code allows you to adjust the sample size ($n$) and the number of parameters ($k$) to observe how the finite-sample rejection rates of the three asymptotic tests (Wald, Likelihood Ratio, and Score) converge toward the ideal Uniform(0,1) calibration.


```{r}
#| label: fig-calibration-tuning
#| fig-cap: "Exact Analytical CDFs of Asymptotic P-values"
#| warning: false
#| fig-width: 12
#| fig-height: 6

library(ggplot2)
library(dplyr)

# --- Tuning Parameters ---
n1 <- 15    # Small sample size for Plot 1
n2 <- 50  # Large sample size for Plot 2
k_val <- 5  # Number of parameters (constant for comparison)
q_val <- 1  # Number of restrictions (constant for comparison)
# -------------------------

generate_calibration_data <- function(n, k, q) {
  alpha_seq <- seq(0.0001, 0.15, length.out = 1000)
  c_alpha <- qchisq(1 - alpha_seq, df = q)
  
  # Wald: 1 - F_W(c_alpha)
  F_val_W <- ((n - k) / (n * q)) * c_alpha
  cdf_W <- 1 - pf(F_val_W, df1 = q, df2 = n - k)
  
  # LR: 1 - F_LR(c_alpha)
  F_val_LR <- ((n - k) / q) * (exp(c_alpha / n) - 1)
  cdf_LR <- 1 - pf(F_val_LR, df1 = q, df2 = n - k)
  
  # Score: 1 - F_S(c_alpha)
  F_val_S <- ifelse(c_alpha < n, (c_alpha * (n - k)) / (q * (n - c_alpha)), Inf)
  cdf_S <- 1 - pf(F_val_S, df1 = q, df2 = n - k)
  
  data.frame(
    Alpha = rep(alpha_seq, 3),
    Rejection_Prob = c(cdf_W, cdf_LR, cdf_S),
    Test = factor(rep(c("Wald", "LR", "Score"), each = length(alpha_seq))),
    n = n,
    k = k
  ) |>
    mutate(Label = paste0("n = ", n, ", k = ", k))
}

# Create combined dataset based on tuning parameters
plot_data <- rbind(
  generate_calibration_data(n = n1, k = k_val, q = q_val),
  generate_calibration_data(n = n2, k = k_val, q = q_val)
)

# Fix factor levels for side-by-side ordering
plot_data$Label <- factor(plot_data$Label, levels = unique(plot_data$Label))

ggplot(plot_data, aes(x = Alpha, y = Rejection_Prob, color = Test)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~ Label) +
  scale_color_manual(values = c("Wald" = "#E41A1C", "LR" = "#4DAF4A", "Score" = "#377EB8")) +
  coord_cartesian(xlim = c(0, 0.12), ylim = c(0, 0.20)) +
  labs(
    x = "Nominal Significance Level (\u03b1)",
    y = "True Rejection Rate: P(p-value < \u03b1)",
    title = "Asymptotic Test Calibration vs. Sample Size",
    subtitle = "Dashed line represents perfect calibration (Rejection Rate = \u03b1)"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )
```

**Summary**

In OLS with finite samples, the $\chi^2$ distribution is an optimistic approximation that leads to **over-rejection** (liberal bias) across all three tests. However:

* The **Wald test** is the most liberal because it does nothing to counteract the too-low $\chi^2$ threshold.
* The **Score test** is the most conservative of the three because its mathematical structure provides a natural downward "correction" that happens to land it closest to the ideal uniform distribution at common significance levels like $\alpha = 0.05$.