
---
title: "Likelihood-based Tests"
engine: knitr
format: 
  html: default
  pdf: default
---



## The Orthogonal Decomposition of Mahalanobis Distance

::: {#thm-orthogonal-decomposition}
### The Orthogonal Decomposition of Mahalanobis Distance

**Definitions and Setup**

Let $x$ be a random vector partitioned into two sub-vectors, $x_0$ and $x_1$, such that $x = (x_0, x_1)^T$.
Let $\mu$ be the corresponding center (mean) vector, partitioned as $\mu = (\mu_0, \mu_1)^T$.
Let $\Sigma$ be a symmetric, positive-definite covariance matrix, partitioned into corresponding blocks:

$$
\Sigma = \begin{pmatrix} \Sigma_{00} & \Sigma_{01} \\ \Sigma_{10} & \Sigma_{11} \end{pmatrix}
$$

Define the squared Mahalanobis distance of any vector $v$ to center $c$ with respect to a positive-definite matrix $S$ as:

$$
D^2(v, c, S) = (v - c)^T S^{-1} (v - c)
$$

Define the conditional expectation (the linear regression surface) of $x_1$ given $x_0$ as $\mu_{1|0}(x_0)$:

$$
\mu_{1|0}(x_0) = \mu_1 + \Sigma_{10}\Sigma_{00}^{-1}(x_0 - \mu_0)
$$

Define the conditional covariance matrix (the Schur complement of $\Sigma_{00}$ in $\Sigma$) as $\Sigma_{1|0}$:

$$
\Sigma_{1|0} = \Sigma_{11} - \Sigma_{10}\Sigma_{00}^{-1}\Sigma_{01}
$$

**Theorem Statement**

For any point $x = (x_0, x_1)$, the total joint squared Mahalanobis distance from $x$ to $\mu$ can be orthogonally decomposed into the marginal distance of $x_0$ plus the conditional distance of $x_1$ given $x_0$:

$$
D^2(x, \mu, \Sigma) = D^2(x_0, \mu_0, \Sigma_{00}) + D^2(x_1, \mu_{1|0}(x_0), \Sigma_{1|0})
$$
:::

::: {#cor-regression-surface}
### The Regression Surface Identity

If we evaluate the joint Mahalanobis distance at the specific point where $x_1$ lies exactly on the conditional expectation surface—meaning the evaluated point is $(x_0, \mu_{1|0}(x_0))$—the total distance simplifies to exactly the marginal distance of $x_0$:

$$
D^2((x_0, \mu_{1|0}(x_0)), \mu, \Sigma) = D^2(x_0, \mu_0, \Sigma_{00})
$$
:::

::: {.proof}
**Algebraic Proof**

By substituting the point $x = (x_0, \mu_{1|0}(x_0))$ directly into the main theorem decomposition, we get:

$$
D^2((x_0, \mu_{1|0}(x_0)), \mu, \Sigma) = D^2(x_0, \mu_0, \Sigma_{00}) + D^2(\mu_{1|0}(x_0), \mu_{1|0}(x_0), \Sigma_{1|0})
$$

Because the distance from any point to itself is strictly zero, the second term vanishes:

$$
D^2(\mu_{1|0}(x_0), \mu_{1|0}(x_0), \Sigma_{1|0}) = (\mu_{1|0}(x_0) - \mu_{1|0}(x_0))^T \Sigma_{1|0}^{-1} (\mu_{1|0}(x_0) - \mu_{1|0}(x_0)) = 0
$$

Leaving only the marginal distance:

$$
D^2((x_0, \mu_{1|0}(x_0)), \mu, \Sigma) = D^2(x_0, \mu_0, \Sigma_{00})
$$
:::

**Geometric Explanation**

In the Mahalanobis geometry, the total squared distance of a point is the sum of its squared "horizontal" position (its marginal distance) and its squared "vertical" drop to the regression line (its conditional residual). 

When you define a point as $(x_0, \mu_{1|0}(x_0))$, you are placing it strictly on the regression line. 

1. It has no residual.
   This means the point lies exactly on the expected vector trajectory, leaving no orthogonal deviation.
2. It does not deviate from the expected value of $x_1$ for that given $x_0$.
   Consequently, no additional distance penalty is incurred in the conditional space.
3. Because the residual is exactly $0$, the orthogonal conditional vector is the zero vector.
   This cleanly zeroes out the conditional quadratic form entirely.

Therefore, 100% of the joint spatial distance between $(x_0, \mu_{1|0}(x_0))$ and the global mean $(\mu_0, \mu_1)$ is driven solely by how far $x_0$ is from $\mu_0$ in the marginal space.

```{shinylive-r}
#| standalone: true
#| viewerHeight: 750
#| echo: false
#| column: screen
#| label: fig-MD-decomp

library(shiny)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Geometry of Mahalanobis Distance Decomposition"),
  
  # A relative container holding the plot and absolute panels
  div(style = "position: relative;",
      
      plotOutput("distPlot", height = "700px", width = "100%"),
      
      # Floating Control Panel (Top Left)
      absolutePanel(
        top = 10, left = 10, width = 320, draggable = TRUE,
        style = "background-color: rgba(255, 255, 255, 0.85); padding: 10px 15px; border-radius: 8px; border: 1px solid #ccc; box-shadow: 2px 2px 10px rgba(0,0,0,0.1);",
        
        h4("Parameters", style = "margin-top: 0; margin-bottom: 10px; font-size: 16px;"),
        
        fluidRow(
          column(6, sliderInput("mu_0", "\\(\\mu_0\\)", min = 0, max = 5, value = 2.2, step = 0.1, ticks = FALSE)),
          column(6, sliderInput("mu_1", "\\(\\mu_1\\)", min = 0, max = 5, value = 2.2, step = 0.1, ticks = FALSE))
        ),
        fluidRow(
          column(6, sliderInput("x_0", "\\(x_0\\)", min = 0, max = 5, value = 0.9, step = 0.1, ticks = FALSE)),
          column(6, sliderInput("x_1", "\\(x_1\\)", min = 0, max = 5, value = 1.9, step = 0.1, ticks = FALSE))
        ),
        hr(style = "margin: 5px 0; border-top: 1px solid #ddd;"),
        fluidRow(
          column(6, sliderInput("sigma_0", "\\(\\sigma_0\\)", min = 0.5, max = 3, value = 1.25, step = 0.05, ticks = FALSE)),
          column(6, sliderInput("sigma_1", "\\(\\sigma_1\\)", min = 0.5, max = 3, value = 1.25, step = 0.05, ticks = FALSE))
        ),
        sliderInput("rho", "Correlation \\(\\rho\\)", min = -0.95, max = 0.95, value = -0.6, step = 0.05, ticks = FALSE)
      ),
      
      # Floating Legend Panel (Bottom Right)
      absolutePanel(
        bottom = 20, right = 20, width = 320, draggable = TRUE,
        style = "background-color: rgba(255, 255, 255, 0.9); padding: 10px; border-radius: 8px; border: 1px solid #ccc; box-shadow: 2px 2px 10px rgba(0,0,0,0.1);",
        uiOutput("legendUI")
      )
  )
)

server <- function(input, output, session) {
  
  # Reactive calculations
  calc_data <- reactive({
    mu <- c(input$mu_0, input$mu_1)
    x <- c(input$x_0, input$x_1)
    
    # Construct Covariance Matrix
    Sigma <- matrix(c(
      input$sigma_0^2, 
      input$rho * input$sigma_0 * input$sigma_1, 
      input$rho * input$sigma_0 * input$sigma_1, 
      input$sigma_1^2
    ), nrow = 2)
    
    Sigma_inv <- solve(Sigma)
    
    # Marginal and Conditional variances
    Sigma_00 <- Sigma[1, 1]
    Sigma_1given0 <- Sigma[2, 2] - (Sigma[1, 2]^2) / Sigma[1, 1]
    
    # Conditional Mean
    mu_1given0 <- mu[2] + (Sigma[1, 2] / Sigma[1, 1]) * (x[1] - mu[1])
    
    # Projected Point p = (x_0, mu_{1|0}(x_0))
    p <- c(x[1], mu_1given0)
    
    # Calculate Distances
    diff_total <- x - mu
    d2_total <- as.numeric(t(diff_total) %*% Sigma_inv %*% diff_total)
    d2_marg <- (x[1] - mu[1])^2 / Sigma_00
    d2_cond <- (x[2] - mu_1given0)^2 / Sigma_1given0
    
    list(mu = mu, x = x, p = p, Sigma = Sigma, Sigma_inv = Sigma_inv, 
         d2_total = d2_total, d2_marg = d2_marg, d2_cond = d2_cond)
  })
  
  output$legendUI <- renderUI({
    res <- calc_data()
    withMathJax(
      HTML(paste0(
        "<div style='font-size: 14px;'>",
        "<b>Distance Legend</b><br><br>",
        "<span style='color: black;'><b>Total:</b> \\( D^2(x, \\mu, \\Sigma) = \\) ", round(res$d2_total, 2), "</span><br>",
        "<span style='color: blue;'><b>Marginal:</b> \\( D^2(x_0, \\mu_0, \\Sigma_{00}) = \\) ", round(res$d2_marg, 2), "</span><br>",
        "<span style='color: red;'><b>Conditional:</b> \\( D^2(x_1, \\mu_{1|0}(x_0), \\Sigma_{1|0}) = \\) ", round(res$d2_cond, 2), "</span>",
        "</div>"
      ))
    )
  })
  
  output$distPlot <- renderPlot({
    res <- calc_data()
    mu <- res$mu; x <- res$x; p <- res$p; Sigma <- res$Sigma; Sigma_inv <- res$Sigma_inv
    
    # Generate grid for quadratic contours
    grid_vals <- seq(-1, 6, length.out = 80)
    grid <- expand.grid(x0 = grid_vals, x1 = grid_vals)
    grid$z <- apply(grid, 1, function(row) {
      diff <- row - mu
      as.numeric(t(diff) %*% Sigma_inv %*% diff)
    })
    
    # Regression surface line parameters
    slope <- Sigma[1, 2] / Sigma[1, 1]
    intercept <- mu[2] - slope * mu[1]
    
    # Midpoints for labels
    mid_marg <- (mu + p) / 2
    mid_cond <- (p + x) / 2
    mid_total <- (mu + x) / 2
    
    ggplot(grid, aes(x = x0, y = x1)) +
      # Contours
      geom_contour(aes(z = z), bins = 20, color = "gray85") +
      
      # The Regression Surface (Light Blue Line extending across plot)
      geom_abline(intercept = intercept, slope = slope, color = "lightblue", linewidth = 2, alpha = 0.5) +
      
      # The Distance Segments
      geom_segment(aes(x = mu[1], y = mu[2], xend = p[1], yend = p[2]), color = "blue", linewidth = 1.2) +
      geom_segment(aes(x = p[1], y = p[2], xend = x[1], yend = x[2]), color = "red", linewidth = 1.2) +
      geom_segment(aes(x = mu[1], y = mu[2], xend = x[1], yend = x[2]), color = "black", linetype = "dashed", linewidth = 1) +
      
      # Labels on Segments (No numbers, just notation)
      annotate("text", x = mid_marg[1] + 0.2, y = mid_marg[2] - 0.2, label = "D^2*(list(x[0], mu[0], Sigma[00]))", parse = TRUE, color = "blue", size = 5) +
      annotate("text", x = mid_cond[1] + 0.25, y = mid_cond[2], label = "D^2*(list(x[1], mu['1|0'](x[0]), Sigma['1|0']))", parse = TRUE, color = "red", size = 5) +
      annotate("text", x = mid_total[1] - 0.2, y = mid_total[2] + 0.2, label = "D^2*(list(x, mu, Sigma))", parse = TRUE, color = "black", size = 5) +
      
      # Points
      annotate("point", x = mu[1], y = mu[2], color = "blue", size = 4) +
      annotate("text", x = mu[1] - 0.2, y = mu[2] + 0.2, label = "mu", parse = TRUE, color = "blue", size = 6) +
      
      annotate("point", x = p[1], y = p[2], color = "purple", shape = 3, size = 4, stroke = 2) +
      annotate("text", x = p[1] + 0.5, y = p[2] - 0.2, label = "list(x[0], mu['1|0'](x[0]))", parse = TRUE, color = "purple", size = 5) +
      
      annotate("point", x = x[1], y = x[2], color = "black", shape = 4, size = 4, stroke = 2) +
      annotate("text", x = x[1] - 0.15, y = x[2] + 0.2, label = "x", parse = TRUE, size = 6) +
      
      coord_fixed(xlim = c(-0.5, 5.5), ylim = c(-0.5, 5.5)) +
      labs(x = expression(x[0]), y = expression(x[1])) +
      theme_minimal(base_size = 16) +
      theme(panel.grid.minor = element_blank())
  })
}

shinyApp(ui, server)
```

## Wilks' Theorem for General Nested Models (Likelihood Ratio Test)

::: {#thm-wilks-general}
### Wilks' Theorem for General Nested Models

Let $\mathbf{Y} = (Y_1, \dots, Y_n)$ be an i.i.d. sample from a distribution $f(\mathbf{y}; \boldsymbol{\theta})$ with $\boldsymbol{\theta} \in \Theta \subseteq \mathbb{R}^p$. Let $\Theta_0$ be a lower-dimensional subspace of $\Theta$ defined by $q$ independent constraints, such that $\dim(\Theta_0) = p - q$. 

Consider testing the nested hypotheses:

$$
H_0: \boldsymbol{\theta} \in \Theta_0 \quad \text{vs} \quad H_1: \boldsymbol{\theta} \in \Theta \setminus \Theta_0
$$

Let $\hat{\boldsymbol{\theta}}$ be the unrestricted Maximum Likelihood Estimator (MLE) over $\Theta$, and let $\tilde{\boldsymbol{\theta}}$ be the restricted MLE over $\Theta_0$. The likelihood ratio test statistic is defined as:

$$
D = 2 \left[ \ell(\hat{\boldsymbol{\theta}}) - \ell(\tilde{\boldsymbol{\theta}}) \right]
$$

Under $H_0$, as $n \to \infty$, the statistic $D$ converges in distribution to a chi-square distribution with degrees of freedom equal to the number of constraints:

$$
D \xrightarrow{d} \chi^2_q
$$
:::

::: {.proof}

1. **Quadratic Taylor Expansion at the Peak** \n
   Under the null hypothesis $H_0$, we perform a second-order Taylor expansion of the log-likelihood function $\ell(\boldsymbol{\theta})$ around the unrestricted MLE $\hat{\boldsymbol{\theta}}$. For sufficiently large $n$, the expansion is given by:
   
   $$
   \ell(\boldsymbol{\theta}) \approx \ell(\hat{\boldsymbol{\theta}}) - \frac{1}{2} (\boldsymbol{\theta} - \hat{\boldsymbol{\theta}})^T \mathcal{I} (\boldsymbol{\theta} - \hat{\boldsymbol{\theta}})
   $$
   
   where $\mathcal{I} = \Sigma^{-1}$ represents the Fisher Information matrix (the asymptotic precision). Using our distance notation $D^2(v, c, S)$, this quadratic approximation is:
   
   $$
   2[\ell(\hat{\boldsymbol{\theta}}) - \ell(\boldsymbol{\theta})] \approx D^2(\boldsymbol{\theta}, \hat{\boldsymbol{\theta}}, \Sigma)
   $$

2. **Evaluation at the Restricted MLE** \n
   The Wilks' statistic $D$ measures the likelihood ratio between the unrestricted peak $\hat{\boldsymbol{\theta}}$ and the restricted peak $\tilde{\boldsymbol{\theta}}$. By plugging $\tilde{\boldsymbol{\theta}}$ into the quadratic approximation above, we obtain:
   
   $$
   D = 2[\ell(\hat{\boldsymbol{\theta}}) - \ell(\tilde{\boldsymbol{\theta}})] \approx D^2(\tilde{\boldsymbol{\theta}}, \hat{\boldsymbol{\theta}}, \Sigma)
   $$

3. **Applying the Mahalanobis Decomposition** \n
   We apply the orthogonal decomposition theorem to partition this joint distance into a marginal component for the parameters of interest ($\boldsymbol{\theta}_0$) and a conditional component for the nuisance parameters ($\boldsymbol{\theta}_1$):
   
   $$
   D^2(\tilde{\boldsymbol{\theta}}, \hat{\boldsymbol{\theta}}, \Sigma) = D^2(\tilde{\boldsymbol{\theta}}_0, \hat{\boldsymbol{\theta}}_0, \Sigma_{00}) + D^2(\tilde{\boldsymbol{\theta}}_1, \mu_{1|0}(\tilde{\boldsymbol{\theta}}_0), \Sigma_{1|0})
   $$

4. **Vanishing of the Nuisance Term** \n
   By the definition of the restricted MLE $\tilde{\boldsymbol{\theta}}$ under the null hypothesis $H_0: \boldsymbol{\theta}_0 = \boldsymbol{\theta}_0^*$:
   
   - The parameters of interest are fixed: $\tilde{\boldsymbol{\theta}}_0 = \boldsymbol{\theta}_0^*$.
   - The restricted nuisance parameters $\tilde{\boldsymbol{\theta}}_1$ are chosen to maximize the restricted likelihood, which is equivalent to minimizing the joint distance $D^2(\tilde{\boldsymbol{\theta}}, \hat{\boldsymbol{\theta}}, \Sigma)$.
   
   As established by the properties of the regression surface, this minimum occurs exactly when $\tilde{\boldsymbol{\theta}}_1 = \mu_{1|0}(\boldsymbol{\theta}_0^*)$. At this point, the conditional term $D^2(\mu_{1|0}(\boldsymbol{\theta}_0^*), \mu_{1|0}(\boldsymbol{\theta}_0^*), \Sigma_{1|0})$ is exactly zero.

5. **Final Marginal Identity and Distribution** \n
   The Wilks' statistic $D$ reduces entirely to the marginal Mahalanobis distance between the estimated and hypothesized parameters of interest:
   
   $$
   D \approx D^2(\boldsymbol{\theta}_0^*, \hat{\boldsymbol{\theta}}_0, \Sigma_{00}) = (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)^T \Sigma_{00}^{-1} (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)
   $$
   
   Under $H_0$, the asymptotic normality of the MLE implies $\sqrt{n}(\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*) \xrightarrow{d} N(\mathbf{0}, \Sigma_{00})$. Therefore, the quadratic form $D$ follows a chi-square distribution with $q$ degrees of freedom:
   
   $$
   D \xrightarrow{d} \chi^2_q
   $$
:::

::: {.remark}
### Geometric Intuition
This proof illustrates that the restricted MLE $\tilde{\boldsymbol{\theta}}$ "slides" along the conditional regression surface $\mu_{1|0}(\boldsymbol{\theta}_0)$ until the conditional component of the discrepancy vanishes. This optimization process effectively "filters out" the noise from the $p-q$ nuisance parameters, leaving only the $q$-dimensional marginal evidence for the hypothesis test.
:::

**Diagram to Illustrate Wilks' Theorem**

The following R code visualizes the likelihood ratio test as a projection. We illustrate the unrestricted MLE $\hat{\boldsymbol{\theta}}$ (the center of the contours), the hypothesized true value $\boldsymbol{\theta}^*$, the geometric center $\hat{\boldsymbol{\theta}}_{1, \Sigma}$, and a jittered point representing the actual restricted MLE $\tilde{\boldsymbol{\theta}}_1$.

```{r}
#| label: fig-lrt-geometry
#| fig-cap: "The geometry of restricted optimization. Blue dot is the unrestricted MLE (center). The red dot represents the projected nuisance parameter center on the null line. The purple cross is the restricted MLE."
library(ggplot2)

# 1. Setup parameters and Fisher Information
theta_hat <- c(2.2, 2.2)      # The Peak (Unrestricted MLE center)
theta_star <- c(0.9, 1.9)     # Hypothesized true value 
I <- matrix(c(1, 0.6, 0.6, 1), nrow = 2) # Curvature

# 2. Calculate the geometric center hat(theta)_{1, Sigma}

# Starting from the center hat(theta)_1, we adjust by the displacement in theta_0
delta_theta_0 <- theta_star[1] - theta_hat[1]
theta_hat_1_Sigma <- theta_hat[2] - (1/I[2,2]) * I[2,1] * delta_theta_0
p_projected <- c(theta_star[1], theta_hat_1_Sigma)

# 3. Create tilde(theta)_1 by adding small non-quadratic jitter to the projected center
set.seed(42)
theta_tilde <- c(theta_star[1], theta_hat_1_Sigma + 0.15) 

# 4. Generate grid for quadratic log-likelihood contours centered at theta_hat
grid_vals <- seq(0, 4, length.out = 100)
grid <- expand.grid(theta0 = grid_vals, theta1 = grid_vals)
grid$z <- apply(grid, 1, function(p) {
  diff <- p - theta_hat
  -0.5 * t(diff) %*% I %*% diff
})

# 5. Plot
ggplot(grid, aes(x = theta0, y = theta1)) +
  geom_contour(aes(z = z), bins = 15, color = "gray70") +
  geom_vline(xintercept = theta_star[1], linetype = "dashed", color = "red") +
  # Unrestricted MLE (The Center)
  annotate("point", x = theta_hat[1], y = theta_hat[2], color = "blue", size = 3) +
  annotate("text", x = theta_hat[1], y = theta_hat[2] + 0.2, 
           label = "hat(theta)", parse = TRUE, color = "blue") +
  # Projected Nuisance Center (hat(theta)_{1, Sigma})
  annotate("point", x = p_projected[1], y = p_projected[2], color = "red", size = 3) +
  annotate("text", x = p_projected[1] - 0.45, y = p_projected[2], 
           label = "hat(theta)[1, Sigma]", parse = TRUE, color = "red") +
  # Restricted MLE (tilde(theta))
  annotate("point", x = theta_tilde[1], y = theta_tilde[2], color = "purple", shape = 3, size = 3) +
  annotate("text", x = theta_tilde[1] + 0.3, y = theta_tilde[2] + 0.1, 
           label = "tilde(theta)", parse = TRUE, color = "purple") +
  # Hypothesized Point (theta^*)
  annotate("point", x = theta_star[1], y = theta_star[2], color = "black", shape = 4, size = 3) +
  annotate("text", x = theta_star[1] - 0.2, y = theta_star[2] - 0.2, 
           label = "theta^'*'", parse = TRUE) +
  labs(title = "Geometry of the Likelihood Ratio Test",
       subtitle = expression(paste("Projection from center ", hat(theta), " onto the null constraint ", theta[0] == theta[0]^"*")),
       x = expression(theta[0]), y = expression(theta[1])) +
  theme_minimal() + 
  coord_fixed()

```



::: {#exm-lrt-normal}

**LRT for Normal Mean with Unknown Variance**

Let $X_1, \dots, X_n \overset{i.i.d.}{\sim} N(\mu, \sigma^2)$ where both $\mu$ and $\sigma^2$ are unknown. The parameter vector is $\theta = (\mu, \sigma^2)$. We wish to test:
$$
H_0: \mu = \mu_0 \quad \text{vs} \quad H_1: \mu \neq \mu_0
$$

1.  **Find the Maximum Likelihood Estimators:** The log-likelihood function is $\ell(\mu, \sigma^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n (x_i - \mu)^2$.

    * **Unrestricted MLEs ($\hat{\theta}$):** $\hat{\mu} = \overline{X}, \quad \hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \overline{X})^2$. 
      The maximized log-likelihood is:
$$
\ell(\hat{\mu}, \hat{\sigma}^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\hat{\sigma}^2) - \frac{n}{2}
$$

    * **Restricted MLEs ($\hat{\theta}_0$):** Under $H_0$, $\hat{\mu}_0 = \mu_0, \quad \hat{\sigma}_0^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \mu_0)^2$.
      The restricted maximized log-likelihood is:
$$
\ell(\hat{\mu}_0, \hat{\sigma}_0^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\hat{\sigma}_0^2) - \frac{n}{2}
$$

2.  **Define the Test Statistic $D$:** The test statistic $D$ is defined as twice the difference between the unrestricted and restricted log-likelihoods:

$$
D = 2 \left[ \ell(\hat{\mu}, \hat{\sigma}^2) - \ell(\hat{\mu}_0, \hat{\sigma}_0^2) \right]
$$
Substituting the expressions above, the constant terms cancel out:
$$
D = 2 \left[ -\frac{n}{2}\ln(\hat{\sigma}^2) - \left( -\frac{n}{2}\ln(\hat{\sigma}_0^2) \right) \right] = n \ln \left( \frac{\hat{\sigma}_0^2}{\hat{\sigma}^2} \right)
$$

3.  **Relate to the t-statistic:** Using the sum of squares decomposition $\hat{\sigma}_0^2 = \hat{\sigma}^2 + (\overline{X} - \mu_0)^2$, we have:

$$
\frac{\hat{\sigma}_0^2}{\hat{\sigma}^2} = 1 + \frac{(\overline{X} - \mu_0)^2}{\hat{\sigma}^2}
$$
Substituting the t-statistic $T = \frac{\overline{X} - \mu_0}{S/\sqrt{n}}$, where $S^2 = \frac{n}{n-1}\hat{\sigma}^2$, we find:
$$
\frac{(\overline{X} - \mu_0)^2}{\hat{\sigma}^2} = \frac{T^2}{n-1}
$$
Thus, $D$ is a monotone increasing function of $T^2$:
$$
D = n \ln \left( 1 + \frac{T^2}{n-1} \right)
$$
Rejecting $H_0$ for large $D$ is equivalent to rejecting for large $|T|$.

4.  **Asymptotic Distribution (Wilks' Theorem):** By Wilks' Theorem, $D \xrightarrow{d} \chi^2_1$. For large $n$, using the approximation $\ln(1+x) \approx x$:

$$
D = n \ln \left( 1 + \frac{T^2}{n-1} \right) \approx n \left( \frac{T^2}{n-1} \right) \approx T^2
$$
    
As $n \to \infty$, $T^2 \xrightarrow{d} \chi^2_1$, confirming the theorem.

:::


## Wald's Theorem for Testing Parameter Restrictions

The Wald Test evaluates the distance between the unrestricted Maximum Likelihood Estimator (MLE) and the null hypothesis value, standardized by the curvature of the log-likelihood at the unrestricted estimate. Unlike the Score test, the Wald test requires estimating the full, unrestricted model.

::: {#thm-wald}

### Asymptotic Distribution of the Wald Statistic

Let $\mathbf{Y} = (Y_1, \dots, Y_n)$ be an i.i.d. sample from a regular family of distributions with parameters $\boldsymbol{\theta} \in \Theta \subseteq \mathbb{R}^{p}$. Partition the parameter vector as $\boldsymbol{\theta} = (\boldsymbol{\theta}_0^T, \boldsymbol{\theta}_1^T)^T$, where $\boldsymbol{\theta}_0$ is the $q$-dimensional parameter of interest and $\boldsymbol{\theta}_1$ is the $(p-q)$-dimensional nuisance parameter. Consider the test:
$$
H_0: \boldsymbol{\theta}_0 = \boldsymbol{\theta}_0^* \quad \text{vs} \quad H_1: \boldsymbol{\theta}_0 \neq \boldsymbol{\theta}_0^*
$$

Let $\hat{\boldsymbol{\theta}} = (\hat{\boldsymbol{\theta}}_0^T, \hat{\boldsymbol{\theta}}_1^T)^T$ be the unrestricted MLE under $H_1$. Let $\Sigma(\boldsymbol{\theta}) = \mathcal{I}^{-1}(\boldsymbol{\theta})$ be the asymptotic covariance matrix of the unrestricted MLE, where $\mathcal{I}(\boldsymbol{\theta})$ is the expected Fisher Information matrix, and let $\Sigma_{00}$ denote its top-left $q \times q$ marginal block.

The Wald statistic is defined as the quadratic form:
$$
W = (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)^T \Sigma_{00}^{-1}(\hat{\boldsymbol{\theta}}) (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)
$$

Under standard regularity conditions, if $H_0$ is true, as $n \to \infty$:
$$
W \xrightarrow{d} \chi^2_q
$$
where $q$ is the dimension of the parameter of interest $\boldsymbol{\theta}_0$.

:::

::: {.proof}

**Algebraic Proof via Mahalanobis Decomposition**

We can derive the Wald statistic directly by borrowing the geometric decomposition used in Wilks' Theorem, isolating the marginal behavior of the unrestricted MLE.

**1. The Total Mahalanobis Distance**
By standard large-sample theory, the unrestricted MLE $\hat{\boldsymbol{\theta}}$ is asymptotically normal. Its asymptotic covariance matrix is $\Sigma = \mathcal{I}^{-1}(\boldsymbol{\theta})$. The total squared Mahalanobis distance between the unrestricted MLE $\hat{\boldsymbol{\theta}}$ and the full hypothesized parameter vector $\boldsymbol{\theta}^* = ({\boldsymbol{\theta}_0^*}^T, {\boldsymbol{\theta}_1^*}^T)^T$ under the precision metric $\Sigma^{-1}$ is:
$$
D_{total} = (\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}^*)^T \Sigma^{-1} (\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}^*)
$$

**2. The Pythagorean Decomposition (From Wilks' Theorem)**
As established in the geometric proof of Wilks' Theorem, any Mahalanobis distance in a partitioned parameter space orthogonally decomposes into a marginal distance for the parameter of interest and a conditional distance for the nuisance parameter:
$$
D_{total} = (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)^T \Sigma_{00}^{-1} (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*) + (\hat{\boldsymbol{\theta}}_{1, \Sigma} - \boldsymbol{\theta}_1^*)^T \mathcal{I}_{11} (\hat{\boldsymbol{\theta}}_{1, \Sigma} - \boldsymbol{\theta}_1^*)
$$
where $\Sigma_{00}$ is the top-left block of $\Sigma$, and $\hat{\boldsymbol{\theta}}_{1, \Sigma}$ is the geometric center of the nuisance parameter constraint.

**3. Isolating the Marginal Distance (The Wald Statistic)**
The Wald test zeroes in exclusively on the first term of this decomposition. Rather than evaluating the full likelihood surface (as Wilks does) or the gradient at the boundary (as the Score test does), the Wald test directly measures the isolated **marginal Mahalanobis distance** of the parameter of interest from its null value:
$$
W_{true} = (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)^T \Sigma_{00}^{-1} (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)
$$
Because the marginal asymptotic distribution of the estimator under the null hypothesis is $\hat{\boldsymbol{\theta}}_0 \overset{a}{\sim} N(\boldsymbol{\theta}_0^*, \Sigma_{00})$, standardizing this deviation by its precision matrix $\Sigma_{00}^{-1}$ produces a quadratic form that follows a chi-square distribution:
$$
W_{true} \xrightarrow{d} \chi^2_q
$$

**4. Slutsky's Substitution for the Unknown Covariance**
In practice, the true marginal covariance $\Sigma_{00}$ is unknown because it depends on the true parameter $\boldsymbol{\theta}$. However, because the Fisher Information matrix is a continuous function of the parameters, and $\hat{\boldsymbol{\theta}}$ is a consistent estimator of $\boldsymbol{\theta}$ under $H_1$, we can invoke Slutsky's Theorem. Substituting the estimated covariance matrix $\Sigma_{00}(\hat{\boldsymbol{\theta}})$ yields the computable Wald statistic without altering its asymptotic distribution:
$$
W = (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*)^T \Sigma_{00}^{-1}(\hat{\boldsymbol{\theta}}) (\hat{\boldsymbol{\theta}}_0 - \boldsymbol{\theta}_0^*) \xrightarrow{d} \chi^2_q
$$

:::

::: {#exm-wald-normal}

**Wald Test for Normal Mean with Unknown Variance**

Let $X_1, \dots, X_n \overset{i.i.d.}{\sim} N(\mu, \sigma^2)$ where both $\mu$ and $\sigma^2$ are unknown. The parameter vector is $\boldsymbol{\theta} = (\mu, \sigma^2)^T$. We wish to test:
$$
H_0: \mu = \mu_0 \quad \text{vs} \quad H_1: \mu \neq \mu_0
$$

1.  **Find the Unrestricted MLEs:** Maximizing the full log-likelihood with respect to both parameters yields the unrestricted estimators:

$$
\hat{\mu} = \overline{X}, \quad \hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \overline{X})^2
$$
    Our unrestricted parameter estimate is $\hat{\boldsymbol{\theta}} = (\hat{\mu}, \hat{\sigma}^2)^T$.

2.  **Determine the Marginal Covariance at the MLE:** The expected Fisher Information matrix for a Normal distribution is diagonal:

$$
\mathcal{I}(\mu, \sigma^2) = \begin{pmatrix} \frac{n}{\sigma^2} & 0 \\ 0 & \frac{n}{2\sigma^4} \end{pmatrix}
$$
    Because the matrix is diagonal, the asymptotic covariance matrix $\Sigma(\boldsymbol{\theta}) = \mathcal{I}^{-1}(\boldsymbol{\theta})$ is simply the matrix of reciprocals. The marginal variance for $\mu$ (the top-left block $\Sigma_{00}$) evaluated at the unrestricted MLE $\hat{\boldsymbol{\theta}}$ is:
$$
\Sigma_{00}(\hat{\boldsymbol{\theta}}) = \frac{\hat{\sigma}^2}{n}
$$

3.  **Compute the Wald Statistic $W$:** We construct the quadratic form using the parameter of interest ($\mu$), its null value ($\mu_0$), and the inverse of its estimated marginal variance ($\Sigma_{00}^{-1}$):

$$
W = (\hat{\mu} - \mu_0)^T \Sigma_{00}^{-1}(\hat{\boldsymbol{\theta}}) (\hat{\mu} - \mu_0)
$$
    Substituting our values:
$$
W = (\overline{X} - \mu_0) \left( \frac{n}{\hat{\sigma}^2} \right) (\overline{X} - \mu_0) = \frac{n(\overline{X} - \mu_0)^2}{\hat{\sigma}^2}
$$

4.  **Contrast with the Score Statistic:**

    By Wald's Theorem, this statistic follows a $\chi^2_1$ distribution asymptotically. Note the critical difference from the Score test derived previously. The Wald statistic uses the unrestricted variance estimator $\hat{\sigma}^2$, whereas the Score statistic uses the restricted variance estimator $\hat{\sigma}_0^2 = \frac{1}{n}\sum (X_i - \mu_0)^2$. Because $\hat{\sigma}_0^2 \ge \hat{\sigma}^2$, it guarantees that mathematically $W \ge S$ for this model.

:::

## Rao's Score (Lagrange Multiplier) Theorem for Restricted Models

The asymptotic distribution of the Score Test statistic relies on the behavior of the gradient of the log-likelihood (the Score vector) evaluated at the restricted Maximum Likelihood Estimator. Because it explicitly evaluates the "force" pushing against the null hypothesis constraint, it is equivalently known as the **Lagrange Multiplier (LM) Test**. This theorem provides a powerful method for hypothesis testing that only requires fitting the model under the null hypothesis.

::: {#thm-score}

### Asymptotic Distribution of the Score Statistic

Let $\mathbf{Y} = (Y_1, \dots, Y_n)$ be an i.i.d. sample from a regular family of distributions with parameters $\boldsymbol{\theta} \in \Theta \subseteq \mathbb{R}^{p}$. Partition the parameter vector as $\boldsymbol{\theta} = (\boldsymbol{\theta}_0^T, \boldsymbol{\theta}_1^T)^T$, where $\boldsymbol{\theta}_0$ is the $q$-dimensional parameter of interest and $\boldsymbol{\theta}_1$ is the $(p-q)$-dimensional nuisance parameter. Consider the test:
$$
H_0: \boldsymbol{\theta}_0 = \boldsymbol{\theta}_0^* \quad \text{vs} \quad H_1: \boldsymbol{\theta}_0 \neq \boldsymbol{\theta}_0^*
$$

Let $\hat{\boldsymbol{\theta}}_0 = ({\boldsymbol{\theta}_0^*}^T, \hat{\boldsymbol{\theta}}_1^T)^T$ be the restricted MLE under $H_0$. Let $\mathbf{U}(\boldsymbol{\theta})$ be the total Score vector partitioned as $\mathbf{U} = [\mathbf{U}_0^T, \mathbf{U}_1^T]^T$, and let $\Sigma = \mathcal{I}^{-1}(\boldsymbol{\theta})$ be the asymptotic covariance matrix of the unrestricted MLE, with $\Sigma_{00}$ representing its $q \times q$ top-left marginal block.

The Score statistic is $S = \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)^T \Sigma_{00}(\hat{\boldsymbol{\theta}}_0) \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)$. Under $H_0$, as $n \to \infty$:
$$
S \xrightarrow{d} \chi^2_q
$$
where $q$ is the dimension of the parameter of interest $\boldsymbol{\theta}_0$.

:::

::: {.proof}

**Algebraic Proof via Joint Distribution and Conditioning**

This proof establishes the behavior of the Score vector using constrained optimization, and then derives its sampling distribution by conditioning the joint asymptotic distribution of the true scores.

1. Constrained Optimization (The Lagrange Multipliers)

    To find the restricted MLE $\hat{\boldsymbol{\theta}}_0$, we maximize the log-likelihood subject to the constraint $\boldsymbol{\theta}_0 = \boldsymbol{\theta}_0^*$. We form the Lagrangian with a $q$-dimensional multiplier vector $\boldsymbol{\lambda}$:
    $$
    \mathcal{L}(\boldsymbol{\theta}, \boldsymbol{\lambda}) = \ell(\boldsymbol{\theta}_0, \boldsymbol{\theta}_1) - \boldsymbol{\lambda}^T (\boldsymbol{\theta}_0 - \boldsymbol{\theta}_0^*)
    $$
    Taking the partial derivatives with respect to the parameters and evaluating them at the restricted optimum $\hat{\boldsymbol{\theta}}_0$ yields the first-order conditions:

    * $\nabla_{\boldsymbol{\theta}_1} \mathcal{L} = \mathbf{U}_1(\hat{\boldsymbol{\theta}}_0) = \mathbf{0}$
    * $\nabla_{\boldsymbol{\theta}_0} \mathcal{L} = \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0) - \hat{\boldsymbol{\lambda}} = \mathbf{0} \implies \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0) = \hat{\boldsymbol{\lambda}}$

    This proves that at the restricted peak, the observed nuisance score $\mathbf{U}_1$ is exactly zero. Furthermore, the evaluated score of interest $\mathbf{U}_0$ is exactly the Lagrange multiplier $\hat{\boldsymbol{\lambda}}$, representing the gradient force pulling toward the unrestricted peak.

2. Joint Asymptotic Distribution of the True Scores

    Let $\mathbf{U}_0$ and $\mathbf{U}_1$ denote the unobservable true score vectors evaluated at the true parameters under the null hypothesis, $\boldsymbol{\theta}^* = ({\boldsymbol{\theta}_0^*}^T, \boldsymbol{\theta}_1^T)^T$. By the Central Limit Theorem, the joint score vector is asymptotically normal, centered at zero, with a covariance matrix given by the expected Fisher Information:
$$
    \begin{pmatrix} \mathbf{U}_0 \\ \mathbf{U}_1 \end{pmatrix} \xrightarrow{d} N \left( \begin{pmatrix} \mathbf{0} \\ \mathbf{0} \end{pmatrix}, \begin{pmatrix} \mathcal{I}_{00} & \mathcal{I}_{01} \\ \mathcal{I}_{10} & \mathcal{I}_{11} \end{pmatrix} \right)
$$

3. The Conditional Distribution

    Using the standard properties of the multivariate normal distribution, the conditional distribution of the true score of interest $\mathbf{U}_0$ given the true nuisance score $\mathbf{U}_1$ is:
$$
    \mathbf{U}_0 \mid \mathbf{U}_1 \sim N\left(\mathcal{I}_{01}\mathcal{I}_{11}^{-1}\mathbf{U}_1, \mathcal{I}_{00} - \mathcal{I}_{01}\mathcal{I}_{11}^{-1}\mathcal{I}_{10}\right)
$$

4. Evaluating at the Restricted MLE (Conditioning on Zero)

    As established in Step 1, evaluating the model at the restricted MLE $\hat{\boldsymbol{\theta}}_0$ structurally forces the nuisance score to zero ($\mathbf{U}_1 = \mathbf{0}$). Therefore, the asymptotic distribution of the evaluated score $\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)$ is equivalent to the conditional distribution of $\mathbf{U}_0$ given $\mathbf{U}_1 = \mathbf{0}$. 

    Substituting $\mathbf{U}_1 = \mathbf{0}$ into the conditional distribution, the mean vanishes entirely:
$$
    \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0) \xrightarrow{d} N\left(\mathbf{0}, \mathcal{I}_{00} - \mathcal{I}_{01}\mathcal{I}_{11}^{-1}\mathcal{I}_{10}\right)
$$

5. Connection to Marginal Covariance and the Final Statistic

    The conditional variance matrix $\mathcal{I}_{00} - \mathcal{I}_{01}\mathcal{I}_{11}^{-1}\mathcal{I}_{10}$ is the Schur complement of the Information matrix. By the formula for the inverse of a partitioned matrix, this is exactly the inverse of the top-left block of the asymptotic covariance matrix $\Sigma = \mathcal{I}^{-1}$:
$$
    \text{Var}(\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)) = \Sigma_{00}^{-1}
$$

    Because the evaluated score $\mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)$ is asymptotically normal with mean zero and variance $\Sigma_{00}^{-1}$, standardizing it by its inverse-variance creates a quadratic form. Since the inverse of the precision is the covariance itself ($(\Sigma_{00}^{-1})^{-1} = \Sigma_{00}$), we obtain the test statistic:
$$
    S = \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0)^T \Sigma_{00} \mathbf{U}_0(\hat{\boldsymbol{\theta}}_0) \xrightarrow{d} \chi^2_q
$$

:::

::: {#exm-score-normal}

**Score Test for Normal Mean with Unknown Variance**

Let $X_1, \dots, X_n \overset{i.i.d.}{\sim} N(\mu, \sigma^2)$ where both $\mu$ and $\sigma^2$ are unknown. The parameter vector is $\boldsymbol{\theta} = (\mu, \sigma^2)^T$. We wish to test:
$$
H_0: \mu = \mu_0 \quad \text{vs} \quad H_1: \mu \neq \mu_0
$$

1.  **Find the Restricted MLEs:** Under $H_0$, we fix $\mu = \mu_0$. Maximizing the restricted log-likelihood with respect to the nuisance parameter $\sigma^2$ yields:

$$
\hat{\sigma}_0^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \mu_0)^2
$$
    The restricted parameter vector is $\hat{\boldsymbol{\theta}}_0 = (\mu_0, \hat{\sigma}_0^2)^T$.

2.  **Evaluate the Full Score Vector:** The full log-likelihood function for the normal sample is:

$$
\ell(\mu, \sigma^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n (X_i - \mu)^2
$$
    
    The total Score vector $\mathbf{U}(\boldsymbol{\theta})$ is the gradient with respect to both parameters:
$$
\mathbf{U}(\mu, \sigma^2) = \begin{pmatrix} U_\mu(\mu, \sigma^2) \\ U_{\sigma^2}(\mu, \sigma^2) \end{pmatrix} = \begin{pmatrix} \frac{n(\overline{X} - \mu)}{\sigma^2} \\ -\frac{n}{2\sigma^2} + \frac{1}{2\sigma^4}\sum_{i=1}^n (X_i - \mu)^2 \end{pmatrix}
$$
    
    Now we evaluate this full vector at the restricted MLE $\hat{\boldsymbol{\theta}}_0 = (\mu_0, \hat{\sigma}_0^2)^T$:
$$
\mathbf{U}(\hat{\boldsymbol{\theta}}_0) = \begin{pmatrix} \frac{n(\overline{X} - \mu_0)}{\hat{\sigma}_0^2} \\ -\frac{n}{2\hat{\sigma}_0^2} + \frac{1}{2\hat{\sigma}_0^4}\sum_{i=1}^n (X_i - \mu_0)^2 \end{pmatrix}
$$
    
    Because $\hat{\sigma}_0^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \mu_0)^2$, the second component perfectly cancels out. By definition, the score for the nuisance parameter evaluated at the restricted MLE is zero:
$$
\mathbf{U}(\hat{\boldsymbol{\theta}}_0) = \begin{pmatrix} \frac{n(\overline{X} - \mu_0)}{\hat{\sigma}_0^2} \\ 0 \end{pmatrix}
$$

3.  **Determine the Marginal Covariance:** The expected Fisher Information matrix for a Normal distribution is diagonal because the mean and variance are orthogonal parameters:

$$
\mathcal{I}(\mu, \sigma^2) = \begin{pmatrix} \frac{n}{\sigma^2} & 0 \\ 0 & \frac{n}{2\sigma^4} \end{pmatrix}
$$
    Because the matrix is diagonal, the asymptotic covariance matrix $\Sigma(\boldsymbol{\theta}) = \mathcal{I}^{-1}(\boldsymbol{\theta})$ is simply the matrix of reciprocals. The marginal variance for $\mu$ evaluated at the restricted MLE $\hat{\boldsymbol{\theta}}_0$ is:
$$
\Sigma_{00}(\hat{\boldsymbol{\theta}}_0) = \frac{\hat{\sigma}_0^2}{n}
$$

4.  **Compute the Score Statistic $S$:** Constructing the quadratic form for the parameter of interest using its marginal variance $\Sigma_{00}$:

$$
S = U_\mu(\hat{\boldsymbol{\theta}}_0)^T \Sigma_{00}(\hat{\boldsymbol{\theta}}_0) U_\mu(\hat{\boldsymbol{\theta}}_0)
$$
$$
S = \left( \frac{n(\overline{X} - \mu_0)}{\hat{\sigma}_0^2} \right) \left( \frac{\hat{\sigma}_0^2}{n} \right) \left( \frac{n(\overline{X} - \mu_0)}{\hat{\sigma}_0^2} \right) = \frac{n(\overline{X} - \mu_0)^2}{\hat{\sigma}_0^2}
$$

5.  **Asymptotic Distribution:** By the Score Theorem, since we are testing $q=1$ parameter restriction, the statistic follows a $\chi^2_1$ distribution:

$$
S = \frac{n(\overline{X} - \mu_0)^2}{\hat{\sigma}_0^2} \xrightarrow{d} \chi^2_1
$$
    This is highly analogous to the squared t-statistic, but specifically uses the variance estimated under the null hypothesis constraint.

:::

## A Comparison of Finite-Sample Distributions of the Three Asymptotic Tests in OLS

#### Exact Finite-Sample Distributions via the F-Statistic

When testing linear restrictions in Ordinary Least Squares (OLS) under the assumption of normal errors, we do not need to rely on asymptotic approximations. The true, exact finite-sample test statistic follows an $F$-distribution. 

Because the Wald ($W$), Likelihood Ratio ($LR$), and Score ($S$) statistics are strictly monotonic algebraic transformations of the exact finite-sample $F$-statistic, we can use the transformation of variables to derive their exact finite-sample Cumulative Distribution Functions (CDFs).

Let us consider testing a linear hypothesis involving $q$ restrictions in an OLS model with $k$ total parameters and $n$ observations. Let $F$ denote the exact finite-sample $F$-statistic, which follows an $F_{q, n-k}$ distribution. Let $G_{q, n-k}(x)$ denote the CDF of this $F$-distribution.

By isolating $F$ in the algebraic definitions of $W$, $LR$, and $S$, we derive their exact finite-sample CDFs:

**Expressing the Test Statistics via Residual Sums of Squares**
 

Let $RSS_1$ denote the residual sum of squares from the **full** (unrestricted) model (estimating all $k$ parameters under $H_1$), and let $RSS_0$ denote the residual sum of squares from the **restricted** model (imposing the $q$ constraints of $H_0$). 

Because imposing constraints can only degrade the fit of the model to the training data, we know that $RSS_0 \ge RSS_1$. The difference $(RSS_0 - RSS_1)$ isolates the exact loss of fit caused by enforcing the null hypothesis.

The exact finite-sample $F$-statistic is the ratio of this loss of fit to the unrestricted noise, scaled by their respective degrees of freedom:
$$
F = \frac{(RSS_0 - RSS_1) / q}{RSS_1 / (n-k)}
$$

By substituting this definition of $F$ into the algebraic transformations derived previously, the "Holy Trinity" of asymptotic tests can be expressed beautifully as functions of the residual sums of squares:

**1. Wald Statistic ($W$)**
The Wald test standardizes the loss of fit using the **unrestricted** variance estimate ($\hat{\sigma}^2 = RSS_1/n$). Substituting the $F$ formula into $W = \frac{nq}{n-k} F$:
$$
W = n \left( \frac{RSS_0 - RSS_1}{RSS_1} \right)
$$

**2. Likelihood Ratio Statistic ($LR$)**
The Likelihood Ratio test evaluates the natural logarithm of the ratio of the two likelihoods, which simplifies directly to the log-ratio of the residual sums of squares. Substituting $F$ into $LR = n \ln \left( 1 + \frac{q}{n-k} F \right)$:
$$
LR = n \ln \left( \frac{RSS_0}{RSS_1} \right)
$$

**3. Score Statistic ($S$)**
The Score test standardizes the loss of fit using the **restricted** variance estimate ($\hat{\sigma}_0^2 = RSS_0/n$). Substituting $F$ into $S = \frac{n q F}{n-k + q F}$:
$$
S = n \left( \frac{RSS_0 - RSS_1}{RSS_0} \right)
$$

#### Asymptotic Equivalence in OLS

Having established the exact algebraic relationship between the three test statistics and the finite-sample $F$-statistic, we can use standard OLS large-sample theory to rigorously prove that all three converge to the exact same $\chi^2_q$ asymptotic distribution.

::: {.proof}

**1. The Asymptotic Limit of the Scaled F-Statistic**
Under the null hypothesis $H_0$ (and assuming standard regularity conditions for OLS), the numerator of the $F$-statistic represents the loss of fit, which asymptotically follows a Chi-square distribution scaled by the true error variance $\sigma^2$:
$$
(RSS_0 - RSS_1) \xrightarrow{d} \sigma^2 \chi^2_q
$$
The denominator of the $F$-statistic is the unbiased estimator of the error variance, which converges in probability to the true variance by the Law of Large Numbers:
$$
\frac{RSS_1}{n-k} \xrightarrow{p} \sigma^2
$$
By Slutsky's Theorem, dividing the convergent numerator by the convergent denominator cancels out the unknown $\sigma^2$, yielding the foundational limit:
$$
qF = \frac{RSS_0 - RSS_1}{RSS_1 / (n-k)} \xrightarrow{d} \chi^2_q
$$

**2. Convergence of the Wald Statistic ($W$)**
We previously defined the Wald statistic algebraically as:
$$
W = \frac{n}{n-k} (qF)
$$
As $n \to \infty$, the ratio of the sample size to the degrees of freedom $\left(\frac{n}{n-k}\right)$ converges to $1$. Applying Slutsky's Theorem again, the Wald statistic converges directly to the scaled $F$-statistic:
$$
W \xrightarrow{d} 1 \cdot \chi^2_q = \chi^2_q
$$

**3. Convergence of the Likelihood Ratio Statistic ($LR$)**
The $LR$ statistic is defined via the natural logarithm:
$$
LR = n \ln \left( 1 + \frac{qF}{n-k} \right)
$$
Because $qF$ converges to a $\chi^2_q$ distribution, it is bounded in probability, meaning $\frac{qF}{n-k} = O_p(n^{-1})$. As $n \to \infty$, this term approaches zero. We can apply the first-order Taylor expansion $\ln(1 + x) \approx x$ for small $x$:
$$
LR \approx n \left( \frac{qF}{n-k} \right) = \frac{n}{n-k}(qF)
$$
This first-order approximation is exactly the Wald statistic. Therefore, the difference between $LR$ and $W$ vanishes asymptotically, giving:
$$
LR \xrightarrow{d} \chi^2_q
$$

**4. Convergence of the Score Statistic ($S$)**
The Score statistic is algebraically defined as:
$$
S = \frac{n q F}{n-k + q F}
$$
Dividing the numerator and denominator by $n-k$ yields:
$$
S = \frac{\frac{n}{n-k}(qF)}{1 + \frac{qF}{n-k}} = \frac{W}{1 + \frac{qF}{n-k}}
$$
As established, $\frac{qF}{n-k} \xrightarrow{p} 0$. The denominator converges in probability to $1$, meaning the Score statistic perfectly tracks the Wald statistic in the limit:
$$
S \xrightarrow{d} \frac{W}{1} \xrightarrow{d} \chi^2_q
$$

**Conclusion:** While the algebraic inequality $W \ge LR \ge S$ holds strictly in any finite sample, the differences between the statistics are of order $O_p(n^{-1})$. As $n \to \infty$, the "gaps" between them shrink to zero, and all three provide asymptotically equivalent tests against the $\chi^2_q$ distribution.
:::

#### Visualizing Finite-Sample Calibration (Size Distortion)

To visualize how badly the asymptotic $\chi^2_q$ approximation distorts the Type I error rates in finite samples, we can plot the true Cumulative Distribution Function (CDF) of the asymptotic p-values. 

If a test is perfectly calibrated, the probability that its p-value is less than a nominal threshold $\alpha$ should be exactly $\alpha$, forming a perfect $y = x$ diagonal line (the Uniform(0,1) CDF). 

The event that the asymptotic p-value is less than $\alpha$ is mathematically equivalent to the test statistic being greater than the $\chi^2_q$ critical value ($c_\alpha$). Therefore, the true rejection probability $P(p < \alpha)$ is exactly $1 - F(c_\alpha)$, where $F$ is the exact finite-sample CDF derived above.

#### Exact Finite-Sample Distribution Calibration Plot

The following code allows you to adjust the sample size ($n$) and the number of parameters ($k$) to observe how the finite-sample rejection rates of the three asymptotic tests (Wald, Likelihood Ratio, and Score) converge toward the ideal Uniform(0,1) calibration.


```{r}
#| label: fig-calibration-tuning
#| fig-cap: "Exact Analytical CDFs of Asymptotic P-values"
#| warning: false
#| fig-width: 12
#| fig-height: 6

library(ggplot2)
library(dplyr)

# --- Tuning Parameters ---
n1 <- 15    # Small sample size for Plot 1
n2 <- 50  # Large sample size for Plot 2
k_val <- 5  # Number of parameters (constant for comparison)
q_val <- 1  # Number of restrictions (constant for comparison)

# -------------------------

generate_calibration_data <- function(n, k, q) {
  alpha_seq <- seq(0.0001, 0.15, length.out = 1000)
  c_alpha <- qchisq(1 - alpha_seq, df = q)
  
  # Wald: 1 - F_W(c_alpha)
  F_val_W <- ((n - k) / (n * q)) * c_alpha
  cdf_W <- 1 - pf(F_val_W, df1 = q, df2 = n - k)
  
  # LR: 1 - F_LR(c_alpha)
  F_val_LR <- ((n - k) / q) * (exp(c_alpha / n) - 1)
  cdf_LR <- 1 - pf(F_val_LR, df1 = q, df2 = n - k)
  
  # Score: 1 - F_S(c_alpha)
  F_val_S <- ifelse(c_alpha < n, (c_alpha * (n - k)) / (q * (n - c_alpha)), Inf)
  cdf_S <- 1 - pf(F_val_S, df1 = q, df2 = n - k)
  
  data.frame(
    Alpha = rep(alpha_seq, 3),
    Rejection_Prob = c(cdf_W, cdf_LR, cdf_S),
    Test = factor(rep(c("Wald", "LR", "Score"), each = length(alpha_seq))),
    n = n,
    k = k
  ) |>
    mutate(Label = paste0("n = ", n, ", k = ", k))
}

# Create combined dataset based on tuning parameters
plot_data <- rbind(
  generate_calibration_data(n = n1, k = k_val, q = q_val),
  generate_calibration_data(n = n2, k = k_val, q = q_val)
)

# Fix factor levels for side-by-side ordering
plot_data$Label <- factor(plot_data$Label, levels = unique(plot_data$Label))

ggplot(plot_data, aes(x = Alpha, y = Rejection_Prob, color = Test)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~ Label) +
  scale_color_manual(values = c("Wald" = "#E41A1C", "LR" = "#4DAF4A", "Score" = "#377EB8")) +
  coord_cartesian(xlim = c(0, 0.12), ylim = c(0, 0.20)) +
  labs(
    x = "Nominal Significance Level (\u03b1)",
    y = "True Rejection Rate: P(p-value < \u03b1)",
    title = "Asymptotic Test Calibration vs. Sample Size",
    subtitle = "Dashed line represents perfect calibration (Rejection Rate = \u03b1)"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )

```

**Summary**

In OLS with finite samples, the $\chi^2$ distribution is an optimistic approximation that leads to **over-rejection** (liberal bias) across all three tests. However:

* The **Wald test** is the most liberal because it does nothing to counteract the too-low $\chi^2$ threshold.
* The **Score test** is the most conservative of the three because its mathematical structure provides a natural downward "correction" that happens to land it closest to the ideal uniform distribution at common significance levels like $\alpha = 0.05$.