---
title: "Maximum Likelihood Estimation"
engine: knitr
format: 
  html: default
  pdf: default
---

## Maximum Likelihood Estimation

### Definitions and Notations

::: {#def-likelihood-mle}

### Maximum Likelihood Estimation

1. **Likelihood Function:**

    Let $f(\mathbf{x}|\boldsymbol{\theta})$ be the joint probability density function of the data $\mathbf{X}$. When viewed as a function of the parameter $\boldsymbol{\theta}$ given fixed data $\mathbf{x}$, it is called the likelihood function:
    $$
    L(\boldsymbol{\theta}; \mathbf{x}) = f(\mathbf{x}|\boldsymbol{\theta})
    $$

2. **Log-likelihood:**

    It is usually easier to maximize the natural logarithm of the likelihood:
    $$
    \ell(\boldsymbol{\theta}; \mathbf{x}) = \log L(\boldsymbol{\theta}; \mathbf{x})
    $$

3. **Maximum Likelihood Estimator (MLE):**

    The MLE $\hat{\boldsymbol{\theta}}_{\text{MLE}}$ is the value that maximizes the likelihood function:
    $$
    \hat{\boldsymbol{\theta}}_{\text{MLE}}(\mathbf{x}) = \operatorname*{argmax}_{\boldsymbol{\theta} \in \Theta} \ell(\boldsymbol{\theta}; \mathbf{x})
    $$

4. **Score Function ($\mathbf{U}$):**

    The score function is defined as the gradient of the log-likelihood with respect to the parameter vector. 
    $$
    \mathbf{U}(\boldsymbol{\theta}; \mathbf{x}) = \nabla_{\boldsymbol{\theta}} \ell(\boldsymbol{\theta}; \mathbf{x})
    $$
    Finding the MLE often involves solving the score equation: 
    $$
    \mathbf{U}(\boldsymbol{\theta}; \mathbf{x}) \equiv 0
    $$

5. **Observed Fisher Information:**

   $$
   \mathbf{J}(\boldsymbol{\theta}; \mathbf{X}) = - \nabla^2 \ell(\boldsymbol{\theta}; \mathbf{X})
   $$

6. **(Expected) Fisher Information Matrix ($\mathcal{I}$):** The covariance matrix of the score vector. It is a deterministic $p \times p$ matrix (for a fixed $\boldsymbol{\theta}$).
    $$
    \mathcal{I}(\boldsymbol{\theta}) = E_{\boldsymbol{\theta}} \left[ \mathbf{J}(\boldsymbol{\theta}; \mathbf{X}) \right]
    $$

:::

### Summary of Key Asymptotic Results for MLE

Under standard regularity conditions (e.g., i.i.d. observations, smooth density, and $\boldsymbol{\theta}$ in the interior of $\Theta$), the MLE exhibits the following asymptotic properties as $n \to \infty$:

1. **Consistency:**
    The estimator converges in probability to the true parameter $\boldsymbol{\theta}_0$.
    $$
    \hat{\boldsymbol{\theta}}_{\text{MLE}} \xrightarrow{p} \boldsymbol{\theta}_0
    $$

2. **Asymptotic Normality of the Score:**
    The Score vector evaluated at the true parameter converges to a Normal distribution with variance equal to the Fisher Information.
    $$
    \frac{1}{\sqrt{n}} \mathbf{U}(\boldsymbol{\theta}_0) \xrightarrow{d} N_p(\mathbf{0}, \mathcal{I}_1(\boldsymbol{\theta}_0))
    $$
    *(Where $\mathcal{I}_1$ is the expected information for a single observation).*

3. **Asymptotic Normality of the Estimator:**
    The MLE is asymptotically normal, unbiased, and efficient (achieving the Cramér-Rao lower bound).
    $$
    \sqrt{n}(\hat{\boldsymbol{\theta}}_{\text{MLE}} - \boldsymbol{\theta}_0) \xrightarrow{d} N_p(\mathbf{0}, \mathcal{I}_1(\boldsymbol{\theta}_0)^{-1})
    $$

4. **Wilks' Theorem (Likelihood Ratio for Nested Models):**
    Consider two nested models: a **Full Model** with parameter space $\Theta_{\text{Full}}$ (dimension $p$) and a **Reduced Model** restricted to a subspace $\Theta_{\text{Reduced}}$ (dimension $q < p$). 
    
    The Likelihood Ratio Statistic (Deviance) is defined as:
    $$
    D = 2 \left( \ell(\hat{\boldsymbol{\theta}}_{\text{Full}}) - \ell(\hat{\boldsymbol{\theta}}_{\text{Reduced}}) \right)
    $$
    Under the null hypothesis that the Reduced Model is true, the statistic converges to a Chi-squared distribution:
    $$
    D \xrightarrow{d} \chi^2_{p-q}
    $$
    *(The degrees of freedom equal the difference in the number of parameters).*

### Examples

#### Example: Normal Distribution (Exact Asymptotics)

::: {#exm-normal-mle-asymptotics}

#### Example: Normal Distribution (Exact Asymptotics)

Let $X_1, \dots, X_n \overset{iid}{\sim} N(\theta, \sigma^2)$ with $\sigma^2$ known. We explicitly derive the key quantities and map them to the four general asymptotic results.

**1. Derivation of Key Quantities**

* **Log-Likelihood:**
    $$
    \ell(\theta) = -\frac{n}{2}\log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n (x_i - \theta)^2
    $$

* **Score Function ($U_n$):**
    Differentiating with respect to $\theta$:
    $$
    U_n(\theta) = \ell'(\theta) = \frac{1}{\sigma^2}\sum_{i=1}^n (x_i - \theta) = \frac{n}{\sigma^2}(\bar{X}_n - \theta)
    $$

* **Maximum Likelihood Estimator ($\hat{\theta}_n$):**
    Setting $U_n(\hat{\theta}) = 0$:
    $$
    \frac{n}{\sigma^2}(\bar{X}_n - \hat{\theta}) = 0 \implies \hat{\theta}_n = \bar{X}_n
    $$

* **Fisher Information ($\mathcal{I}_n$):**
    Taking the negative second derivative:
    $$
    \mathcal{I}_n(\theta) = -E[\ell''(\theta)] = -E\left[ -\frac{n}{\sigma^2} \right] = \frac{n}{\sigma^2}
    $$
    *(Note: For a single observation, $\mathcal{I}_1 = 1/\sigma^2$).*

**2. Asymptotic Properties**

We now illustrate the four main theorems using these derived forms:

1.  **Consistency**
    
    The MLE is $\hat{\theta}_n = \bar{X}_n$. By the Weak Law of Large Numbers, we know that:
    $$
    \bar{X}_n \xrightarrow{p} \theta
    $$
    Thus, the estimator is consistent.

2.  **Asymptotic Normality of the Score**
    
    The score function is $U_n(\theta) = \frac{n}{\sigma^2}(\bar{X}_n - \theta)$. Since $\bar{X}_n$ is normally distributed, the Score is **exactly** Normal for any $n$:
    $$
    U_n(\theta) \sim N\left(0, \frac{n}{\sigma^2}\right) = N(0, \mathcal{I}_n)
    $$
    Scaling by $1/\sqrt{n}$, we recover the asymptotic result:
    $$
    \frac{1}{\sqrt{n}} U_n(\theta) \sim N(0, \mathcal{I}_1)
    $$

3.  **Asymptotic Normality of the Estimator**
    
    The distribution of the standardized MLE is derived from the properties of the sample mean:
    $$
    \sqrt{n}(\hat{\theta}_n - \theta) = \sqrt{n}(\bar{X}_n - \theta) \sim N(0, \sigma^2)
    $$
    Note that $\sigma^2 = 1/\mathcal{I}_1$. Therefore:
    $$
    \sqrt{n}(\hat{\theta}_n - \theta) \sim N(0, \mathcal{I}_1^{-1})
    $$
    Again, the asymptotic approximation holds exactly for finite sample sizes here.

4.  **Wilks' Theorem (Deviance Approximation)**
    
    We test $H_0: \theta = \theta_0$. The Deviance is $D = 2[\ell(\hat{\theta}) - \ell(\theta_0)]$.
    
    **Step A: Taylor Expansion**
    We expand the log-likelihood around the MLE. In general, this is an approximation, but for the Normal distribution (where the log-likelihood is exactly quadratic), the expansion is **exact**:
    $$
    \ell(\theta_0) = \ell(\hat{\theta}) + (\theta_0 - \hat{\theta}) \underbrace{\ell'(\hat{\theta})}_{0} + \frac{1}{2}(\theta_0 - \hat{\theta})^2 \ell''(\hat{\theta})
    $$
    Substituting the constant curvature $\ell''(\theta) = -n/\sigma^2$:
    $$
    2[\ell(\hat{\theta}) - \ell(\theta_0)] = -(\theta_0 - \hat{\theta})^2 \left( -\frac{n}{\sigma^2} \right) = \frac{n}{\sigma^2}(\hat{\theta} - \theta_0)^2
    $$
    
    **Step B: Exactness in the Normal Case**
    Because the log-likelihood for the Normal mean is a perfect quadratic (a parabola), the third derivatives are zero. Therefore, the Taylor approximation is **exact**:
    $$
    D = \frac{n}{\sigma^2}(\hat{\theta} - \theta_0)^2
    $$

    **Step C: Equivalence of Forms ($D = \text{Wald} = \text{Score}$)**
    
    Let $\Delta \hat \theta = \hat{\theta} - \theta_0$ be the estimation error. We can rewrite the exact Deviance derived above in terms of the Information ($\mathcal{I}_n = n/\sigma^2$) and the Score at the null ($U_0 = \mathcal{I}_n \Delta \hat \theta$).

    * **Deviance Form:** From Step B, we see directly:
        $$
        D = \left( \frac{n}{\sigma^2} \right) (\hat{\theta} - \theta_0)^2 = \mathcal{I}_n (\Delta \hat \theta)^2
        $$
        
    * **Score Statistic Form:** The Score at the null hypothesis is $U(\theta_0) = \mathcal{I}_n (\hat{\theta} - \theta_0)$.
        $$
        U^T \mathcal{I}^{-1} U = [\mathcal{I}_n \Delta \hat \theta] \cdot [\mathcal{I}_n^{-1}] \cdot [\mathcal{I}_n \Delta \hat \theta] = \Delta \hat \theta \cdot \mathcal{I}_n \cdot \Delta \hat \theta = \mathcal{I}_n (\Delta \hat \theta)^2
        $$

    * **Result:** In the Normal case (perfect quadratic likelihood), the three statistics are identical:
        $$
        \boxed{ D = \Delta \hat \theta^T \mathcal{I}_n \Delta \hat \theta = U_n(\theta_0)^T \mathcal{I}_n^{-1} U_n(\theta_0) }
        $$
    
    **Step D: Conclusion**
    The term inside the Deviance is the standard Z-statistic squared ($Z \sim N(0,1)$):
    $$
    D = \left( \frac{\bar{X}_n - \theta_0}{\sigma/\sqrt{n}} \right)^2 = Z^2 \sim \chi^2_1
    $$
    This illustrates Wilks' theorem: the deviance statistic converges to a Chi-squared distribution (degrees of freedom = 1).

:::

#### Uniform Distribution (Boundary Case)

::: {#exm-uniform-mle-boundary}


Let $X_1, \dots, X_n \overset{iid}{\sim} \operatorname{\text{Unif}}(0, \theta)$. The likelihood is:
$$
L(\theta; \mathbf{x}) = \frac{1}{\theta^n} I(x_{(n)} \le \theta)
$$
This function strictly decreases for $\theta \ge x_{(n)}$ and is zero otherwise. Thus:
$$
\hat{\theta}_{\text{MLE}} = x_{(n)}
$$
Note that the Score equation approach fails here because the support depends on $\theta$, making the log-likelihood discontinuous at the boundary.

:::

## Optimization Algorithms for Finding MLE

When the score equation $\mathbf{U}(\boldsymbol{\theta}) = \mathbf{0}$ does not have a closed-form solution, we must rely on iterative numerical methods to find the Maximum Likelihood Estimator. The Newton-Raphson algorithm is the most common approach, utilizing the curvature of the log-likelihood surface to guide the search for the maximum.


### Newton-Raphson Iteration

::: {.algorithm}

Starting from an initial guess $\boldsymbol{\theta}^{(0)}$, the algorithm updates the estimate of the parameter vector at each step $t$ using the following recurrence relation:

$$
\boldsymbol{\theta}^{(t+1)} = \boldsymbol{\theta}^{(t)} + \left[ \mathbf{J}(\boldsymbol{\theta}^{(t)}) \right]^{-1} \mathbf{U}(\boldsymbol{\theta}^{(t)})
$$

where:

1. **$\mathbf{U}(\boldsymbol{\theta}^{(t)})$** is the score vector (gradient) evaluated at the current estimate.

2. **$\mathbf{J}(\boldsymbol{\theta}^{(t)})$** is the **Observed Fisher Information Matrix**, defined as the negative Hessian of the log-likelihood:

   
   $$
   \mathbf{J}(\boldsymbol{\theta}) = - \nabla^2 \ell(\boldsymbol{\theta}) = - \frac{\partial^2 \ell(\boldsymbol{\theta})}{\partial \boldsymbol{\theta} \partial \boldsymbol{\theta}^T}
   $$

:::


::: proof

### Derivation via Taylor Expansion

The algorithm is derived by approximating the score function $\mathbf{U}(\boldsymbol{\theta})$ with a first-order Taylor expansion around the current estimate $\boldsymbol{\theta}^{(t)}$:

1. **Linear Approximation** \n

   We approximate the score at the next step as:
   
   $$
   \mathbf{U}(\boldsymbol{\theta}^{(t+1)}) \approx \mathbf{U}(\boldsymbol{\theta}^{(t)}) + \nabla \mathbf{U}(\boldsymbol{\theta}^{(t)}) (\boldsymbol{\theta}^{(t+1)} - \boldsymbol{\theta}^{(t)})
   $$

2. **Setting to Zero** \n

   Since we want to find the root where $\mathbf{U}(\hat{\boldsymbol{\theta}}) = \mathbf{0}$, we set the left side to zero:
   
   $$
   \mathbf{0} = \mathbf{U}(\boldsymbol{\theta}^{(t)}) - \mathbf{J}(\boldsymbol{\theta}^{(t)}) (\boldsymbol{\theta}^{(t+1)} - \boldsymbol{\theta}^{(t)})
   $$
   
   Note that $\nabla \mathbf{U}(\boldsymbol{\theta}) = \nabla^2 \ell(\boldsymbol{\theta}) = -\mathbf{J}(\boldsymbol{\theta})$.

3. **Solving for the Update** \n

   Rearranging the terms gives the update rule:
   
   $$
   \mathbf{J}(\boldsymbol{\theta}^{(t)}) (\boldsymbol{\theta}^{(t+1)} - \boldsymbol{\theta}^{(t)}) = \mathbf{U}(\boldsymbol{\theta}^{(t)})
   $$
   
   $$
   \boldsymbol{\theta}^{(t+1)} = \boldsymbol{\theta}^{(t)} + \mathbf{J}(\boldsymbol{\theta}^{(t)})^{-1} \mathbf{U}(\boldsymbol{\theta}^{(t)})
   $$

:::




### Fisher Scoring

::: algorithm
The Newton-Raphson algorithm specifically uses the **Observed Information** $\mathbf{J}(\boldsymbol{\theta})$. If the expected Fisher Information $\mathcal{I}(\boldsymbol{\theta}) = E[\mathbf{J}(\boldsymbol{\theta})]$ is used instead, the method is known as **Fisher Scoring**.


1. **Convergence Speed** \n

   Newton-Raphson typically achieves quadratic convergence near the maximum, making it very fast when the initial guess is good.

2. **Stability** \n

   Because $\mathbf{J}(\boldsymbol{\theta})$ depends on the specific data observed, it can occasionally be non-positive definite in regions far from the MLE, which may cause the algorithm to diverge. Fisher Scoring is often more stable in these cases because $\mathcal{I}(\boldsymbol{\theta})$ is always positive semi-definite for regular families.

:::

### Example: Poisson Regression (MLE)

::: {#exm-poisson-glm}
Consider a GLM where $Y_i \sim \text{Poisson}(\lambda_i)$. We use the canonical link $\log(\lambda_i(\boldsymbol{\beta})) = \mathbf{x}_i^\top \boldsymbol{\beta}$, which implies the mean function is $\lambda_i(\boldsymbol{\beta}) = \exp(\mathbf{x}_i^\top \boldsymbol{\beta})$.

1. **Exponential Family Representation**

    We rewrite the total log-likelihood by grouping terms associated with each $\beta_j$:
     
    $$
    \begin{aligned}
    \ell(\boldsymbol{\beta}) &= \sum_{i=1}^n \left( y_i (\mathbf{x}_i^\top \boldsymbol{\beta}) - \lambda_i(\boldsymbol{\beta}) - \log(y_i!) \right) \\
    &= \sum_{i=1}^n \sum_{j=0}^p y_i x_{ij} \beta_j - \sum_{i=1}^n \exp(\mathbf{x}_i^\top \boldsymbol{\beta}) - \sum_{i=1}^n \log(y_i!) \\
    &= \sum_{j=0}^p \beta_j \underbrace{\left( \sum_{i=1}^n y_i x_{ij} \right)}_{T_j(\mathbf{y})} - \underbrace{\sum_{i=1}^n \lambda_i(\boldsymbol{\beta})}_{A(\boldsymbol{\beta})} + \text{const}
    \end{aligned}
    $$

    This is a multivariate exponential family in **canonical form** where:

    * **Natural Parameters:** The regression coefficients $\boldsymbol{\beta} = (\beta_0, \dots, \beta_p)^\top$.
    * **Sufficient Statistics:** $\mathbf{T}(\mathbf{y}) = \mathbf{X}^\top \mathbf{y}$.
    * **Log-Partition Function:** $A(\boldsymbol{\beta}) = \sum_{i=1}^n \lambda_i(\boldsymbol{\beta})$.

2. **Derivations via Moments of Sufficient Statistics**

    Since we are in canonical form, the Score and Information are simply the derivatives of the log-partition function $A(\boldsymbol{\beta})$.

    * **Score Vector ($\mathbf{U}$):**
        $$
        \mathbf{U}(\boldsymbol{\beta}) = \mathbf{T}(\mathbf{y}) - \nabla A(\boldsymbol{\beta}) = \mathbf{X}^\top \mathbf{y} - \mathbf{X}^\top \boldsymbol{\lambda}(\boldsymbol{\beta}) = \mathbf{X}^\top (\mathbf{y} - \boldsymbol{\lambda}(\boldsymbol{\beta}))
        $$
        where $\boldsymbol{\lambda}(\boldsymbol{\beta}) = (\lambda_1(\boldsymbol{\beta}), \dots, \lambda_n(\boldsymbol{\beta}))^\top$.

    * **Fisher Information ($J$):**
        The information is the Hessian of the log-partition function:
        $$
        J(\boldsymbol{\beta}) = \nabla^2 A(\boldsymbol{\beta}) = \mathbf{X}^\top \mathbf{W}(\boldsymbol{\beta}) \mathbf{X}
        $$
        where $\mathbf{W}(\boldsymbol{\beta}) = \text{diag}(\lambda_1(\boldsymbol{\beta}), \dots, \lambda_n(\boldsymbol{\beta}))$.

   The details of procedure for deriving the above formula for score and fisher information matrix using scalar derivative is given @exm-poisson-score-fisher. 

3. **Newton-Raphson Algorithm for Poisson MLE**

    To find the MLE $\hat{\boldsymbol{\beta}}$, we iterate until convergence:
     
    $$
    \boldsymbol{\beta}^{(t+1)} = \boldsymbol{\beta}^{(t)} + [J(\boldsymbol{\beta}^{(t)})]^{-1} \mathbf{U}(\boldsymbol{\beta}^{(t)})
    $$
     
    Substituting the Poisson-specific forms:
     
    $$
    \boldsymbol{\beta}^{(t+1)} = \boldsymbol{\beta}^{(t)} + (\mathbf{X}^\top \mathbf{W}(\boldsymbol{\beta}^{(t)}) \mathbf{X})^{-1} \mathbf{X}^\top (\mathbf{y} - \boldsymbol{\lambda}(\boldsymbol{\beta}^{(t)}))
    $$
     
    This specific iteration is also known as **Iteratively Reweighted Least Squares (IRLS)**.

:::



```{r}
#| label: fig-poisson-nr-and-fit
#| fig-cap: " Left: Observed data with the true and estimated mean functions; Right: Newton-Raphson path on the log-likelihood surface."
#| echo: true
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5
#| dpi: 300

library(latex2exp)

# 1. Data Generation
set.seed(42)
n <- 50
beta_true <- c(0.5, 1.5)
x_vals <- runif(n, -1, 1)
X_mat <- cbind(1, x_vals)
y_obs <- rpois(n, lambda = exp(X_mat %*% beta_true))

# 2. Log-Likelihood Function for Contours
log_lik <- function(b0, b1) {
  eta <- b0 + b1 * x_vals
  sum(y_obs * eta - exp(eta))
}

# 3. Newton-Raphson Implementation
b_curr <- c(2, 3) # Initial guess
path <- matrix(NA, nrow = 10, ncol = 2)
path[1, ] <- b_curr

for (t in 2:10) {
  lambda <- as.vector(exp(X_mat %*% b_curr))
  W <- diag(lambda)
  U <- t(X_mat) %*% (y_obs - lambda)
  J <- t(X_mat) %*% W %*% X_mat
   
  b_new <- b_curr + solve(J) %*% U
  path[t, ] <- b_new
  b_curr <- b_new
}
path <- na.omit(path)
beta_mle <- path[nrow(path), ]

# --- Plotting Side-by-Side ---
par(mfrow = c(1, 2))


# Plot B: Data and Fitted Curves
plot(x_vals, y_obs, pch = 21, bg = "gray80", col = "gray50",
     xlab = "x", ylab = "y", main = "Poisson Regression Fit")

# Generate smooth lines for lambda
x_smooth <- seq(-1, 1, length.out = 100)
lambda_true <- exp(beta_true[1] + beta_true[2] * x_smooth)
lambda_mle  <- exp(beta_mle[1] + beta_mle[2] * x_smooth)

lines(x_smooth, lambda_true, col = "blue", lwd = 2, lty = 2)
lines(x_smooth, lambda_mle, col = "red", lwd = 2)

legend("topleft", legend = c("True Mean", "MLE Mean", "Observed Data"),
       col = c("blue", "red", "gray50"), lty = c(2, 1, NA), 
       pch = c(NA, NA, 21), pt.bg = "gray80", bty = "n", cex = 0.8)

# Plot A: NR Optimization Path
b0_grid <- seq(-0.5, 2.5, length.out = 50)
b1_grid <- seq(0.5, 3.5, length.out = 50)
z_vals <- outer(b0_grid, b1_grid, Vectorize(log_lik))

contour(b0_grid, b1_grid, z_vals, nlevels = 20, 
        xlab = TeX("$\\beta_0$"), ylab = TeX("$\\beta_1$"),
        main = "Newton-Raphson Path")
lines(path[,1], path[,2], type = "b", col = "red", pch = 19, lwd = 1.5)
points(beta_true[1], beta_true[2], col = "blue", pch = 4, lwd = 2, cex = 1.2)
legend("bottomright", legend = c("NR Path", "True Beta"), 
       col = c("red", "blue"), pch = c(19, 4), bty = "n", cex = 0.8)

```


## Review of Convergence Theorems for Probability



::: {#thm-lln}

### Weak Law of Large Numbers (WLLN)
Let $X_1, \dots, X_n$ be independent and identically distributed (i.i.d.) random variables with mean $E[X_i] = \mu$ and finite variance $\text{Var}(X_i) = \sigma^2 < \infty$.

Then, the sample mean $\bar{X}_n = \frac{1}{n}\sum_{i=1}^n X_i$ converges in probability to $\mu$:
$$
\bar{X}_n \xrightarrow{p} \mu \quad \text{as } n \to \infty
$$
Formal definition: For any $\epsilon > 0$, $\lim_{n \to \infty} P(|\bar{X}_n - \mu| > \epsilon) = 0$.

:::



::: {#thm-clt}

### Central Limit Theorem for IID Cases
Let $X_1, \dots, X_n$ be i.i.d. random variables with mean $E[X_i] = \mu$ and finite variance $0 < \text{Var}(X_i) = \sigma^2 < \infty$.

Then, the random variable $\sqrt{n}(\bar{X}_n - \mu)$ converges in distribution to a normal distribution with mean 0 and variance $\sigma^2$:
$$
\sqrt{n}(\bar{X}_n - \mu) \xrightarrow{d} N(0, \sigma^2)
$$

:::


::: {#thm-lindeberg-feller}

### Lindeberg-Feller CLT (For Non-Identical Distributions)
This variation is crucial for regression analysis (e.g., OLS properties with fixed regressors) where variables are independent but **not** identically distributed.

Let $X_1, \dots, X_n$ be independent random variables with $E[X_i] = \mu_i$ and $\text{Var}(X_i) = \sigma_i^2$.
Define the **average variance** $\tilde{\sigma}_n^2$:
$$
\tilde{\sigma}_n^2 = \frac{1}{n} \sum_{i=1}^n \sigma_i^2
$$

If the **Lindeberg Condition** holds:
For every $\epsilon > 0$,
$$
\lim_{n \to \infty} \frac{1}{n \tilde{\sigma}_n^2} \sum_{i=1}^n E\left[ (X_i - \mu_i)^2 \cdot I\left( |X_i - \mu_i| > \epsilon \sqrt{n \tilde{\sigma}_n^2} \right) \right] = 0
$$

Then the standardized sum converges to a standard normal:
$$
\frac{\sum_{i=1}^n (X_i - \mu_i)}{\sqrt{n \tilde{\sigma}_n^2}} \xrightarrow{d} N(0, 1)
$$

:::

::: {#cor-lindeberg-feller-approx}

### Approximating Distribution for Sample Mean (Non-i.i.d.)

Under the conditions of the Lindeberg-Feller CLT, we can derive the asymptotic distribution for the sample mean $\bar{X}_n = \frac{1}{n}\sum_{i=1}^n X_i$.

Let $\bar{\mu}_n = \frac{1}{n}\sum_{i=1}^n \mu_i$ be the average mean. Note that the denominator in the CLT is simply $\sqrt{n} \tilde{\sigma}_n$.

The standardized sum converges to $N(0,1)$:
$$
\frac{\sum (X_i - \mu_i)}{\sqrt{n \tilde{\sigma}_n^2}} = \frac{n(\bar{X}_n - \bar{\mu}_n)}{\sqrt{n} \tilde{\sigma}_n} = \frac{\sqrt{n}(\bar{X}_n - \bar{\mu}_n)}{\tilde{\sigma}_n} \xrightarrow{d} N(0, 1)
$$

This implies the following **approximate distributions** for large $n$:

1. **For the Sample Mean:**

    $$
    \bar{X}_n \overset{\cdot}{\sim} N\left(\bar{\mu}_n, \frac{\tilde{\sigma}_n^2}{n}\right)
    $$

2. **For the Scaled Difference (Root-n consistency):**

    $$
    \sqrt{n}(\bar{X}_n - \bar{\mu}_n) \overset{\cdot}{\sim} N\left(0, \tilde{\sigma}_n^2\right)
    $$

*Note: If all $X_i$ share the same mean $\mu$, simply replace $\bar{\mu}_n$ with $\mu$.*

:::

::: {#thm-slutsky}

### Slutsky's Theorem
Let $X_n$ and $Y_n$ be sequences of random variables. If $X_n \xrightarrow{d} X$ and $Y_n \xrightarrow{p} c$, where $c$ is a constant, then:

1. **Sum:** $X_n + Y_n \xrightarrow{d} X + c$

2. **Product:** $X_n Y_n \xrightarrow{d} cX$

3. **Quotient:** $X_n / Y_n \xrightarrow{d} X/c$ (provided $c \ne 0$)

:::

::: {#thm-generalized-slutsky}

### Generalized Slutsky's Theorem (Continuous Mapping)

The arithmetic operations in Slutsky's theorem are special cases of a broader property.

Let $X_n \xrightarrow{d} X$ and $Y_n \xrightarrow{p} c$, where $c$ is a constant.
Let $g: \mathbb{R}^2 \to \mathbb{R}^k$ be a function that is **continuous** at every point $(x, c)$ where $x$ is in the support of $X$.

Then:
$$
g(X_n, Y_n) \xrightarrow{d} g(X, c)
$$

This implies that for any "well-behaved" algebraic combination (polynomials, exponentials, etc.) of a sequence converging in distribution and a sequence converging in probability to a constant, the limit behaves as if the constant were substituted directly.

:::

::: {#exm-asymptotic-normal-variance}

### Asymptotic Normality of Sample Variance

Let $X_1, \dots, X_n \overset{iid}{\sim} N(\mu, \sigma^2)$. We wish to derive the asymptotic distribution of the MLE for variance, 

$$\hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \bar{X})^2.$$

1. Algebraic Expansion

    We rewrite the estimator by adding and subtracting the true mean $\mu$:
    $$
    \sum_{i=1}^n (X_i - \bar{X})^2 = \sum_{i=1}^n ((X_i - \mu) - (\bar{X} - \mu))^2
    $$
    $$
    = \sum_{i=1}^n (X_i - \mu)^2 - 2(\bar{X} - \mu)\sum_{i=1}^n(X_i - \mu) + n(\bar{X} - \mu)^2
    $$
    Since $\sum(X_i - \mu) = n(\bar{X} - \mu)$:
    $$
    = \sum_{i=1}^n (X_i - \mu)^2 - n(\bar{X} - \mu)^2
    $$
    Dividing by $n$, we get:
    $$
    \hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \mu)^2 - (\bar{X} - \mu)^2
    $$

2. Scaling by $\sqrt{n}$

    We rearrange to look at the pivotal quantity $\sqrt{n}(\hat{\sigma}^2 - \sigma^2)$:
    $$
    \sqrt{n}(\hat{\sigma}^2 - \sigma^2) = \sqrt{n}\left( \frac{1}{n}\sum_{i=1}^n (X_i - \mu)^2 - \sigma^2 \right) - \sqrt{n}(\bar{X} - \mu)^2
    $$

3. Applying Convergence Theorems


    * **Term 1 (CLT):**
     
        Let $W_i = (X_i - \mu)^2$. Since $X_i \sim N(\mu, \sigma^2)$, $W_i/\sigma^2 \sim \chi^2_1$.
        moments of $W_i$: $E[W_i] = \sigma^2$ and $\text{Var}(W_i) = 2\sigma^4$.
        By the standard CLT:
        $$
        \sqrt{n}(\bar{W} - E[W]) = \sqrt{n}\left( \frac{1}{n}\sum (X_i - \mu)^2 - \sigma^2 \right) \xrightarrow{d} N(0, 2\sigma^4)
        $$

    * **Term 2 (Slutsky):**
     
        Consider the term $\sqrt{n}(\bar{X} - \mu)^2$. We can rewrite this as:
        $$
        \underbrace{\sqrt{n}(\bar{X} - \mu)}_{\xrightarrow{d} N(0, \sigma^2)} \cdot \underbrace{(\bar{X} - \mu)}_{\xrightarrow{p} 0 \text{ by LLN}}
        $$
        By Slutsky's Theorem, the product converges to $Z \cdot 0 = 0$.

**Conclusion:**

Combining the terms:
$$
\sqrt{n}(\hat{\sigma}^2 - \sigma^2) \xrightarrow{d} N(0, 2\sigma^4) - 0 = N(0, 2\sigma^4)
$$

:::

## Asymptotic Theory of Maximized Likelihood

### Consistency of the MLE

#### IID Cases

Consistency establishes that as the sample size grows, the estimator converges in probability to the true parameter value.

::: {#lem-kl-divergence-mle}

### Minimization of KL Divergence at the True Density

Let $f(y|\boldsymbol{\theta}_0)$ be the true density of a random variable $Y$, and let $f(y|\boldsymbol{\theta})$ be any other density in the family. The **Kullback-Leibler (KL) Divergence** (or relative entropy) between the true distribution and a candidate distribution is defined as:

$$
D_{KL}(f(\cdot|\boldsymbol{\theta}_0) \| f(\cdot|\boldsymbol{\theta})) = E_{\boldsymbol{\theta}_0} \left[ \log \left( \frac{f(Y|\boldsymbol{\theta}_0)}{f(Y|\boldsymbol{\theta})} \right) \right]
$$

This quantity satisfies the **Information Inequality**:

$$
D_{KL}(f(\cdot|\boldsymbol{\theta}_0) \| f(\cdot|\boldsymbol{\theta})) \ge 0
$$

with equality if and only if $f(y|\boldsymbol{\theta}) = f(y|\boldsymbol{\theta}_0)$ almost everywhere. This implies that the expected log-likelihood $M(\boldsymbol{\theta}) = E_{\boldsymbol{\theta}_0}[\log f(Y|\boldsymbol{\theta})]$ is uniquely maximized at the true parameter $\boldsymbol{\theta}_0$ (or true density for $Y$).

:::

::: {.proof}

1. **Information Inequality via Jensen's Inequality**

   
   Consider the negative KL divergence, which is the expected log-likelihood ratio. By **Jensen's Inequality**, since the logarithm is a strictly concave function:
   
   $$
   -D_{KL}(f(\cdot|\boldsymbol{\theta}_0) \| f(\cdot|\boldsymbol{\theta})) = E_{\boldsymbol{\theta}_0} \left[ \log \left( \frac{f(Y|\boldsymbol{\theta})}{f(Y|\boldsymbol{\theta}_0)} \right) \right] \le \log E_{\boldsymbol{\theta}_0} \left[ \frac{f(Y|\boldsymbol{\theta})}{f(Y|\boldsymbol{\theta}_0)} \right]
   $$

2. **Evaluating the Expectation**

   
   We evaluate the expectation on the right-hand side:
   
   $$
   E_{\boldsymbol{\theta}_0} \left[ \frac{f(Y|\boldsymbol{\theta})}{f(Y|\boldsymbol{\theta}_0)} \right] = \int_{\mathcal{Y}} \left( \frac{f(y|\boldsymbol{\theta})}{f(y|\boldsymbol{\theta}_0)} \right) f(y|\boldsymbol{\theta}_0) \, dy = \int_{\mathcal{Y}} f(y|\boldsymbol{\theta}) \, dy = 1
   $$

3. **Conclusion**

   
   Substituting the result into the inequality:
   
   $$
   -D_{KL}(f(\cdot|\boldsymbol{\theta}_0) \| f(\cdot|\boldsymbol{\theta})) \le \log(1) = 0 \implies D_{KL}(f(\cdot|\boldsymbol{\theta}_0) \| f(\cdot|\boldsymbol{\theta})) \ge 0
   $$
   
   Because the log function is strictly concave, the equality $E[\log Z] = \log E[Z]$ holds if and only if $Z$ is a constant with probability 1. Here, $Z = f(Y|\boldsymbol{\theta})/f(Y|\boldsymbol{\theta}_0)$, so equality holds if and only if $f(y|\boldsymbol{\theta}) = f(y|\boldsymbol{\theta}_0)$ for almost all $y$.

:::



::: {#thm-mle-consistency}

### Consistency of MLE

Let $Y_1, \dots, Y_n$ be i.i.d. with density $f(y|\boldsymbol{\theta})$. Let $\boldsymbol{\theta}_0$ be the true parameter. Assume the following **Identifiability** condition holds:
   
$$
f(y|\boldsymbol{\theta}) = f(y|\boldsymbol{\theta}_0) \text{ for almost all } y \implies \boldsymbol{\theta} = \boldsymbol{\theta}_0
$$

Under this and other standard regularity conditions (such as compactness of the parameter space $\Theta$), the Maximum Likelihood Estimator $\hat{\boldsymbol{\theta}}_n$ satisfies:

$$
\hat{\boldsymbol{\theta}}_n \xrightarrow{p} \boldsymbol{\theta}_0 \quad \text{as } n \to \infty
$$

:::



::: {#rem-identifiability}

### The Role of Identifiability

Identifiability is a fundamental requirement for consistency because it ensures that different parameter values correspond to different probability distributions.

1. **Uniqueness of the Mapping**

   Mathematically, it means the mapping $\boldsymbol{\theta} \to f(\cdot|\boldsymbol{\theta})$ is one-to-one. If the model were not identifiable, there would exist at least one $\boldsymbol{\theta}_1 \neq \boldsymbol{\theta}_0$ such that $f(y|\boldsymbol{\theta}_1) = f(y|\boldsymbol{\theta}_0)$.
   
2. **Uniqueness of the Maximum**

   In the context of the proof, @lem-kl-divergence-mle shows that the KL divergence is zero if and only if the densities are identical. Identifiability allows us to move from the space of "densities" back to the space of "parameters," ensuring that $\boldsymbol{\theta}_0$ is the **unique** global maximum of the expected log-likelihood function $M(\boldsymbol{\theta})$.
   
3. **Failure of Consistency**

   Without identifiability, the data cannot distinguish between $\boldsymbol{\theta}_0$ and $\boldsymbol{\theta}_1$. The MLE would not converge to a single point but rather to the set of all parameters that produce the true density.

:::

::: {.proof}

1. **The Objective Function**

   
   The MLE maximizes the average log-likelihood:
   
   $$
   M_n(\boldsymbol{\theta}) = \frac{1}{n} \sum_{i=1}^n \log f(Y_i|\boldsymbol{\theta})
   $$
   
   By the **Weak Law of Large Numbers (WLLN)**, for any fixed $\boldsymbol{\theta}$, $M_n(\boldsymbol{\theta})$ converges in probability to its expectation:
   
   $$
   M_n(\boldsymbol{\theta}) \xrightarrow{p} M(\boldsymbol{\theta}) = E_{\boldsymbol{\theta}_0} [\log f(Y|\boldsymbol{\theta})]
   $$

2. **Convergence to the Unique Maximum**

   
   According to @lem-kl-divergence-mle, the limit function $M(\boldsymbol{\theta})$ is maximized when the KL divergence is zero. This happens uniquely at $\boldsymbol{\theta}_0$ due to identifiability. 
   
   Since $\hat{\boldsymbol{\theta}}_n$ is the maximizer of $M_n(\boldsymbol{\theta})$, and $M_n(\boldsymbol{\theta})$ converges uniformly to $M(\boldsymbol{\theta})$, the sequence of maximizers must converge to the maximizer of the limit function:
   
   $$
   \hat{\boldsymbol{\theta}}_n \xrightarrow{p} \boldsymbol{\theta}_0
   $$

:::

```{r}
#| label: fig-likelihood-concentration-normal
#| fig-cap: "Consistency: Concentration of the Average Log-Likelihood for $N(\\theta, 1)$"
#| echo: false
#| fig-align: center
#| fig-width: 7
#| fig-height: 5
#| dpi: 300

library(latex2exp)

# Increase right margin for labels
par(mar = c(5, 5, 2, 8) + 0.1, xpd = TRUE)

# 1. Setup Simulation Parameters
set.seed(123)
theta_0 <- 0
theta_seq <- seq(-3, 3, length.out = 200)

# Expected log-likelihood M(theta) for N(theta, 1) when true theta = 0

# M(theta) = -0.5 * log(2*pi) - 0.5 * E[(X - theta)^2]

# Since E[X^2]=1 and E[X]=0, E[(X-theta)^2] = 1 + theta^2
m_theta <- -0.5 * log(2 * pi) - 0.5 * (1 + theta_seq^2)

# 2. Setup Plot (No axes, no box)
plot(theta_seq, m_theta, type = "n", 
     xlab = "", ylab = "", 
     axes = FALSE, ylim = c(-6, -0.5), xlim = c(-3, 3.5))

# 3. Draw Axis Lines with Arrows
arrows(x0 = -3, y0 = -5.8, x1 = 3.5, y1 = -5.8, length = 0.1, lwd = 1) # x-axis
arrows(x0 = -3, y0 = -5.8, x1 = -3, y1 = -0.8, length = 0.1, lwd = 1) # y-axis

# 4. Generate and Plot Sample Average Log-Likelihoods M_n(theta)
n_val <- c(5, 50)
cols <- c("#EF9A9A", "#B71C1C") # Light red for small n, dark red for larger n

for(i in 1:length(n_val)){
   # Generate data from N(0, 1)
   x_samp <- rnorm(n_val[i], mean = theta_0, sd = 1)
   
   # Calculate M_n(theta) for each theta in the sequence
   # M_n(theta) = (1/n) * sum(log f(x_i | theta))
   m_n <- sapply(theta_seq, function(t) {
      mean(-0.5 * log(2 * pi) - 0.5 * (x_samp - t)^2)
   })
   
   lines(theta_seq, m_n, col = cols[i], lwd = 1.5)
}

# 5. Plot the True Limit M(theta)
lines(theta_seq, m_theta, col = "blue", lwd = 2.5)

# 6. Add Reference Line for True Parameter
segments(theta_0, -5.8, theta_0, m_theta[which.min(abs(theta_seq - theta_0))], lty = 3)

# 7. Labels and Legend
text(3.5, -6.1, TeX("$\\theta$"))
text(-3.8, -0.8, TeX("$M_n(\\theta)$"), adj = 1)
text(theta_0, -6.1, TeX("$\\theta_0$"))

legend(x = 3.2, y = -2,
       legend = c(TeX("$M(\\theta)$"), 
                 TeX("$M_n(\\theta), n=5$"),
                 TeX("$M_n(\\theta), n=50$")),
       col = c("blue", cols[1], cols[2]),
       lty = 1,
       lwd = c(2.5, 1.5, 1.5),
       bty = "n")

```

#### Non-IID Cases

In regression settings, observations are independent but not identically distributed (i.n.i.d.). We generalize the consistency result by requiring that the "average" information accumulates sufficiently.

::: {#thm-mle-consistency-general}

#### Consistency of MLE (General Case)
Let $Y_1, \dots, Y_n$ be independent observations with densities $f_i(y|\boldsymbol{\theta})$ (e.g., depending on covariates $x_i$). Let $\boldsymbol{\theta}_0$ be the true parameter.

Under the following conditions:

1. **Parameter Space:** Compact parameter space $\Theta$.

2. **Identification:** For any $\boldsymbol{\theta} \neq \boldsymbol{\theta}_0$, the average Kullback-Leibler divergence is strictly positive in the limit:

    $$\liminf_{n \to \infty} \frac{1}{n} \sum_{i=1}^n \text{KL}(f_i(\cdot|\boldsymbol{\theta}_0) || f_i(\cdot|\boldsymbol{\theta})) > 0$$

3. **Uniform Convergence:** The log-likelihood satisfies a Uniform Law of Large Numbers (ULLN).

Then $\hat{\boldsymbol{\theta}}_n \xrightarrow{p} \boldsymbol{\theta}_0$.

:::

::: {.proof}

1. **The Objective Function**

   The MLE maximizes the average log-likelihood:
   $$
   M_n(\boldsymbol{\theta}) = \frac{1}{n} \sum_{i=1}^n \log f_i(Y_i|\boldsymbol{\theta})
   $$
   Unlike the i.i.d. case, the terms in the sum have different expectations. We rely on a **WLLN for Independent Variables** (e.g., Chebyshev's WLLN). Provided the variances are bounded, $M_n(\boldsymbol{\theta})$ converges in probability to the **average expectation**:
   $$
   M_n(\boldsymbol{\theta}) \xrightarrow{p} \bar{M}_n(\boldsymbol{\theta}) = \frac{1}{n} \sum_{i=1}^n E_{\boldsymbol{\theta}_0} [\log f_i(Y_i|\boldsymbol{\theta})]
   $$
   *Note: For consistency of the maximizer, we strictly require **Uniform Convergence** over $\Theta$, not just pointwise convergence.*

2. **Identifying the Maximum (Average KL Divergence)**

   We compare the limit function at $\boldsymbol{\theta}$ versus $\boldsymbol{\theta}_0$. Consider the average difference:
   $$
   \bar{M}_n(\boldsymbol{\theta}) - \bar{M}_n(\boldsymbol{\theta}_0) = \frac{1}{n} \sum_{i=1}^n E_{\boldsymbol{\theta}_0} \left[ \log \left( \frac{f_i(Y_i|\boldsymbol{\theta})}{f_i(Y_i|\boldsymbol{\theta}_0)} \right) \right]
   $$
   By Jensen's Inequality applied to *each* term in the sum:
   $$
   E_{\boldsymbol{\theta}_0} \left[ \log \frac{f_i(Y_i|\boldsymbol{\theta})}{f_i(Y_i|\boldsymbol{\theta}_0)} \right] \le \log E_{\boldsymbol{\theta}_0} \left[ \frac{f_i(Y_i|\boldsymbol{\theta})}{f_i(Y_i|\boldsymbol{\theta}_0)} \right] = \log(1) = 0
   $$
   Summing these inequalities implies that $\bar{M}_n(\boldsymbol{\theta})$ is uniquely maximized at $\boldsymbol{\theta}_0$ (provided the identification condition holds). Since the sample function $M_n(\boldsymbol{\theta})$ converges uniformly to this maximized limit function, the argmax $\hat{\boldsymbol{\theta}}_n$ converges to $\boldsymbol{\theta}_0$.

:::

### Asymptotic Normality of the Score Vector

The Score function acts as the "engine" for the normality of the MLE. We treat the I.I.D. and non-I.I.D. cases separately.

#### IID Cases

::: {#thm-score-normality-iid}

#### Normality of Score (I.I.D.)
Let $Y_1, \dots, Y_n$ be i.i.d. with density $f(y|\boldsymbol{\theta})$.
Define the **Fisher Information matrix** for a single observation as the expected outer product of the score:
$$
\mathcal{I}_1(\boldsymbol{\theta}) = E\left[ \left( \nabla_{\boldsymbol{\theta}} \log f(Y_1|\boldsymbol{\theta}) \right) \left( \nabla_{\boldsymbol{\theta}} \log f(Y_1|\boldsymbol{\theta}) \right)^T \right]
$$
Then, the scaled total score vector converges to a Normal distribution:
$$
\frac{1}{\sqrt{n}} \mathbf{U}_n(\boldsymbol{\theta}_0; \mathbf{Y}) \xrightarrow{d} N(\mathbf{0}, \mathcal{I}_1(\boldsymbol{\theta}_0))
$$

:::

::: {.proof}
The total score is the sum of independent score contributions:
$$
\mathbf{U}_n(\boldsymbol{\theta}_0; \mathbf{Y}) = \sum_{i=1}^n \nabla \log f(Y_i|\boldsymbol{\theta}_0) = \sum_{i=1}^n \mathbf{u}_i
$$
From **Bartlett's Identities**, for each observation $i$:

1. **Mean**

   $E[\mathbf{u}_i] = \mathbf{0}$.

2. **Variance**

   $\text{Var}(\mathbf{u}_i) = E[\mathbf{u}_i \mathbf{u}_i^T] = \mathcal{I}_1(\boldsymbol{\theta}_0)$.

Since the terms $\mathbf{u}_i$ are i.i.d. with finite variance, we apply the **Multivariate Central Limit Theorem (Lindeberg-Lévy)**:
$$
\frac{1}{\sqrt{n}} \sum_{i=1}^n \mathbf{u}_i \xrightarrow{d} N(\mathbf{0}, \text{Var}(\mathbf{u}_i)) = N(\mathbf{0}, \mathcal{I}_1(\boldsymbol{\theta}_0))
$$

:::

#### Case B: The Non-I.I.D. Case (Regression Setting)

::: {#thm-score-normality-noniid}

#### Normality of Score (Independent, Non-Identical)
Let $Y_1, \dots, Y_n$ be independent but not necessarily identically distributed (e.g., due to different covariates). Let $\mathcal{I}_n(\boldsymbol{\theta}_0) = \sum_{i=1}^n \text{Var}(\mathbf{u}_i)$ be the total information.

Define the **Average Fisher Information Matrix**:
$$
\bar{\mathcal{I}}_n(\boldsymbol{\theta}_0) = \frac{1}{n} \mathcal{I}_n(\boldsymbol{\theta}_0) = \frac{1}{n} \sum_{i=1}^n E[\mathbf{u}_i \mathbf{u}_i^T]
$$

If the **Lindeberg Condition** is satisfied for the sequence of score vectors, then the standardized score converges to a standard Normal:
$$
\left( \mathcal{I}_n(\boldsymbol{\theta}_0) \right)^{-1/2} \mathbf{U}_n(\boldsymbol{\theta}_0; \mathbf{Y}) \xrightarrow{d} N(\mathbf{0}, \mathbf{I}_p)
$$
*(Here $\mathbf{I}_p$ denotes the Identity matrix).*

**Approximating Distribution:**
If the average information converges to a positive definite limit $\bar{\mathcal{I}}_n \to \mathcal{I}_{\infty}$, we have the following approximation for the scaled average score $\sqrt{n} \bar{\mathbf{U}}_n(\boldsymbol{\theta}_0; \mathbf{Y})$:
$$
\sqrt{n} \bar{\mathbf{U}}_n(\boldsymbol{\theta}_0; \mathbf{Y}) = \frac{1}{\sqrt{n}} \mathbf{U}_n(\boldsymbol{\theta}_0; \mathbf{Y}) \overset{\cdot}{\sim} N(\mathbf{0}, \bar{\mathcal{I}}_n(\boldsymbol{\theta}_0))
$$

:::

::: {.proof}
This is a direct application of the **Lindeberg-Feller Central Limit Theorem**.
The total score $\mathbf{U}_n(\boldsymbol{\theta}_0; \mathbf{Y}) = \sum \mathbf{u}_i$ is a sum of independent random vectors with mean $\mathbf{0}$ and variances $\text{Var}(\mathbf{u}_i)$.
Provided that no single observation's score contribution dominates the sum (the Lindeberg condition), the standardized sum converges to a standard Normal distribution.

:::

### Asymptotic Normality of the MLE

We now transfer the normality from the Score function to the estimator $\hat{\boldsymbol{\theta}}$ using a Taylor expansion (the Delta Method logic).

::: {#def-regularity-conditions}

#### Regularity Conditions for Asymptotic Normality
The following conditions are required to ensure the validity of the Taylor expansion and the convergence of the remainder term:

1. **Interiority:** The true parameter $\boldsymbol{\theta}_0$ lies in the interior of the parameter space $\Theta$.

2. **Smoothness:** The log-likelihood function $\log f(y|\boldsymbol{\theta})$ is three times continuously differentiable with respect to $\boldsymbol{\theta}$.

3. **Boundedness:** The third derivatives are bounded by an integrable function $M(y)$ (to control the Taylor remainder).

4. **Positive Information:** The Fisher Information Matrix $\mathcal{I}_1(\boldsymbol{\theta}_0)$ exists and is non-singular (positive definite).

5. **Interchangeability:** Differentiation and integration can be interchanged for the density $f(y|\boldsymbol{\theta})$.

:::

::: {#thm-mle-normality}

#### Asymptotic Normality of MLE

Under the regularity conditions above, the MLE $\hat{\boldsymbol{\theta}}_n$ satisfies:
$$
\sqrt{n}(\hat{\boldsymbol{\theta}}_n - \boldsymbol{\theta}_0) \xrightarrow{d} N\left(\mathbf{0}, [\mathcal{I}_1(\boldsymbol{\theta}_0)]^{-1}\right)
$$

:::

::: {.proof}

1. **Taylor Expansion of Score Equation**

   We expand the Score function $\mathbf{U}_n(\boldsymbol{\theta})$ around the true parameter $\boldsymbol{\theta}_0$. Since $\hat{\boldsymbol{\theta}}_n$ is the MLE, $\mathbf{U}_n(\hat{\boldsymbol{\theta}}_n) = \mathbf{0}$.
   
   Define the **Observed Information Matrix** as the negative Hessian:
   $$
   \mathbf{J}_n(\boldsymbol{\theta}) = -\nabla^2 \ell_n(\boldsymbol{\theta})
   $$

   ::: {.callout-note appearance="simple"}
   **The Linear Link (Score vs. Parameter Error)**
   The fundamental approximation linking the random Score vector to the estimation error is:
   $$
   \underbrace{\mathbf{U}_n(\hat{\boldsymbol{\theta}}_n)}_{=\mathbf{0}} \approx \mathbf{U}_n(\boldsymbol{\theta}_0) - \mathbf{J}_n(\boldsymbol{\theta}_0) (\hat{\boldsymbol{\theta}}_n - \boldsymbol{\theta}_0)
   $$
   $$
   \implies \quad \mathbf{U}_n(\boldsymbol{\theta}_0) \approx \mathbf{J}_n(\boldsymbol{\theta}_0) (\hat{\boldsymbol{\theta}}_n - \boldsymbol{\theta}_0)
   $$
   :::

2. **Scaling and Inversion**

   Multiply the link equation by $\frac{1}{\sqrt{n}}$ and introduce $n$ to the Observed Information term to stabilize the limits:
   $$
   \frac{1}{\sqrt{n}} \mathbf{U}_n(\boldsymbol{\theta}_0) \approx \left( \frac{1}{n} \mathbf{J}_n(\boldsymbol{\theta}_0) \right) \sqrt{n}(\hat{\boldsymbol{\theta}}_n - \boldsymbol{\theta}_0)
   $$
   Rearranging to solve for the estimator's distribution:
   $$
   \sqrt{n}(\hat{\boldsymbol{\theta}}_n - \boldsymbol{\theta}_0) \approx \left[ \frac{1}{n} \mathbf{J}_n(\boldsymbol{\theta}_0) \right]^{-1} \left[ \frac{1}{\sqrt{n}} \mathbf{U}_n(\boldsymbol{\theta}_0) \right]
   $$

3. **Convergence of Components**

   * **Numerator (Score):** From the I.I.D. Score Normality theorem:
       $$
       \frac{1}{\sqrt{n}} \mathbf{U}_n(\boldsymbol{\theta}_0) \xrightarrow{d} Z \sim N(\mathbf{0}, \mathcal{I}_1(\boldsymbol{\theta}_0))
       $$

   * **Denominator (Observed Info):** By the **WLLN**, the average observed information converges to the expected information:
       $$
       \frac{1}{n} \mathbf{J}_n(\boldsymbol{\theta}_0) = -\frac{1}{n} \sum_{i=1}^n \nabla^2 \log f(Y_i|\boldsymbol{\theta}_0) \xrightarrow{p} -E[\nabla^2 \log f] = \mathcal{I}_1(\boldsymbol{\theta}_0)
       $$

4. **Slutsky's Theorem**

   Combining these via Slutsky's Theorem (matrix inverse is a continuous mapping):
   $$
   \sqrt{n}(\hat{\boldsymbol{\theta}}_n - \boldsymbol{\theta}_0) \xrightarrow{d} [\mathcal{I}_1(\boldsymbol{\theta}_0)]^{-1} Z
   $$
   The variance of this limiting distribution is:
   $$
   \text{Var}([\mathcal{I}_1]^{-1} Z) = \mathcal{I}_1^{-1} \text{Var}(Z) (\mathcal{I}_1^{-1})^T = \mathcal{I}_1^{-1} \mathcal{I}_1 \mathcal{I}_1^{-1} = \mathcal{I}_1^{-1}
   $$
   Thus:
   $$
   \sqrt{n}(\hat{\boldsymbol{\theta}}_n - \boldsymbol{\theta}_0) \xrightarrow{d} N\left( \mathbf{0}, [\mathcal{I}_1(\boldsymbol{\theta}_0)]^{-1} \right)
   $$

:::

```{r}
#| label: fig-score-approx
#| fig-cap: "Linear Approximation of the Score Function: The vertical purple segment represents the Score at the true parameter ($\\Delta U$), while the horizontal green segment shows the estimation error ($\\hat{\\theta}_n - \\theta_0$). The red dashed line indicates the linear approximation ($-J(\\theta_0)$)."
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)

# --- 1. Settings and Data Generation (Cauchy) ---
set.seed(42)
n_sim <- 10
n_obs <- 10
theta_0 <- 10 
gamma <- 1    

# Generate Datasets
datasets <- matrix(rcauchy(n_sim * n_obs, location = theta_0, scale = gamma), 
                   nrow = n_sim, ncol = n_obs)

# Define Score and Hessian Functions
score_cauchy <- function(y, theta) {
  sum( (2 * (y - theta)) / (1 + (y - theta)^2) )
}
hessian_cauchy <- function(y, theta) {
  u <- y - theta
  sum( (2 * (u^2 - 1)) / (1 + u^2)^2 )
}

# --- 2. Analyze Main Dataset ---
y_main <- datasets[1, ]

# Find MLE numerically
mle_res <- optimize(function(th) sum(dcauchy(y_main, location = th, log = TRUE)),
                    interval = c(theta_0 - 5, theta_0 + 5), maximum = TRUE)
theta_hat <- mle_res$maximum

# Calculate Geometry at theta_0
u_true <- score_cauchy(y_main, theta_0)
slope_true <- hessian_cauchy(y_main, theta_0)

# --- 3. Prepare Plot Data ---
theta_seq <- seq(8, 12, length.out = 300)

# Curves Data
plot_data <- data.frame()
for(i in 1:n_sim) {
  y_curr <- datasets[i, ]
  u_vals <- sapply(theta_seq, function(th) score_cauchy(y_curr, th))
  plot_data <- rbind(plot_data, data.frame(theta = theta_seq, U = u_vals, sim_id = as.factor(i), is_main = (i == 1)))
}

# Linear Approximation Data (Tangent Line)
tangent_data <- data.frame(
  theta = theta_seq,
  U_lin = u_true + slope_true * (theta_seq - theta_0)
)

# --- 4. Plotting ---
ggplot() +
  # Axes Reference
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  geom_vline(xintercept = theta_0, color = "black", linetype = "dotted") +
  
  # A. Background Curves (Grey)
  geom_line(data = subset(plot_data, !is_main), 
            aes(x = theta, y = U, group = sim_id), 
            color = "grey80", alpha = 0.5) +
  
  # B. Main Curve (Blue) & Tangent (Red Dashed)
  geom_line(data = subset(plot_data, is_main), aes(x = theta, y = U), 
            color = "blue", size = 1.2) +
  geom_line(data = tangent_data, aes(x = theta, y = U_lin), 
            color = "red", linetype = "dashed", size = 1) +
  
  # C. THICK SEGMENTS for Delta U and Delta Theta
  # Vertical Segment (Delta U): From axis to curve at theta_0
  # Color: Purple
  annotate("segment", x = theta_0, xend = theta_0, y = 0, yend = u_true,
           color = "purple", size = 2, alpha = 0.8) +
  
  # Horizontal Segment (Delta Theta): From theta_0 to theta_hat along the axis
  # Color: Dark Green
  annotate("segment", x = theta_0, xend = theta_hat, y = 0, yend = 0,
           color = "darkgreen", size = 2, alpha = 0.8) +
  
  # D. Points
  geom_point(aes(x = theta_0, y = u_true), color = "red", size = 3) +
  geom_point(aes(x = theta_hat, y = 0), color = "darkgreen", size = 3) +
  
  # E. Labels (Delta labels)
  annotate("text", x = theta_0 + 0.15, y = u_true / 2, 
           label = "Delta~U", parse = TRUE, color = "purple", fontface = "bold", hjust = 0) +
  
  annotate("text", x = (theta_0 + theta_hat)/2, y = -1.0, 
           label = "hat(theta)[n] - theta[0]", parse = TRUE, color = "darkgreen", fontface = "bold") +
  
  # F. Parameter Labels (Revised Position)
  annotate("text", x = theta_hat, y = 1, label = "hat(theta)[n]", parse = TRUE, color = "darkgreen", size = 5) +
  annotate("text", x = theta_0, y = 1, label = "theta[0]", parse = TRUE, color = "red", size = 5) +

  # G. Styling
  coord_cartesian(ylim = c(-10, 10), xlim = c(8.5, 11.5)) +
  labs(title = expression(paste("Score Function Approximation (Cauchy)")),
       subtitle = expression(paste("Linear Link: ", Delta, "U ", approx, " -J ", Delta, "theta")),
       x = expression(theta), y = expression(U(theta))) +
  theme_minimal()

```


### Wilks' Theorem 

#### Wilks' Theorem (Simple Null Hypothesis)

::: {#thm-wilks-simple}

#### Wilks' Theorem (Simple Null Hypothesis)
Consider testing $H_0: \boldsymbol{\theta} = \boldsymbol{\theta}_0$ against $H_1: \boldsymbol{\theta} \ne \boldsymbol{\theta}_0$.

Define the **Likelihood Ratio** $\Lambda_n$ as:
$$
\Lambda_n = \frac{L(\boldsymbol{\theta}_0)}{L(\hat{\boldsymbol{\theta}}_{\text{MLE}})}
$$

Define the **Deviance Statistic** $D_n$ as:
$$
D_n = -2 \log \Lambda_n = 2 [\ell_n(\hat{\boldsymbol{\theta}}) - \ell_n(\boldsymbol{\theta}_0)]
$$

Under the null hypothesis, the Deviance converges to a Chi-squared distribution:
$$
D_n \xrightarrow{d} \chi^2_p
$$
where $p$ is the dimension of $\boldsymbol{\theta}$.

:::

::: {.proof}
We express the likelihood difference $\Delta \ell$ in terms of the Score vector $\mathbf{U}_n(\boldsymbol{\theta}_0)$ and apply the Central Limit Theorem directly to the Score.

1. **Quadratic Approximation of Deviance**

    Expand $\ell_n(\boldsymbol{\theta}_0)$ around the MLE $\hat{\boldsymbol{\theta}}$. Since the gradient at the MLE is zero ($\mathbf{U}_n(\hat{\boldsymbol{\theta}}) = \mathbf{0}$), the first-order term vanishes:
    $$
    \ell_n(\boldsymbol{\theta}_0) \approx \ell_n(\hat{\boldsymbol{\theta}}) - \frac{1}{2} (\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}_0)^T \mathbf{J}_n(\hat{\boldsymbol{\theta}}) (\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}_0)
    $$
    Rearranging for the Deviance ($D_n$):
    $$
    D_n = 2[\ell_n(\hat{\boldsymbol{\theta}}) - \ell_n(\boldsymbol{\theta}_0)] \approx (\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}_0)^T \mathbf{J}_n(\hat{\boldsymbol{\theta}}) (\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}_0)
    $$

2. **Substituting the Score**

    Recall the linear link we established in the Normality proof: $\mathbf{U}_n(\boldsymbol{\theta}_0) \approx \mathbf{J}_n(\boldsymbol{\theta}_0)(\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}_0)$.
    Inverting this relationship gives the parameter error in terms of the Score:
    $$
    (\hat{\boldsymbol{\theta}} - \boldsymbol{\theta}_0) \approx \mathbf{J}_n^{-1} \mathbf{U}_n(\boldsymbol{\theta}_0)
    $$
    Substitute this back into the Deviance equation:
    $$
    D_n \approx (\mathbf{J}_n^{-1} \mathbf{U}_n)^T \mathbf{J}_n (\mathbf{J}_n^{-1} \mathbf{U}_n)
    $$
    $$
    \boxed{D_n \approx \mathbf{U}_n(\boldsymbol{\theta}_0)^T \mathbf{J}_n^{-1} \mathbf{U}_n(\boldsymbol{\theta}_0)}
    $$
    This form (the Score Statistic) shows that the deviance is essentially the squared length of the Score vector, standardized by the Information.

3. **Asymptotic Convergence**

    We rewrite the expression using normalized quantities to apply the limit theorems.

    * **Score:** $\frac{1}{\sqrt{n}} \mathbf{U}_n(\boldsymbol{\theta}_0) \xrightarrow{d} \mathbf{Z} \sim N(\mathbf{0}, \mathcal{I}_1)$.
    * **Information:** $\frac{1}{n} \mathbf{J}_n \xrightarrow{p} \mathcal{I}_1$.

    $$
    D_n \approx \left( \frac{1}{\sqrt{n}} \mathbf{U}_n \right)^T \left( \frac{1}{n} \mathbf{J}_n \right)^{-1} \left( \frac{1}{\sqrt{n}} \mathbf{U}_n \right)
    $$
    Taking the limit:
    $$
    D_n \xrightarrow{d} \mathbf{Z}^T \mathcal{I}_1^{-1} \mathbf{Z}
    $$
    Since $\mathbf{Z} \sim N(\mathbf{0}, \mathcal{I}_1)$, the quadratic form follows a Chi-squared distribution:
    $$
    \mathbf{Z}^T \mathcal{I}_1^{-1} \mathbf{Z} \sim \chi^2_p
    $$

:::


### Summary of Asymptotic Approximations

### Summary of Asymptotic Approximations

In the limit as $n \to \infty$ (or exactly for the Normal distribution), the log-likelihood behaves like a quadratic function. This leads to the following relationships between the Deviance ($D$), Score ($\mathbf{U}$), Error ($\Delta \boldsymbol{\theta}$), and Information ($\mathcal{I}$).

| Relationship | Formula | Intuition (Geometry of Parabola) |
| :--- | :--- | :--- |
| **Deviance vs. Error** | $D \approx \Delta \boldsymbol{\theta}^T \mathcal{I} \Delta \boldsymbol{\theta}$ | The vertical drop ($D$) is proportional to the squared horizontal distance ($\Delta \boldsymbol{\theta}^2$), scaled by curvature ($\mathcal{I}$). This is the **Wald Statistic**. |
| **Score vs. Error** | $\mathbf{U} \approx \mathcal{I} \Delta \boldsymbol{\theta}$ | The slope ($\mathbf{U}$) increases linearly with horizontal distance ($\Delta \boldsymbol{\theta}$). The rate of increase is the curvature ($\mathcal{I}$). |
| **Deviance vs. Score** | $D \approx \mathbf{U}^T \mathcal{I}^{-1} \mathbf{U}$ | The vertical drop ($D$) is proportional to the squared slope ($\mathbf{U}^2$), *divided* by curvature ($\mathcal{I}$). This is the **Score Statistic**. |
| **Information Definition** | $\mathcal{I} \approx -\nabla \mathbf{U}$ | The curvature ($\mathcal{I}$) is the rate of change of the slope ($\mathbf{U}$) (negative Hessian). |
| **Information Approximation** | $\mathcal{I} \approx \sum_{i=1}^n \mathbf{U}_i \mathbf{U}_i^T$ | The curvature ($\mathcal{I}$) is estimated by the "Sum of Squared Gradients" (Outer Products). This relies on Bartlett's 2nd Identity: $E[\mathcal{I}] = \text{Var}(\mathbf{U})$. |


```{tikz}
%| label: fig-asymptotic-relationships-final
%| fig-cap: "Asymptotic Relationships: The Information matrix $\\mathcal{I}$ bridges the spaces. Note the use of the empirical sum of outer products $\\sum \\mathbf{U}_i \\mathbf{U}_i^T$ to estimate curvature."
%| echo: false
%| fig-align: "center"
%| engine.opts:
%|   extra.preamble: "\\usepackage{amsmath}"

\begin{tikzpicture}[
    node distance=4cm,
    auto,
    block/.style={
        rectangle,
        draw=blue!60,
        fill=blue!5,
        very thick,
        minimum size=1.5cm,
        text width=3.5cm,
        align=center,
        rounded corners
    },
    infoblock/.style={
        rectangle,
        draw=red!60,
        fill=red!5,
        very thick,
        minimum size=1.5cm,
        text width=3.5cm,
        align=center,
        rounded corners,
        dashed
    },
    arrow/.style={
        ->,
        >=stealth,
        thick,
        color=black!70
    },
    dashed_arrow/.style={
        ->,
        >=stealth,
        thick,
        dashed,
        color=red!70
    }
]

    % Nodes
    \node[block] (error) at (0, 4) {
        \textbf{Parameter Space} \\
        Error ($\Delta \boldsymbol{\theta}$)
    };

    \node[block] (score) at (7, 4) {
        \textbf{Gradient Space} \\
        Total Score ($\mathbf{U} = \sum \mathbf{U}_i$)
    };

    \node[block] (dev) at (3.5, 0) {
        \textbf{Likelihood Space} \\
        Deviance ($D$)
    };

    \node[infoblock] (info) at (3.5, 7.5) {
        \textbf{Information Space} \\
        Curvature ($\mathcal{I}$)
    };

    % Edges - Main Cycle
    
    % Error -> Score
    \draw[arrow] (error) -- node[midway, above, align=center] {
        $\mathbf{U} \approx \mathcal{I} \Delta \boldsymbol{\theta}$ \\
        (\textcolor{blue}{Multiply by $\mathcal{I}$})
    } (score);

    % Score -> Deviance
    \draw[arrow] (score) to[bend left=20] node[midway, right=0.2cm, align=left] {
        $D \approx \mathbf{U}^T \mathcal{I}^{-1} \mathbf{U}$ \\
        (\textcolor{blue}{Square \& Divide})
    } (dev);

    % Error -> Deviance
    \draw[arrow] (error) to[bend right=20] node[midway, left=0.2cm, align=right] {
        $D \approx \Delta \boldsymbol{\theta}^T \mathcal{I} \Delta \boldsymbol{\theta}$ \\
        (\textcolor{blue}{Square \& Multiply})
    } (dev);

    % Edges - Information Approximations
    
    % Score -> Info
    % Connecting from the "Gradient Space" to "Information" using the sum of squares logic
    \draw[dashed_arrow] (score) to[bend right=30] node[midway, right=0.1cm, align=left] {
        $\mathcal{I} \approx \sum_{i} \mathbf{U}_i \mathbf{U}_i^T$ \\
        (\textcolor{red}{Sum of Squares})
    } (info);

    % Parameter -> Info (Bartlett Identity Concept)
    \draw[dashed_arrow] (error) to[bend left=30] node[midway, left=0.1cm, align=right] {
        Bartlett Identity \\
        $E[\mathcal{I}] = \text{Var}(\mathbf{U})$
    } (info);

\end{tikzpicture}
```

### Generalized Wilks' Theorem (Parameter Partitioning)


::: {#lem-score-decomp}
**Quadratic Form Decomposition of the Score**

Let $\mathbf{U}$ be a partitioned vector and $\mathbf{J}$ be a partitioned symmetric positive-definite matrix:
$$
\mathbf{U} = \begin{pmatrix} \mathbf{U}_{\psi} \\ \mathbf{U}_{\lambda} \end{pmatrix}, \quad
\mathbf{J} = \begin{pmatrix} \mathbf{J}_{\psi\psi} & \mathbf{J}_{\psi\lambda} \\ \mathbf{J}_{\lambda\psi} & \mathbf{J}_{\lambda\lambda} \end{pmatrix}
$$
The quadratic form $\mathbf{U}^T \mathbf{J}^{-1} \mathbf{U}$ can be decomposed into the sum of a marginal term (involving only $\lambda$) and a conditional term (involving $\psi$ adjusted for $\lambda$):
$$
\mathbf{U}^T \mathbf{J}^{-1} \mathbf{U} = \mathbf{U}_{\lambda}^T \mathbf{J}_{\lambda\lambda}^{-1} \mathbf{U}_{\lambda} + \tilde{\mathbf{U}}_{\psi}^T (\mathbf{J}^{\psi\psi}) \tilde{\mathbf{U}}_{\psi}
$$
where:

1. $\mathbf{J}^{\psi\psi} = (\mathbf{J}_{\psi\psi} - \mathbf{J}_{\psi\lambda} \mathbf{J}_{\lambda\lambda}^{-1} \mathbf{J}_{\lambda\psi})^{-1}$ is the top-left block of the inverse matrix $\mathbf{J}^{-1}$ (the inverse of the Schur Complement).

2. $\tilde{\mathbf{U}}_{\psi} = \mathbf{U}_{\psi} - \mathbf{J}_{\psi\lambda} \mathbf{J}_{\lambda\lambda}^{-1} \mathbf{U}_{\lambda}$ is the "effective" score for $\psi$, orthogonalized with respect to $\lambda$.

:::

::: {.proof}

1. Block Matrix Inversion

Using the Banachiewicz inversion formula, the blocks of the inverse matrix $\mathbf{J}^{-1}$ are expressed in terms of the Schur complement $S = \mathbf{J}_{\psi\psi} - \mathbf{J}_{\psi\lambda} \mathbf{J}_{\lambda\lambda}^{-1} \mathbf{J}_{\lambda\psi}$:

$$
\mathbf{J}^{-1} = \begin{pmatrix} \mathbf{J}^{\psi\psi} & \mathbf{J}^{\psi\lambda} \\ \mathbf{J}^{\lambda\psi} & \mathbf{J}^{\lambda\lambda} \end{pmatrix} = \begin{pmatrix} S^{-1} & -S^{-1} \mathbf{K} \\ -\mathbf{K}^T S^{-1} & \mathbf{J}_{\lambda\lambda}^{-1} + \mathbf{K}^T S^{-1} \mathbf{K} \end{pmatrix}
$$
where we define the projection operator $\mathbf{K} = \mathbf{J}_{\psi\lambda} \mathbf{J}_{\lambda\lambda}^{-1}$ for brevity.

2. Algebraic Expansion

Expand the quadratic form $\mathbf{U}^T \mathbf{J}^{-1} \mathbf{U}$ using these blocks:
$$
\begin{aligned}
\mathbf{U}^T \mathbf{J}^{-1} \mathbf{U} &= \mathbf{U}_{\psi}^T S^{-1} \mathbf{U}_{\psi} + 2\mathbf{U}_{\psi}^T (-S^{-1} \mathbf{K}) \mathbf{U}_{\lambda} + \mathbf{U}_{\lambda}^T (\mathbf{J}_{\lambda\lambda}^{-1} + \mathbf{K}^T S^{-1} \mathbf{K}) \mathbf{U}_{\lambda}
\end{aligned}
$$

3. Regrouping

Separate the term that does not involve $S^{-1}$:
$$
\mathbf{U}^T \mathbf{J}^{-1} \mathbf{U} = \mathbf{U}_{\lambda}^T \mathbf{J}_{\lambda\lambda}^{-1} \mathbf{U}_{\lambda} + \left[ \mathbf{U}_{\psi}^T S^{-1} \mathbf{U}_{\psi} - 2\mathbf{U}_{\psi}^T S^{-1} \mathbf{K} \mathbf{U}_{\lambda} + \mathbf{U}_{\lambda}^T \mathbf{K}^T S^{-1} \mathbf{K} \mathbf{U}_{\lambda} \right]
$$

4. Completing the Square

The term in the brackets is a perfect quadratic form $(\mathbf{a} - \mathbf{b})^T S^{-1} (\mathbf{a} - \mathbf{b})$:
$$
\left[ \dots \right] = (\mathbf{U}_{\psi} - \mathbf{K} \mathbf{U}_{\lambda})^T S^{-1} (\mathbf{U}_{\psi} - \mathbf{K} \mathbf{U}_{\lambda})
$$
Substituting back the definitions of $\mathbf{K}$ and $\tilde{\mathbf{U}}_{\psi}$:
$$
(\mathbf{U}_{\psi} - \mathbf{J}_{\psi\lambda} \mathbf{J}_{\lambda\lambda}^{-1} \mathbf{U}_{\lambda})^T S^{-1} (\mathbf{U}_{\psi} - \mathbf{J}_{\psi\lambda} \mathbf{J}_{\lambda\lambda}^{-1} \mathbf{U}_{\lambda}) = \tilde{\mathbf{U}}_{\psi}^T (\mathbf{J}^{\psi\psi}) \tilde{\mathbf{U}}_{\psi}
$$

Thus, the identity is proven.

:::

::: {#thm-wilks-general}

#### Wilks' Theorem for Composite Hypotheses
Let the parameter vector $\boldsymbol{\theta}$ of dimension $p$ be partitioned into parameters of interest $\boldsymbol{\psi}$ (dimension $k$) and nuisance parameters $\boldsymbol{\lambda}$ (dimension $p-k$):
$$
\boldsymbol{\theta} = \begin{pmatrix} \boldsymbol{\psi} \\ \boldsymbol{\lambda} \end{pmatrix}
$$
Consider testing $H_0: \boldsymbol{\psi} = \boldsymbol{\psi}_0$ against $H_1: \boldsymbol{\psi} \neq \boldsymbol{\psi}_0$. The nuisance parameters $\boldsymbol{\lambda}$ are unspecified under both hypotheses.

The Likelihood Ratio Test statistic (Deviance) converges to a Chi-squared distribution with degrees of freedom equal to the dimension of $\boldsymbol{\psi}$:
$$
D_n = 2 [\ell_n(\hat{\boldsymbol{\theta}}) - \ell_n(\hat{\boldsymbol{\theta}}_0)] \xrightarrow{d} \chi^2_k
$$

:::

::: {.proof}

**Overview of the proof:** We expand the quadratic approximation of the likelihood and partition the information matrix to isolate the contribution of the parameters of interest.

1. **Quadratic Approximation**

    From the simple Wilks' theorem, we know that twice the log-likelihood difference approximates a quadratic form based on the Score vector $\mathbf{U}$ and Information matrix $J$ at the true parameter:
    $$
    2[\ell_n(\hat{\boldsymbol{\theta}}) - \ell_n(\boldsymbol{\theta}^*)] \approx \mathbf{U}^T J^{-1} \mathbf{U}
    $$

2. **Partitioning**

    Partition the Score vector and Information matrix according to $(\boldsymbol{\psi}, \boldsymbol{\lambda})$:
    $$
    \mathbf{U} = \begin{pmatrix} \mathbf{U}_{\psi} \\ \mathbf{U}_{\lambda} \end{pmatrix}, \quad
    J = \begin{pmatrix} J_{\psi\psi} & J_{\psi\lambda} \\ J_{\lambda\psi} & J_{\lambda\lambda} \end{pmatrix}
    $$

3. **Decomposition of the Full Model ($H_1$)**

    The fit of the full model uses the inverse of the full information matrix. Using the **Schur Complement** identity for block inversion, the total quadratic form decomposes into two orthogonal terms:
    $$
    \underbrace{\mathbf{U}^T J^{-1} \mathbf{U}}_{\text{Total Fit}} = \underbrace{\mathbf{U}_{\lambda}^T J_{\lambda\lambda}^{-1} \mathbf{U}_{\lambda}}_{\text{Nuisance Fit}} + \underbrace{\tilde{\mathbf{U}}_{\psi}^T (J^{\psi\psi}) \tilde{\mathbf{U}}_{\psi}}_{\text{Interest Fit (Adjusted)}}
    $$

    * The first term represents the variation explained by $\boldsymbol{\lambda}$ ignoring $\boldsymbol{\psi}$.
    * The second term represents the variation explained by $\boldsymbol{\psi}$ after accounting for $\boldsymbol{\lambda}$.

4. **Fit of the Reduced Model ($H_0$)**

    Under $H_0$, $\boldsymbol{\psi}$ is fixed. The model only maximizes over $\boldsymbol{\lambda}$. Consequently, the quadratic approximation for the reduced model only involves the $\lambda$-block:
    $$
    2[\ell_n(\hat{\boldsymbol{\theta}}_0) - \ell_n(\boldsymbol{\theta}^*)] \approx \mathbf{U}_{\lambda}^T J_{\lambda\lambda}^{-1} \mathbf{U}_{\lambda}
    $$

5. **The Difference (Deviance)**

    Subtracting the reduced fit from the full fit cancels out the nuisance term:
    $$
    \begin{aligned}
    D_n &= 2[\ell_n(\hat{\boldsymbol{\theta}}) - \ell_n(\boldsymbol{\theta}^*)] - 2[\ell_n(\hat{\boldsymbol{\theta}}_0) - \ell_n(\boldsymbol{\theta}^*)] \\
    &\approx \left( \text{Nuisance Fit} + \text{Interest Fit} \right) - \left( \text{Nuisance Fit} \right) \\
    &= \tilde{\mathbf{U}}_{\psi}^T (J^{\psi\psi}) \tilde{\mathbf{U}}_{\psi}
    \end{aligned}
    $$

6. **Distribution**

    The remaining term is a quadratic form of the $k$-dimensional vector $\tilde{\mathbf{U}}_{\psi}$. Since $\mathbf{U}$ is asymptotically Normal, this quadratic form follows a Chi-squared distribution with degrees of freedom equal to the dimension of $\boldsymbol{\psi}$:
    $$
    D_n \xrightarrow{d} \chi^2_k
    $$

:::


## Examples


### Asympotic Distributions of MLE of Normal Sample

::: {#exm-normal-mle-unknown-variance}

#### Example: Normal Distribution with Unknown Variance ($N(\mu, \sigma^2)$)

Let $X_1, \dots, X_n \overset{iid}{\sim} N(\mu, \sigma^2)$. The parameter vector is $\boldsymbol{\theta} = (\mu, \sigma^2)^T$. We assume both parameters are unknown.

**1. Derivation of Key Quantities**

* **Log-Likelihood:**
    $$
    \ell(\boldsymbol{\theta}) = -\frac{n}{2}\log(2\pi) - \frac{n}{2}\log(\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n (x_i - \mu)^2
    $$

* **Score Vector $\mathbf{U}(\boldsymbol{\theta})$:**
    Gradient w.r.t $\mu$ and $\sigma^2$:
    $$
    \mathbf{U}(\boldsymbol{\theta}) = \begin{pmatrix} \frac{\partial \ell}{\partial \mu} \\ \frac{\partial \ell}{\partial \sigma^2} \end{pmatrix} = \begin{pmatrix} \frac{1}{\sigma^2}\sum (x_i - \mu) \\ -\frac{n}{2\sigma^2} + \frac{1}{2\sigma^4}\sum (x_i - \mu)^2 \end{pmatrix}
    $$

* **Maximum Likelihood Estimator ($\hat{\boldsymbol{\theta}}$):**
    Solving $\mathbf{U}(\hat{\boldsymbol{\theta}}) = \mathbf{0}$:
    $$
    \hat{\mu} = \bar{X}, \quad \hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \bar{X})^2
    $$

* **Fisher Information Matrix $\mathcal{I}_n(\boldsymbol{\theta})$:**
    The matrix of negative expected second derivatives (note that the cross-terms are zero, implying orthogonality of $\mu$ and $\sigma^2$):
    $$
    \mathcal{I}_n(\boldsymbol{\theta}) = \begin{pmatrix} \frac{n}{\sigma^2} & 0 \\ 0 & \frac{n}{2\sigma^4} \end{pmatrix}
    $$

**2. Asymptotic Properties**

Based on the general theory, we state the asymptotic distributions for the MLE, Score, and Deviance (testing the full vector $H_0: \boldsymbol{\theta} = \boldsymbol{\theta}_0$).

1.  **Estimator Normality:**
    $$
    \hat{\boldsymbol{\theta}} \overset{a}{\sim} N_2\left( \boldsymbol{\theta}, \mathcal{I}_n^{-1} \right) = N_2\left( \begin{pmatrix} \mu \\ \sigma^2 \end{pmatrix}, \begin{pmatrix} \frac{\sigma^2}{n} & 0 \\ 0 & \frac{2\sigma^4}{n} \end{pmatrix} \right)
    $$

2.  **Score Normality:**
    $$
    \mathbf{U}(\boldsymbol{\theta}) \overset{a}{\sim} N_2\left( \mathbf{0}, \mathcal{I}_n \right) = N_2\left( \begin{pmatrix} 0 \\ 0 \end{pmatrix}, \begin{pmatrix} \frac{n}{\sigma^2} & 0 \\ 0 & \frac{n}{2\sigma^4} \end{pmatrix} \right)
    $$

3.  **Deviance (Wilks' Theorem):**
    For testing $H_0: (\mu, \sigma^2) = (\mu_0, \sigma_0^2)$ vs $H_1: \text{unrestricted}$:
    $$
    D = 2[\ell(\hat{\boldsymbol{\theta}}) - \ell(\boldsymbol{\theta}_0)] \xrightarrow{d} \chi^2_2
    $$

**3. Analytical Verification**

We check if these asymptotic approximations match the exact finite-sample properties.

* **For the Mean ($\mu$):**
    * **Score:** $U_\mu = \frac{n}{\sigma^2}(\bar{X} - \mu)$. Since $\bar{X}$ is Normal, $U_\mu$ is **exactly Normal** for any $n$.
    * **MLE:** $\hat{\mu} = \bar{X}$ is **exactly Normal** for any $n$.

* **For the Variance ($\sigma^2$):**
    * **Score:** $U_{\sigma^2} = \frac{1}{2\sigma^4} \left( \sum (X_i - \mu)^2 - n\sigma^2 \right)$. The term $\sum (X_i - \mu)^2$ follows a scaled Chi-squared distribution ($\sigma^2 \chi^2_n$).
    * **Verification:** The Score is **not exactly Normal** (it is a shifted Gamma/Chi-squared). However, by the Central Limit Theorem, as $n \to \infty$, a Chi-squared random variable converges to Normality. Thus, the property holds **asymptotically**.

* **For the Deviance ($D$):**
    * The Deviance involves sums of squared errors and logarithms of sums of squares. It is **not exactly Chi-squared** for finite $n$ (unlike the known-variance case). It converges to $\chi^2_2$ as $n$ increases.

**4. Simulation Verification**

We simulate 5000 datasets with $n=100$. We calculate the standardized MLEs, standardized Scores, and the Deviance to check convergence.

```{r}
#| label: fig-normal-unknown-var-sim
#| fig-cap: "Simulation of Normal (Unknown Variance) Asymptotics (n=100). Left/Center-Left: Standardized MLEs. Center-Right: Standardized Score for Variance. Right: Deviance (2 df)."
#| warning: false
#| message: false

set.seed(2024)
n_sim <- 5000
n <- 100
mu_true <- 5
sigma_sq_true <- 4 # sigma = 2

# Expected Fisher Information Terms
I_mu <- n / sigma_sq_true
I_sigma2 <- n / (2 * sigma_sq_true^2)

# Storage
z_mle_mu <- numeric(n_sim)
z_mle_sigma2 <- numeric(n_sim)
z_score_sigma2 <- numeric(n_sim)
deviance_vals <- numeric(n_sim)

for(i in 1:n_sim) {
  x <- rnorm(n, mean = mu_true, sd = sqrt(sigma_sq_true))
  
  # 1. Calculate MLEs
  mu_hat <- mean(x)
  # MLE for variance uses n, not n-1
  sigma2_hat <- mean((x - mu_hat)^2) 
  
  # 2. Standardized MLEs (Wald Statistics)
  z_mle_mu[i] <- (mu_hat - mu_true) * sqrt(I_mu)
  z_mle_sigma2[i] <- (sigma2_hat - sigma_sq_true) * sqrt(I_sigma2)
  
  # 3. Score for Sigma^2 at Truth
  # U_sigma2 = -n/(2*sigma^2) + sum(x-mu)^2 / (2*sigma^4)
  rss_true <- sum((x - mu_true)^2)
  u_sigma2 <- -n/(2*sigma_sq_true) + rss_true/(2*sigma_sq_true^2)
  z_score_sigma2[i] <- u_sigma2 / sqrt(I_sigma2)
  
  # 4. Deviance (Likelihood Ratio)
  # ll(theta) = -n/2*log(2pi) - n/2*log(sigma^2) - RSS/(2*sigma^2)
  ll_hat <- -n/2*log(2*pi) - n/2*log(sigma2_hat) - n/2 # Second term simplifies
  ll_true <- -n/2*log(2*pi) - n/2*log(sigma_sq_true) - rss_true/(2*sigma_sq_true)
  
  deviance_vals[i] <- 2 * (ll_hat - ll_true)
}

# --- Plotting ---
par(mfrow = c(2, 2))

# Plot 1: MLE for Mu (Exact Normal)
hist(z_mle_mu, breaks = 30, freq = FALSE, col = "lightblue",
     main = expression(Std. ~ hat(mu)), xlab = "Z", border = "white")
curve(dnorm(x), add = TRUE, col = "red", lwd = 2)

# Plot 2: MLE for Sigma^2 (Asymptotic Normal)
hist(z_mle_sigma2, breaks = 30, freq = FALSE, col = "lightgreen",
     main = expression(Std. ~ hat(sigma)^2), xlab = "Z", border = "white")
curve(dnorm(x), add = TRUE, col = "blue", lwd = 2)

# Plot 3: Score for Sigma^2 (Asymptotic Normal)
hist(z_score_sigma2, breaks = 30, freq = FALSE, col = "plum1",
     main = expression(Std. ~ U[sigma^2]), xlab = "Z", border = "white")
curve(dnorm(x), add = TRUE, col = "darkred", lwd = 2)

# Plot 4: Deviance (Asymptotic Chi-sq with 2 df)
hist(deviance_vals, breaks = 30, freq = FALSE, col = "wheat",
     main = "Deviance", xlab = "D", border = "white")
curve(dchisq(x, df = 2), add = TRUE, col = "darkgreen", lwd = 2)
```

:::

### Asymptotic Distributions of Poisson Regression

::: {#exm-pr-asym}

#### Simulation Study: Poisson Regression Asymptotics

In this example, we verify the asymptotic normality of the Maximum Likelihood Estimator (MLE) and the Chi-squared distribution of the Deviance statistic using a simulation study.

**1. Simulation Setup**

We consider a Generalized Linear Model (GLM) with a Poisson response and a canonical log-link function:
$$
Y_i \sim \text{Poisson}(\lambda_i), \quad \log(\lambda_i) = \beta_0 + \beta_1 x_i
$$

* **True Parameters:** We set the true coefficient vector to $\boldsymbol{\beta} = (0.5, 1.5)^\top$.
* **Covariates:** We generate $n=200$ covariate values $x_i$ from a Uniform(0, 1) distribution. These remain fixed across all simulations.
* **Data Generation:** For each simulation iteration, we generate response vector $\mathbf{y}$ using the true mean $\lambda_i = \exp(0.5 + 1.5 x_i)$.

**2. Model Fitting and Statistics**

For each of the `n_sim = 2000` generated datasets, we perform the following:

1.  **Fit the Model:** We estimate $\hat{\boldsymbol{\beta}}$ using the `glm()` function in R.
2.  **Calculate Standardized MLEs (Wald Statistic):**
    $$
    Z_{\hat{\beta}_j} = \frac{\hat{\beta}_j - \beta_{j, \text{true}}}{\text{SE}(\hat{\beta}_j)}
    $$
    The standard errors are derived from the inverse of the Fisher Information matrix, $\mathcal{I}^{-1} = (\mathbf{X}^\top \mathbf{W} \mathbf{X})^{-1}$.
3.  **Calculate Standardized Score:**
    $$
    Z_{U_j} = \frac{U_j(\boldsymbol{\beta}_{\text{true}})}{\sqrt{\text{Var}(U_j)}}
    $$
    The score is evaluated at the *true* parameter values.
4.  **Calculate Deviance:**
    $$
    D = 2 \left( \ell(\hat{\boldsymbol{\beta}}) - \ell(\boldsymbol{\beta}_{\text{true}}) \right)
    $$
    This statistic compares the fitted model to the true model. According to Wilks' theorem, this should follow a $\chi^2_p$ distribution (where $p=2$ is the number of parameters).

**3. Simulation Code and Results**

The histograms below overlay the theoretical asymptotic densities (Red/Blue lines) on top of the empirical histograms. The close alignment confirms that even for a moderate sample size of $n=200$, the asymptotic approximations are highly accurate.

```{r}
#| label: fig-poisson-sim-asymptotics
#| fig-cap: "Asymptotic distributions for Poisson Regression (n=200). Top Row: Standardized MLEs for Intercept and Slope (Normal). Bottom Left: Standardized Score for Slope (Normal). Bottom Right: Deviance Statistic (Chi-squared with 2 df)."
#| echo: true
#| warning: false

library(latex2exp)

# 1. Setup
set.seed(42)
n_sim <- 2000
n <- 200
beta_true <- c(0.5, 1.5)

# Generate fixed covariates
x_cov <- runif(n, 0, 1)
X_mat <- cbind(1, x_cov) # Design matrix

# True mean vector and Fisher Information at Truth
lambda_true <- exp(X_mat %*% beta_true)
W <- diag(as.vector(lambda_true))
I_fisher <- t(X_mat) %*% W %*% X_mat
inv_I <- solve(I_fisher)

# Storage vectors
z_mle_b0 <- numeric(n_sim)
z_mle_b1 <- numeric(n_sim)
z_score_b1 <- numeric(n_sim)
deviance_vals <- numeric(n_sim)

# 2. Simulation Loop
for(i in 1:n_sim) {
  # Generate response from true model
  y_sim <- rpois(n, lambda = lambda_true)
  
  # Fit GLM
  fit <- glm(y_sim ~ x_cov, family = poisson)
  beta_hat <- coef(fit)
  
  # A. Standardized MLEs (Wald)
  # Uses the TRUE information matrix for standardization to check theory strictly
  z_mle_b0[i] <- (beta_hat[1] - beta_true[1]) / sqrt(inv_I[1,1])
  z_mle_b1[i] <- (beta_hat[2] - beta_true[2]) / sqrt(inv_I[2,2])
  
  # B. Standardized Score at Truth
  # U = X^T (Y - lambda)
  U <- t(X_mat) %*% (y_sim - lambda_true)
  z_score_b1[i] <- U[2] / sqrt(I_fisher[2,2])
  
  # C. Deviance (Likelihood Ratio)
  # D = 2 * (ll_hat - ll_true)
  ll_hat <- as.numeric(logLik(fit))
  ll_true <- sum(dpois(y_sim, lambda_true, log = TRUE))
  deviance_vals[i] <- 2 * (ll_hat - ll_true)
}

# 3. Visualization (2x2 Grid)
par(mfrow = c(2, 2))

# Plot 1: MLE Beta 0
hist(z_mle_b0, breaks = 30, freq = FALSE, col = "lightblue", 
     main = TeX("Std. MLE $\\hat{\\beta}_0$"), xlab = "Z", border = "white")
curve(dnorm(x), add = TRUE, col = "red", lwd = 2)

# Plot 2: MLE Beta 1
hist(z_mle_b1, breaks = 30, freq = FALSE, col = "lightblue", 
     main = TeX("Std. MLE $\\hat{\\beta}_1$"), xlab = "Z", border = "white")
curve(dnorm(x), add = TRUE, col = "red", lwd = 2)

# Plot 3: Score U_1 (Slope)
hist(z_score_b1, breaks = 30, freq = FALSE, col = "lightgreen", 
     main = TeX("Std. Score $U(\\beta_1)$"), xlab = "Z", border = "white")
curve(dnorm(x), add = TRUE, col = "blue", lwd = 2, lty = 2)

# Plot 4: Deviance
hist(deviance_vals, breaks = 30, freq = FALSE, col = "lightpink", 
     main = "Deviance Statistic", xlab = "D", border = "white")
curve(dchisq(x, df = 2), add = TRUE, col = "darkgreen", lwd = 2)
```

:::

## MLE and Penalized Likelihood in Deep Learning

In deep learning, the objective function is typically the negative log-likelihood of the dataset, often supplemented by a penalty term (regularization) to prevent overfitting. While the theoretical foundations of Maximum Likelihood Estimation remain the same, the practical application in high-dimensional neural networks faces significant computational challenges that necessitate specialized optimization strategies.

### The Optimization Challenge

In traditional statistical models, $\ell(\boldsymbol{\theta})$ is often globally concave, allowing algorithms like Newton-Raphson to converge reliably. Deep learning models, however, present a different landscape:

1. **Non-convexity** \n

   The log-likelihood surface $\ell(\boldsymbol{\theta})$ is highly non-convex with numerous local minima, saddle points, and plateaus.
   
   
2. **High Dimensionality** \n

   The parameter vector $\boldsymbol{\theta}$ can contain millions or billions of weights ($p \gg 10^6$). This makes the storage and inversion of the Observed Information Matrix $\mathbf{J}(\boldsymbol{\theta})$ (the $p \times p$ Hessian) computationally intractable, as its storage requires $O(p^2)$ memory.

### Penalized Likelihood (Regularization)

To improve generalization, we often optimize a **penalized log-likelihood**:

$$
\ell_{pen}(\boldsymbol{\theta}) = \ell(\boldsymbol{\beta}) - \Omega(\boldsymbol{\theta})
$$

where $\Omega(\boldsymbol{\theta})$ is a penalty function. Common examples include:

1. **$L_2$ Regularization (Weight Decay):** $\Omega(\boldsymbol{\theta}) = \frac{\lambda}{2} \|\boldsymbol{\theta}\|^2$, which corresponds to a Gaussian prior in a Bayesian framework.
2. **$L_1$ Regularization (Lasso):** $\Omega(\boldsymbol{\theta}) = \lambda \|\boldsymbol{\theta}\|_1$, which encourages sparsity in the network weights.

### Purely Gradient-Based Methods (O(p) Complexity)

Because the Hessian is too large to compute, the most common approach is to use methods that only require the Score vector $\mathbf{U}(\boldsymbol{\theta})$. These methods have a memory complexity of $O(p)$, making them feasible for large-scale problems.

1. **Stochastic Gradient Descent (SGD)** \n

   Instead of the full score, SGD uses a "mini-batch" estimate $\mathbf{U}_{batch}$. The update is:
   $$
   \boldsymbol{\theta}^{(t+1)} = \boldsymbol{\theta}^{(t)} + \alpha \mathbf{U}_{batch}(\boldsymbol{\theta}^{(t)})
   $$
   While simple, SGD treats all dimensions equally, which is inefficient if the likelihood surface is much steeper in some directions than others.

2. **Momentum-Based Methods** \n

   These maintain a moving average of past scores to dampen oscillations and accelerate progress in consistent directions, effectively acting as a low-pass filter on the noisy stochastic gradients.

### Avoiding the Hessian via Score Covariance

Modern optimizers leverage the **Second Moment Identity** (Bartlett's Identity) to capture curvature information without the $O(p^2)$ cost of the Hessian.

::: {#thm-fisher-approximation}

### Fisher Information Approximation and the Empirical Fisher

In the context of high-dimensional optimization, the identity $\mathcal{I}(\boldsymbol{\theta}) = \text{Cov}_{\boldsymbol{\theta}_0}(\mathbf{U}) = E_{\boldsymbol{\theta}_0}[\mathbf{J}]$ is not just a theoretical curiosity, but a computational lifeline. 

1. **Justification via the Second Moment** \n

   Recall that for the score vector $\mathbf{U}$, we have $E_{\boldsymbol{\theta}_0}[\mathbf{U}] = \mathbf{0}$. Therefore, the covariance matrix simplifies to the second moment:
   
   $$
   \text{Cov}(\mathbf{U}) = E_{\boldsymbol{\theta}_0}[\mathbf{U}\mathbf{U}^\top] - E_{\boldsymbol{\theta}_0}[\mathbf{U}]E_{\boldsymbol{\theta}_0}[\mathbf{U}]^\top = E_{\boldsymbol{\theta}_0}[\mathbf{U}\mathbf{U}^\top]
   $$
   
   Under Bartlett's Second Identity, this expectation is exactly equal to the expected Observed Information (the expected negative Hessian), $E_{\boldsymbol{\theta}_0}[\mathbf{J}]$.

2. **The Empirical Fisher Estimate** \n

   In deep learning, we cannot compute the expectation over all possible data. Instead, we use the **Empirical Fisher Information Matrix** ($\hat{\mathcal{I}}$), which replaces the population expectation with the sample average across $n$ observations:
   
   $$
   \hat{\mathcal{I}} = \frac{1}{n} \sum_{i=1}^n \left( \nabla_{\boldsymbol{\theta}} \log f(y_i | \boldsymbol{\theta}) \right) \left( \nabla_{\boldsymbol{\theta}} \log f(y_i | \boldsymbol{\theta}) \right)^\top
   $$
   
   This estimate is constructed solely from **first-order gradients** (the score for each data point).

3. **Handling High Dimensionality ($p$)** \n

   Since $\hat{\mathcal{I}}$ is a $p \times p$ matrix, storing it is $O(p^2)$. For a model with 1 million parameters, the matrix would have $10^{12}$ elements. To handle this, deep learning optimizers typically use the **Diagonal Empirical Fisher**:
   
   $$
   \text{diag}(\hat{\mathcal{I}}) = \frac{1}{n} \sum_{i=1}^n [ \mathbf{U}_i \odot \mathbf{U}_i ]
   $$
   
   where $\odot$ is the Hadamard (element-wise) product. This reduces the complexity to $O(p)$.

4. **Why use the Score Covariance instead of the Hessian?** \n

   Computing the diagonal of the Hessian $\mathbf{J}$ directly (the second derivatives $\frac{\partial^2 \ell}{\partial \theta_j^2}$) is significantly more expensive than computing the gradient. Most deep learning frameworks are optimized for "backpropagation" (first derivatives). By using the squared gradients to approximate the diagonal of the Hessian, we get "second-order-like" curvature information for the price of first-order computation.

:::



::: {#rem-empirical-fisher-caveat}

### A Note on the Approximation
While $\text{Cov}(\mathbf{U}) = E[\mathbf{J}]$ holds at the true parameter $\boldsymbol{\theta}_0$, it may not hold perfectly during the early stages of training when $\boldsymbol{\theta}$ is far from the truth. In these regions, the empirical Fisher is technically an approximation of the *variability* of the gradients, which often serves as a robust proxy for the *sensitivity* (curvature) of the loss surface.

:::

### Adaptive Methods: Diagonal Covariance Scaling

Algorithms like **RMSProp** and **Adam** avoid the $p \times p$ matrix entirely by assuming the Fisher Information is approximately diagonal. This reduces the memory cost back to $O(p)$.

1. **RMSProp** \n

   It maintains a running average of the squared scores for each parameter $j$:
   $$
   v_j^{(t)} = \gamma v_j^{(t-1)} + (1-\gamma) (U_j^{(t)})^2
   $$
   Since $(U_j)^2$ is a stochastic estimate of the variance $\text{Var}(U_j) = \mathcal{I}_{jj}$, the term $\sqrt{v_j}$ approximates the local curvature (sensitivity) along that coordinate.

2. **Adam (Adaptive Moment Estimation)** \n

   Adam combines the concepts of momentum and RMSProp by maintaining moving averages of both the first and second moments of the score vector. It scales the update by the inverse square root of the estimated variance:
   
   $$
   \theta_j^{(t+1)} = \theta_j^{(t)} + \alpha \frac{\hat{m}_j}{\sqrt{\hat{v}_j} + \epsilon}
   $$
   
   Here, the terms $m_j$ and $v_j$ are defined as:
   
   * **The First Moment ($m_j$):** An exponentially decaying average of the score $U_j$, which acts as momentum to smooth out noisy gradients:
     $$
     m_j^{(t)} = \beta_1 m_j^{(t-1)} + (1 - \beta_1) U_j^{(t)}
     $$
   
   * **The Second Moment ($v_j$):** An exponentially decaying average of the squared score $(U_j)^2$:
     $$
     v_j^{(t)} = \beta_2 v_j^{(t-1)} + (1 - \beta_2) (U_j^{(t)})^2
     $$
     Following the logic of @thm-fisher-approximation, $v_j$ serves as an estimate of the diagonal of the Fisher Information Matrix ($\mathcal{I}_{jj}$).
   
   * **Bias Correction:** Because $m$ and $v$ are typically initialized at zero, they are biased toward zero during the initial steps. Adam uses the bias-corrected estimates $\hat{m}_j = m_j / (1 - \beta_1^t)$ and $\hat{v}_j = v_j / (1 - \beta_2^t)$ for the final update.

   This "preconditioning" flattens the likelihood surface by normalizing the step size for each parameter. Dimensions with high gradient variance (high $\mathcal{I}_{jj}$) receive smaller updates, while dimensions with low variance receive larger updates, effectively performing a coordinate-wise Newton-Raphson step using the Score Covariance as a proxy for the Hessian.



::: {#rem-adam-interpretation}

### Adam as Quasi-Newton
From a statistical perspective, Adam can be viewed as an approximation of the Fisher Scoring algorithm. Instead of calculating the full expected information matrix and its inverse, Adam uses the moving average of squared gradients to estimate the diagonal entries of the information matrix in an online fashion. This allows the algorithm to adapt to the local geometry of the log-likelihood surface at a minimal $O(p)$ computational cost.

:::

::: {#rem-natural-gradient}

### Natural Gradient and K-FAC
The most sophisticated application of this theory is **Natural Gradient Descent**. To handle large $p$, methods like **K-FAC (Kronecker-factored Approximate Curvature)** approximate the Fisher Information Matrix as a product of smaller matrices. This allows for a more "Hessian-like" update that accounts for correlations between parameters while remaining computationally efficient.

:::

## Appendix: Derivation of Score and Information in Poisson Regression

:::: {#exm-poisson-score-fisher}

In a Poisson Generalized Linear Model with a canonical link, the mean is modeled as $\lambda_i(\boldsymbol{\beta}) = \exp(\mathbf{x}_i^\top \boldsymbol{\beta})$. The derivation of the Score vector and the Fisher Information matrix relies on the chain rule and the properties of the exponential function.

### Derivation of the Score Vector $\mathbf{U}(\boldsymbol{\beta})$

The Score vector is defined as the gradient of the log-likelihood with respect to the parameter vector $\boldsymbol{\beta}$.


1. **Differentiate the Log-Likelihood** \n

   Starting from the log-likelihood function:
   $$
   \ell(\boldsymbol{\beta}) = \sum_{i=1}^n \left( y_i (\mathbf{x}_i^\top \boldsymbol{\beta}) - \exp(\mathbf{x}_i^\top \boldsymbol{\beta}) - \log(y_i!) \right)
   $$
   We take the partial derivative with respect to a single component $\beta_j$:
   $$
   \frac{\partial \ell}{\partial \beta_j} = \sum_{i=1}^n \left( y_i \frac{\partial (\mathbf{x}_i^\top \boldsymbol{\beta})}{\partial \beta_j} - \frac{\partial \exp(\mathbf{x}_i^\top \boldsymbol{\beta})}{\partial \beta_j} \right)
   $$

2. **Apply the Chain Rule** \n

   Note that $\frac{\partial (\mathbf{x}_i^\top \boldsymbol{\beta})}{\partial \beta_j} = x_{ij}$. Using the chain rule on the exponential term:
   $$
   \frac{\partial \exp(\mathbf{x}_i^\top \boldsymbol{\beta})}{\partial \beta_j} = \exp(\mathbf{x}_i^\top \boldsymbol{\beta}) \cdot x_{ij} = \lambda_i(\boldsymbol{\beta}) x_{ij}
   $$

3. **Form the Vector** \n

   Substituting these back, we get:
   $$
   \frac{\partial \ell}{\partial \beta_j} = \sum_{i=1}^n (y_i - \lambda_i(\boldsymbol{\beta})) x_{ij}
   $$
   In matrix notation, this corresponds to:
   $$
   \mathbf{U}(\boldsymbol{\beta}) = \mathbf{X}^\top (\mathbf{y} - \boldsymbol{\lambda}(\boldsymbol{\beta}))
   $$




### Derivation of the Observed Fisher Information $J(\boldsymbol{\beta})$

The Observed Information is the negative Hessian matrix of the log-likelihood.


1. **Differentiate the Score** \n

   We take the derivative of the $j$-th component of the score, $U_j = \sum_{i=1}^n (y_i - \lambda_i(\boldsymbol{\beta})) x_{ij}$, with respect to $\beta_k$:
   $$
   \frac{\partial^2 \ell}{\partial \beta_j \partial \beta_k} = \frac{\partial}{\partial \beta_k} \sum_{i=1}^n (y_i - \lambda_i(\boldsymbol{\beta})) x_{ij} = \sum_{i=1}^n \left( -\frac{\partial \lambda_i(\boldsymbol{\beta})}{\partial \beta_k} \right) x_{ij}
   $$

2. **Differentiate the Mean Function** \n

   Recall $\lambda_i(\boldsymbol{\beta}) = \exp(\mathbf{x}_i^\top \boldsymbol{\beta})$. Its derivative with respect to $\beta_k$ is:
   $$
   \frac{\partial \lambda_i(\boldsymbol{\beta})}{\partial \beta_k} = \lambda_i(\boldsymbol{\beta}) x_{ik}
   $$

3. **Construct the Information Matrix** \n

   The $(j, k)$ element of the Hessian is:
   $$
   H_{jk} = -\sum_{i=1}^n \lambda_i(\boldsymbol{\beta}) x_{ij} x_{ik}
   $$
   The Observed Information $J(\boldsymbol{\beta}) = -H$ is therefore:
   $$
   J(\boldsymbol{\beta})_{jk} = \sum_{i=1}^n \lambda_i(\boldsymbol{\beta}) x_{ij} x_{ik}
   $$
   In matrix form, let $\mathbf{W}(\boldsymbol{\beta})$ be a diagonal matrix where $W_{ii} = \lambda_i(\boldsymbol{\beta})$. Then:
   $$
   J(\boldsymbol{\beta}) = \mathbf{X}^\top \mathbf{W}(\boldsymbol{\beta}) \mathbf{X}
   $$




::: {#rem-canonical-simplicity}

### Canonical Link Advantage
Notice that because we used the canonical link ($\log \lambda = \eta$), the second derivative of the log-likelihood (the Hessian) does not depend on the observed data $y_i$. It only depends on the predicted means $\lambda_i(\boldsymbol{\beta})$. Consequently, the **Observed Information** and the **Expected Information** are identical:
$$
J(\boldsymbol{\beta}) = \mathcal{I}(\boldsymbol{\beta})
$$
This is a unique property of canonical links in Generalized Linear Models.

:::

::::