---
title: "Minimum Variance Estimators"
engine: knitr
format: 
  html: default
  pdf: default
---

## Completeness

### Complete Statistic

::: {#def-completeness}

### Complete Statistic
A statistic $T$ is said to be **complete** if for any real-valued function $g$,
$$E[g(T)|\theta] = 0 \quad \text{for all } \theta$$
implies
$$P(g(T) = 0 | \theta) = 1 \quad \text{for all } \theta$$

:::

Significance: If $T$ is complete, then there exists at most one unbiased estimator for $\theta$ that is a function of $T$.

::: {#exm-uniform-incomplete}

### Uniform Distribution (Not Complete)
Let $X_1, \dots, X_n \sim \text{Unif}(\theta-1, \theta+1)$. The statistic $T(X) = (X_{(1)}, X_{(n)})$ is a Minimal Sufficient Statistic. However, it is **not complete**.

Consider the range $R = X_{(n)} - X_{(1)}$. The distribution of $R$ does not depend on $\theta$ (it is an ancillary statistic). Let $g(T) = X_{(n)} - X_{(1)} - c$, where $c = E[X_{(n)} - X_{(1)}]$.
Then $E[g(T)] = 0$ for all $\theta$, but $g(T)$ is not identically zero.

:::

### Exponential Family Completeness

::: {#lem-exponential-complete}

### Exponential Family Completeness
If $T = (T_1, \dots, T_k)$ is the natural statistic of an exponential family that contains an open rectangle in the parameter space, then $T$ is complete.

:::

::: {.proof}
**Proof of Lemma @lem-exponential-complete:**

Consider a $k$-parameter exponential family in canonical form with density:
$$f(y|\boldsymbol{\eta}) = h(y) \exp\left( \sum_{j=1}^k \eta_j T_j(y) - A(\boldsymbol{\eta}) \right)$$
where $\boldsymbol{\eta}$ belongs to a natural parameter space $\Xi$ that contains a $k$-dimensional open rectangle. 

Let $g(T)$ be a function such that $E_{\boldsymbol{\eta}}[g(T)] = 0$ for all $\boldsymbol{\eta} \in \Xi$. The expectation is defined as:
$$\int g(t) h(t) \exp\left( \sum_{j=1}^k \eta_j t_j - A(\boldsymbol{\eta}) \right) dt = 0$$

Since $\exp(-A(\boldsymbol{\eta}))$ is a strictly positive constant with respect to $t$, we can divide it out:
$$\int g(t) h(t) \exp\left( \sum_{j=1}^k \eta_j t_j \right) dt = 0$$

We can decompose $g(t)$ into its positive and negative parts, $g(t) = g^+(t) - g^-(t)$. Substituting this in gives:
$$\int g^+(t) h(t) e^{\sum \eta_j t_j} dt = \int g^-(t) h(t) e^{\sum \eta_j t_j} dt$$

These integrals represent the **multivariate Laplace transforms** of the measures $m^+(t) = g^+(t)h(t)$ and $m^-(t) = g^-(t)h(t)$. A fundamental property of Laplace transforms is that if two transforms are equal over an open set (in this case, the open rectangle in $\Xi$), then the underlying measures must be identical almost everywhere. 

Therefore, $g^+(t)h(t) = g^-(t)h(t)$, which implies $g(t) = 0$ almost everywhere with respect to the distribution of $T$. Thus, $T$ is complete.

:::

### Relationship Between Completeness and Minimality

While sufficiency ensures that a statistic captures all information about $\theta$, and minimality ensures it does so without redundancy, completeness is a stronger property that guarantees the statistic is "uniquely informative" for unbiased estimation.

::: {#thm-complete-minimal}

### Completeness to Minimal Sufficiency
If $T$ is a **complete sufficient statistic** for a family of distributions $\{f(x|\theta) : \theta \in \Theta\}$, then $T$ is also a **minimal sufficient statistic**.

:::

::: {.proof}
To prove $T$ is minimal sufficient, we must show that for any other sufficient statistic $S$, $T$ is a function of $S$.

1.  **Define a Conditional Expectation:**

    Let $S$ be any sufficient statistic. Define the statistic $\phi(S) = E[T|S]$. Because $S$ is sufficient, $\phi(S)$ is a valid statistic (it does not depend on $\theta$).

2.  **Compare Expectations:**

    By the Law of Iterated Expectations:
    $$E[\phi(S)] = E[E[T|S]] = E[T]$$
    This implies that $E[\phi(S) - T] = 0$ for all $\theta \in \Theta$.

3.  **Construct a Function of T:**

    Define $h(T) = E[\phi(S)|T]$. Since $T$ is sufficient, $h(T)$ is a statistic. By the properties of expectation:
    $$E[h(T)] = E[E[\phi(S)|T]] = E[\phi(S)] = E[T]$$
    Therefore, $E[h(T) - T] = 0$ for all $\theta \in \Theta$.

4.  **Apply Completeness:**

    Since $g(T) = h(T) - T$ is a function of $T$ whose expectation is zero for all $\theta$, and $T$ is **complete**, it follows that:
    $$g(T) = 0 \implies h(T) = T \quad \text{with probability 1.}$$

5.  **Conclusion:**

    Since $T$ is a function of $\phi(S)$ (which is a function of $S$), $T$ must be a function of $S$. Since this holds for any sufficient statistic $S$, $T$ is minimal sufficient.

:::

```{r fig-proof-diagram-r}
#| fig-cap: "Logical flow of the proof: Showing T is a function of any sufficient statistic S."
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 4
#| out-extra: 'style="width: 70% !important; display: block; margin-left: auto; margin-right: auto;"'

par(mar = c(1, 1, 1, 1))
plot(0, 0, type = "n", xlim = c(0, 10), ylim = c(0, 10), axes = FALSE, xlab = "", ylab = "")

# Box coordinates and text
text(2, 8, "Sufficient Stat S", cex = 1.2, font = 2)
text(8, 8, expression(phi(S) == E(T* "|" *S)), cex = 1.1)
text(8, 2, expression(h(T) == E(phi(S)* "|" *T)), cex = 1.1)
text(2, 2, "Complete Stat T", cex = 1.2, font = 2)

# Arrows
arrows(3.5, 8, 6.5, 8, length = 0.1, lwd = 2) # S -> phi(S)
arrows(8, 7, 8, 3, length = 0.1, lwd = 2)     # phi(S) -> h(T)
arrows(6.5, 2, 3.5, 2, length = 0.1, lwd = 2, lty = 2) # h(T) -> T (Completeness)
arrows(2, 7, 2, 3, length = 0.1, lwd = 2, col = "blue") # S -> T (Final Result)

# Annotations
text(5, 8.5, "Extract T from S", cex = 0.8)
text(9.2, 5, "Condition on T", cex = 0.8, srt = 270)
text(5, 1.5, "Completeness says h(T) = T", cex = 0.8, col = "red")
text(1, 5, "T is a function of S", cex = 0.8, col = "blue", srt = 90)

```

::: {.remark}
It is important to note that the converse is not generally true. A **minimal sufficient statistic is not necessarily complete**. A standard counterexample is the $\text{Unif}(\theta-1, \theta+1)$ distribution, where the minimal sufficient statistic $(X_{(1)}, X_{(n)})$ is not complete because the range $X_{(n)} - X_{(1)}$ is an ancillary statistic.

:::

The relationship between statistics types is visualized by @fig-stats-hierarchy:

```{r fig-stats-hierarchy}
#| fig-cap: "Hierarchy of Statistics: Completeness implies Minimal Sufficiency under regularity."
#| echo: false
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 4
#| out-extra: 'style="width: 70% !important; display: block; margin-left: auto; margin-right: auto;"'

# Tighten margins
par(mar = c(0.1, 0.1, 0.1, 0.1))

# Setup empty plot
plot(0, 0, type = "n", xlim = c(-2.6, 2.6), ylim = c(-2.5, 2.6), 
     axes = FALSE, xlab = "", ylab = "", asp = 1)

# Function to draw circles
draw_circle <- function(x, y, r) {
  theta <- seq(0, 2 * pi, length.out = 100)
  lines(x + r * cos(theta), y + r * sin(theta), lwd = 1.5)
}

# Drawing hierarchy
draw_circle(0, 0, 2.5)      # Sufficient
draw_circle(0, -0.3, 1.8)   # Minimal SS
draw_circle(0, -0.8, 1.0)   # Complete

# Labels
text(0, 1.8, "Sufficient", font = 2)
text(0, 0.8, "Minimal SS", font = 2)
text(0, -0.8, "Complete", font = 2)

```

## Uniformly Minimum Variance Unbiased Estimator (UMVUE)

### Definition

::: {#def-umvue}

### Uniformly Minimum Variance Unbiased Estimator (UMVUE)
A statistic $T(x)$ is a UMVUE for $\theta$ if:

1.  $E(T(x)|\theta) = \theta$ for all $\theta$ (Unbiased).
2.  $\text{Var}(T(x)|\theta) \le \text{Var}(d(x)|\theta)$ for all $\theta$ and for all other unbiased estimators $d(x)$.

:::



### Rao-Blackwell Theorem

The Rao-Blackwell theorem provides a mechanism for improving an estimator by utilizing a sufficient statistic.

::: {#thm-rao-blackwell}

### Rao-Blackwell Theorem
Given that $T$ is a sufficient statistic and $d_1(x)$ is an unbiased estimator ($E[d_1(x)] = \theta$). Define $g(T) = E[d_1(x) | T]$. Then:

1.  $g(T)$ is a statistic (independent of $\theta$).
2.  $E[g(T)] = \theta$ (Unbiased).
3.  $\text{Var}(g(T)) \le \text{Var}(d_1(x))$.

:::

::: {.proof}

1.  **Statistic Property:** By the definition of sufficiency, the conditional distribution of $X$ given $T$ does not depend on $\theta$. Thus, $g(T) = E[d_1(x)|T]$ does not involve $\theta$ and is a valid statistic.
2.  **Unbiasedness:** By the Law of Iterated Expectations:

    $$E[g(T)] = E_T[ E_X(d_1(X)|T) ] = E_X[d_1(X)] = \theta$$

3.  **Variance Reduction:** Using the Law of Total Variance:

    $$\text{Var}(d_1(X)) = \text{Var}(E[d_1(X)|T]) + E[\text{Var}(d_1(X)|T)] = \text{Var}(g(T)) + E[\text{Var}(d_1(X)|T)]$$
    Since $\text{Var}(d_1(X)|T) \ge 0$, it follows that $E[\text{Var}(d_1(X)|T)] \ge 0$. Therefore, $\text{Var}(g(T)) \le \text{Var}(d_1(X))$. Equality holds only if $d_1(X) = g(T)$ almost surely.

:::



### Lehmann-Scheffé Theorem

While Rao-Blackwell improves an estimator, Lehmann-Scheffé ensures that we have found the *best* one.

::: {#thm-lehmann-scheffe}

### Lehmann-Scheffé Theorem
If $T$ is a complete and sufficient statistic, and there is an unbiased estimator $d(X)$ such that $E[d(X)] = \theta$, then $\phi(T) = E[d(X)|T]$ is the unique UMVUE for $\theta$.

:::

::: {.proof}

1.  **Existence:** From Rao-Blackwell, $\phi(T)$ is an unbiased estimator with variance at most that of any $d(X)$.
2.  **Uniqueness:** Suppose there exists another unbiased estimator $\psi(T)$ that is also a function of the same complete sufficient statistic $T$. Then:

    $$E[\phi(T) - \psi(T) | \theta] = E[\phi(T)|\theta] - E[\psi(T)|\theta] = \theta - \theta = 0$$
    By the property of **completeness** for $T$, $E[g(T)] = 0$ implies $P(g(T)=0) = 1$. Thus, $\phi(T) = \psi(T)$ almost surely. This proves that $\phi(T)$ is the unique unbiased estimator based on $T$, and therefore the unique UMVUE.

:::

## Connection to Likelihood Theory

In the previous chapters, we defined the **Score function** $U(\theta) = \frac{\partial}{\partial \theta} \ell(\theta)$. By the **Fisher-Neyman Factorization Theorem**, if $T$ is sufficient, then $U(\theta)$ is a function of $T$:
$$\ell(\theta) = \log h(x) + \log g(T(x), \theta) \implies U(\theta) = \frac{\partial}{\partial \theta} \log g(T(x), \theta)$$
This confirms that the curvature and gradients of the log-likelihood are entirely contained within the sufficient statistic $T$, explaining why the UMVUE must be a function of $T$.



## Methods for Finding UMVUE

To find the UMVUE for a parameter $\theta$:

1.  **Find a Complete Sufficient Statistic:** Identify $T$ (usually via Exponential Family natural statistics).
2.  **Find an Unbiased Estimator:** Find any simple $d(X)$ such that $E[d(X)] = \theta$.
3.  **Rao-Blackwellize:** Compute $g(T) = E[d(X)|T]$. 

### Example: Poisson UMVUE for $\lambda^2$

::: {#exm-poisson-umvue}

### Poisson UMVUE for $\lambda^2$
Let $X_1, \dots, X_n \sim \text{Poisson}(\lambda)$. $T = \sum X_i$ is complete sufficient.
$E[X_1^2 - X_1] = E[X_1^2] - E[X_1] = (\lambda^2 + \lambda) - \lambda = \lambda^2$.
So $d(X) = X_1^2 - X_1$ is unbiased.
The UMVUE is $g(T) = E[X_1(X_1-1) | T]$. Using the moments of $X_1|T \sim \text{Bin}(T, 1/n)$:
$$g(T) = \frac{T(T-1)}{n^2}$$

:::

### Example: Joint UMVUE in the Normal Family

The power of the Exponential Family lemma is most evident when dealing with distributions involving multiple parameters, such as the Normal distribution.

::: {#exm-normal-joint-umvue}

### Joint UMVUE for $\mu$ and $\sigma^2$
Let $X_1, \dots, X_n \sim N(\mu, \sigma^2)$ where both parameters are unknown. The joint density can be written in the canonical exponential form:

$$f(\mathbf{x}|\mu, \sigma^2) = \underbrace{(2\pi\sigma^2)^{-n/2} \exp\left(-\frac{n\mu^2}{2\sigma^2}\right)}_{C(\boldsymbol{\eta})} \exp\left( \frac{\mu}{\sigma^2}\sum x_i - \frac{1}{2\sigma^2}\sum x_i^2 \right)$$

1.  **Identify Natural Statistics:**

    The natural sufficient statistics are $T_1 = \sum X_i$ and $T_2 = \sum X_i^2$. Because the natural parameter space contains an open rectangle in $\mathbb{R} \times \mathbb{R}^-$, the vector $(T_1, T_2)$ is **complete and sufficient**.

2.  **Transformation to Common Estimators:**

    Any one-to-one mapping of a complete sufficient statistic is also complete and sufficient. We can define:
    $$\bar{X} = \frac{T_1}{n} \quad \text{and} \quad S^2 = \frac{1}{n-1}\left( T_2 - \frac{T_1^2}{n} \right)$$
    Since $(\bar{X}, S^2)$ is a one-to-one function of $(T_1, T_2)$, it is also a **complete sufficient statistic**.

3.  **Applying Lehmann-Scheffé:**

    * Since $E[\bar{X}] = \mu$ and $\bar{X}$ is a function of the complete sufficient statistic, $\bar{X}$ is the **UMVUE for $\mu$**.
    * Since $E[S^2] = \sigma^2$ and $S^2$ is a function of the complete sufficient statistic, $S^2$ is the **UMVUE for $\sigma^2$**.

:::

### UMVUE for $\log(\sigma^2)$ in the Normal Family
:::: {#exm-normal-log-variance}

### UMVUE for $\log(\sigma^2)$ in the Normal Family
While we saw that no UMVUE exists for $\log(\lambda)$ in the Poisson case, an unbiased estimator for the log-variance *does* exist in the Normal distribution $N(\mu, \sigma^2)$.

1. **The Target:** We want to estimate $g(\sigma^2) = \log(\sigma^2)$.
2. **Distributional Fact:** Recall that for a sample of size $n$, the statistic $Y = \frac{(n-1)S^2}{\sigma^2}$ follows a $\chi^2_{n-1}$ distribution (which is a $\text{Gamma}(\alpha = \frac{n-1}{2}, \beta = 2)$ distribution).
3. **Log-Moments of Gamma:** For a variable $W \sim \text{Gamma}(\alpha, \beta)$, the expected value of its logarithm is:

   $$E[\log(W)] = \psi(\alpha) + \log(\beta)$$
   where $\psi(x) = \frac{d}{dx} \log \Gamma(x)$ is the **digamma function**.

4. **Deriving the Estimator:**

   Expanding the expectation of $Y$ and rearranging to isolate $\log(\sigma^2)$:
   $$E\left[ \log(S^2) + \left( \log\left(\frac{n-1}{2}\right) - \psi\left(\frac{n-1}{2}\right) \right) \right] = \log(\sigma^2)$$

5. **Conclusion:**

   Since $S^2$ is a complete sufficient statistic, the **Lehmann-Scheffé Theorem** implies:
   $$\widehat{\log(\sigma^2)}_{\text{UMVUE}} = \log(S^2) + C(n)$$
   where $C(n) = \log\left(\frac{n-1}{2}\right) - \psi\left(\frac{n-1}{2}\right)$ is the **positive** correction term.





Because the logarithm is a concave function, Jensen's Inequality tells us that $E[\log(S^2)] < \log(E[S^2]) = \log(\sigma^2)$. Therefore, $\log(S^2)$ is "too small," and we must add a positive value $C(n)$ to reach unbiasedness.

**Numerical Values of the Correction**

As seen in @tbl-log-bias-correction, the bias is quite severe for small samples but vanishes as $n \to \infty$.

```{r}
#| label: tbl-log-bias-correction
#| tbl-cap: "Positive correction factors $C(n)$ to be added to $\\log(S^2)$"
#| echo: false

n_values <- c(2, 5, 10, 30, 100)
alpha <- (n_values - 1) / 2

# Now calculated as log(alpha) - digamma(alpha) to make it positive
correction <- log(alpha) - digamma(alpha)

results <- data.frame(
  n = n_values,
  alpha = alpha,
  c1 = round(correction, 4)
)

knitr::kable(
  results, 
  col.names = c("$n$", "$\\alpha = \\frac{n-1}{2}$", "Positive Correction $C(n)$"),
  escape = FALSE,
  align = "ccc"
)

```

::::