---
title: "Introduction to Statistical Methods"
subtitle: "Confidence Interval"
author: "Longhai Li"
date: "`r Sys.Date()`"
output: 
   html_document:
       toc: true
       number_sections: true
       highlight: tango
       fig_width: 10
       fig_height: 8
---


# Simulating Confidence Intervals
## Generate a population data

```{r }
X <- rgamma (1000000, 10, .08)
# use another sample
# X <- rbinom (100000, 1, 0.3)

mu <- mean (X) # population mean
mu
sigma <- sd (X) # population sd
sigma
hist (X, main = "Population Distribution")
abline(v=mu,col="green")
```

## Draw sample of sample mean


```{r }
n <- 30
n_rep <- 2000
CI_rep <- matrix (0, n_rep, 3) 

xbar <- replicate (n_rep, mean(sample (X, size = n, replace = T) ) )
data.frame(xbar = xbar[1:10])
```

Look at the distribution of sample mean in comparison with the population


```{r }
xbar.hist <- hist (xbar, nclass=20, xlim = range (X))
abline(v=mu,col="green")
```

## Find 95% confidence intervals for each sample mean 

```{r }
# find margin error (me)
alpha <- 0.05
z <- - qnorm (alpha/2) # or using z <- qnorm (alpha/2, lower = F)
me <- z * sigma/sqrt (n)

# find a CI with the first sample
ci <- c(xbar[1] - me, xbar[1] + me)
mu.is.covered <- ci[1] < mu & mu < ci[2] 

# repeat n_rep times
for (i in 1:n_rep)
{
    CI_rep [i,1:2] <- c(xbar[i] - me, xbar[i] + me)
    CI_rep [i, 3] <- CI_rep [i,1] < mu & mu < CI_rep [i,2] 
}
library("rmarkdown")
paged_table(
round(
  data.frame("true mean"= mu,
             "sample mean"=xbar[1:50], 
             "lower bound" = CI_rep[1:50,1],
             "upper bound" = CI_rep[1:50,2],
             "true mean covered"=CI_rep[1:50,3])
,2)
)
```

## Plot CI 


```{r }
n_plot <- 50

plot (0,0,type = "n", ylab = "Replication ID", xlab = "x",  
      xlim = c(90, 160), #xlim = range (CI_rep[,1:2]),
      ylim = c(0,n_plot+1))
title (main = "Illustration of Confidence Intervals")
abline (v = mu, col = "green", lwd = 3)
abline (v = mu + c(-me, +me), col = "green")
for (i in 1:n_plot)
{
    lines (CI_rep[i,1:2], c(i,i), col =  2 - CI_rep[i, 3] )
    points (xbar[i], i, pch = 4, col = 2 - CI_rep[i,3])
}

## actual cover rate
mean (CI_rep [,3])
```

# Finding Confidence Interval for a Population Mean with t Quantile
```{r}

survey <- read.csv("survey.csv")
colnames(survey)
t.test(survey$Height) # for two-sided symmetric CI
# alternatively
confint (lm(survey$Height~1))
```

# Finding Confidence Interval for a Population Proportion with Z Quantile
```{r}
survey <- read.csv("survey.csv")
freq.sex <- table(survey$Sex); freq.sex
# Confidence interval for probability of female
prop.test (freq.sex[1], sum(freq.sex), conf.level = 0.95, correct = FALSE)
# conputing it directly
p.female <- freq.sex[1]/sum(freq.sex)
ME <- sqrt(p.female*(1 - p.female)/sum(freq.sex))*qnorm(0.975)
# CI 
p.female + c(-ME, ME)

```
Note that for 95\% CI, should use 0.975 lower tail quantile or 0.025 upper tail quantile of standard normal

# Margin Errors of Proportions

```{r }
p<- seq (0,1, by=0.05)

n<- 1200

data.frame(p=p, me=sqrt(p*(1-p)/n)*1.96)

n<- 2070

data.frame(p=p, me=sqrt(p*(1-p)/n)*1.96)
```

