---
title: "Linear Regression with a Categorical Variable"
subtitle: "Revisiting ANOVA with Linear Modelling"
author: "Longhai Li"
date: "September 2019"
output: 
   html_document:
       toc: true
       number_sections: true
       highlight: tango
       fig_width: 10
       fig_height: 8
---
```{r setup, include=FALSE, echo=FALSE}
options(digits = 4)
require("knitr")
opts_knit$set(root.dir = "~/OneDrive - University of Saskatchewan/teaching/stat245_1909/rdemo")
```


# Read a dataset of animal coagulation time vs diet 

```{r }
coagulation <- read.table("data/coagulation.txt", header=T)

coagulation

coag <- coagulation$coag
diet <- coagulation$diet
```

# Visualize the data using scatterplot and boxplot

```{r }
plot (as.numeric(diet), coag, 
      col = diet, 
      pch = as.numeric(diet), 
      xaxt="n", xlab = "Diet")
axis (1, 1:4, LETTERS[1:4])

boxplot(coag~diet)
```

# Fit a Linear Model

## Fit a linear model  with no intercept:

```{r }

# 
g1 <- lm(coag~0+diet)
summary(g1)
# What model is fitted?
data.frame(model.matrix(g1))


```
This model is written as:
$$
y_i = \mu_A I(x_i=A)+\mu_B I(x_i=B)+\mu_C I(x_i=C)+\mu_D I(x_i=D)+\epsilon_i
$$

## Fit a linear model with an intercept:

```{r }

# 
g <- lm(coag~diet)
summary(g)
# What model is fitted?
data.frame(model.matrix(g))


```

This model is written as:
$$
y_i = \beta_0+\beta_B I(x_i=B)+\beta_CI(x_i=C)+\beta_D I(x_i=D)+\epsilon_i
$$
where, $\beta_0=\mu_A, \beta_B=\mu_B-\mu_A, \beta_C=\mu_C-\mu_A, \beta_D=\mu_D-\mu_A$.

# Predictive (fitted) values and Residuals
This section is for illustrating prediction and residuals 

```{r }
y <- coag
y_hat_h0 <- rep (mean (y), length (y)); y_hat_h0
y_hat_h1 <- predict (lm (y~diet)); y_hat_h1;
resid_h0 <- y - y_hat_h0 # residual (error) from h0 model
resid_h1 <- y - y_hat_h1 # residual (error) from h1 model
data.frame (diet, y=y, y_hat_h0, resid_h0, y_hat_h1, resid_h1, y_hat_h1-y_hat_h0 )

# Visulize the predictive values and residuals
plot (y, col = diet,pch = as.numeric(diet),
      main="Coaglution time versus Run Order (Sample ID)")
lines (1:4,g$fitted.values[1:4], col = 1)
lines (5:10,g$fitted.values[5:10], col = 2)
lines (11:16,g$fitted.values[11:16], col = 3)
lines (17:24,g$fitted.values[17:24], col = 4)
abline(h=mean(y))

# compute SS
SSt <-sum((y - y_hat_h0)^2); SSt
SSe <- sum((y - y_hat_h1)^2); SSe
SStr <- SSt - SSe; SStr # alternatively we can compute SStr by
SStr2 <- sum((y_hat_h0-y_hat_h1)^2); SStr2
```

# Short-cut Formulae for Computing Sum Squares 

```{r }
ybar_i. <- tapply (y, mean, INDEX = diet); ybar_i.
ybar_.. <- mean (y); ybar_..
n_i <- table (diet); n_i
n <- length (y)
SSy <- sum (y^2)
# the above values will be provided in test question

# compute SSt,SSe, SStr
SSt <- SSy-n*ybar_..^2; SSt
SStr <- sum(n_i*ybar_i.^2) - n*ybar_..^2; SStr
SSe <- SSt - SStr; SSe

# compute F and output anova table
k <- length (unique (diet));k
n <- length (y);n
df1 <- k - 1
df2 <- n - k
MStr <- SStr/df1
MSe <- SSe/df2
f <- MStr /MSe
pvalue <- 1-pf (f, df1=df1, df2=df2); pvalue

anova_table <- data.frame (Df = c(df1,df2), 
                           SS = c(SStr,SSe), 
                           MS = c(MStr, MSe), 
                           F = c(f,NA), 
                           pvalue = c(pvalue, NA)); 
row.names(anova_table) <- c("diet", "Residuals")
anova_table
```

# R function for ANOVA with Linear Model

Compute ANOVA with the fitting result with intercept
```{r}
g <- lm(coag~diet)
anova (g)
```

Do not compute ANOVA with the fitting result w/o intercept
```{r}
g1 <- lm(coag~0+diet)
anova (g1)
```