---
title: "R Functions for Computing Z-Residuals <code>survreg</code> and <code>coxph</code> Objects"
author: "[Longhai Li](https://longhaisk.github.io/)"
date: "`r format(Sys.Date(),'%B %d, %Y')`"
output: 
  html_document:
    toc: false
    self_contained: false
    number_sections: FALSE
    code_folding: hide
    df_print: paged
    math_method: katex
---

```{css, echo=FALSE}
  
    .main-container {
      max-width: 1200px;
      margin-left: 10px;
      margin-right: 10px;
    }
    
    a{
        font-weight: bold
        color: blue
    }
    body{ /* Normal  */
          font-size: 16px;
    }
    td {  /* Table  */
      font-size: 12px;
    }
    h1.title {
      font-size: 32px;
      color:  Blue;
    }
    h1 { /* Header 1 */
      font-size: 32px;
      color: RoyalBlue;
    }
    h2 { /* Header 2 */
        font-size: 28px;
        color:RoyalBlue;
    }
    h3 { /* Header 3 */
      font-size: 25px;
      font-family: serif;
      color: RoyalBlue;
    }
    code.r{ /* Code block */
        font-size: 12px;
    }
    pre { /* Code block - determines code spacing between lines */
        font-size: 12px;
        /*max-height: 250px;
        overflow-y: auto;*/
    
    }
    
    /*
      Targeting all code-folding buttons to be horizontally centered
      in their normal position in the document flow.
    */
    button.code-folding-btn {
      /* --- Positioning & Centering --- */
      float: none !important;             /* Remove any existing floats like pull-right */
      display: block !important;          /* Make the button a block-level element so auto margins work for centering.
                                            Alternatively, 'table' can work if 'block' causes width issues. */
      margin-left: auto !important;       /* Center the block element horizontally */
      margin-right: auto !important;      /* Center the block element horizontally */
      margin-top: 4px !important;         /* Restore or add some top/bottom margin if needed */
      margin-bottom: 4px !important;      /* e.g., your original margin-bottom: 4px; */
    
      /* --- Optional: Set a max-width if you don't want the button to be tiny --- */
      /* By default, a button's width is based on its content.
         If display:block makes it too narrow or if you want a consistent width: */
      /* width: auto; /* This is usually default for buttons, letting content decide */
      /* min-width: 100px; /* Example: ensure it's at least 100px wide */
      /* max-width: 200px; /* Example: ensure it's not more than 200px wide */
    
      /* --- Keep other desired visual styles from Bootstrap or your previous CSS --- */
      /* (Bootstrap .btn classes will still apply most visual styling) */
      /* You usually don't need to re-declare all visual properties unless
         you want to specifically override Bootstrap's .btn styling. */
    
      /* Example visual properties if needed (many are covered by .btn): */
      padding: 6px 12px; /* Adjust padding if Bootstrap's -xs or -sm is too small for a centered button */
      font-size: 14px;   /* Adjust font size if needed */
      /* color: #333; */
      /* background-color: #fff; */
      /* border: 1px solid #ccc; */
      /* border-radius: 3px; */
      text-align: center; /* Ensures text inside button is centered if button has fixed width */
      cursor: pointer;
      /* ... other styles you want to ensure ... */
    }


```



```{r setup, include=TRUE, echo=FALSE}
library(knitr)

opts_chunk$set(
  echo = TRUE,     # Show code by default
  message = FALSE, # Hide messages by default
  warning = FALSE  # Hide warnings by default
)

nrep <- 50
```






# {.tabset .tabset-fade .tabset-pills}

## R Functions for Z-Residuals





### What is Z-Residual (NRSP Residual)?

Let $t_i$ be a possibly right-censored observation, $\delta_i$ be the indicator for being uncensored, and $\hat{S}_i(\cdot)$ be an estimated survival function of original uncensored $T^*_i$ given covariate $x_i$ and parameters.  The randomized survival probability (**RSP**)  for $t_i$ is defined as:
\begin{equation}
S_i^{R}(t_i, \delta_i, U_i) =
\left\{
\begin{array}{rl}
\hat S_{i}(t_i), & \text{if $t_i$ is uncensored, i.e., $\delta_{i}=1$,}\\
U_i\, \hat S_{i}(t_i), & \text{if $t_i$ is censored, i.e., $\delta_i=0$.} 
\end{array}
\right.
\end{equation}
where $U_i$ is a uniform random number on $(0, 1)$. Li, L., Wu, T., & Feng, C. (2021) shows that RSPs are uniformly distributed on $(0,1)$ under the true model for $T^*_i$. Therefore, they can be transformed into residuals with any desired distribution. Z-residuals are transformed RSPs with the standard normal quantile: 
\begin{equation}
r_{i}^{\text{Z}}(t_i, \delta_i, U_i)=-\Phi^{-1} (S_{i}^R(t_i, \delta_i, U_i))
\end{equation}

Z-residuals are approximately standard normally distributed under the true model with estimated parameters. The approximate normality holds for any censoring distribution. Z-residuals can be used to check the goodness-of-fit of $S_i(\cdot)$ quantitatively by applying SW normality test to Z-residuals, and graphically by visualizing their QQplot. Scatterplots of Z-residuals may reveal the model departure nature. 


Please note that in the original publication by Li et al. (2021), a negative sign was not incorporated into the definition of the Z-residual. Upon further consideration, it was determined that the addition of a negative sign is more appropriate. This modification ensures that larger residuals correspond to larger observed values, which is consistent with the convention for Pearson's residuals.

### R Functions for Computing Z-residuals
```{r zresid}
##input: survreg_fit is a survreg object
resid_survreg<-function(survreg_fit)
{
  distr<-survreg_fit$dist
  y<- survreg_fit$y
  m <- nrow (y)
  parms<- as.numeric(survreg_fit[["parms"]])
  alpha_hat<-1/survreg_fit$scale
  
  if (distr %in% c("weibull","exponential","logistic","lognormal",
                   "loglogistic","gaussian", "loggaussian","rayleigh"))
  {
    SP<-1-(psurvreg(as.data.frame(as.matrix(y))[,-2], 
                    survreg_fit[["linear.predictors"]], 
                    scale=1/alpha_hat, distribution=distr))
    haz <- dsurvreg(as.data.frame(as.matrix(y))[,-2],
                    survreg_fit[["linear.predictors"]],
                    scale=1/alpha_hat, distribution=distr)/SP
  }else if  (distr %in% "t") 
  {
    SP<-1-(psurvreg(as.data.frame(as.matrix(y))[,-2], 
                    survreg_fit[["linear.predictors"]], 
                    scale=1/alpha_hat, distribution=distr,
                    parms = parms))
    haz <- dsurvreg(as.data.frame(as.matrix(y))[,-2], 
                    survreg_fit[["linear.predictors"]],
                    scale=1/alpha_hat, distribution=distr,
                    parms = parms)/SP
  }else stop ("The distribution is not supported")
  censored <- which(as.data.frame(as.matrix(y))[,-1]==0)
  n.censored <- length(censored)
  ## Z-residuals
  RSP <- SP
  RSP[censored] <- RSP[censored]*runif(n.censored)
  zresid <- -qnorm(RSP)
  
  ## Unmodified CS residual
  ucs<- -log(SP)
  ## Modified CS residual
  MSP<- SP
  MSP[censored] <- SP[censored]/exp(1)
  mcs <- -log(MSP)
  nmsp <- -qnorm (MSP)
  ##Martingale Residual
  martg<- as.data.frame(as.matrix(survreg_fit$y))[,-1]- ucs
  ##Deviance Residual
  dev<- sign(martg)* sqrt((-2)*(martg+as.data.frame(as.matrix(survreg_fit$y))[,-1]*
                            log(as.data.frame(as.matrix(survreg_fit$y))[,-1]-martg)))
  zresid.sw.pvalue<-shapiro.test(zresid)$p.value
  ##hazard function
  haz_fn<- haz
  
  list(RSP=RSP,zresid=zresid,zresid.sw.pvalue=zresid.sw.pvalue,
       ucs=ucs, mcs=mcs,martg=martg, dev=dev, nmsp=nmsp,haz_fn=haz_fn)
}
## outputs:
### RSP --- Randomized Survival Probabilities
### zresid --- Normalized RSP 
### zresid.sw.pvalue --- GOF test p-values by applying SW test to Z-residuals
### ucs --- unmodified CS residuals
### mcs --- modified CS residuals
### nmsp --- normalized modified SP
### martg --- Martingale residuals
### dev --- Deviance residuals
### haz_fn --- hazard function of cs residuals


##input: coxfit_fit is a coxph object
resid_coxph<-function(coxfit_fit)
{
  y<- coxfit_fit$y
  m <- nrow (y)
  mre <- resid(coxfit_fit, type="martingale")
  dre <- resid(coxfit_fit, type="deviance")
  ##Unmodified Cox-Snell residual
  ucs <- as.data.frame(as.matrix(y))[,-1] - mre
  ##Survival Function
  SP<- exp(-ucs)
  censored <- which(as.data.frame(as.matrix(y))[,-1]==0)
  n.censored <- length(censored)
  ##NMSP residual (normal-transformed modified survival prob)
  MSP<- SP
  MSP[censored] <- SP[censored]/exp(1)
  mcs <- -log(MSP)
  nmsp<- -qnorm(MSP)
  ##Z-residuals 
  RSP <- SP
  RSP[censored] <- RSP[censored]*runif(n.censored)
  zresid <- -qnorm(RSP)
  zresid.sw.pvalue<-shapiro.test(zresid)$p.value
  list(zresid=zresid,RSP=RSP,zresid.sw.pvalue=zresid.sw.pvalue,
       ucs=ucs,mcs=mcs,nmsp=nmsp)
}
## outputs:
### RSP --- Randomized Survival Probabilities
### zresid --- Normalized RSP 
### zresid.sw.pvalue --- GOF test p-values by applying SW test to Z-residuals
### ucs --- unmodified CS residuals
### mcs --- modified CS residuals
### nmsp --- normalized modified SP
### martg --- Martingale residuals
### dev --- Deviance residuals
### haz_fn --- hazard function of cs residuals

```

### Installation

Copy the above R functions into your R enviroment.

### NRSP vs Z-residuals

This page serves as supplementary material to the research paper by Li, L., Wu, T., & Feng, C. (2021). Within the aforementioned publication, we introduced a residual termed "normalized randomized survival probabilities  residuals (NRSP residual)." Subsequently, for the purposes of brevity and enhanced interpretability, this nomenclature was revised. The residuals are now referred to as "Z-residuals",  reflecting the conventional use of the letter <em>Z</em> to denote a random variable following a standard normal distribution.


The R functions available on this page may still be employed for the computation of Z-residuals for <code>survreg</code> and <code>coxph</code> objects. However, the authors have also released an R package on GitHub called <a href="https://github.com/tiw150/Zresidual"><b>Zresidual</b></a>, which is designed to calculate Z-residuals for a broader array of statistical models including generalized linear models and provide more diagnostic tools for Z-residuals. Further details are referred to the package documentation and <a href="https://tiw150.github.io/Zresidual_demo.html">a demonstration for diagnosing non-linearity</a>. 

### References

<ul>
<li> 
	<b>Li, L.</b>, Wu, T., Feng, C. (2021). Model diagnostics for censored regression via randomized survival probabilities. <a href="https://doi.org/10.1002/sim.8852">Statistics in Medicine 40, 1482–1497</a>.
	<a href="/doc/1911.00198v4.pdf">[<b>PDF</b>]</a>;
	<a href="/software/NRSP"><b>[R Functions and Demonstration]</b></a>; 
	<a href="https://api.zotero.org/users/1693946/publications/items/Z7WFLYHB/file/view">[<b>slides</b>]</a>

<li> 
	Wu, T., <b>Li, L.</b>, Feng, C. (2025). Z-residual diagnostic tool for assessing covariate functional form in shared frailty models. 
	<a href="https://doi.org/10.1080/02664763.2024.2355551">Journal of Applied Statistics 52(1), 28–58.</a> 
	<a href="/doc/jas_z_residual_nolinear.pdf">[<b>PDF</b>]</a>;
	<a href="https://github.com/tiw150/Zresidual"><b>[Z-residual on Github]</b></a>;
	<a href="https://tiw150.github.io/Zresidual_demo.html"><b>[Demo]</b></a>
	<a href="https://api.zotero.org/users/1693946/publications/items/WRQUGMIR/file/view"><b>[slides]</b></a>

	
<li> 
	Wu, T., Feng, C., <b>Li, L.</b> (2025). Cross-validatory Z-Residual for Diagnosing Shared Frailty Models. 
	<a href="https://doi.org/10.1080/00031305.2024.2421370">The American Statistician 79(2), 198–211.</a> <a href="/doc/cv_zresidual_final.pdf">[<b>PDF</b>]</a>;
	<a href="https://www.tandfonline.com/eprint/9CQD3QCVP56MCVKFZWEY/full?target=10.1080/00031305.2024.2421370"><b>[Free Reprints]</b></a>;
	<a href="https://api.zotero.org/users/1693946/publications/items/7FZTDQ3Z/file/view"><b>[slides]</b></a>;
	<a href="https://github.com/tiw150/Zresidual"><b>[Z-residual on Github]</b></a>;
	<a href="https://tiw150.github.io/CV_Zresidual_demo.html"><b>[Demo]</b></a> 

	

</ul>


## Illustration
### Data Generation Function
```{r}
##simulated data:t from weibull distrbution,c from exponentail distrbutions
set.seed(1) 
rexp2 <- function(n, rate){ if (rate==0) rep(Inf,n) else rexp(n=n, rate = rate)}
simulated_data<- function(n,beta0 , beta1 , alpha, mean.censor)
{   
  x <- rbinom(n, size = 1, p = 0.5)
  t0<- rexp2(n, rate= 1/mean.censor)
  survreg_sim_data <- rsurvreg( n, mean = beta0 + beta1 * x,
                                scale=1/alpha, distribution='weibull')
  t <- pmin(survreg_sim_data, t0)
  d <- as.numeric(t0>= survreg_sim_data )
  data_form<- data.frame(survreg_sim_data,t0,t,d,x) 
  out_r<-list(data_form=data_form, alpha=alpha, beta0=beta0, beta1=beta1)
  return (out_r) 
}
```
### Animated Randomized Survival Probabilities of `r nrep` Simulated Datasets

```{r rsp, fig.show="animate", animation.hook='gifski', interval=1, aniopts="controls,loop",cache= TRUE, fig.width=9, fig.height=6.5, echo=TRUE, results="hide"}
library("foreach")
library("survival")

n<-2000
beta0<-2
beta1<-1
alpha<-1.7
mean.censor<-14

##nrep is preset to a number
foreach (j = 1:nrep) %do%
{
   ## simulating a dataset
   out_r<- simulated_data(n=n,beta0=beta0,beta1=beta1,
                          alpha=alpha, mean.censor=mean.censor)
   simulated_data_random<-out_r$data_form
   
   ##checking censoring rate
   table(simulated_data_random$d)
   
   ##fit AFT model
   survreg_fit_t <- survreg(Surv(out_r[[1]]$t, 
                                 out_r[[1]]$d) ~ out_r[[1]]$x,dist="weibull")
   survreg_fit_w <- survreg(Surv(out_r[[1]]$t, 
                                 out_r[[1]]$d) ~ out_r[[1]]$x,dist="lognormal")

   ## compute residuals 
   true_model <- resid_survreg(survreg_fit_t)
   wrong_model<- resid_survreg(survreg_fit_w)

   rsp.t <- true_model$RSP
   zresid.t<- true_model$zresid
   rsp.w <- wrong_model$RSP 
   zresid.w<- wrong_model$zresid

   t<-seq(0, 50, length=n)
   gp1<- which(out_r[[1]]$x==1&out_r[[1]]$d==1)
   gp2<- which(out_r[[1]]$x==1&out_r[[1]]$d==0)
   gp3<- which(out_r[[1]]$x==0&out_r[[1]]$d==1)
   gp4<- which(out_r[[1]]$x==0&out_r[[1]]$d==0)

   s1.t<-1-(psurvreg(t, mean=sum(survreg_fit_t$coefficients),
                     scale=survreg_fit_t$scale, distribution="weibull"))
   s0.t<-1-(psurvreg(t, mean=survreg_fit_t$coefficients[[1]],
                     scale=survreg_fit_t$scale, distribution="weibull"))

   s1.w<-1-(psurvreg(t, mean=sum(survreg_fit_w$coefficients),
                     scale=survreg_fit_w$scale, distribution="lognormal"))
   s0.w<-1-(psurvreg(t, mean=survreg_fit_w$coefficients[[1]], 
                     scale=survreg_fit_w$scale, distribution="lognormal"))
   
   par(mfrow = c(2,2),mar=c(4,4,2,2))
   
   ##The plot of true model
   plot(t, s1.t,type="l", ylab="S(t)",xlab="Time",
        main="RSPs, True Model",ylim=c(0,1.0),xlim=c(0,40))
   lines(t, s0.t)
   ### choose a random sample of observations for display
   sp <- sample(1:n,400)
   s.gp1 <- gp1[which(gp1 %in% sp)]
   s.gp2 <- gp2[which(gp2 %in% sp)]
   s.gp3 <- gp3[which(gp3 %in% sp)]
   s.gp4 <- gp4[which(gp4 %in% sp)]

   points(out_r[[1]]$t[s.gp1], rsp.t[s.gp1],col="green",pch=0)
   points(out_r[[1]]$t[s.gp2],rsp.t[s.gp2],col="red",pch=8)
   points(out_r[[1]]$t[s.gp3],rsp.t[s.gp3],pch=2,col="green")
   points(out_r[[1]]$t[s.gp4],rsp.t[s.gp4],pch=1,col="red")
   legend ("topright", 
           legend = c(expression('Uncensored,x'['i']*"=1"),
                      expression('Censored,x'['i']*'=1'),
                      expression('Uncensored,x'['i']*'=0'),
                      expression('Censored,x'['i']*'=0')),
           pch=c(0,8,2,1), col = c(3,2,3,2))
   hist(rsp.t,xlab="Randomized Survival Probability",main="Histogram of RSPs, True Model")
   
   ##The plot of wrong model
   plot(t, s1.w,type="l",ylab="S(t)",xlab="Time",main="RSPs, Wrong Model",
        ylim=c(0,1.0),xlim=c(0,40))
   lines(t, s0.w)
   
   sp <- sample(1:n,400)
   s.gp1 <- gp1[which(gp1 %in% sp)]
   s.gp2 <- gp2[which(gp2 %in% sp)]
   s.gp3 <- gp3[which(gp3 %in% sp)]
   s.gp4 <- gp4[which(gp4 %in% sp)]

   points(out_r[[1]]$t[s.gp1],rsp.w[s.gp1],col="green",pch=0)
   points(out_r[[1]]$t[s.gp2],rsp.w[s.gp2],col="red",pch=8)
   points(out_r[[1]]$t[s.gp3],rsp.w[s.gp3],pch=2,col="green")
   points(out_r[[1]]$t[s.gp4],rsp.w[s.gp4],pch=1,col="red")

   legend ("topright", legend = c(expression('Uncensored,x'['i']*"=1"),
                                  expression('Censored,x'['i']*'=1'),
                                  expression('Uncensored,x'['i']*'=0'),
                                  expression('Censored,x'['i']*'=0')),
           pch=c(0,8,2,1), col = c(3,2,3,2)
          )
   hist(rsp.w,xlab="Randomized Survival Probability",
        main="Histogram of RSPs, Wrong Model")
}
```

### References
Detailed explanations about this illustration can be found from this paper:

<ul>
<li> 
	<b>Li, L.</b>, Wu, T., Feng, C. (2021). Model diagnostics for censored regression via randomized survival probabilities. <a href="https://doi.org/10.1002/sim.8852">Statistics in Medicine 40, 1482–1497</a>.
	<a href="/doc/1911.00198v4.pdf">[<b>PDF</b>]</a>;
	<a href="/software/NRSP"><b>[R Functions and Demonstration]</b></a>; 
	<a href="https://api.zotero.org/users/1693946/publications/items/Z7WFLYHB/file/view">[<b>slides</b>]</a>
<ul>

## Wrong Family 
### Data Generation Function

```{r}
##simulated data:t from weibull distrbution,c from exponentail distrbutions
set.seed(1) 
rexp2 <- function(n, rate){ if (rate==0) rep(Inf,n) else rexp(n=n, rate = rate)}
simulated_data<- function(n,beta0 , beta1 , alpha, mean.censor)
{   
  x <- rbinom(n, size = 1, p = 0.5)
  t0<- rexp2(n, rate= 1/mean.censor)
  survreg_sim_data <- rsurvreg( n, mean = beta0 + beta1 * x,
                                scale=1/alpha, distribution='weibull')
  t <- pmin(survreg_sim_data, t0)
  d <- as.numeric(t0>= survreg_sim_data )
  data_form<- data.frame(survreg_sim_data,t0,t,d,x) 
  out_r<-list(data_form=data_form, alpha=alpha, beta0=beta0, beta1=beta1)
  return (out_r) 
}
```

### Animated Z-residuals and Cumuative Hazard of CS Residuals of `r nrep` Simulated Datasets

```{r WrongFamily, fig.show="animate",animation.hook='gifski', interval=1, aniopts="controls,loop",cache= TRUE, fig.width=10, fig.height=6.5, echo=TRUE, results="hide"}
library("foreach")
library("survival")


n<- 800
beta0<-2
beta1<-1
alpha<-2
mean.censor<-14.6

##nrep is preset to a number
foreach (j = 1:nrep) %do%
{
    ## simulating a dataset
    out_r<- simulated_data(n=n,beta0=beta0,beta1=beta1,
                           alpha=alpha, mean.censor=mean.censor)
    simulated_data_random<-out_r$data_form
    ##checking censoring rate
    table(simulated_data_random$d)
    
    ##fit AFT model 
    true_model <- survreg(Surv(out_r[[1]]$t, out_r[[1]]$d) ~ out_r[[1]]$x,dist="weibull")
    wrong_model <- survreg(Surv(out_r[[1]]$t, out_r[[1]]$d) ~ out_r[[1]]$x,dist="lognormal")
    zresid.t<-resid_survreg(true_model)
    zresid.w<-resid_survreg(wrong_model)
    
    ##the cumulative hazard function estimated by Kaplan-Meier method of CS residuals
    km.ucs.t <- survfit(Surv(zresid.t$ucs, out_r[[1]]$d)~1,type='fleming')
    id.ucs.t<-order(zresid.t$ucs)
    km.ucs.w <- survfit(Surv(zresid.w$ucs, out_r[[1]]$d)~1,type='fleming')
    id.ucs.w<-order(zresid.w$ucs)
    ##The plot of true model
    par(mfrow = c(2,3),mar=c(4,4,2,2))     
    resid.lim <- c(-1,1) * max(range(abs(zresid.t$zresid),abs(zresid.w$zresid)),5)     
    cs.lim <- c(0,max(6,range(km.ucs.t$cumhaz,km.ucs.w$cumhaz)))
    
    plot(zresid.t$zresid,ylab="Z-residuals",col=out_r[[1]]$d+2,pch=out_r[[1]]$d+1,
         ylim=resid.lim, main="True Model, Z-residuals  plot") 
    legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),
            col = c(3,2), cex=1)
    
    qqnorm(zresid.t$zresid,xlab="Theoretical Quantiles", ylab="Sample Quantiles",
           main= paste0("True Model, QQ plot, SW p-value=", 
                       sprintf("%4.3f",zresid.t$zresid.sw.pvalue)), 
           xlim=resid.lim, ylim = resid.lim)
    abline(a=0, b=1)
    
    
    plot(km.ucs.t, fun="cumhaz", xlab="Unmodified Cox-Snell Residuals",
         ylab="Cumulative Hazard Function", 
         main="True Model, Cum. Hazard of CS Residuals",
         xlim=c(0,6), ylim = c(0,6))
    abline(a=0, b=1, col="blue", lty=2)
    
    points(km.ucs.t$time, -log(km.ucs.t$surv), col =out_r[[1]]$d[id.ucs.t]+2,
           pch=out_r[[1]]$d[id.ucs.t]+1)
    legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),
            col = c(3,2), cex=1)
    

    ##The plot of wrong model
    
    plot(zresid.w$zresid,ylab="Z-residuals",col=out_r[[1]]$d+2,pch=out_r[[1]]$d+1,
         ylim=resid.lim, main="Wrong Model, Z-residuals plot") 
    legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),col = c(3,2), cex=1)
    
    qqnorm(zresid.w$zresid,xlab="Theoretical Quantiles", ylab="Sample Quantiles",
           main=paste0("Wrong Model, QQ plot, SW p-value=",
                       sprintf("%4.3f",zresid.w$zresid.sw.pvalue)), 
           xlim=resid.lim, ylim = resid.lim)
    abline(a=0,b=1)
    

    plot(km.ucs.w, fun="cumhaz", xlab=("Unmodified Cox-Snell Residuals"),
         ylab=("Cumulative Hazard Function"),
         main="Wrong Model, Cum. Hazard of CS Residuals",  
         xlim=c(0,6), ylim = c(0,6))
    abline(a=0, b=1, col="blue", lty=2)
    
    points(km.ucs.w$time, -log(km.ucs.w$surv), col =out_r[[1]]$d[id.ucs.w]+2,
           pch=out_r[[1]]$d[id.ucs.w]+1)
    legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),col = c(3,2), cex=1)
}
```

### References
Detailed explanations about this simulation can be found from this paper:

<ul>
<li> 
	<b>Li, L.</b>, Wu, T., Feng, C. (2021). Model diagnostics for censored regression via randomized survival probabilities. <a href="https://doi.org/10.1002/sim.8852">Statistics in Medicine 40, 1482–1497</a>.
	<a href="/doc/1911.00198v4.pdf">[<b>PDF</b>]</a>;
	<a href="/software/NRSP"><b>[R Functions and Demonstration]</b></a>; 
	<a href="https://api.zotero.org/users/1693946/publications/items/Z7WFLYHB/file/view">[<b>slides</b>]</a>
<ul>


## Non-linearity
### Data Generation Function

```{r}
##simulated data:t from weibull distrbution,c from exponentail distrbutions
set.seed(1) 
rexp2 <- function(n, rate){ if (rate==0) rep(Inf,n) else rexp(n=n, rate = rate)}
simulated_data<- function(n, beta0, beta1, alpha, mean.censor)
{
  x <- runif(n,0,(3*pi)/2)
  fx <-  sin(2*x)
  t0<- rexp2(n,rate= 1/mean.censor)
  ##gamma is shape
  survreg_sim_data <- rsurvreg( n, mean = beta0 + beta1 * fx,
                                scale=1/alpha, distribution='weibull')
  t <- pmin(survreg_sim_data, t0)
  d <- as.numeric(t0>= survreg_sim_data )
  data_form<- data.frame(survreg_sim_data,t0,t,d,x, fx) 
  out_r<-list(data_form=data_form, alpha=alpha, beta0=beta0, beta1=beta1)
  return (out_r) 
}
```

### Animated Z-residuals and Cumuative Hazard of CS Residuals of `r nrep` Simulated Datasets


```{r Nonlinearity, fig.show="animate", animation.hook='gifski', interval=1, aniopts="controls,loop",cache= TRUE, fig.width=10, fig.height=6.5, echo=TRUE, results="hide"}
library("foreach")
library("survival")

n<-800
beta0<-2
beta1<-2
alpha<-1.8
mean.censor<-16

##nrep is preset to a number

foreach (j = 1:nrep) %do%
{
    ## simulating a dataset
    out_r<- simulated_data(n=n,beta0=beta0,beta1=beta1,
                           alpha=alpha,mean.censor=mean.censor)
    simulated_data_random<-out_r$data_form
    ##checking censoring rate
    table(simulated_data_random$d)
    ##fit AFT model 
    true_model <- survreg(Surv(out_r[[1]]$t , out_r[[1]]$d) ~ out_r[[1]]$fx,dist="weibull")
    wrong_model <- survreg(Surv(out_r[[1]]$t , out_r[[1]]$d) ~ out_r[[1]]$x,dist="weibull")
    zresid.t<-resid_survreg(true_model)
    zresid.w<-resid_survreg(wrong_model)
    ##the cumulative hazard function estimated by Kaplan-Meier method of CS residuals
    km.ucs.t <- survfit(Surv(zresid.t$ucs, out_r[[1]]$d)~1,type='fleming')
    id.ucs.t<-order(zresid.t$ucs)
    km.ucs.w <- survfit(Surv(zresid.w$ucs, out_r[[1]]$d)~1,type='fleming')
    id.ucs.w<-order(zresid.w$ucs)
    
    par(mfrow = c(2,3),mar=c(4,4,2,2))     
    resid.lim <- c(-1,1) * max(range(abs(zresid.t$zresid),abs(zresid.w$zresid)),5)     
    cs.lim <- c(0,max(6,range(km.ucs.t$cumhaz,km.ucs.w$cumhaz)))
    ##The plot of true model
    plot(simulated_data_random$x,zresid.t$zresid,xlab="X",ylab="Z-residuals",col=out_r[[1]]$d+2,
         pch=out_r[[1]]$d+1,main="True Model, Z-residuals plot",ylim=resid.lim) 
    legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),col = c(3,2), cex=1)
    
    qqnorm(zresid.t$zresid,xlab="Theoretical Quantiles", ylab="Sample Quantiles",
           main=paste0("True Model, QQ plot, SW p-value=", 
                       sprintf("%4.3f",zresid.t$zresid.sw.pvalue)), 
           xlim=resid.lim, ylim = resid.lim);abline(a=0,b=1)
    
    plot(km.ucs.t, fun="cumhaz", xlab="Unmodified Cox-Snell Residuals",
         ylab="Cumulative Hazard Function",
         main="True Model, Cum. Hazard of CS Residuals",
         xlim=c(0,6), ylim = c(0,6))
    abline(0, 1, col="blue", lty=2)
    points(km.ucs.t$time, -log(km.ucs.t$surv), col =out_r[[1]]$d[id.ucs.t]+2,
           pch=out_r[[1]]$d[id.ucs.t]+1)
    legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),col = c(3,2), cex=1)
    
    ##The plot of wrong model
    plot(simulated_data_random$x,zresid.w$zresid,xlab="X",ylab="Z-residuals",col=out_r[[1]]$d+2,
         pch=out_r[[1]]$d+1,main="Wrong Model, Z-residuals plot",ylim=resid.lim)  
    legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),col = c(3,2), cex=1)
    
    qqnorm(zresid.w$zresid,xlab="Theoretical Quantiles", ylab="Sample Quantiles",
           main=paste0("Wrong Model, QQ plot, SW p-value=", 
                       sprintf("%4.3f",zresid.w$zresid.sw.pvalue)), 
           xlim=resid.lim, ylim = resid.lim);abline(a=0,b=1)
    
    plot(km.ucs.w, fun="cumhaz", xlab=("Unmodified Cox-Snell Residuals"),
         ylab=("Cumulative Hazard Function"),
         main="Wrong Model, Cum. Hazard of CS Residuals", 
         xlim=c(0,6), ylim = c(0,6))
    abline(0, 1, col="blue", lty=2)
    points(km.ucs.w$time, -log(km.ucs.w$surv), col =out_r[[1]]$d[id.ucs.w]+2,
           pch=out_r[[1]]$d[id.ucs.w]+1)
    legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),col = c(3,2), cex=1)
}

```


### References
Detailed explanations about this simulation can be found from this paper:

<ul>
<li> 
	<b>Li, L.</b>, Wu, T., Feng, C. (2021). Model diagnostics for censored regression via randomized survival probabilities. <a href="https://doi.org/10.1002/sim.8852">Statistics in Medicine 40, 1482–1497</a>.
	<a href="/doc/1911.00198v4.pdf">[<b>PDF</b>]</a>;
	<a href="/software/NRSP"><b>[R Functions and Demonstration]</b></a>; 
	<a href="https://api.zotero.org/users/1693946/publications/items/Z7WFLYHB/file/view">[<b>slides</b>]</a>
<ul>


## Real Data

### Load a dataset 
```{r  echo=TRUE}
library(survival)
real.data <- read.csv("real_data.csv")
real.data$grade <- as.factor(real.data$grade)
real.data
```

### Fit three AFT Models
```{r  echo=TRUE, results="hide"}
##fit three AFT models 
fit_weibull     <- survreg(Surv(real.data$time , real.data$status)~.,
                           data=real.data[,-c(1:3)],dist="weibull")
fit_lognormal   <- survreg(Surv(real.data$time , real.data$status)~.,
                           data=real.data[,-c(1:3)],dist="lognormal")
fit_loglogistic <- survreg(Surv(real.data$time , real.data$status) ~.,
                           data=real.data[,-c(1:3)],dist="loglogistic")

```

### `r nrep` Sets of Replicated Z-residuals

```{r realdata, fig.show="animate",animation.hook='gifski', interval=1,aniopts="controls,loop",cache= TRUE, fig.width=10, fig.height=9, echo=TRUE, results="hide"}
library("foreach")
##nrep is preset for controlling the number of replications of Z-residuals
foreach (j = 1:nrep) %do%
{
  ##compute residuals (including Z-residuals)
  resid.wb<-resid_survreg(fit_weibull)
  resid.ln<-resid_survreg(fit_lognormal)
  resid.ll<-resid_survreg(fit_loglogistic)
  
  ##the cumulative hazard function estimated by Kaplan-Meier method of CS residuals
  km.wb <- survfit(Surv(resid.wb$ucs, real.data$status)~1,type='fleming')
  id.wb<-order(resid.wb$ucs)
  km.ln <- survfit(Surv(resid.ln$ucs, real.data$status)~1,type='fleming')
  id.ln<-order(resid.ln$ucs)
  km.ll <- survfit(Surv(resid.ll$ucs, real.data$status)~1,type='fleming')
  id.ll<-order(resid.ll$ucs)
  
  par(mfrow = c(3,3),mar=c(4,4,2,2))
  
  ##The plot of Weibull model
  plot(resid.wb$zresid,xlab="Index", ylab="Z-residuals",main="Weibull,Z-residuals plot", 
       col=real.data$status+2, pch=real.data$status+1,ylim=c(-5,5))
  legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),
          col = c(3,2), cex=1)
  
  qqnorm(resid.wb$zresid,xlab="Theoretical Quantiles",ylab="Sample Quantiles", 
         main=paste0("Weibull, QQ plot, SW p-value=", 
                     sprintf("%4.3f", resid.wb$zresid.sw.pvalue)), 
         xlim=c(-5,5), ylim = c(-5,5))
  abline(a=0,b=1)

  
  plot(km.wb, fun="cumhaz", xlab="Cox-Snell Residuals",
       ylab="Cumulative Hazard Function",
       main="Weibull, Cum. Hazard of CS Residuals",
       ylim= c(0,4),xlim=c(0,4))  
  abline(0, 1, col="blue", lty=2)
  points(km.wb$time,-log(km.wb$surv), col=real.data$status[id.wb]+2,
         pch=real.data$status[id.wb]+1 )
  legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),
          col = c(3,2), cex=1)
  
  ##The plot of Log-normal model
  plot(resid.ln$zresid,xlab="Index",ylab="Z-residuals",main="Log-normal,Z-residuals plot",
       col=real.data$status+2, pch=real.data$status+1,ylim=c(-5,5))
  legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),
          col = c(3,2), cex=1)
  
  qqnorm(resid.ln$zresid,xlab="Theoretical Quantiles", ylab="Sample Quantiles",
         main=paste0("Log-normal, QQ plot, SW p-value=", 
                     sprintf("%4.3f", resid.ln$zresid.sw.pvalue)), 
         xlim=c(-5,5), ylim = c(-5,5))
  abline(a=0,b=1)

  
  plot(km.ln, fun="cumhaz", xlab=("Cox-Snell Residuals"),
       ylab=("Cumulative Hazard Function"),
       main="Log-normal, Cum. Hazard of CS Residuals",
       ylim= c(0,4),xlim=c(0,4))  
  abline(0, 1, col="blue", lty=2)
  points(km.ln$time, -log(km.ln$surv),col=real.data$status[id.ln]+2,
         pch=real.data$status[id.ln]+1 )
  legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),
          col = c(3,2), cex=1)
  
  
  ##The plot of Log-logistic model
  plot(resid.ll$zresid,xlab="Index",ylab="Z-residuals",main="Log-logistic,Z-residuals plot",
       col=real.data$status+2, pch=real.data$status+1,ylim=c(-5,5))
  legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),
          col = c(3,2), cex=1)
  
  qqnorm(resid.ll$zresid,xlab="Theoretical Quantiles", ylab="Sample Quantiles",
         main=paste0("Log-logistic, QQ plot, SW p-value=", 
                     sprintf("%4.3f", resid.ll$zresid.sw.pvalue)), 
         xlim=c(-5,5), ylim = c(-5,5))
  abline(a=0,b=1)
  
  plot(km.ll, fun="cumhaz", xlab=("Cox-Snell Residuals"), 
       ylab=("Cumulative Hazard Function"),
       main="Log-logistic, Cum. Hazard of CS Residuals",
       ylim= c(0,4),xlim=c(0,4))   
  abline(0, 1, col="blue", lty=2)
  points(km.ll$time, -log(km.ll$surv), col=real.data$status[id.ll]+2,
         pch=real.data$status[id.ll]+1)
  legend ("topleft", pch = c(2,1), legend = c("Uncensored", "Censored"),
          col = c(3,2), cex=1)
}
```

### References
Detailed explanations about this data analysis can be found from this paper:

<ul>
<li> 
	<b>Li, L.</b>, Wu, T., Feng, C. (2021). Model diagnostics for censored regression via randomized survival probabilities. <a href="https://doi.org/10.1002/sim.8852">Statistics in Medicine 40, 1482–1497</a>.
	<a href="/doc/1911.00198v4.pdf">[<b>PDF</b>]</a>;
	<a href="/software/NRSP"><b>[R Functions and Demonstration]</b></a>; 
	<a href="https://api.zotero.org/users/1693946/publications/items/Z7WFLYHB/file/view">[<b>slides</b>]</a>
<ul>


```{=html}
<!-- modifying "show" to "show code" --->
<script>
document.addEventListener('DOMContentLoaded', function() {
  setTimeout(function() {
    // Target buttons that have 'code-folding-btn' and 'data-toggle="collapse"'
    var codeFoldingButtons = document.querySelectorAll('button.code-folding-btn[data-toggle="collapse"], a.code-folding-btn[data-toggle="collapse"]');

    codeFoldingButtons.forEach(function(button) {
      var targetSelector = button.getAttribute('href') || button.getAttribute('data-target');
      if (!targetSelector) return;

      var targetElement = null;
      try {
        if (targetSelector.startsWith("#")) {
          targetElement = document.querySelector(targetSelector);
        }
      } catch (e) { return; }

      if (!targetElement || !targetElement.querySelector('pre')) {
        return;
      }

      function updateButtonText(btn) {
        var currentText = btn.textContent.trim();
        if (currentText === "Show") {
          btn.textContent = "Show Code";
        }
        // Optional: to change "Hide" to "Hide Code"
        // else if (currentText === "Hide") {
        //   btn.textContent = "Hide Code";
        // }
      }

      updateButtonText(button);

      var observer = new MutationObserver(function() {
        updateButtonText(button);
      });
      observer.observe(button, { childList: true, characterData: true, subtree: true });
    });
  }, 500); // Delay
});
</script>
```