---
title: "Unequal Probability Sampling"
author: "Longhai Li"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: no
    toc_depth: 2
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
    df_print: paged
  word_document:
    toc: yes
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
  pdf_document: 
    toc: yes
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---


\section{Used Functions or Packages}
# Functions and Packages
```{r}
## ydata --- observations of the variable of interest
## xdata --- observations of the auxilliary variable
## N --- population size

## the output is the ratio of ybarU/xbarU
srs_ratio_est <- function (ydata, xdata, N = Inf)
{	
  n <- length (xdata)
  xbar <- mean (xdata)
  ybar <- mean (ydata)
  B_hat <- ybar / xbar
  d <- ydata - B_hat * xdata
  var_d <- sum (d^2) / (n - 1)
  sd_B_hat <- sqrt ((1 - n/N) * var_d / n) / xbar
  mem <- qt (0.975, df = n - 1) * sd_B_hat
  output <- c (B_hat, sd_B_hat, B_hat - mem, B_hat + mem )
  
  
  names (output) <- c("Est.", "S.E.", "ci.low", "ci.upp" )
  output
}

## ydata --- observations of the variable of interest
## xdata --- observations of the auxilliary variable
## N --- population size
## xbarU --- population mean of auxilliary variable

## the output is the estimate mean or total (est.total=TRUE)
srs_reg_est <- function (ydata, xdata, xbarU, N = Inf, est.total = FALSE)
{
  n <- length (ydata)
  lmfit <- lm (ydata ~ xdata)
  Bhat <- lmfit$coefficients
  efit <- lmfit$residuals
  SSe <- sum (efit^2) / (n - 2) 
  yhat_reg <- Bhat[1] + Bhat[2] * xbarU
  se_yhat_reg <- sqrt ((1-n/N) * SSe / n)
  mem <- qt (0.975, df = n - 2) * se_yhat_reg
  output <- c(yhat_reg, se_yhat_reg, yhat_reg - mem, yhat_reg + mem)
  
  if (est.total) {
      if(!is.finite(N)) stop("N must be finite for estimating population total" )
      output <- output * N
  }

  names (output) <- c("Est.", "S.E.", "ci.low", "ci.upp" )
  output
}

## sdata --- a vector of original survey data
## N --- population size
## to find total, multiply N to the estimate returned by this function

srs_mean_est <- function (sdata, N = Inf)
{
	n <- length (sdata)
	ybar <- mean (sdata)
	se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
	mem <- qt (0.975, df = n - 1) * se.ybar
	c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}

# total estimation for datasets collected with UPS with Replacement
upswr_total_est <- function (total, psi)
{	
  srs_mean_est (total/psi, N = Inf)
}
# ratio estimation for datasets collected with UPS with Replacement
upswr_ratio_est <- function (total, M, psi)
{
  srs_ratio_est (total/psi, M/psi, N = Inf)
}

```

# Hours on Studying (analysis of a dataset collected by UPSWR)
```{r}

psi <- c(24, 100, 100, 76, 44) / 647
Mi <- c(24, 100, 100, 76, 44)
hrs <- c(75, 203, 203, 191, 168)

plot(hrs~Mi)
summary (lm(hrs~Mi))

## total hrs spent by all students

upswr_total_est (total = hrs, psi = psi)

## mean hrs spent by each student
upswr_ratio_est (total = hrs, M = Mi, psi = psi)

```

# Analysis of statepop.csv data (collected with UPSWR)

The file "statepop.csv" is a dataset from an unequal-probability sample of 100 counties in the United States. Counties were chosen using the cumulative-size method from the listings in the County and City Data Book (U.S. Census Bureau, 1994) with probabilities proportional to their populations. The total population for all counties is $M_0 = \sum_{i=1} M_i = 255,077,536$. Sampling was done with replacement, so very large counties occur multiple times in the sample. For example, Los Angeles County, with the largest population in the United States, occurs four times.

```{r}
statepop <- read.csv ("data/statepop.csv")

M0 <- 255077536 ## pop of USA in 1992
popn <- statepop$popn ## pop per county in 1992
psi <- popn / M0
phys <- statepop$phys

plot (popn, phys, log="xy")

summary(lm (phys~psi))
# estimate the total number of physicians in USA
upswr_total_est (phys, psi)

# estimate number of physicians per capita
upswr_ratio_est (phys, popn, psi)
```

**Compare with an estimate based on an SRS sample (Assignment 3 Question 3)**

```{r}
# read srs sample of size 100
counties <- read.csv ("data/counties.csv", na = "-99")
plot(physician~totpop, data=counties, log="xy")

summary (lm(physician~totpop, data=counties))
totpop <-  255077536
N <- 3141 # total number of counties
# naive estimate 
srs_mean_est (counties$physician, N = 3141) * 3141

# ratio estimate
srs_ratio_est (counties$physician, counties$totpop, N = 3141) * totpop

# regression estimate
srs_reg_est (counties$physician, counties$totpop, totpop/N, N = N) * N 

# From the textbook by Lohr, the true value is: 
total_physician <- 532638
```

