---
title: "Data Analysis and Simulation for Ratio and Regression Estimation"
author: "Longhai Li"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: no
    toc_depth: 2
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
    df_print: paged
  word_document:
    toc: yes
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
  pdf_document: 
    toc: yes
    number_sections: yes
    highlight: tango
    fig_width: 10
    fig_height: 8
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---


# Functions and packages for Analyzing Data
```{r}
## ydata --- observations of the variable of interest
## xdata --- observations of the auxilliary variable
## N --- population size
## xbarU --- population mean of auxilliary variable

## the output is the estimate mean or total (est.total=TRUE)
srs_reg_est <- function (ydata, xdata, xbarU, N = Inf, est.total = FALSE)
{
  n <- length (ydata)
  lmfit <- lm (ydata ~ xdata)
  Bhat <- lmfit$coefficients
  efit <- lmfit$residuals
  SSe <- sum (efit^2) / (n - 2) 
  yhat_reg <- Bhat[1] + Bhat[2] * xbarU
  se_yhat_reg <- sqrt ((1-n/N) * SSe / n)
  mem <- qt (0.975, df = n - 2) * se_yhat_reg
  output <- c(yhat_reg, se_yhat_reg, yhat_reg - mem, yhat_reg + mem)
  
  if (est.total) {
      if(!is.finite(N)) stop("N must be finite for estimating population total" )
      output <- output * N
  }

  names (output) <- c("Est.", "S.E.", "ci.low", "ci.upp" )
  output
}


## ydata --- observations of the variable of interest
## xdata --- observations of the auxilliary variable
## N --- population size

## the output is the ratio of ybarU/xbarU
srs_ratio_est <- function (ydata, xdata, N = Inf)
{	
  n <- length (xdata)
  xbar <- mean (xdata)
  ybar <- mean (ydata)
  B_hat <- ybar / xbar
  d <- ydata - B_hat * xdata
  var_d <- sum (d^2) / (n - 1)
  sd_B_hat <- sqrt ((1 - n/N) * var_d / n) / xbar
  mem <- qt (0.975, df = n - 1) * sd_B_hat
  output <- c (B_hat, sd_B_hat, B_hat - mem, B_hat + mem )
  
  
  names (output) <- c("Est.", "S.E.", "ci.low", "ci.upp" )
  output
}

## sdata --- a vector of original survey data in a domain
## N --- population size
## n --- total sample size (not the sample size in the domain)
## to find total, multiply domain size N_d to the estimate returned by this function
srs_domain_mean_est <- function (sdata, n, N = Inf)
{
	n_d <- length (sdata)
	ybar <- mean (sdata)
	se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n_d)  
	mem <- qt (0.975, df = n_d - 1) * se.ybar
	c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}

## sdata --- a vector of original survey data
## N --- population size
## to find total, multiply N to the estimate returned by this function

srs_mean_est <- function (sdata, N = Inf)
{
	n <- length (sdata)
	ybar <- mean (sdata)
	se.ybar <- sqrt((1 - n / N)) * sd (sdata) / sqrt(n)  
	mem <- qt (0.975, df = n - 1) * se.ybar
	c (Est. = ybar, S.E. = se.ybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}

## to find total, multiply N to the estimate returned by this function
## for poststratification, use nh = n * Nh/N
str_mean_estimate <- function (ybarh, sh, nh, Nh)
{
    N <- sum (Nh)
    Pi_h <- Nh/N
    ybar <- sum(ybarh * Pi_h)
    seybar <- sqrt(sum((1-nh/Nh)*Pi_h^2*sh^2/nh))
    mem <- 1.96 * seybar
    c(Est. = ybar, S.E. = seybar, ci.low = ybar - mem, ci.upp = ybar + mem)
}


```




# Ratio  Estimation for cherry.csv dataset 
## Importing data
```{r}
cherry <- read.csv ("data/cherry.csv", header = T)
cherry
plot (cherry$volume ~ cherry$diameter)
summary (lm(cherry$volume ~ 0+cherry$diameter))
```


## SRS estimate 

```{r}
N <- 2967
## estimating the mean of volume
srs_mean_volume <- srs_mean_est(cherry$volume, N = N)
srs_mean_volume
## estimating the total of volume
srs_total_volume <- srs_mean_est(cherry$volume, N = N) * N
srs_total_volume
```

## Step-by-step calculation with Ratio Estimation

### Estimating B and calculating residuals
```{r}
## input
ydata <- cherry$volume
xdata <- cherry$diameter
N <- 2967

## calculation
n <- length (xdata)
xbar <- mean (xdata)
ybar <- mean (ydata)
B_hat <- ybar / xbar ## ratio estimate


plot (cherry$volume ~ cherry$diameter)
abline (a = 0, b = B_hat)

d <- ydata - B_hat * xdata ## errors
data.frame (cherry, d = d)
```

### Estimating SE of B
```{r}
## estimating S^2_e
var_d <- sum (d^2) / (n - 1) ## variance of errors
sd_B_hat <- sqrt ((1 - n/N) * var_d / n) / xbar ## SE for B
mem <- qt (0.975, df = n - 1) * sd_B_hat ## margin error for B

## output
output_B <- c (B_hat, sd_B_hat, B_hat - mem, B_hat + mem )
names (output_B) <- c("Est.", "S.E.", "ci.low", "ci.upp" )
output_B

```

### Estimating the mean volume of wood
```{r}

mean_diameters <- 41835/N
output_B * mean_diameters

```

### Estimating the total volume of wood
```{r}

t_diameters <- 41835
output_B * t_diameters

```


## Ratio estimation with a function 
### Estimating the ratio of volume to diameter
```{r}
B_v2d <- srs_ratio_est (ydata = cherry$volume, xdata = cherry$diameter,  N = 2967)
B_v2d

```


### Estimating the mean of volume
```{r}
xbarU <- 41835/N
ratio_mean_volume <- srs_ratio_est (ydata = cherry$volume, xdata = cherry$diameter, N = 2967) * xbarU
ratio_mean_volume


```

### Estimating the total of volume
```{r}
total_diameters <- 41835
srs_ratio_est (ydata = cherry$volume, xdata = cherry$diameter, N = 2967) * total_diameters

```

### Percentage of Variance Reduction
```{r}
cat(1-(ratio_mean_volume[2]/srs_mean_volume[2])^2)
```

# Simulation study with agpop.csv

## Information of population data 
```{r }
agpop <- read.csv ("data/agpop.csv")
agpop <- agpop[agpop$acres92 != -99, ] ## remove those counties with na

# sample size
n <- 300
# population size
N <- nrow (agpop)

# true values that we want to estimate
tyU <- sum (agpop [,"acres92"])
# suppose known for ratio estimate
txU <- sum (agpop [,"acres87"]) 
B <- tyU/txU
plot (agpop [, "acres87"], agpop [, "acres92"])
abline(a=0,b=B)
# expected reduction of variance of ratio estimate to srs

1-var(agpop$acres92-B*agpop$acres87)/var(agpop$acres92)

# linear model output

summary(lm(agpop$acres92 ~ agpop$acres87))
```

## Simulation Studies
```{r cache=TRUE}

nsim <- 2000
sim_rat <- data.frame (
	simple = rep (0, nsim), ratio = rep (0, nsim), B = rep (0, nsim))

for (i in 1:nsim)
{
  srs <- sample (N,n)
  y_srs <- agpop[srs, "acres92"]
  x_srs <- agpop[srs, "acres87"]

  # SRS estimate
  sim_rat$simple [i] <- mean (y_srs) * N 
  # ratio estimate
  sim_rat$B [i] <- mean (y_srs) / mean (x_srs)
  sim_rat$ratio [i] <-  sim_rat$B [i] * txU
}

sim_rat[1:20,]
boxplot (sim_rat [, 1:2])
abline (h = tyU, col = "red")

sim_var <- sapply (sim_rat[,1:2], var)

ratio_sim_var <- sim_var/sim_var[1]
percentage_reduction <- 1-ratio_sim_var
data.frame (sim_var, ratio_sim_var,percentage_reduction)

```

# Estimating Domain Means and Post-stratification with agsrs.csv example
## Estimating Domain Means
```{r}
agsrs <- read.csv("data/agsrs.csv")
n <- nrow (agsrs)
n
## naively applying SRS estimation to each domain
region_agsrs_domain <- data.frame(rbind(
    NC = srs_domain_mean_est(agsrs$acres92[agsrs$region=="NC"], n=n, N = 3078),
    NE = srs_domain_mean_est(agsrs$acres92[agsrs$region=="NE"], n=n,N = 3078),
    S = srs_domain_mean_est(agsrs$acres92[agsrs$region=="S"], n=n,N = 3078),
    W = srs_domain_mean_est(agsrs$acres92[agsrs$region=="W"], n=n,N = 3078))
)
region_agsrs_domain

## naively applying SRS estimation to each domain gives similar results
region_agsrs_srs <- data.frame(rbind(
    NC = srs_mean_est(agsrs$acres92[agsrs$region=="NC"], N = 3078),
    NE = srs_mean_est(agsrs$acres92[agsrs$region=="NE"], N = 3078),
    S = srs_mean_est(agsrs$acres92[agsrs$region=="S"], N = 3078),
    W = srs_mean_est(agsrs$acres92[agsrs$region=="W"], N = 3078))
)
region_agsrs_srs

```
## Post-stratification Analysis
```{r}
nh <- tapply (1:nrow(agsrs), agsrs[,"region"], length)
n <- sum (nh)
n
sh <- tapply (agsrs[, "acres92"], agsrs[,"region"], sd)
ybarh <- tapply (agsrs[, "acres92"], agsrs[,"region"], mean)
# create a vector with external information
Nh <- c(NC = 1054, NE = 220, S= 1382, W = 422)
```

**Create nh proportional to Nh, instead of using observed nh**

```{r}
nh_post <- Nh/sum(Nh)* n

## regional summary for stratified sampling estimation
data.frame(cbind(ybarh, sh, nh, prop_obs=nh/sum(nh), Nh, nh_post, prop_post = nh_post/sum(nh_post)))

## strata mean estimate
str_mean_estimate (ybarh, sh, nh_post, Nh)

## compare with SRS estimate

srs_mean_est(agsrs$acres92, N = 3087)

## true mean
mean(agpop$acres92)
```


# Estimating Domain Means and Post-stratification with teacher example

## Importing Data
```{r}
teacher <- read.csv ("data/college_teacher.csv")
teacher
table(teacher[,2:1])
addmargins(table(teacher[,2:1]), 2)
prop.table(table(teacher[,2:1]), margin = "gender")
```

## Domain Summary
```{r}
n <- nrow (teacher)
yh <- tapply (teacher$teacher, INDEX = teacher$gender, FUN = mean); yh
sh <- tapply (teacher$teacher, INDEX = teacher$gender, FUN = sd); sh
nh_obs <- tapply (teacher$teacher, INDEX = teacher$gender, FUN = length)
prop_h_obs <- nh_obs/sum (nh_obs); prop_h_obs
data.frame(yh, sh, nh_obs, prop_h_obs)
```

## Estimating Domain Means

```{r}
Nh <- c(3000, 1000)
n <- nrow(teacher)
rbind(
    female=srs_domain_mean_est(subset(teacher, gender==1)$teacher, n=n, N=3000),
    male = srs_domain_mean_est(subset(teacher, gender==2)$teacher, n=n, N=1000)
)
```

## Post-stratification Analysis
```{r}
Nh <- c(3000, 1000)
prop_h_post <- Nh/sum (Nh); prop_h_post
```

**Create nh proportional to Nh, instead of using observed nh**

```{r}
nh_post <- Nh/sum (Nh) * n

## show the differences
data.frame(p_hat = yh, sh, nh_obs, prop_obs=prop_h_obs, nh_post, prop_post=prop_h_post)
```


```{r}
## poststratification estimation of mean
str_mean_estimate (yh, sh, nh_post, Nh)
```

## Analysis with SRS
```{r}
srs_mean_est (teacher$teacher, N=sum(Nh))

```



# Regression Estimation for the cherry.csv dataset 

## Importing data
```{r}
cherry <- read.csv ("data/cherry.csv", header = T)
ydata <- cherry$volume 
xdata <- cherry$diameter
t_diameters <- 41835
xbarU <- t_diameters/2967 
N <- 2967
```

## Step-by-step calculation

### Fitting a linear regression model
```{r}
n <- length (ydata)
lmfit <- lm (ydata ~ xdata)
summary (lmfit)
plot (xdata, ydata)
abline (lmfit)
Bhat <- lmfit$coefficients
efit <- ydata - (Bhat[1] + Bhat[2] * xdata)
data.frame (cherry, residual=efit) ## for visualization

SSe <- sum (efit^2) / (n - 2) 
SSe
## SSe can be obtained directly from lm fitting output
anova(lmfit)
```

### Estimate the mean
```{r}

yhat_reg <- Bhat[1] + Bhat[2] * xbarU
se_yhat_reg <- sqrt ((1-n/N) * SSe / n)
mem <- qt (0.975, df = n - 2) * se_yhat_reg
output <- c(yhat_reg, se_yhat_reg, yhat_reg - mem, yhat_reg + mem)
names (output) <- c("Est.", "S.E.", "ci.low", "ci.upp" )
output
```
### Estimate the total

```{r}
output * N 

```

## Regression estimation Using the function 
### Estimating the mean
```{r}

reg_mean_volume <- srs_reg_est(ydata = cherry$volume, xdata = cherry$diameter, 
                 xbarU=t_diameters/2967, N = 2967) 
reg_mean_volume
```

### Estimating the total
```{r}

t_diameters <- 41835
srs_reg_est(ydata = cherry$volume, xdata = cherry$diameter, 
                 xbarU=t_diameters/2967, N = 2967, est.total = TRUE)

```

### Percentage of Variance Reduction 
```{r}
cat(1-(reg_mean_volume[2]/srs_mean_volume[2])^2)
```


# Regression estimation for photo counts of dead trees
To estimate the number of dead trees in an area, we divide the area into 100 square plots and count the number of dead trees on a photograph of each plot. Photo counts can be made quickly, but sometimes a tree is misclassiﬁed or not detected. So we select an SRS of 25 of the plots for ﬁeld counts of dead trees. We know that the population mean number of dead trees per plot from the photo count is 11.3.

```{r}
photocounts <- c (10,12,7, 13,13, 6,17, 16, 15, 10, 14, 12, 10, 5,12, 10,
10, 9, 6, 11, 7, 9, 11, 10, 10)

fieldcounts <- c (15, 14, 9, 14, 8, 5, 18, 15, 13, 15, 11, 15, 12, 8, 13,
9, 11, 12, 9, 12, 13, 11, 10, 9, 8)


lmfit <- lm (fieldcounts ~ photocounts)
summary (lmfit)
plot (photocounts, fieldcounts)
abline (lmfit)

# estimate for the mean number of dead trees per plot
srs_reg_est (ydata = fieldcounts, xdata = photocounts, xbarU = 11.3, N = 100)

# estimate for the total number of dead trees in the area
srs_reg_est (ydata = fieldcounts, xdata = photocounts, xbarU = 11.3, N = 100, est.total = TRUE)




# compare to simple estimate
srs_mean_est (sdata = fieldcounts, N = 100)

# compare to ratio estimate 
srs_ratio_est (ydata = fieldcounts, xdata = photocounts, N = 100) * 11.3

```


